### 2022-01-26 pset2/solutions.tex
add tex file

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 507; Problem Set 2}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\allowdisplaybreaks

\begin{document}

\maketitle

\question{\textbf{1. Heat Capacities}}
\answer{
    The definition of heat capacity gives that
    \begin{equation}
        C \equiv \dv{Q}{T}
    \end{equation}
    $C_P$ and $C_V$ are $C$ but with the pressure, $P$, and volume, $V$, each held constant respectively. Applying the first law of thermodynamics, this gives
    \begin{equation}
        C_V = \dv{U}{T},\qquad C_P = \dv{U}{T} + P \dv{V}{T}
    \end{equation}
    Therefore the difference of the two is found as follows (applying the ideal gas law).
    \begin{align}
        C_P - C_V &= P \dv{V}{T} \\
                  &= P \dv{T} \qty(\frac{N k_B T}{P}) \\
                  &= P \qty(\frac{N k_B}{P}) \\
        \Aboxed{ C_P - C_V &= N k_B }
    \end{align}
    Now we can additionally find the ratio of the two
    \begin{align}
        \frac{C_P}{C_V} &= \frac{\dv{U}{T} + P \dv{V}{T}}{\dv{U}{T}} \\
                        &= 1 + P \dv{V}{T} \dv{T}{U} \\
                        &= 1 + P \dv{V}{U}
    \end{align}
    \todo{not sure where to go now}
}

\pagebreak

\question{\textbf{2. The Fastest Electron}}
\answer{
    We are given the following information:
    \begin{equation}
        V = (40 \unit{kpc})^3, \qquad n = 0.01 \unit{cm^{-3}}, \qquad T = 10^6 \unit{K}
    \end{equation}
    This means that the total number of particles in this volume is approximately
    \begin{equation}
        N = n V = 1.88 \times 10^{67}
    \end{equation}
    I think that this means that the fastest particle would have a probability of $1 / N$. We of course know that the Maxwell-Boltzmann distribution is
    \begin{equation}
        f(v) = \qty( \frac{m}{2 \pi k_B T} )^{3/2} 4 \pi v^2 \exp \qty(- \frac{m v^2}{2 k_B T})
    \end{equation}
    In order to find the maximum velocity we can set the following integral equal to $1 / N$ since we need to find the velocity above which the probability of finding a particle is $1 / N$ (such that we expect to find a single particle).
    \begin{align}
        \frac{1}{N} &= \int_{v_{\rm max}}^{\infty} f(v) \dd{v} \\
                    &= \int_{v_{\rm max}}^{\infty} \dd{v} \qty( \frac{m}{2 \pi k_B T} )^{3/2} 4 \pi v^2 \exp \qty(- \frac{m v^2}{2 k_B T}) \\
                    &= \sqrt{\frac{2}{\pi}} \int_{v_{\rm max}}^{\infty} \dd{v} \qty( \frac{m}{2 k_B T} )^{3/2} v^2 \exp \qty(- \frac{m v^2}{2 k_B T}) \\
                    &= \sqrt{\frac{2}{\pi}} \int_{v_{\rm max}}^{\infty} \dd{v} A^{3/2} v^2 \exp \qty(- A v^2) \\
                    &\approx \sqrt{\frac{2}{\pi}} \cdot \frac{\sqrt{A}}{2} v_{\rm max} \exp \qty(-A v_{\rm max}^2) \tag{apply hint} \\
        \frac{1}{N} &\approx \sqrt{\frac{A}{2 \pi}} v_{\rm max} \exp \qty(-A v_{\rm max}^2)
    \end{align}
    At this point I wasn't entirely sure how to solve this so I plugged the whole lot into mathematica (using the values for $N$ and $T$ above plus assuming $m = m_e$). After also asserting that $v_{\rm max}$ is large, this gives
    \begin{equation}
        \boxed{ v_{\rm max} = 6.9 \times 10^7 \unit{m}{s^{-1}} }
    \end{equation}
    Given that $0.99 c \approx 3 \times 10^8$, $v_{\rm max}$ is approximately 23\% of the speed of cosmic rays. This implies to me that cosmic rays \textbf{cannot} be thermal in origin.
}

\pagebreak

\question{\textbf{3. Atmospheric Escape}}\\

\question{3a. Compute number density}
\answer{
    We know that the mean free path is defined as
    \begin{equation}
        \lambda_{\rm esc} \equiv \frac{1}{n_{\rm esc} \sigma},
    \end{equation}
    thus we can simply set this equal to the scale height and solve for the number density.
    \begin{align}
        \lambda_{\rm esc} &= H \\
        \frac{1}{n_{\rm esc} \sigma} &= \frac{k_B T}{m g} \\
        \Aboxed{ n_{\rm esc} &= \frac{m g}{\sigma k_B T} }
    \end{align}
}

\question{3b. Find particle flux}
\answer{
    Integration time!
    \begin{align}
        \phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta}
                       &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta}
    \end{align}
}

\end{document}

### 2022-01-27 pset2/solutions.tex
update tex and add pdf

@@
 \newcommand{\avg}[1]{\left\langle #1 \right\rangle}
+\newcommand{\angstrom}{\mbox{\normalfont\AA}}
 \allowdisplaybreaks
@@
 
-\question{3b. Find particle flux}
+\question{3b. Particle flux expression}
 \answer{
-    Integration time!
+    Integration time! Basically just applying the double angle rule and doing some integrals.
     \begin{align}
-        \phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta}
-                       &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta}
+        \phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta} \\
+                       &= \frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \frac{1}{2} \int_0^{\pi/2} \sin (2 \theta) \dd{\theta} \\
+                       &= -\frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} -\frac{1}{4} \qty[\cos (2 \theta)]_0^{\pi/2} \\
+                       &= -\frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} -\frac{1}{4} \cdot \qty(-1 - 1) \\
+                       &= \frac{v f(v)}{8 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \\
+        \Aboxed{ \phi(v) \dd{v}  &= \frac{v f(v)}{4} \dd{v} }
     \end{align}
@@
 
+\question{3c. Total particle flux}
+\answer{
+    Right, let's see if I can remember how to do integrals! Let's start by performing a substitution.
+    \begin{align}
+        \phi &= \int_{\rm v_{\rm esc}}^{\infty} \phi(v) \dd{v} \\
+             &= \frac{1}{4} \int_{\rm v_{\rm esc}}^{\infty} v f(v) \dd{v} \\
+             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} \qty( \frac{m}{2 k_B T} )^{3/2} v^3 \exp \qty(- \frac{m v^2}{2 k_B T}) \dd{v} \\
+             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \dd{v} \\
+           u &= v^2, \qquad \dv{u}{v} = 2 v \\
+        \phi &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
+             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} \frac{1}{2} A^{3/2} u \exp \qty(- A u) \dd{u} \\
+             &= \sqrt{\frac{A^3}{32 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} u \exp \qty(- A u) \dd{u}
+    \end{align}
+    Okay let's pause for a second.
+That covers the substitution part, now we need to do integration by parts.
+Since I'm already using $u$ let's call the two variables in integration by parts $a$ and $b$ such that
+    \begin{equation}
+        \int a \dd{b} = a b - \int b \dd{a}
+    \end{equation}
+    In our case, we have that
+    \begin{equation}
+        a = u,\qquad \dd{b} = \exp(- A u)
+    \end{equation}
+    and thus we can also write that
+    \begin{equation}
+        \dd{a} = 1,\qquad b = - \frac{1}{A} \exp(- A u)
+    \end{equation}
+    So now putting that all together, we can write that the integral in full is
+    \begin{align}
+        &= \qty[ u \cdot - \frac{1}{A} \exp(- A u)]_{\rm v_{\rm esc}^2}^{\infty} - \int_{\rm v_{\rm esc}^2}^{\infty} - \frac{1}{A} \exp(- A u) \dd{u} \\
+        &= \qty[ - \frac{u}{A} \exp(- A u)]_{\rm v_{\rm esc}^2}^{\infty} + \frac{1}{A} \int_{\rm v_{\rm esc}^2}^{\infty} \exp(- A u) \dd{u} \\
+        &= \qty(0 + \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2)) - \qty[\frac{\exp(- A u)}{A^2}]_{\rm v_{\rm esc}^2}^{\infty} \\
+        &= \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2}
+    \end{align}
+    Okay phew, we made it. Now let's go and plug that back in!
+    \begin{align}
+        \phi &= \sqrt{\frac{A^3}{32 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
+             &= \frac{(1 + A v_{\rm esc}^2) e^{- A v_{\rm esc}^2}}{\sqrt{32 \pi A}}
+    \end{align}
+    Now let's put this into the format that you asked for.
+My constant $A$ is related to $v_s$ as $A = v_s^{-2}$ and thus we can write that $A v_{\rm esc}^2 = \lambda_{\rm esc}$.
+Additionally, we want this in terms of the total number of particles and so we multiply by the number density.
+    \begin{equation}
+        \boxed{ \phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{\sqrt{32 \pi}} }
+    \end{equation}
+}
+
+\question{3d. Earth Hydrogen Loss}
+\answer{
+    Let's just quickly write down the constants that we've been given:
+    \begin{equation}
+        m \approx 2 \unit{amu} = 3.32 \times 10^{-27} \unit{kg},\quad T = 1000 K,\quad \sigma = \pi \angstrom^2 = \pi \times 10^{-20} \unit{m^2}
+    \end{equation}
+    where the mass has come from them being hydrogen molecules. Now let's evaluate $\lambda_{\rm esc}, n_{\rm esc}, v_s$ so that we can find $\phi$.
+    \begin{align}
+        v_{\rm esc} &= 11200 \unit{m}{s^{-1}} \\
+        n_{\rm esc} &= \frac{m_{\rm H2} \cdot g}{\sigma k_B T} = \frac{3.35 \times 10^{-27} \unit{kg} \cdot 10 \unit{m}{s^{-2}}}{\pi \times 10^{-20} \unit{m^2} \cdot 1.38 \times 10^{-23} \unit{m^2}{kg}{s^{-2}}{K^{-1}} \cdot 1000 \unit{K}} = 7.66 \times 10^{13} \unit{m^{-3}} \\
+        v_s &= \sqrt{\frac{2 k_B T}{m_{\rm H2}}} = \sqrt{\frac{2 \cdot 1.38 \times 10^{-23} \unit{m^2}{kg}{s^{-2}}{K^{-1}} \cdot 1000 \unit{K}}{3.35 \times 10^{-27} \unit{kg}}} = 2883 \unit{m}{s^{-1}} \\
+        \lambda_{\rm esc} &= \qty(\frac{11200}{2883})^2 = 15.1
+    \end{align}
+    Cool, now it's time to just plug those numbers in to find about $\phi$.
+    \begin{equation}
+        \phi_{\rm H2} = 9.91 \times 10^{10} \unit{m^{-2} s^{-1}}
+    \end{equation}
+    Finally, we need to multiply this by the area through which the particles move.
+Since we are given that $H \ll R$, we can say the surface at this height is essentially the same as the surface area of the Earth.
+    \begin{align}
+        \mathcal{R}_{\rm H2} &= \phi \cdot 4 \pi R_{\oplus}^2 \\
+                    &= 9.91 \times 10^{10} \unit{m^{-2} s^{-1}} \cdot 4 \pi (6.4 \times 10^{6} \unit{m})^2 \\
+        \Aboxed{ \mathcal{R}_{\rm H2} &= 5.1 \times 10^{25} \unit{molecules}{s^{-1}} }
+    \end{align}
+    Therefore, if we were to evolve forwards for $1 \unit{Gyr}$, the total number of molecules that would be lost are
+    \begin{equation}
+        \boxed{ N_{\rm H2, lost} = 1.6 \times 10^{42} }
+    \end{equation}
+    and so this is comparable to the current hydrogen content of the Earth's atmosphere, approximately a factor of 6 lower.
+}
+
+\question{3d. Earth Oxygen Loss}
+\answer{
+    This calculation will be exactly the same except we now change the mass from $m \approx 2\unit{amu}$ to $m \approx 32\unit{amu}$.
+This drastically changes things due to the exponential terms and so the rate of oxygen loss is an extremely low
+    \begin{equation}
+        \boxed{ \mathcal{R}_{\rm O2} = 1.6 \times 10^{-71} \unit{molecules}{s^{-1}} }
+    \end{equation}
+    Meaning that over a billion years we wouldn't expect the loss of \textit{any} oxygen.
+I therefore conclude that over time the abundance of atmospheric oxygen relative to hydrogen will increase (assuming no other sources or sinks).
+}
+
 \end{document}

### 2022-01-28 pset2/solutions.tex
minor changes
reset q1
use capital phi
finish off question 1

@@
     \end{equation}
-    $C_P$ and $C_V$ are $C$ but with the pressure, $P$, and volume, $V$, each held constant respectively. Applying the first law of thermodynamics, this gives
+    $C_P$ and $C_V$ are the same as $C$ but with the pressure, $P$, and volume, $V$, each held constant respectively. Applying the first law of thermodynamics, this gives
     \begin{equation}
-        C_V = \dv{U}{T},\qquad C_P = \dv{U}{T} + P \dv{V}{T}
+        C_V = \qty(\dv{U}{T})_{V, N}\qquad C_P = \qty(\dv{U}{T})_{P} + P \qty(\dv{V}{T})_{P}
     \end{equation}
-    Therefore the difference of the two is found as follows (applying the ideal gas law).
+    Now let's apply the equipartition theorem and therefore write that
+    \begin{equation}
+        U = \frac{1}{2} N f k_B T
+    \end{equation}
+    where $f$ is the number of degrees of freedom. Now let's do some derivatives and apply the ideal gas law ($V = N k_B T / P$)
     \begin{align}
-        C_P - C_V &= P \dv{V}{T} \\
-                  &= P \dv{T} \qty(\frac{N k_B T}{P}) \\
-                  &= P \qty(\frac{N k_B}{P}) \\
-        \Aboxed{ C_P - C_V &= N k_B }
+        C_{V} &= \dv{T} \qty(\frac{1}{2} N f k_B T)\\
+              &= N k_B \cdot \frac{f}{2} \\
+        C_{P} &= \dv{T} \qty(\frac{1}{2} N f k_B T) + P \dv{T} \qty(\frac{N k_B T}{P})\\
+              &= \frac{1}{2} N f k_B + N k_B \\
+              &= N k_B \qty(1 + \frac{f}{2}) \\
     \end{align}
-    Now we can additionally find the ratio of the two
-    \begin{align}
-        \frac{C_P}{C_V} &= \frac{\dv{U}{T} + P \dv{V}{T}}{\dv{U}{T}} \\
-                        &= 1 + P \dv{V}{T} \dv{T}{U} \\
-                        &= 1 + P \dv{V}{U}
-    \end{align}
-    \todo{not sure where to go now}
+    Therefore, the difference between the two is
+    \begin{equation}
+        \boxed{ C_P - C_V = N k_B }
+    \end{equation}
+    And dividing the two gives the adiabatic index
+    \begin{equation}
+        \boxed{ \frac{C_P}{C_V} = 1 + \frac{2}{f} = \gamma }
+    \end{equation}
 }
@@
     \begin{align}
-        \phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta} \\
+        \Phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta} \\
                        &= \frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \frac{1}{2} \int_0^{\pi/2} \sin (2 \theta) \dd{\theta} \\
@@
                        &= \frac{v f(v)}{8 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \\
-        \Aboxed{ \phi(v) \dd{v}  &= \frac{v f(v)}{4} \dd{v} }
+        \Aboxed{ \Phi(v) \dd{v}  &= \frac{v f(v)}{4} \dd{v} }
     \end{align}
@@
     \begin{align}
-        \phi &= \int_{\rm v_{\rm esc}}^{\infty} \phi(v) \dd{v} \\
+        \Phi &= \int_{\rm v_{\rm esc}}^{\infty} \Phi(v) \dd{v} \\
              &= \frac{1}{4} \int_{\rm v_{\rm esc}}^{\infty} v f(v) \dd{v} \\
@@
            u &= v^2, \qquad \dv{u}{v} = 2 v \\
-        \phi &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
+        \Phi &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
              &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} \frac{1}{2} A^{3/2} u \exp \qty(- A u) \dd{u} \\
@@
     \begin{align}
-        \phi &= \sqrt{\frac{A^3}{32 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
+        \Phi &= \sqrt{\frac{A^3}{32 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
              &= \frac{(1 + A v_{\rm esc}^2) e^{- A v_{\rm esc}^2}}{\sqrt{32 \pi A}}
@@
     \begin{equation}
-        \boxed{ \phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{\sqrt{32 \pi}} }
+        \boxed{ \Phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{\sqrt{32 \pi}} }
     \end{equation}

### 2022-02-18 pset3/solutions.tex
start reading the data

--- 2022-01-26 pset2/solutions.tex
@@
 
-\title{Astro 507; Problem Set 2}
+\title{Astro 507; Problem Set 3}
 \author{\textbf{Tom Wagg}}
@@
 
-\question{\textbf{1. Heat Capacities}}
-\answer{
-    The definition of heat capacity gives that
-    \begin{equation}
-        C \equiv \dv{Q}{T}
-    \end{equation}
-    $C_P$ and $C_V$ are $C$ but with the pressure, $P$, and volume, $V$, each held constant respectively. Applying the first law of thermodynamics, this gives
-    \begin{equation}
-        C_V = \dv{U}{T},\qquad C_P = \dv{U}{T} + P \dv{V}{T}
-    \end{equation}
-    Therefore the difference of the two is found as follows (applying the ideal gas law).
-    \begin{align}
-        C_P - C_V &= P \dv{V}{T} \\
-                  &= P \dv{T} \qty(\frac{N k_B T}{P}) \\
-                  &= P \qty(\frac{N k_B}{P}) \\
-        \Aboxed{ C_P - C_V &= N k_B }
-    \end{align}
-    Now we can additionally find the ratio of the two
-    \begin{align}
-        \frac{C_P}{C_V} &= \frac{\dv{U}{T} + P \dv{V}{T}}{\dv{U}{T}} \\
-                        &= 1 + P \dv{V}{T} \dv{T}{U} \\
-                        &= 1 + P \dv{V}{U}
-    \end{align}
-    \todo{not sure where to go now}
-}
+\question{\textbf{1. COBE}}
 
-\pagebreak
+\question{\textbf{2. Entropy}}
 
-\question{\textbf{2. The Fastest Electron}}
-\answer{
-    We are given the following information:
-    \begin{equation}
-        V = (40 \unit{kpc})^3, \qquad n = 0.01 \unit{cm^{-3}}, \qquad T = 10^6 \unit{K}
-    \end{equation}
-    This means that the total number of particles in this volume is approximately
-    \begin{equation}
-        N = n V = 1.88 \times 10^{67}
-    \end{equation}
-    I think that this means that the fastest particle would have a probability of $1 / N$. We of course know that the Maxwell-Boltzmann distribution is
-    \begin{equation}
-        f(v) = \qty( \frac{m}{2 \pi k_B T} )^{3/2} 4 \pi v^2 \exp \qty(- \frac{m v^2}{2 k_B T})
-    \end{equation}
-    In order to find the maximum velocity we can set the following integral equal to $1 / N$ since we need to find the velocity above which the probability of finding a particle is $1 / N$ (such that we expect to find a single particle).
-    \begin{align}
-        \frac{1}{N} &= \int_{v_{\rm max}}^{\infty} f(v) \dd{v} \\
-                    &= \int_{v_{\rm max}}^{\infty} \dd{v} \qty( \frac{m}{2 \pi k_B T} )^{3/2} 4 \pi v^2 \exp \qty(- \frac{m v^2}{2 k_B T}) \\
-                    &= \sqrt{\frac{2}{\pi}} \int_{v_{\rm max}}^{\infty} \dd{v} \qty( \frac{m}{2 k_B T} )^{3/2} v^2 \exp \qty(- \frac{m v^2}{2 k_B T}) \\
-                    &= \sqrt{\frac{2}{\pi}} \int_{v_{\rm max}}^{\infty} \dd{v} A^{3/2} v^2 \exp \qty(- A v^2) \\
-                    &\approx \sqrt{\frac{2}{\pi}} \cdot \frac{\sqrt{A}}{2} v_{\rm max} \exp \qty(-A v_{\rm max}^2) \tag{apply hint} \\
-        \frac{1}{N} &\approx \sqrt{\frac{A}{2 \pi}} v_{\rm max} \exp \qty(-A v_{\rm max}^2)
-    \end{align}
-    At this point I wasn't entirely sure how to solve this so I plugged the whole lot into mathematica (using the values for $N$ and $T$ above plus assuming $m = m_e$).
-After also asserting that $v_{\rm max}$ is large, this gives
-    \begin{equation}
-        \boxed{ v_{\rm max} = 6.9 \times 10^7 \unit{m}{s^{-1}} }
-    \end{equation}
-    Given that $0.99 c \approx 3 \times 10^8$, $v_{\rm max}$ is approximately 23\% of the speed of cosmic rays. This implies to me that cosmic rays \textbf{cannot} be thermal in origin.
-}
-
-\pagebreak
-
-\question{\textbf{3. Atmospheric Escape}}\\
-
-\question{3a. Compute number density}
-\answer{
-    We know that the mean free path is defined as
-    \begin{equation}
-        \lambda_{\rm esc} \equiv \frac{1}{n_{\rm esc} \sigma},
-    \end{equation}
-    thus we can simply set this equal to the scale height and solve for the number density.
-    \begin{align}
-        \lambda_{\rm esc} &= H \\
-        \frac{1}{n_{\rm esc} \sigma} &= \frac{k_B T}{m g} \\
-        \Aboxed{ n_{\rm esc} &= \frac{m g}{\sigma k_B T} }
-    \end{align}
-}
-
-\question{3b. Find particle flux}
-\answer{
-    Integration time!
-    \begin{align}
-        \phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta}
-                       &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta}
-    \end{align}
-}
+\question{\textbf{2. Entropy}}
 

### 2022-02-23 pset3/solutions.tex
2a complete

@@
 
-\question{\textbf{1. COBE}}
+\question{1. \textbf{FIRAS}}
 
-\question{\textbf{2. Entropy}}
+\answer{}
 
-\question{\textbf{2. Entropy}}
+\question{2a. \textbf{Entropy}}
+\answer{
+    Let's derive the entropy for an ideal, non-relativistic Fermi gas (in terms of $V$, $z$ and $T$). For this I will follow Lecture 10. The entropy is defined as
+    \begin{equation}
+        S = \frac{U + PV - N \mu}{T},
+    \end{equation}
+    so we just need to use definitions of $U$, $P$, $N$ and $\mu$. The definitions of $\mu$ is trivial when using the fugacity
+    \begin{equation}
+        \mu = k_B T \ln z
+    \end{equation}
+    For the others, I refer to lecture 10 where we can find the definition of $n$ on slide 5
+    \begin{equation}
+        N = n V = \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V F_{1/2}(z)
+    \end{equation}
+    and the definition of $P$ on slide 7
+    \begin{equation}
+        P = \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z)
+    \end{equation}
+    For the total energy, we don't have it from the slides but we can derive it in the same way using Fermi-Dirac integrals.
+It starts in the same way as $n$ but with an extra factor of $\epsilon$ in the numerator, then we use the substitution $w = \epsilon / k_B T$
+    \begin{align}
+        U &= (2s + 1) \frac{V}{h^3} \int \frac{\epsilon}{e^{\frac{\epsilon - \mu}{k_B T}} + 1} \dd{\va{p}} \\
+        U &= \frac{4\pi (2s + 1)}{2 h^3} (2 m k_B T)^{3/2} k_B T \int_0^\infty \dd{w} \frac{w^{3/2}}{e^w z^{-1} + 1} \\
+        U &= \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T F_{3/2}(z)
+    \end{align}
+    Now it's simply a matter of plugging all of these into the entropy expression.
+    \begin{align}
+        S &= \frac{U + PV - N \mu}{T} \\
+          &= \frac{1}{T} \qty[ \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T F_{3/2}(z) + \frac{4(2s + 1)}{3 \pi^{1/2}} V k_B T \lambda^{-3} F_{3/2}(z) + \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T \ln z F_{1/2}(z)] \\
+          &= \frac{2(2s + 1) V k_B T}{\pi^{1/2} \lambda^3} \qty[ F_{3/2}(z) + \frac{2}{3} F_{3/2}(z) + \ln z F_{1/2}(z)] \\
+        \Aboxed{ S &= \frac{2(2s + 1) V k_B T}{\pi^{1/2} \lambda^3} \qty[ \frac{5}{3} F_{3/2}(z) + F_{1/2}(z) \ln z] }
+    \end{align}
+}
+
+\question{2b. \textbf{Expanding pressure}}
+\answer{
+    \begin{align}
+        \frac{P}{n k_B T} &= \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z) \qty[\frac{2(2s + 1)}{\pi^{1/2} \lambda^3} k_B T F_{1/2}(z)]^{-1} \\
+                          &= \frac{2 F_{3/2}(z)}{3 F_{1/2}(z)}
+    \end{align}
+    \todo{come back to this on Thursday}
+}
+
+\question{\textbf{3. Brown Dwarf}}
+
+\answer{}
 

### 2022-02-23 pset3/solutions.tex
3a done

@@
 
-\answer{}
+\question{3a. \textbf{Fugacity}}
+\answer{
+    The general goal here is to find the number density based on the central temperature and density and then solve for the Fermi-Dirac integral and we can then invert that to find the fugacity.
+To start, we know that
+    \begin{equation}
+        \rho = m_e n_e + m_{\rm H} n_{\rm H} + m_{\rm He} n_{\rm He}
+    \end{equation}
+    We can rewrite this (and neglect the electron term) as
+    \begin{equation}
+        \rho = n_{\rm H} \qty(m_{\rm H} + m_{\rm He} \frac{n_{\rm He}}{n_{\rm H}})
+    \end{equation}
+    Now we can plug in the masses of hydrogen and helium as well as their relative abundance (given as 0.1 in the problem description).
+    \begin{equation}
+        n_{\rm H} = \frac{\rho_c}{1.4 m_p}
+    \end{equation}
+    Now we just relate this to the electron number density simply as
+    \begin{equation}
+        n = \qty(2 \frac{n_{\rm He}}{n_{\rm H}} + 1) n_{\rm H}
+    \end{equation}
+    And combining the two gives an expression for the number density in terms of the central density
+    \begin{equation}
+        n = \frac{1.2 \rho_c}{1.4 m_p}
+    \end{equation}
+    The other part that we need is the de-Broglie wavelength which depends on the central temperature as
+    \begin{equation}
+        \lambda = \frac{h}{\sqrt{2 \pi m k_B T_c}}
+    \end{equation}
+    Now let's use this with the equation for the number density from Lecture 10 and solve for the Fermi-Dirac integral (plugging in values in the final equation).
+    \begin{align}
+        n &= \frac{2(2s + 1)}{n \pi^{1/2} \lambda^3} F_{1/2}(z) \\
+        F_{1/2}(z) &= \frac{n \pi^{1/2} \lambda^3}{2(2s + 1)} \\
+        F_{1/2}(z) &= \frac{\pi^{1/2}}{2(2s + 1)} \frac{1.2 \rho_c}{1.4 m_p} \qty(\frac{h}{\sqrt{2 \pi m k_B T_c}})^3 \\
+        F_{1/2}(z) &= 2.08
+    \end{align}
+    I took this value and inverted it using the approximate analytic expressions from the paper (Aymerich-Humet+1981) to find the fugacity.
+    \begin{equation}
+        \boxed{ z = 1.65 }
+    \end{equation}
+}
 

### 2022-02-23 pset3/solutions.tex
3b
3c

@@
     \begin{equation}
-        \lambda = \frac{h}{\sqrt{2 \pi m k_B T_c}}
+        \lambda = \frac{h}{\sqrt{2 \pi m_e k_B T_c}}
     \end{equation}
@@
         F_{1/2}(z) &= \frac{n \pi^{1/2} \lambda^3}{2(2s + 1)} \\
-        F_{1/2}(z) &= \frac{\pi^{1/2}}{2(2s + 1)} \frac{1.2 \rho_c}{1.4 m_p} \qty(\frac{h}{\sqrt{2 \pi m k_B T_c}})^3 \\
+        F_{1/2}(z) &= \frac{\pi^{1/2}}{2(2s + 1)} \frac{1.2 \rho_c}{1.4 m_p} \qty(\frac{h}{\sqrt{2 \pi m_e k_B T_c}})^3 \\
         F_{1/2}(z) &= 2.08
@@
 
+\question{3b. \textbf{Pressure}}
+\answer{
+    In order to calculate the pressure we need to apply the same pressure equation as 2a
+    \begin{equation}
+        P = \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z)
+    \end{equation}
+    In this specific case, $s = 1/2$, $T = 6 \times 10^6 \unit{K}$, $\lambda$ is calculated in the same way as part a and we found that $z = 1.65$ in part a. This gives the pressure as
+    \begin{equation}
+        \boxed{ P = 1.92 \times 10^{16} \unit{Pa} }
+    \end{equation}
+}
+
+\question{3c. \textbf{Relativistic?}}
+\answer{
+    In lecture we found that electrons are relativistic when their fermi momentum is such that
+    \begin{equation}
+        p_{\rm F} \sim m_e c
+    \end{equation}
+    We showed that in terms of density this was equivalent to
+    \begin{equation}
+        \rho_{\rm rel} \sim 2 \times 10^6 \unit{g}{cm^{-3}} \cdot \frac{\mu_e}{2 m_p}
+    \end{equation}
+    The last term with $\mu_e$ is going to be approximately of order unity and so, since the central density is only $325 \unit{g}{cm^{-3}}$ we have that the electrons \textbf{are} relativistic
+    \begin{equation}
+        \boxed{ \rho_c \ll \rho_{\rm rel} \implies \text{non-relativistic} }
+    \end{equation}
+}
+
 \end{document}

### 2022-02-23 pset3/solutions.tex
3d

@@
     \begin{align}
-        n &= \frac{2(2s + 1)}{n \pi^{1/2} \lambda^3} F_{1/2}(z) \\
+        n &= \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} F_{1/2}(z) \\
         F_{1/2}(z) &= \frac{n \pi^{1/2} \lambda^3}{2(2s + 1)} \\
@@
 
+\question{3d. \textbf{Ideal Gas}}
+\answer{
+    For this part we simply need to calculate the pressure assuming that it was an ideal gas. In this case
+    \begin{equation}
+        P_{\rm ideal} = n k_B T
+    \end{equation}
+    We found $n$ in part a and $T = 6 \times 10^6 \unit{K}$ and thus we can quickly plug these in to find that
+    \begin{equation}
+        \boxed{P_{\rm ideal} = 1.38 \times 10^{16} \unit{Pa}}
+    \end{equation}
+    We can therefore see that the degeneracy contributes significantly since the pressure is about 40\% larger when including degeneracy effects.
+}
+
 \end{document}

### 2022-02-24 pset3/solutions.tex
3e

@@
 }
+\usepackage{empheq}
 
@@
 
+\question{3e. \textbf{Polytropes}}
+\answer{
+    We are given the two following equations
+    \begin{align}
+        \rho_c &= 8.44 \unit{g}{cm^{-3}} \frac{M}{R^3} \\
+        R M^{1/3} &= \frac{0.13}{(\mu_e F_{1/2}(z))^{2/3}}
+    \end{align}
+    First, let's quickly find the mass per electron in units of the proton mass, $\mu_e$, knowing that the gas is 90\% hydrogen (1 nucleon, 1 electron) and 10\% helium (4 nucleons, 2 electrons).
+    \begin{equation}
+        \mu_e = 0.9 \cdot 1 + 0.1 \cdot 2 = 1.1
+    \end{equation}
+    Now we can combine these equations to instead write that
+    \begin{align}
+        M &= \frac{\rho_c R^3}{8.44 \unit{g}{cm^{-3}}} \\
+        R &= \qty[\frac{0.13}{(\mu_e F_{1/2}(z))^{2/3}} \qty(\frac{\rho_c}{8.44 \unit{g}{cm^{-3}}})^{-1/3}]^{1/2}
+    \end{align}
+    Plugging in our numbers $\mu_e = 1.1$, $z = 1.65$ and $\rho_c = 325 \unit{g}{cm^{-3}}$ to get R (and then M) gives
+    \begin{empheq}[box=\fbox]{align}
+        R &= 0.15 \unit{R_{\odot}} \\
+        M &= 0.13 \unit{M_{\odot}}
+    \end{empheq}
+}
+
 \end{document}

### 2022-02-25 pset3/solutions.tex
2b

@@
 \question{1. \textbf{FIRAS}}
-
-\answer{}
+\answer{
+    \todo{copy over stuff from jupyter notebook}
+}
 
@@
 \answer{
+    To start, we can right a simplified expression for $P / n k_B T$ using the expressions from part a.
     \begin{align}
         \frac{P}{n k_B T} &= \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z) \qty[\frac{2(2s + 1)}{\pi^{1/2} \lambda^3} k_B T F_{1/2}(z)]^{-1} \\
-                          &= \frac{2 F_{3/2}(z)}{3 F_{1/2}(z)}
+        \frac{P}{n k_B T} &= \frac{2 F_{3/2}(z)}{3 F_{1/2}(z)}
     \end{align}
-    \todo{come back to this on Thursday}
+    Now we need to expand this in terms of $z$. Let's start by just expanding the Fermi-Dirac integral to start with. We know this is defined as
+    \begin{equation}
+        F_v(z) = \int_0^\infty \dd{w} \frac{w^v}{e^w z^{-1} + 1}
+    \end{equation}
+    If we perform a Taylor expansion on this we find
+    \begin{equation}
+        F_v(z) \approx \int_0^\infty \qty(\sum_{i = 1}^{\infty} e^{- i w} w^v z^i) \dd{w}
+    \end{equation}
+    We can switch the order of this sum and integral and also pull out the $z$ terms to find that
+    \begin{equation}
+        F_v(z) \approx \sum_{i = 1}^{\infty} \qty( z_i \int_0^\infty e^{- i w} w^v \dd{w} )
+    \end{equation}
+    Let's keep the first two terms of this (since just keeping the first results in just getting 1 as the answer haha). This gives the ratio as
+    \begin{equation}
+        \frac{P}{n k_B T} \approx \frac{2}{3} \frac{\frac{3 \sqrt{\pi}}{4} z + \frac{3 \sqrt{\pi}}{16 \sqrt{2}} z^2}{\frac{\sqrt{\pi}}{2} z + \frac{\sqrt{\pi}}{4 \sqrt{2}} z^2}
+    \end{equation}
+    Also of those factors of $\sqrt{\pi}$ cancel and we can also taylor expand again to get a nice simple expression as a function of the fugacity
+    \begin{equation}
+        \boxed{ \frac{P}{n k_B T} \approx 1 + \frac{z}{4 \sqrt{2}} }
+    \end{equation}
 }

### 2022-03-05 pset4/solutions.tex
start hw4, q1 done

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 507; Problem Set 4}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\maketitle

\question{1. \textbf{Impossible Astronomy}}

\question{1a. \textbf{Dense Planet}}
\answer{
    Let's operate under the assumption that the planet is made entirely from a single element. The density of the planet is
    \begin{align}
        \rho &= \frac{M}{\frac{4}{3} \pi R^3} = \frac{3 M_{\rm jup}}{\frac{4}{3} \pi R_{\rm earth}^3} \\
        \Aboxed{ \rho &= 5239 \unit{g}{cm^{-3}} }
    \end{align}
    However, we know that the density of iron is $\rho_{\rm iron} \approx 8 \unit{g}{cm^{-3}}$. Therefore, the planet is much more dense than iron. Since iron is the densest common element that could make up this planet, that means that it must be impossible.
}

\question{1b. \textbf{Cold planet}}
\answer{
    Looking at the plot on Slide 9 of Lecture 12, we see that in the low temperature limit, for a planet with mass $3 M_{\rm jup}$, the maximum possible radius is $R_{\rm max} = R_{\rm jup}$. Therefore the radius of $5 M_{\rm jup}$ is not possible.
}

\question{1c. \textbf{Overgrown Neutron Star}}
\answer{
    We showed in class that the absolute upper bound on the mass of a neutron star is $2.9 \unit{M_{\odot}}$ based on the setting that the sound speed must be less that the speed of light. Therefore, a neutron star of mass $4 \unit{M_{\odot}}$ cannot possibly exist.
}

\question{1d. \textbf{Overgrown White Dwarf}}
\answer{
    The maximum white dwarf mass is the Chandrasekhar mass, $M_{\rm ch} = 1.44 \unit{M_{|odot}}$. So twice the mass of the sun is not possible.
}

\question{1e. \textbf{Chilly White Dwarf}}
\answer{
    Using the white dwarf cooling relation and that the age of the high-$\alpha$ disc is around $12 \unit{Gyr}$, we know that the coolest white dwarf that can exist is around $1500 \unit{K}$. Therefore this white dwarf is too cold and hasn't had enough time to cool.
}

\question{1f. \textbf{Baby black hole}}
\answer{
    If the black hole is a stellar remnant then it must have collapsed in on itself and overcome both electron and neutron degeneracy pressure. Since the ``black hole'' is less than the mass of the sun, it is below the Chandrasekhar mass and so couldn't have overcome this pressure.
}

\end{document}

### 2022-03-09 pset4/solutions.tex
progress on question 2
setup rest of the questions
pset4 2a added

@@
 
+\question{\textbf{Water/Ice Mass-Radius Relation}}
+
+\question{2a. \textbf{Sketch}}
+\answer{
+    I struggled with sketching this so I instead just plotted the functions directly. In the plot below, the blue curve shows the approximate equation of state which is given by
+    \begin{equation}
+        \rho(P) = \rho_0 + K P^{n}
+    \end{equation}
+    The orange curve is in the regime in which $P$ is very large and so we can assume that the first term can be neglected such that
+    \begin{equation}
+        \rho(P) = K P^n
+    \end{equation}
+    Rearranging this into a polytrope form gives
+    \begin{equation}
+        \boxed{ P(\rho) = \qty(\frac{\rho}{K})^{1/n} } 
+    \end{equation}
+    and so the adiabatic index in this case is $\Gamma = 1 / n$.
+Since $n = 0.513$, this means that the water is stiffer than an adiabatic ideal gas at high pressure since $\Gamma = 1 / 0.513 = 1.95$ is larger than $5/3$.
+    \begin{center}
+        \includegraphics[width=\textwidth]{figures/2a.pdf}
+    \end{center}
+}
+
+\question{2b. \textbf{Transition pressure}}
+\answer{
+
+}
+
+\question{2c. \textbf{Temperature Assumptions}}
+\answer{
+
+}
+
+\question{2d. \textbf{Maximum Mass}}
+\answer{
+
+}
+
+\question{\textbf{Radiation-dominated star}}
+
+\question{3a. \textbf{Temperature expression}}
+\answer{
+    See notebook but $1/n$ and so stiffer.
+}
+
+\question{3b. \textbf{Lower-limit on mass}}
+\answer{
+
+}
+
+\question{3c. \textbf{Evaluate mass}}
+\answer{
+
+}
+
+\question{3d. \textbf{Polytropes}}
+\answer{
+
+}
+
+\question{3e. \textbf{Lane-Emden equation}}
+\answer{
+
+}
+
 \end{document}

### 2022-03-09 pset4/solutions.tex
pset4 2b

@@
 \answer{
-
+    Just estimated by eye from the plot, the transition pressure between constant density and rising density occurs at approximately $10^{11} \unit{Pa}$.
+We can also verify this by finding that the density exceeds twice the initial density at a pressure of
+    \begin{equation}
+        \boxed{ P_{\rm trans} \approx 1.17 \times 10^{11} \unit{Pa} }
+    \end{equation}
+    which agrees with our initial estimate.
+The density is constant below this transition pressure as Coulomb forces are still able to withstand the pressure (such that the structure remained and matter is not ``squashed'').
+Once the transition pressure is reached, the Coulomb forces are no longer enough to withstand the pressure and the water/ice is then slowly crushed, increasing the density.
 }

### 2022-03-09 pset4/solutions.tex
pset4 q2c

@@
 \answer{
-
+    It is a good approximation that there is no temperature dependence in this equation because the degeneracy pressure plays a much more important role at these low temperatures.
+The water/ice is not changing phase around these temperatures and the microscopic velocities are not very high, thus the effect on the overall pressure/density is comparitvely negigible.
 }

### 2022-03-09 pset4/solutions.tex
pset4 q2d

@@
 \answer{
-
+    In order to find the maximum mass that a constant density is a good approximation, we can use the equations for the conservation of mass
+    \begin{equation}
+        \dv{M}{r} = 4 \pi r^2 \rho,
+    \end{equation}
+    and hydrostatic equilibrium
+    \begin{equation}
+        \dv{P}{r} = - \frac{G M}{r^2} \rho
+    \end{equation}
+    in the one-zone model, applying the boundary conditions that
+    \begin{align}
+        P_{\rm centre} &= P_{\rm trans},\quad P_{\rm edge} = 0\\
+        \rho_{\rm centre} &= \rho_{\rm edge} = \rho_0\\
+        r_{\rm centre} &= 0
+    \end{align}
+    In class we showed that for the innermost bin of the planet, the hydrostatic equilibrium equation becomes
+    \begin{equation}
+        P_{\rm centre} - P_{\rm edge} = \frac{2 \pi}{3} G \rho_{\rm centre}^2 r_{\rm edge}^2
+    \end{equation}
+    which, using the definitions of the various $P$ and $\rho$s, we can write as
+    \begin{equation}
+        r_{\rm edge} = \sqrt{\frac{3}{2} \frac{P_{\rm trans}}{G \pi \rho_0^2}}
+    \end{equation}
+    Now we can use this in the conservation of mass equation. In this equation we find that
+    \begin{align}
+        \dd{M} &= M_{\rm edge} - M_{\rm centre} = M \\
+        M &= 4 \pi \rho_0 r_{\rm edge}^2 (r_{\rm edge} - r_{\rm centre}) 
+    \end{align}
+    And therefore putting the two together gives that the maximum mass is
+    \begin{equation}
+        M_{\rm max} = 4 \pi \rho_0 \qty(\frac{3}{2} \frac{P_{\rm trans}}{G \pi \rho_0^2})^{\frac{3}{2}}
+    \end{equation}
+    which, using $P_{\rm trans} = 1.17 \times 10^{11} \unit{Pa}$ and $\rho_0 = 1.46 \unit{g}{cm^{-3}}$, gives
+    \begin{equation}
+        \boxed{ M_{\rm max} = 23.9 \unit{M_{\oplus}} }
+    \end{equation}
+    This is therefore much larger than the mass of the Earth, implying that the constant density assumption for the Earth is reasonable.
+I expect that the power low slope would flatten out beyond the transition pressure since the material would start to be crushed by the pressure, thus increasing less in radius for the same increase in mass.
 }

### 2022-03-09 pset4/solutions.tex
pset4 q3b

@@
 \answer{
-    See notebook but $1/n$ and so stiffer.
+    Applying hydrostatic equilibrium, we know that
+    \begin{equation}
+        \frac{P_{\rm centre}}{R} \approx \frac{G M^2}{R^5}
+    \end{equation}
+    If we assume that $P_{\rm centre} = P_{\rm rad} = a_{\rm rad} T^4$ then we can find the expression for the temperature.
+    \begin{align}
+        \frac{a_{\rm rad} T^4}{R} &\approx \frac{G M^2}{R^5}\\
+        \Aboxed{ T &\approx \qty(\frac{G M^2}{a_{\rm rad} R^4})^{1/4} }
+    \end{align}
 }
@@
 \answer{
-
+    Okay strap in folks, it's time for lots of equations! Let's set up the inequality between the radiation and gas pressure first of all (using the ideal gas law and results from part a)
+    \begin{align}
+        P_{\rm rad} &> P_{\rm gas} \\
+        a_{\rm rad} T^4 &> n k_B T \\
+        a_{\rm rad} &> n k_B T^{-3}
+    \end{align}
+    For the next steps we'll need to apply the expression for temperature from the first part.
+But this expression contains the radius and we don't want that!
+Never fear, we can simply write the radius in terms of density since $\rho = 4/3 \pi R^3$ (dropping factors of order unity)
+    \begin{equation}
+        T \approx \qty(\frac{G M^2}{a_{\rm rad}})^{1/4} \qty(\frac{M}{\rho})^{-1/3}
+    \end{equation}
+    Great!
+Now one last thing that would be useful here is to get rid of that pesky $n$ term.
+We could instead turn this number density into a regular density (and once again neglecting factors of order unity)
+    \begin{equation}
+        n \approx \frac{\rho}{m_p}
+    \end{equation}
+    You're sitting there thinking, wait a moment Tom, you've just added $\rho$ terms which we \textit{also} don't know!?
+Don't worry dear reader, as for my next trick I will make them disappear (with some other fun algebra along the way!)
+    \begin{align}
+        a_{\rm rad} &> \frac{\rho}{m_p} k_B \qty(\frac{G M^2}{a_{\rm rad}})^{-3/4} \qty(\frac{M}{\rho}) \\
+        a_{\rm rad} &> \frac{M}{m_p} k_B \qty(\frac{G M^2}{a_{\rm rad}})^{-3/4} \\
+        M^{1/2} a_{\rm rad} &> \frac{k_B}{m_p} \qty(\frac{G}{a_{\rm rad}})^{-3/4} \\
+        M^{1/2} &> \frac{k_B G^{-3/4}}{m_p} a_{\rm rad}^{-1/4} \\
+        M &> \frac{k_B^2}{m_p^2} G^{-3/2} a_{\rm rad}^{-1/2}
+    \end{align}
+    What a journey it's been, now in our final step we can simply apply a definition of $a_{\rm rad}$ that we are given in the question,
+    \begin{equation}
+        a_{\rm rad} = \frac{k_B^4}{c^3 h^3},
+    \end{equation}
+    in order to get our final expression for the lower limit on the mass.
+    \begin{align}
+        M &> \frac{k_B^2}{m_p^2} G^{-3/2} \frac{c^{3/2} h^{3/2}}{k_B^2} \\
+          &> \frac{1}{m_p^2} \qty(\frac{G}{c h})^{-3/2} \\
+        \Aboxed{ M_{\rm rad} &> m_p \qty(\frac{G m_p^2}{c h})^{-3/2} }
+    \end{align}
 }

### 2022-03-09 pset4/solutions.tex
pset4 q3c
pset4 q3d+e

@@
 \answer{
+    Plugging these numbers in we find that the lower limit corresponds to
+    \begin{equation}
+        \boxed{ M_{\rm rad} > 29.2 \unit{M_{\odot}} }
+    \end{equation}
+    which is fairly close to what we would expect for radiation-dominated stars, how festive!
+}
 
-}
+\pagebreak
 
@@
 \answer{
-
+    If we assume that all of the pressure in the star comes from radiation pressure then, from hydrostatic equilibrium, we have that
+    \begin{align}
+        \frac{P}{R} &\approx \frac{G M^2}{R^5} \\
+        P &\approx G M^2 \qty(\frac{M}{\rho})^{-4/3} \\
+        \implies P &\propto \rho^{4/3}
+    \end{align}
+    Therefore we find that $\Gamma = 4/3$ in this case and so the polytrope index is $n = 3$, like a relativistic gas, which I think makes sense for radiation-dominated stars!
+I \textit{think} the assumption we made that results in the pressure just being a function of density and not temperature is just that we are in hydrostatic equilibrium and that all of the pressure comes from radiation but I'm honestly not sure.
 }
@@
 \answer{
-
+    If we could make these assumptions then we could use the Lane-Emden Equation in the same way as in lecture, specifically with $n = 3$.
+This would give us the density of the star as a function of the radius like in the plot in lecture 11 slide 20 and so we'd know the interal structure of stars above the radiation-dominated mass limit.
+    \todo{need to check with Eric what he actually wants from us on this one}
 }

### 2022-03-09 pset4/solutions.tex
proofread pset4

@@
 
-\question{1b. \textbf{Cold planet}}
+\question{1b. \textbf{Overgrown planet}}
 \answer{
-    Looking at the plot on Slide 9 of Lecture 12, we see that in the low temperature limit, for a planet with mass $3 M_{\rm jup}$, the maximum possible radius is $R_{\rm max} = R_{\rm jup}$.
+    Looking at the Slide 9 of Lecture 12, we see that in the low temperature limit, for a planet with mass $3 M_{\rm jup}$, the maximum possible radius is $R_{\rm max} = R_{\rm jup}$.
 Therefore the radius of $5 M_{\rm jup}$ is not possible.
@@
 
-\question{1c. \textbf{Overgrown Neutron Star}}
+\question{1c. \textbf{Overweight Neutron Star}}
 \answer{
-    We showed in class that the absolute upper bound on the mass of a neutron star is $2.9 \unit{M_{\odot}}$ based on the setting that the sound speed must be less that the speed of light.
+    We showed in class that the absolute upper bound on the mass of a neutron star is $2.9 \unit{M_{\odot}}$ based on the requirement that the sound speed must be less that the speed of light.
 Therefore, a neutron star of mass $4 \unit{M_{\odot}}$ cannot possibly exist.
@@
 
-\question{1d. \textbf{Overgrown White Dwarf}}
+\question{1d. \textbf{Overweight White Dwarf}}
 \answer{
-    The maximum white dwarf mass is the Chandrasekhar mass, $M_{\rm ch} = 1.44 \unit{M_{|odot}}$. So twice the mass of the sun is not possible.
+    The maximum white dwarf mass is the Chandrasekhar mass, $M_{\rm ch} = 1.44 \unit{M_{\odot}}$. So twice the mass of the sun is not possible.
 }
 
-\question{1e. \textbf{Chilly White Dwarf}}
+\question{1e. \textbf{White Dwarfs just aren't that cool}}
 \answer{
-    Using the white dwarf cooling relation and that the age of the high-$\alpha$ disc is around $12 \unit{Gyr}$, we know that the coolest white dwarf that can exist is around $1500 \unit{K}$.
+    Using the white dwarf cooling relation and that the age of the high-$\alpha$ disc is around $12 \unit{Gyr}$ old, we know that the coolest white dwarf that can exist is around $1500 \unit{K}$.
 Therefore this white dwarf is too cold and hasn't had enough time to cool.
@@
 Since the ``black hole'' is less than the mass of the sun, it is below the Chandrasekhar mass and so couldn't have overcome this pressure.
+Therefore it is not possible that this is a black hole.
 }
@@
 \answer{
-    Just estimated by eye from the plot, the transition pressure between constant density and rising density occurs at approximately $10^{11} \unit{Pa}$.
+    Just estimating by eye from the plot, the transition pressure between constant density and rising density occurs at approximately $10^{11} \unit{Pa}$.
 We can also verify this by finding that the density exceeds twice the initial density at a pressure of
@@
     \end{equation}
-    which agrees with our initial estimate.
-The density is constant below this transition pressure as Coulomb forces are still able to withstand the pressure (such that the structure remained and matter is not ``squashed'').
+    which agrees with our initial guess.
+The density is constant below this transition pressure as Coulomb forces are still able to withstand the pressure at this point (such that the structure remains and the water/ice is not ``squashed'').
 Once the transition pressure is reached, the Coulomb forces are no longer enough to withstand the pressure and the water/ice is then slowly crushed, increasing the density.
@@
     It is a good approximation that there is no temperature dependence in this equation because the degeneracy pressure plays a much more important role at these low temperatures.
-The water/ice is not changing phase around these temperatures and the microscopic velocities are not very high, thus the effect on the overall pressure/density is comparitvely negigible.
+The water/ice is not changing phase around these temperatures and the microscopic velocities are not very high, thus the effect on the overall pressure/density is comparitively negigible.
 }
@@
     \end{equation}
-    which, using the definitions of the various $P$ and $\rho$s, we can write as
+    which, using the definitions of the various $P$'s and $\rho$'s, we can rearrange into
     \begin{equation}
@@
 Now one last thing that would be useful here is to get rid of that pesky $n$ term.
-We could instead turn this number density into a regular density (and once again neglecting factors of order unity)
+We could instead turn this number density into a regular density (and once again neglect factors of order unity)
     \begin{equation}

### 2022-03-15 507_thermo/pset1/solutions.tex
move all of thermo into a folder

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 507; Problem Set 1}
\author{\textbf{Tom Wagg}}
\date{January 20, 2022}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\allowdisplaybreaks

\begin{document}

\maketitle

\noindent\textbf{Links for your convenience}\\
The implementation of my simulation can be found here: \url{https://github.com/TomWagg/thermo_winter22/blob/tom/pset1}. The simulation code specifically is in \texttt{simulator.py}. All code for reproducing figures can be found in this folder under files called \texttt{problem\_x.py}, where \texttt{x} is the problem number. Additionally, full size figure files are in the \texttt{figures} directory.\\

\question{1a. \textbf{Code Summary}}
\answer{
    I initially wrote this simulation in a more object-oriented way but found that it was a tad slow and so switched to keeping track of everything in arrays and vectorising the code. The \texttt{Simulation} class handles pretty much everything and you can input the particle/box details in this Class.

    I use cgs units for my variables: sizes are in centimetres, masses are in grams, times in seconds, and energies in ergs. I visualise the code using \texttt{tkinter} which a group of us set up with a lot of help from David! This can be toggled on or off by using the \texttt{visualise} parameter in the initialisation function of the class.

    I determine whether a collision has occurred by checking whether these two conditions hold
    \begin{align}
        \norm{\vb{x_1} - \vb{x_2}} &<= r_1 + r_2 \\
        \vb{v_1} \cdot \vb{v_2} &< 0
    \end{align}
    where $\vb{x}$ is the position of a particle, $r$ is the radius of a particle, $\vb{v}$ is the velocity of a particle and the subscript correspond to the different particles. These conditions ensure that the particles are in contact and moving towards one another.

    For resolving particle collisions I solve these equations (which I got from \href{https://en.wikipedia.org/wiki/Elastic_collision\#Two-dimensional_collision_with_two_moving_objects}{here})
    \begin{align}
        \mathbf{v}_{1}^{\prime} &= \mathbf{v}_{1}-\frac{2 m_{2}}{m_{1}+m_{2}} \frac{\left\langle\mathbf{v}_{1}-\mathbf{v}_{2}, \mathbf{x}_{1}-\mathbf{x}_{2}\right\rangle}{\left\|\mathbf{x}_{1}-\mathbf{x}_{2}\right\|^{2}}\left(\mathbf{x}_{1}-\mathbf{x}_{2}\right) \\
        \mathbf{v}_{2}^{\prime} &= \mathbf{v}_{2}-\frac{2 m_{1}}{m_{1}+m_{2}} \frac{\left\langle\mathbf{v}_{2}-\mathbf{v}_{1}, \mathbf{x}_{2}-\mathbf{x}_{1}\right\rangle}{\left\|\mathbf{x}_{2}-\mathbf{x}_{1}\right\|^{2}}\left(\mathbf{x}_{2}-\mathbf{x}_{1}\right)
    \end{align}
    These equations essentially redistribute the velocity along the contact angle based on the initial velocities of each particle and their masses, $m$.

    Finally, for wall collisions, I check whether a particle is within a radius of one of the edges of the box. If this is the case, I reflect the velocity component that is perpendicular to the wall.
}

\pagebreak

\question{1b. \textbf{Snapshot Plots}}
\answer{
    The initial and final distributions of particles are shown in Figure~\ref{fig:1b}.
    \begin{figure}[htbp]
        \centering
        \includegraphics[width=\textwidth]{figures/1b.png}
        \caption{Initial and final distributions of the positions and velocities of the particles. The setting for this simulation were a box of size 1000cm, particles of size 20cm, initial energy of each particle of 0.1 erg and each with a mass of 1g.}
        \label{fig:1b}
    \end{figure}
}

\pagebreak

\question{2a. \textbf{2D Maxwellian}}
\answer{
    Okay let's start with the given function
    \begin{equation}
        f(q, p) \propto \exp \qty( - \frac{E}{k_B T} )
    \end{equation}
    We can first use this to find the speed distribution $\dv{N}{v}$, where $v = \abs{\vb{v}}$.
    \begin{align}
        \dd{N} &= A \exp \qty( - \frac{E}{k_B T} ) \dd{\vb{v}} \\
                  &= A v \exp \qty( - \frac{E}{k_B T} ) \dd{v} \dd{\theta} \\
                  &= A v \exp \qty( - \frac{m v^2}{2 k_B T} ) \dd{v} \dd{\theta}
    \end{align}
    We can normalise this distribution since we know it must be equal to 1 and assume it is isotropic.
    \begin{align}
        \int_0^{\infty} \int_0^{2\pi} A v \exp \qty( - \frac{m v^2}{2 k_B T} ) \dd{v} \dd{\theta} &= 1 \\
        2 \pi A \int_0^{\infty} v \exp \qty( - \frac{m v^2}{2 k_B T} ) \dd{v} &= 1 \\
        2 \pi A \cdot \frac{k_B T}{m} &= 1 \\
        A &= \frac{m}{2 \pi k_B T}
    \end{align}
    This implies that the final distribution (after integrating out the angles) is
    \begin{equation}\label{eq:v_dist}
        \boxed{ \dv{N}{v} = \frac{m v}{k_B T} \exp \qty( - \frac{m v^2}{2 k_B T} ) }
    \end{equation}
    Now I can be lazy and do a change of variables instead of going through that whole rigmarole again since we know that
    \begin{align}
        v &= \qty(\frac{2 E}{m})^{1/2} \\
        \dv{v}{E} &= \qty(\frac{2}{m})^{1/2} \cdot \frac{1}{2} E^{-1/2} \\
                  &= \qty(\frac{1}{2 m E})^{1/2}
    \end{align}
    This means that we can write the energy distribution as
    \begin{align}
        \dv{N}{E} &= \dv{N}{v} \dv{v}{E} \\
                  &= \qty(\frac{1}{2 m E})^{1/2} \frac{m \qty(\frac{2 E}{m})^{1/2}}{k_B T} \exp \qty( - \frac{m \qty(\frac{2 E}{m})}{2 k_B T} ) \\
        \Aboxed{ \dv{N}{E} &= \frac{1}{k_B T} \exp \qty( - \frac{E}{k_B T} ) }\label{eq:e_dist}
    \end{align}
}

\question{2b. \textbf{Show simulation approaches distribution}}
\answer{
    In Figure~\ref{fig:2b}, I show that my simulated speed and energy distributions approaches the distributions given in Equations~\ref{eq:v_dist} and \ref{eq:e_dist}. To make this plot I use a box of 300 particles of radius 3cm and mass 1g, where the box is 750cm in length and each particle starts with 1 erg of energy.

    I then evolve the simulation for 10,000 seconds and save the distributions at this point. Next, to increase the number of samples I evolve it 100 more times for 100 seconds each and save the speeds and energies in each of these cases. In this way, I effectively have a sample of 30,000 particles instead of 300 and so the histogram looks nice and smooth (IMHO) :D

    One can see that the simulation reproduces the analytic distributions quite nicely, fun!
    \begin{figure}[htb]
        \centering
        \includegraphics[width=\textwidth]{figures/2b.pdf}
        \caption{Plot showing the speed and energy distributions of the simulation converging to the expected analytic result. For the analytic distributions I use Equations~\ref{eq:v_dist} and \ref{eq:e_dist}.}
        \label{fig:2b}
    \end{figure}
}

\question{2c. \textbf{Multiple masses}}
\answer{
    In Figure~\ref{fig:2c_velocity}, I show the new distributions for when half of the masses are 10 times higher.

    Let's talk about the speed distribution first (left side). The distribution no longer follows a single Maxwellian, but instead follows a \textit{mixture} of two Maxwellians. The total distribution is found by summing the distribution using the $v_{\rm rms}$ from the lower mass and higher mass particles separately (and then normalising). This makes sense because the particles of different masses will move at different speeds when given the same kinetic energy and so they form two distributions.

    The distribution for the energy is found in the same way. However, the distribution still resembles a single Maxwellian since the functional form doesn't explicitly depend on the mass. It also makes sense intuitively since each particle is given the same initial kinetic energy regardless of mass.

    \begin{figure}[htb]
        \centering
        \includegraphics[width=\textwidth]{figures/2c.pdf}
        \caption{As Figure~\ref{fig:2b}, except now half of the particles have masses of 10 g. For the speed distribution I overlay individual Maxwellians for each of the subpopulations as well as their normalised sum.}
        \label{fig:2c_velocity}
    \end{figure}
}

\question{3a. \textbf{Relaxation Criterion}}
\answer{
    For a numerical criterion I used a K-S test on both the speed and the energy. I convert the distributions into CDFs and then used \texttt{scipy.stats.kstest} to attain p-values for both tests. Once both tests attain a p-value of at least 0.05 I assume that the system has relaxed to the steady state.
}

\question{3b. \textbf{Relaxation time derivation}}
\answer{
    The relaxation timescale is essentially the amount of time it takes the system to forget its initial conditions. This loss of information will occur during collisions and thus we can write the relaxation timescale as a function of the number of collisions.

    So, let's first derive the average time between collisions for a particle. We are given that the initial speed of the particle is $v_0$, its cross section is $\sigma$ and the number density of particles is $n$. Given these values, we know that the area that the particle moves through in a time $t$ is
    \begin{equation}
        A = \sigma v_0 t.
    \end{equation}
    Then the number of collisions that occurs is this area multiplied by the number of density and it is then trivial to convert this to an average time until a collision.
    \begin{align}
        n_{\rm collision} = n \sigma v_0 t,\\
        t_{\rm one-collision} = \frac{1}{n \sigma v_0}
    \end{align}
    Finally, we can write that the relaxation timescale is the time for $X$ collisions to occur and so it is
    \begin{equation}\label{eq:t_relax}
        \boxed{ \tau_{\rm relax} = \frac{X}{n \sigma v_0} }
    \end{equation}
    Deciding on the exact value of $X$ is a more difficult decision since we would need to know how many collisions every particles needs to have to forget their initial conditions and so for now I simply state that $\tau_{\rm relax}$ scales in this way.
}

\question{3c. \textbf{Comparison with criterion}}
\answer{
    I compare my answers in 3a and 3b in Figure~\ref{fig:3c}. I found that value of $X = 1.2$ gave a reasonably good fit and so I used that in the plots. For the simulations I used the same default values as problem 1 and varied the number of particles and their size.

    For each respective number of particles and radius, I ran the simulation 10 times and I plot these results as error bars spanning the full range with a point at the median of the values. You can see that the results approximately agree!
    \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{figures/3c.pdf}
        \caption{A comparison of the numerical criterion from 3a with the analytic formula in 3b. I use a value of $X = 1.2$ in Equation~\ref{eq:t_relax} since that appears to give a reasonable fit.}
        \label{fig:3c}
    \end{figure}
}

\question{\textbf{4. Pressure (Extra Credit)}}
\answer{
    In Figure~\ref{fig:4}, I show that the relation $P = n k_B T$ holds for my simulation. For right-hand side we can write that
    \begin{align}
        n k_B T &= \frac{N}{A} \cdot \frac{1}{2} m v_{\rm rms}^2 \\
                &= \frac{N m v_{\rm rms}^2}{2 L^2}
    \end{align}
    where $N$ is the number of particles, $L$ is the length of one side of the box, $m$ is the mass of particles and $v_{\rm rms}$ is the root-mean-square velocity. Then for the left-hand side we can write that the average pressure on the wall of the box over the whole simulation is simply
    \begin{align}
        P &= \frac{\Delta p}{4 L \cdot \Delta t} \\
          &= \frac{\sum_{i=1}^{N_{\rm collide}} 2 m \abs{v_{\perp, i}}}{4 L \cdot \Delta t}
    \end{align}
    where $N_{\rm collide}$ is the number of collisions with the walls, $v_{\perp}$ is the speed in the direction perpendicular to the wall with which the particle collided and $\Delta t$ is the length of time for which the simulation was run.

    You can see from the bottom panel of Figure~\ref{fig:4} that my simulation shows strong agreement between the value for $P$ and $n k_B T$ across many different values of $N$ (and therefore $n$) and thus shows that this relation holds. For reference, I created this figure using a box of size 1000cm and particles with initial KE of 1 erg, masses of 1g and radii 5cm.
    \begin{figure}[htb]
        \centering
        \includegraphics[width=\textwidth]{figures/4.pdf}
        \caption{Demonstration that $P = n k_B T$ holds for my simulation. The top panel plots the values of $P$ and $n k_B T$ separately for different values of $N$. The bottom panel shows their ratio with a line through unity and shading to indicate a 20\% difference.}
        \label{fig:4}
    \end{figure}
}

\end{document}

### 2022-03-15 507_thermo/pset2/solutions.tex
move all of thermo into a folder

--- 2022-01-28 pset2/solutions.tex
@@
              &= \frac{1}{4} \int_{\rm v_{\rm esc}}^{\infty} v f(v) \dd{v} \\
-             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} \qty( \frac{m}{2 k_B T} )^{3/2} v^3 \exp \qty(- \frac{m v^2}{2 k_B T}) \dd{v} \\
-             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \dd{v} \\
+             &= \sqrt{\frac{1}{\pi}} \int_{\rm v_{\rm esc}}^{\infty} \qty( \frac{m}{2 k_B T} )^{3/2} v^3 \exp \qty(- \frac{m v^2}{2 k_B T}) \dd{v} \\
+             &= \sqrt{\frac{1}{\pi}} \int_{\rm v_{\rm esc}}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \dd{v} \\
            u &= v^2, \qquad \dv{u}{v} = 2 v \\
-        \Phi &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
-             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} \frac{1}{2} A^{3/2} u \exp \qty(- A u) \dd{u} \\
-             &= \sqrt{\frac{A^3}{32 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} u \exp \qty(- A u) \dd{u}
+        \Phi &= \sqrt{\frac{1}{\pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
+             &= \sqrt{\frac{1}{\pi}} \int_{\rm v_{\rm esc}^2}^{\infty} \frac{1}{2} A^{3/2} u \exp \qty(- A u) \dd{u} \\
+             &= \sqrt{\frac{A^3}{4 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} u \exp \qty(- A u) \dd{u}
     \end{align}
@@
     \begin{align}
-        \Phi &= \sqrt{\frac{A^3}{32 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
-             &= \frac{(1 + A v_{\rm esc}^2) e^{- A v_{\rm esc}^2}}{\sqrt{32 \pi A}}
+        \Phi &= \sqrt{\frac{A^3}{4 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
+             &= \frac{(1 + A v_{\rm esc}^2) e^{- A v_{\rm esc}^2}}{2 \sqrt{\pi A}}
     \end{align}
@@
     \begin{equation}
-        \boxed{ \Phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{\sqrt{32 \pi}} }
+        \boxed{ \Phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{2 \sqrt{\pi}} }
     \end{equation}
@@
     \end{equation}
-    where the mass has come from them being hydrogen molecules. Now let's evaluate $\lambda_{\rm esc}, n_{\rm esc}, v_s$ so that we can find $\phi$.
+    where the mass has come from them being hydrogen molecules. Now let's evaluate $\lambda_{\rm esc}, n_{\rm esc}, v_s$ so that we can find $\Phi$.
     \begin{align}
@@
     \end{align}
-    Cool, now it's time to just plug those numbers in to find about $\phi$.
+    Cool, now it's time to just plug those numbers in to find about $\Phi$.
     \begin{equation}
-        \phi_{\rm H2} = 9.91 \times 10^{10} \unit{m^{-2} s^{-1}}
+        \Phi_{\rm H2} = 2.80 \times 10^{10} \unit{m^{-2} s^{-1}}
     \end{equation}
@@
     \begin{align}
-        \mathcal{R}_{\rm H2} &= \phi \cdot 4 \pi R_{\oplus}^2 \\
-                    &= 9.91 \times 10^{10} \unit{m^{-2} s^{-1}} \cdot 4 \pi (6.4 \times 10^{6} \unit{m})^2 \\
-        \Aboxed{ \mathcal{R}_{\rm H2} &= 5.1 \times 10^{25} \unit{molecules}{s^{-1}} }
+        \mathcal{R}_{\rm H2} &= \Phi \cdot 4 \pi R_{\oplus}^2 \\
+                    &= 2.80 \times 10^{11} \unit{m^{-2} s^{-1}} \cdot 4 \pi (6.4 \times 10^{6} \unit{m})^2 \\
+        \Aboxed{ \mathcal{R}_{\rm H2} &= 1.4 \times 10^{26} \unit{molecules}{s^{-1}} }
     \end{align}
@@
     \begin{equation}
-        \boxed{ N_{\rm H2, lost} = 1.6 \times 10^{42} }
+        \boxed{ N_{\rm H2, lost} = 4.5 \times 10^{42} }
     \end{equation}
-    and so this is comparable to the current hydrogen content of the Earth's atmosphere, approximately a factor of 6 lower.
+    and so this is approximately within a factor of two of the current hydrogen content of the Earth's atmosphere!
 }
@@
     \begin{equation}
-        \boxed{ \mathcal{R}_{\rm O2} = 1.6 \times 10^{-71} \unit{molecules}{s^{-1}} }
+        \boxed{ \mathcal{R}_{\rm O2} = 4.4 \times 10^{-71} \unit{molecules}{s^{-1}} }
     \end{equation}

### 2022-03-15 507_thermo/pset3/solutions.tex
move all of thermo into a folder

--- 2022-02-25 pset3/solutions.tex
@@
 
-\question{1. \textbf{FIRAS}}
+\question{1. \textbf{Maximum photon chemical potential}}
 \answer{
-    \todo{copy over stuff from jupyter notebook}
+    For this problem I used a couple of simple things. First the blackbody equation with a nonzero chemical potential
+    \begin{equation}
+        B_\nu (T) = \frac{2 h \nu^3}{c^2} \frac{1}{\exp(\frac{h \nu - \mu}{k_B T}) + 1}
+    \end{equation}
+    where we've just inserted the $\mu$ term in the exponential. We also need to calculate $\chi^2$ for this and we do this by
+    \begin{equation}
+        \chi^2 = \sum_i \frac{(\text{measured}_i - \text{model}_i)^2}{\text{uncertainty}_i^2}
+    \end{equation}
+    So then it's just a matter of reading in the FIRAS data file, converting the (honestly baffling) units into more familiar things and calculating the $\chi^2$ over a grid of temperatures and chemical potentials.
+Then the maximum allowed chemical potential with a 3-$\sigma$ uncertainty is just the maximum chemical potential within $\Delta \chi^2 \approx 6.63$ of $\chi^2_{\rm min}$ (as the reading says).
+
+    \noindent You can find the code that I used to calculate this \href{https://github.com/TomWagg/thermo_winter22/blob/tom/pset3/code/problem_1.ipynb}{in my GitHub repo} and I found that
+    \begin{equation}
+        \boxed{ \mu_{\rm max} \approx 3.55 \times 10^{-20} \unit{erg} \approx 1.1 \times 10^{-4} \, k_B T_{\rm CMB} }
+    \end{equation}
+    I also visualised this on a contour plot which you can see in the plot on the next page.
 }
@@
     \end{equation}
+    \begin{center}
+        \includegraphics[width=\textwidth]{figures/max_mu.pdf}
+    \end{center}
     For the total energy, we don't have it from the slides but we can derive it in the same way using Fermi-Dirac integrals.
@@
           &= \frac{1}{T} \qty[ \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T F_{3/2}(z) + \frac{4(2s + 1)}{3 \pi^{1/2}} V k_B T \lambda^{-3} F_{3/2}(z) + \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T \ln z F_{1/2}(z)] \\
-          &= \frac{2(2s + 1) V k_B T}{\pi^{1/2} \lambda^3} \qty[ F_{3/2}(z) + \frac{2}{3} F_{3/2}(z) + \ln z F_{1/2}(z)] \\
-        \Aboxed{ S &= \frac{2(2s + 1) V k_B T}{\pi^{1/2} \lambda^3} \qty[ \frac{5}{3} F_{3/2}(z) + F_{1/2}(z) \ln z] }
+          &= \frac{2(2s + 1) V k_B}{\pi^{1/2} \lambda^3} \qty[ F_{3/2}(z) + \frac{2}{3} F_{3/2}(z) + \ln z F_{1/2}(z)] \\
+        \Aboxed{ S &= \frac{2(2s + 1) V k_B}{\pi^{1/2} \lambda^3} \qty[ \frac{5}{3} F_{3/2}(z) + F_{1/2}(z) \ln z] }
     \end{align}
@@
     \begin{equation}
-        F_v(z) \approx \int_0^\infty \qty(\sum_{i = 1}^{\infty} e^{- i w} w^v z^i) \dd{w}
+        F_v(z) \approx \int_0^\infty \qty(\sum_{i = 1}^{\infty} (-1)^{i - 1} e^{- i w} w^v z^i) \dd{w}
     \end{equation}
@@
     \begin{equation}
-        F_v(z) \approx \sum_{i = 1}^{\infty} \qty( z_i \int_0^\infty e^{- i w} w^v \dd{w} )
+        F_v(z) \approx \sum_{i = 1}^{\infty} \qty( (-1)^{i - 1} z_i \int_0^\infty e^{- i w} w^v \dd{w} )
     \end{equation}
@@
     \begin{equation}
-        \frac{P}{n k_B T} \approx \frac{2}{3} \frac{\frac{3 \sqrt{\pi}}{4} z + \frac{3 \sqrt{\pi}}{16 \sqrt{2}} z^2}{\frac{\sqrt{\pi}}{2} z + \frac{\sqrt{\pi}}{4 \sqrt{2}} z^2}
+        \frac{P}{n k_B T} \approx \frac{2}{3} \frac{\frac{3 \sqrt{\pi}}{4} z - \frac{3 \sqrt{\pi}}{16 \sqrt{2}} z^2}{\frac{\sqrt{\pi}}{2} z - \frac{\sqrt{\pi}}{4 \sqrt{2}} z^2}
     \end{equation}
@@
 }
+
+\pagebreak
 
@@
 
+\pagebreak
+
 \question{3b. \textbf{Pressure}}
@@
 
+\pagebreak
+
 \question{3e. \textbf{Polytropes}}

### 2022-03-15 507_thermo/pset4/solutions.tex
move all of thermo into a folder

--- 2022-03-09 pset4/solutions.tex
@@
 But this expression contains the radius and we don't want that!
-Never fear, we can simply write the radius in terms of density since $\rho = 4/3 \pi R^3$ (dropping factors of order unity)
+Never fear, we can simply write the radius in terms of density since $\rho = M / (\frac{4}{3} \pi R^3)$ (dropping factors of order unity)
     \begin{equation}
@@
     \end{align}
-    Therefore we find that $\Gamma = 4/3$ in this case and so the polytrope index is $n = 3$, like a relativistic gas, which I think makes sense for radiation-dominated stars!
+    Therefore we find that $\Gamma = 4/3$ in this case and so the polytrope index is $n = 3$, like a relativistic gas, which makes sense for a radiation-dominated star!
 I \textit{think} the assumption we made that results in the pressure just being a function of density and not temperature is just that we are in hydrostatic equilibrium and that all of the pressure comes from radiation but I'm honestly not sure.
@@
 This would give us the density of the star as a function of the radius like in the plot in lecture 11 slide 20 and so we'd know the interal structure of stars above the radiation-dominated mass limit.
-    \todo{need to check with Eric what he actually wants from us on this one}
 }

### 2022-03-15 519_radiative/pset1/solutions.tex
upload radiative stuff too

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 519A; Problem Set 1 - \textbf{Tom Wagg}}
% \author{Due Oct~22 by 5pm (in McQuinn's mailbox)}

\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}\\#1\\
}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\begin{document}

\maketitle

Please use cgs units for all calculations.

{\it \noindent 1.  {\it (10/5 class)} The visible light emitted by planets/moons is star-light diffusely reflected from their surfaces.  To calculate this accurately can be complicated, but some insight can be gained by using a simple approximation know as ``Lambert's law''.  In this approximation, a fraction $a$ (the ``albedo'') of the radiation incident on a small area element of the surface is assumed to be diffusely reflected with uniform brightness in the outward hemisphere.

 a)Show that the diffusely reflected flux seen at earth due to one such surface element $dA$ on the planet is
$$ dF = \frac{L a \cos \theta_0 \cos \theta_1 dA}{4\pi^2 D^2 d^2},$$
where $D$ is the distance of the planet from the star, and $\theta_0$ is the angle of the star's rays w.r.t. the outward normal of $dA$.  Likewise $d$ is the distance of the planet from earth, and $\theta_1$ is the angle between the earth-planet direction and the outward normal of $dA$.  $L$ is the stellar luminosity.}

\answer{
    We can show this result in several steps. First, it is useful to find the flux onto on the small surface element $\dd{A}$ on the planet. This is given by
    \begin{equation}
        \dd{F_{p, {\rm in}}} = \frac{L \cos \theta_0}{4 \pi D^2},
    \end{equation}
    where the factor of $\cos \theta_0$ accounts for the fact that only a project surface area $\dd{A} \cos \theta_0$ receives the flux. Next we can write that the flux \textit{off} the small surface element is
    \begin{equation}
        \dd{F_{p, {\rm out}}} = \frac{a L \cos \theta_0}{4 \pi D^2},
    \end{equation}
    as only a fraction $a$ of the flux is reflected. Since we know that the flux at a surface of uniform brightness is $F = \pi I$, we can write that
    \begin{equation}
        \dd{I_{p, {\rm out}}} = \frac{a L \cos \theta_0}{4 \pi^2 D^2},
    \end{equation}
    where we have divided by $\pi$. Since intensity is conserved over rays, this is also the intensity at the observer. Therefore, we can convert this to the flux at the observer by multiplying by the solid angle
    \begin{align}
        \dd{F_{o}} &= \frac{a L \cos \theta_0}{4 \pi^2 D^2} \dd{\Omega},\\
        &= \frac{a L \cos \theta_0}{4 \pi^2 D^2} \frac{\dd{A_{\rm eff}}}{d^2},\\
        &= \frac{a L \cos \theta_0}{4 \pi^2 D^2} \frac{\dd{A} \cos \theta_1}{d^2},
    \end{align}
    which gives us the final desired expression:
    \begin{equation}
        \boxed{ \dd{F} = \frac{L a \cos \theta_0 \cos \theta_1 \dd{A}}{4 \pi^2 D^2 d^2} }
    \end{equation}
}

{\it b) Let's take the case for the full moon, and it's fine to evaluate in the limit that the sun and earth are very far so that $\cos \theta_0  \approx \cos \theta_1$.  It's also best to choose a coordinate system where these angles become the polar angle when integrating over the moon's surface. Integrate $dF$ over the exposed surface of the moon to calculate the total reflected flux seen on earth.  Then, use that the moon's albedo is $a=0.1$ and its radius is $R_m = 1.7\times10^8$cm to calculate the reflected flux seen on earth in cgs units.  The other values are easy to look up: $D=1.5\times10^{13}$cm,  $d=3.8\times10^{10}$cm, $L_\odot = 3.9\times10^{33}$erg/s.  How much smaller is the flux of the moon's than the sun's direct flux?   }

\answer{
    To start, we need to integrate the flux over the surface of the moon. To get the total flux we therefore write (assuming $\theta_0 = \theta_1$)
    \begin{align}
        F_m &= \int \frac{L a \cos \theta^2 \dd{A}}{4 \pi^2 D^2 d^2}, \\
        F_m &= \frac{L a}{4 \pi^2 D^2 d^2} \int_{\rm moon} \cos \theta^2 \dd{A}.
    \end{align}
    Next, we need to integrate the surface element over the moon. For this purpose, since the distances are much greater than the sizes of the sun and moon, $\theta$ also represents the azimuthal angle in spherical coordinates. We therefore need to integrate over the half of the moon that is visible to us. This can be written as
    \begin{align}
        F_m &= \frac{L a}{4 \pi^2 D^2 d^2} \int_0^{2\pi} \dd{\phi} \int_0^{\pi / 2} R_m^2 \sin \theta \cos \theta^2 \dd{\theta},\\
        &= \frac{L a R_m^2}{4 \pi^2 D^2 d^2} \cdot 2 \pi \int_0^{\pi / 2} \sin \theta \cos \theta^2 \dd{\theta},\\
        &= \frac{L a R_m^2}{4 \pi^2 D^2 d^2} \cdot \frac{2 \pi}{3},\\
        F_m &= \frac{L a}{6 \pi} \qty(\frac{R_m}{D d})^2.
    \end{align}
    At this point it is simply a matter of plugging in numbers to find the flux:
    \begin{align}
        F_m &= \frac{3.9 \times 10^{33} \cdot 0.1}{6 \pi} \qty(\frac{1.7 \times 10^8}{1.5 \times 10^{13} \cdot 3.8 \times 10^{10}})^2,\\
        \Aboxed{ F_m &= 1.8 \unit{erg}{s^{-1}}{cm^{-2}} }
    \end{align}
    Then we know that the flux from the sun is simply
    \begin{equation}
        F_{\odot} = \frac{L_\odot}{4 \pi D^2},
    \end{equation}
    where we have assumed that $D$ is also approximately the distance to the Earth. Therefore we can write that the ratio of the two is
    \begin{equation}
        \boxed{ \frac{F_m}{F_{\odot}} = \frac{2 a}{3} \qty(\frac{R_m}{d})^2 = 1.3 \times 10^{-6} }
    \end{equation}
}

{\it c) How much less intense on average is the moon than the sun using that they both span the same angular angular area on the sky? (Note that `intense' here refers to `intensity'.)  Ignoring Limb darkening which primarily affects the edge of the sun,\footnote{We will consider Limb darkening towards the end of the class when we discuss atmospheres.  This darkening owes to the sun's emission not coming from a single surface.} to good approximation the Sun is the same intensity at all points. How is the intensity profile of the moon different?}

\answer{
    The intensity is given by the flux integrated over the solid angle. However, since the angular area of the sun and the moon is equal, the ratio of the intensity is equivalent to the ratio of the flux and thus
    \begin{equation}
        \boxed{ \frac{I_m}{I_\odot} = \frac{F_m}{F_\odot} = 1.3 \times 10^{-6} }    
    \end{equation}
    The intensity profile of the moon is different because the albedo of the surface is not equal in all areas, the highlands can have $a \approx 0.3$ whilst the mares can have $a \approx 0.08$. In contrast, the sun is essentially uniform except for sunspots and of course is radiating directly rather than reflecting.\\
}

{\it 2. {\it (10/7 class)} Imagine a homogeneous disk with height $H$ and radius $R$ that is emitting thermally with a single temperature $T$.  A skewer with impact parameter $r<R$ parallel to the disk axis has optical depth $\tau_\nu$ .  The disk is observed at an angle $\theta$ w.r.t. the disk axis at a distance $d \gg R$.  Calculate the following: \\
a)  the observed $I_\nu$ for the sightline that intersects the disk center}

\answer{
    To start, we can write that the solution to the radiative transfer equation is
    \begin{equation}
        I_\nu = I_\nu(0) e^{-\tau_\nu} + \int_0^{\tau_\nu(s)} \dd{\tau^\prime} S_{\nu} (\tau^\prime) e^{-(\tau_\nu - \tau^{\prime})}.
    \end{equation}
    Since the disc is thermally emitting, we can assume that the source function if $S_\nu = B_\nu (T)$ and assume that it is constant. This allows us to instead write that the intensity is
    \begin{equation}
        I_\nu = I_\nu(0) e^{-\tau_\nu} + B_\nu(T) (1 - e^{-\tau}).
    \end{equation}
    There is no emission from behind the galaxy and so we can drop the first term to get
    \begin{equation}
        I_\nu = B_\nu(T) (1 - e^{-\tau}).
    \end{equation}
    Finally, since the sightline is inclined at an angle $\theta$, this means that the effective optical depth is larger than the optical depth when going perpendicular to the disc and this gives the solution
    \begin{equation}
        \boxed{ I_\nu = B_\nu(T) \qty(1 - e^{-\tau / \cos \theta}) }
    \end{equation}
}

{\it b)  the observed $F_\nu$ of the disk  (assume $H\ll R$ so you can ignore edge effects)}

\answer{
    The flux is the intensity integrated over the solid angle, which gives
    \begin{equation}
        F_\nu = \int I_\nu \dd{\Omega},
    \end{equation}
    which since $d \gg R$ can simply be written as
    \begin{align}
        F_\nu &= I_\nu \frac{\pi R^2}{d^2},\\
        \Aboxed{ F_\nu &= B_\nu(T) \qty(1 - e^{-\tau / \cos \theta}) \frac{\pi R^2}{d^2} }
    \end{align}
}

{\it c)  the optically thick and thin limits of $F_\nu$ from part (b).  {\it For the optically thin limit, keep the linear order term in $\tau_\nu$ -- the correct answer is not zero!}}

\answer{
    Let's explore each limit. In the optically thick limit we have $\tau \gg 1$ and therefore the exponential term ($e^{-\tau / \cos \theta}$) goes to zero and so we have
    \begin{equation}
        \boxed{ \text{optically thick: } F_\nu = B_\nu(T) \frac{\pi R^2}{d^2} }
    \end{equation}
    In the optically thin limit $\tau \ll 1$ and so we can Taylor expand the exponential term
    \begin{align}
        F_\nu &= B_\nu(T) \qty(1 - e^{-\tau / \cos \theta}) \frac{\pi R^2}{d^2} \\
        &= B_\nu(T) \qty(1 - \qty[1 - \frac{\tau}{\cos \theta}]) \frac{\pi R^2}{d^2} \\
        \Aboxed{ \text{optically thin: } F_\nu &= B_\nu(T) \frac{\tau}{\cos \theta} \frac{\pi R^2}{d^2} }
    \end{align}
}

{\it 3. {\it   (10/7 class)} A graduate student was upset about the quality of the pizza at Friday Pizza Lunch.  He proceeded to throw all the pizza slices on the ground in a manner that was more or less random so that any area on the ground had an equal probability to be covered in pizza.  Each pizza slice has area $P$, and the student threw down $N$ slices in a room of area $A$.  Assume that, once a pizza slice is on the floor, it does not affect the probability of a pizza slice landing on the same area.  You can also take $P \ll A$.  \\
a) What is the ``optical depth'' in pizza slices between the ground and someone looking down?}

\answer{
    The optical depth (assuming we have integrated over all frequencies) is defined as
    \begin{equation}
        \tau = \int \dd{s} \alpha,
    \end{equation}
    where $\alpha$ is the absorption coefficient and it can be written as
    \begin{equation}
        \alpha = n \sigma.
    \end{equation}
    In this problem the number density of pizza slices is $n = N / (A \dd{s})$ and the cross section is $\sigma = P$. Thus we can simply write that $\tau$ is
    \begin{align}
        \tau &= \frac{N}{A \dd{s}} \cdot P \cdot \dd{s},\\
        \Aboxed{ \tau &= \frac{N P}{A} }
    \end{align}
}

{\it b) Using mathematics similar to the radiative transfer equation, calculate the fraction of the ground that is still visible and not covered by a pizza slice? Please justify your approach.}

\answer{
    The ``fraction of the ground that is still visible'' could instead be thought of as the probability that ``radiation'' will pass through the medium without being absorbed. In this case the fraction is simply
    \begin{equation}
        \boxed{ \text{fraction still visible} = e^{-\tau} = \exp(-\frac{NP}{A}) }
    \end{equation}
}

{\it c) What fraction of the ground is covered by exactly two pizza slices? {\it (Hint: First derive the part b result using the Poisson probability distribution $P(\tau | \lambda) = \lambda^\tau \exp[-\lambda]/\tau!$, where $\lambda = P N/A$ is the average number of slices at every point.)}}

\answer{
    For clarity, I will instead write the Poisson probability distribution as $P(n | \tau) = \tau^n \exp(-\tau) / n!$. In this case, we want to find the probability that a light ray would hit exactly two pizza slices given that the average number of slices at each point is $\tau$. This probability is simply
    \begin{equation}
        \boxed{ P(2 | \tau) = \frac{1}{2} \tau^2 \exp(-\tau),\qquad \text{where } \tau = \frac{NP}{A} }
    \end{equation}
}

{\it 4.)   {\it (10/12 class)} Imagine a spherical dust-free HII region in which the ionized gas has electron/proton density $n=10^4 \;$cm$^{-3}$ out to a radius $1$ pc, where the HII bubble ends.  This HII region is at a distance of $1\,$kpc from the earth.

\noindent a) What is the flux of Ly$\alpha$ photons from the region [photons cm$^{-2}$ s$^{-1}$]?   You may use that the recombination rate coefficient of hydrogen is $\alpha_B= 4\times10^{-13}$cm$^3$~s$^{-1}$ such that the rate density of recombinations is $\alpha_b n^2$ and that every recombination results in $0.66$ of a Ly$\alpha$ photon. Note that in the absence of dust, Ly$\alpha$ photons can be scattered but they cannot be destroyed.}

\answer{
    The flux can be written as
    \begin{equation}
        F = \frac{L}{4 \pi d^2},
    \end{equation}
    where $L$ is the luminosity and $d$ is the distance. We can find the luminosity by multiplying the rate density of the HII bubble by its volume and the number of photons produced per recombination
    \begin{equation}
        L = \alpha_B n^2 \cdot \frac{4 \pi}{3} r^3 \cdot \frac{2}{3},
    \end{equation}
    where I've assumed that $0.66 = 2/3$ is what you mean in the question. Overall, we can therefore write that the flux is
    \begin{equation}
        F = \frac{2}{9} \frac{\alpha_B n^2 r^3}{d^2},
    \end{equation}
    where we have the values of each of these quantities and thus we find that the flux is
    \begin{equation}
        \boxed{ F = 2.74 \times 10^7 \unit{photons}{cm^{-2}}{s^{-1}} }
    \end{equation}
}

{\it \noindent b) The gas in the HII region is at temperature $T=10^4$K.  In the optical through radio, thermal free-free emission dominates the emission from the HII region (AKA bremsstrahlung emission).  This is emission from free electrons scattering off of free ions.  We will discuss free free emission later in the class, but it suffices to note that it's specific emissivity from gas is $\epsilon_\nu \approx 7\times10^{38} n^2 T^{-1/2} \exp[h\nu/kT]$ in cgs units, ignoring an order unity `Gaunt factor' (eqn 5.14 in RL).  Remember that this emission is thermal so $B_\nu(T) = \epsilon_\nu/(4\pi \alpha_\nu)$.

What is the brightness temperature of the free-free emission through the cloud center?  You can assume that we are observing $h \nu \ll k T$ such that the Rayleigh-Jeans law applies (but that the emission is still optically thin), and that the sightline does not intersect a dense star.}

\answer{
    Let's first apply RL Eq.~1.62 (a solution to the RT equation with temperature brightness) which gives
    \begin{equation}
        T_b = T_b(0) e^{-\tau_\nu} + T (1 - e^{-\tau_\nu}).
    \end{equation}
    In the case of this problem we are assuming there is nothing behind the gas and thus $T_b(0) = 0$ meaning that
    \begin{equation}
        T_b = T (1 - e^{-\tau_\nu}).
    \end{equation}
    Since we are given $T$, the last unknown is therefore $\tau_\nu$. We know that the optical depth can be found as
    \begin{equation}
        \tau_\nu = \int \dd{s} \alpha_\nu,
    \end{equation}
    and since we assume that the gas is uniformly dense then this integral is trivial and so
    \begin{equation}
        \tau_\nu = 2 r \alpha_\nu,
    \end{equation}
    where $r$ is the radius of the bubble. In order to find the absorption coefficient, $\alpha_\nu$, we can apply the fact that the emission is thermal and thus $B_\nu(T) = \epsilon_\nu / (4 \pi \alpha_\nu)$ and that $B_\nu(T)$ is a blackbody (in Rayleigh-Jeans limit) and so solve for $\alpha_\nu$
    \begin{align}
        B_\nu = \frac{2 \nu^2}{c^2} k_B T &= \frac{\epsilon_\nu}{4 \pi \alpha_\nu} \\
        \alpha_\nu &= \frac{\epsilon_\nu}{4 \pi} \frac{c^2}{2 \nu^2} \frac{1}{k_B T}
    \end{align}
    We can then plug \textit{alllll} of this in to get that the brightness temperature is
    \begin{align}
        T_b &= T \qty(1 - \exp[- 2 r \alpha_\nu]) \\
        T_b &= T \qty(1 - \exp[- 2 r \frac{\epsilon_\nu}{4 \pi} \frac{c^2}{2 \nu^2} \frac{1}{k_B T}]) \\
        T_b &= T \qty(1 - \exp[- (1 \unit{pc}) \frac{7 \times 10^{-38} \unit{cm^3}{erg}{K^{1/2}} (10^4 \unit{cm^{-3}})^2 (10^4 \unit{K})^{-1/2})}{4 \pi k_B (10^4 \unit{K})} \frac{c^2}{\nu^2}]) \\
        \Aboxed{ T_b &= 10^4 \qty(1 - e^{-1.1 \times 10^{19} (\nu / \unit{Hz})^{-2}}) \unit{K} }
    \end{align}
}

{\it \noindent c) How does the brightness temperature change if the HII region were at ten times the distance?}

\answer{
    Since the brightness temperature is analogous to an intensity it is not affected by distance. \textbf{Therefore the brightness temperature would stay the same.}
}

{\it \noindent d) Below some frequency the emission is not optically thin.  What frequency?  What is the brightness temperature well below this critical frequency?}

\answer{
    The switch between optically thin and thick occurs at $\tau_\nu = 1$. Therefore since the quantity in the exponential is $\tau_\nu$ we can write that
    \begin{align}
        (\nu / \unit{Hz})^2 &= 1.1 \times 10^{19} \\
        \Aboxed{ \nu &= 3.31 \unit{GHz} }
    \end{align}
    At this frequency, the brightness temperature is equal to
    \begin{align}
        T_b &= T(1 - e^{-1}) \\
        \Aboxed{ T_b &= 6300 \unit{K} }
    \end{align}
}

\end{document}

### 2022-03-15 519_radiative/pset2/solutions.tex
upload radiative stuff too

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 519A; Problem Set 2}
\author{\textbf{Tom Wagg}}

\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}\\#1\\
}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\allowdisplaybreaks

\begin{document}
\maketitle

\noindent {\it 1) [10/19 class, beginning] {\bf Sobolev approximation and HI 21cm from the galactic disk.}

a) Remember the optical depth we see in terms of the Einstein B-coefficients is given by 
$$
\tau_\nu = \int ds \frac{h \nu_0}{4\pi} n_1 B_{12} \left(1 - \frac{g_1 n_2}{ g_2 n_1} \right) \phi(\nu -\nu_{\rm line-center})  
$$
where the densities can depend on $s$ and I've centered $\phi(\Delta \nu)$ so that it peaks at $\Delta \nu=0$ zero, which is different than in class. Doppler shift results in the line center redshifting with respect to us at $s=0$ so that
$$
\nu_{\rm line-center} = \nu_0 (1+ \frac{dv}{ c ds} s),
$$
where $\nu_0$ is the frequency of the transition.   Use this setup and that the densities vary slowly compared to $\phi$ with $s$ at fixed $\nu$ (i.e. assume they are constant) to derive the Sobolev approximation for the optical depth 
$$
\tau_\nu = \frac{h c }{4\pi dv/ds} n_1 B_{12} \left(1 - \frac{g_1 n_2}{ g_2 n_1} \right).,
$$
where the density here is the one at $s$ such that $\nu = \nu_0 (1+ \frac{dv}{ c ds} s)$.\\

{\it Hint: It may be helpful to stick in a Gaussian for $\phi$ or even a Dirac delta function to derive this noting that $\int ds \delta^D( y s) =  |y|^{-1}$.}\\}

\answer{
    As an initial step, we can move the quantities that are constant across $\dd{s}$ out of the integral
    \begin{equation}\label{eq:q1_init}
        \tau_\nu = \frac{h \nu_0}{4 \pi} n_1 B_{12} \qty(1 - \frac{g_1 n_2}{g_2 n_1}) \int \dd{s} \phi(\nu - \nu_{\rm line-center}).
    \end{equation}
    Now let's focus on that integral (which I'll refer to as $X$ for brevity) and substitute in the expression that we are given for $\nu_{\rm line-center}$. We can also follow the hint and assume that the linewidth function is a Dirac delta function.
    \begin{equation}
        X \equiv \int \dd{s} \phi\qty[\nu - \nu_0 \qty(1 + \frac{1}{c} \dv{v}{s} s)] = \int \dd{s} \delta^D\qty[\nu - \nu_0 \qty(1 + \frac{1}{c} \dv{v}{s} s)]
    \end{equation}
    Since the integral of a Dirac delta function will be the same independent of how far it is shifted, we can ignore any terms that don't involve $s$ within the delta function. This allows us to then apply the hint!
    \begin{align}
        X &= \int \dd{s} \delta^D\qty[\frac{\nu_0}{c} \dv{v}{s} s] \\
        &= \abs{ \frac{\nu_0}{c} \dv{v}{s}}^{-1}\\
        X &= \frac{c}{\nu_0} \frac{1}{\dd{v}/\dd{s}}
    \end{align}
    Now that we've got an expression for the integral we can return to Eq.~\ref{eq:q1_init} and substitute the integral with $X$ in order to obtain the Sobolev approximation
    \begin{align}
        \tau_\nu &= \frac{h \nu_0}{4 \pi} n_1 B_{12} \qty(1 - \frac{g_1 n_2}{g_2 n_1}) \cdot X \\
        &= \frac{h \nu_0}{4 \pi} n_1 B_{12} \qty(1 - \frac{g_1 n_2}{g_2 n_1}) \cdot \frac{c}{\nu_0} \frac{1}{\dd{v}/\dd{s}} \\
        \Aboxed{ \tau_\nu&= \frac{h c}{4 \pi \dd{v}/\dd{s}} n_1 B_{12} \qty(1 - \frac{g_1 n_2}{g_2 n_1}) }
    \end{align}
}

{\it b.)  Let's apply this formula to calculate 21cm emission from neutral hydrogen in the Galactic Disk. Assume the temperature $T_S$ of 21cm transition is $100$K, which sets the ratio in the upper and lower levels $g_1 n_2/ g_2 n_1 = \exp[- h \nu_0/ k_bT_S]$. Using your favorite equation $T_b = T_S \left(1 -\exp(-\tau_\nu) \right)$ and this plot, what would you estimate for the optical depth of the Milky Way to this transition at the frequencies that are most absorbed?}

\answer{
    Let's rearrange the given equation to solve for $\tau_\nu$
    \begin{align}
        T_b &= T_S \qty(1 -\exp(-\tau_\nu)) \\
        \frac{T_b}{T_S} &= 1 -\exp(-\tau_\nu) \\
        \exp(-\tau_\nu) &= 1 - \frac{T_b}{T_S} \\
        -\tau_\nu &= \ln \qty(1 - \frac{T_b}{T_S}) \\
        \tau_\nu &= - \ln \qty(1 - \frac{T_b}{T_S})
    \end{align}
    And so we arrive at an expression for the optical depth. We are given that $T_S = 100 \unit{K}$ and we are also told that the linewidth function is centred on the peak optical depth and thus from inspection of the plot we see that $T_b = 70 \unit{K}$ at this peak. This gives the optical depth
    \begin{equation}
        \boxed{ \tau_\nu = -\ln(1 - 70/100) \approx 1.2 }
    \end{equation}
}

\newpage

{\it c.)  Let's now try to estimate the typical optical depth for a sightline through the Milky Way disk. First, $\nu_0 = 1.4$GHz and $A_{21} = 2.85\times10^{-15}$s$^{-1}$ or equivalently the f-value is $5.85\times10^{-21}$.  It takes $A_{21}^{-1} = 10$Myr for hydrogen to transition spontaneously from the excited hyperfine state to the ground state.  Also, for the Milky Way disk, roughly a sightline has $dv/ds = 10$ km/s /kpc extending over $\sim 10$kpc (or spanning a velocity of 100 km/s), and let's assume an average density of hydrogen is $n=1$cm$^{-3}$ characteristic of the average ISM.

To simplify our expression for the Sobolev optical depth, take the limit that $h \nu_0/ k_bT_S \ll 1$ as $h \nu_0/k_b = 0.07\,$K. Plugging in all the factors, what optical depth do you find? Is it of the same order of magnitude with the one you inferred from the plot?}

\answer{
    First, it is convenient to convert $A_{21}$ to $B_{12}$ before continuing
    \begin{align*}
        A_{21} = \frac{2 h \nu^3}{c^2} B_{21} &= \frac{2 h \nu^3}{c^2} \frac{g_1}{g_2} B_{12} \tag{RL 1.72ab} \\
        B_{12} &= \frac{c^2}{2 h \nu^3} \frac{g_2}{g_1} A_{21} \\
               &= \frac{3 c^2}{2 h \nu^3} A_{21} \tag{$g_2 = 3, g_1 = 1$} \\
        B_{12} &= 2 \times 10^5 \unit{s}{g^{-1}}
    \end{align*}
    Then we can go through and plug in all of the factors to find the optical depth
    \begin{align*}
        \tau_\nu &= \frac{h c}{4 \pi \dd{v}/\dd{s}} n B_{12} \qty(1 - \frac{g_1 n_2}{g_2 n_1}) \\
        &= \frac{h c}{4 \pi \dd{v}/\dd{s}} \frac{n_1}{4} B_{12} \qty(1 - \exp(- h \nu_0 / k_b T_S)) \tag{assume fixed $T_S$} \\
        &= \frac{h c}{4 \pi \dd{v}/\dd{s}} \frac{n_1}{4} B_{12} \qty(1 - (1 - h \nu_0 / k_b T_S)) \tag{$h \nu_0 / k_b T_S \ll 1$} \\
        &= \frac{h c}{4 \pi \dd{v}/\dd{s}} \frac{n_1}{4} B_{12} \frac{0.07 \unit{K}}{T_S} \tag{$h \nu_0 / k_b = 0.07 \unit{K}$} \\
        &= \frac{(2 \times 10^{-16} \unit{erg}{cm})}{4 \pi (10 \unit{km}{s^{-1}}{kpc^{-1}})} \frac{(1 \unit{cm^{-3}})}{4} (2 \times 10^5)  \frac{0.07 \unit{K}}{100 \unit{K}} \\
        \Aboxed{ \tau_\nu &= 1.7 }
    \end{align*}
    This is the same order of magnitude as the optical depth that I inferred from the plot!
}

{\it 2) [10/21 class] {\bf energy densities}
a) the cosmic microwave background, which is a spatially uniform black body with $T=2.7$K. Outside of galaxies, this energy density likely dominates over all of the below energy densities.}

\answer{
    We can find the energy density by applying the Stefan Boltzmann law:
    \begin{align}
        u = a T^4 &= \frac{4 \sigma T^4}{c} \\
        &= \frac{4 \cdot 5.67 \times 10^{-8} \cdot 2.7^4}{3 \times 10^8} \unit{J}{m^{-3}} \\
        &= \frac{4 \cdot 5.67 \times 10^{-8} \cdot 2.7^4}{3 \times 10^8} \frac{10^{-6}}{1.6 \times 10^{-19}} \unit{eV}{cm^{-3}} \\
        \Aboxed{ u &= 0.25 \unit{eV}{cm^{-3}} }
    \end{align}
}

{\it b) the Milky Way magnetic field (assuming $B \approx 10\mu$ G)}

\answer{
    \begin{align}
        u &= \frac{B_0^2}{8 \pi} \\
        &= \frac{10^{-10}}{8 \pi} \unit{G^2} \\
        &= \frac{10^{-10}}{8 \pi} \unit{cm^{-1}}{g}{s^{-2}} = \frac{10^{-10}}{8 \pi} \unit{erg}{cm^{-3}} \\
        &= \frac{10^{-17}}{8 \pi} \unit{J}{cm^{-3}} = \frac{10^{-17}}{8 \pi \cdot 1.6 \times 10^{-19}} \unit{eV}{cm^{-3}} \\
        \Aboxed{ u &= 2.5 \unit{eV}{cm^{-3}} }
    \end{align}
}

{\it c)  starlight in the Milky Way assuming a luminosity of $10^{10}L_\odot$. To simplify, assume a uniform disk of radius $R=10$kpc and height $H=100$pc.  A further simplification is that the $\Delta \Omega \approx 2\pi \times H/R$ of the sky you see through a distance of $R$ of the disk. The average distance through the disk over full sky is $\sim H$ times an order-one prefactor that we don't care about since this is a rough estimate. Use this to estimate the average intensity and then the energy density.}

\answer{
    We worked out in an earlier quiz that the intensity when looking through a disc of this nature is
    \begin{equation}
        I = \frac{n L H}{4 \pi},
    \end{equation}
    where we are still using $H$ since the hint in the question explains that the average distance through the disc is $H$. Then we know that the number density times a luminosity is the \textit{total} luminosity over the whole volume, meaning
    \begin{equation}
        n L = \frac{L_{\rm tot}}{\pi R^2 H},
    \end{equation}
    where $L_{\rm tot} = 10^{10} L_{\odot}$. We can then write that the intensity is
    \begin{equation}
        I = \frac{L_{\rm tot}}{4 \pi^2 R^2}.
    \end{equation}
    Next we want to find the mean intensity, which using RL Eq.~1.8 is
    \begin{equation}
        J = \frac{1}{4 \pi } \int I \dd{\Omega} = \frac{1}{4\pi} I \cdot 4 \pi = I
    \end{equation}
    Then to find the energy density we simply apply RL Eq.~1.9
    \begin{align}
        u &= \frac{4 \pi}{c} J \\
        &= \frac{4 \pi}{c} \frac{L_{\rm tot}}{4 \pi^2 R^2} \\
        &= \frac{L_{\rm tot}}{\pi R^2c } \\
        \Aboxed{ u &= 0.27 \unit{eV}{cm^{-3}}}
    \end{align}
}

{\it d) $10^4$K warm gas in the ISM at a characteristic number density of $n=1$cm$^{-3}$  {\it (okay, this one wasn't covered in class, but you should know the formula!)}.}

\answer{
    \begin{align}
        u &= E \cdot n \\
        u &= \frac{3}{2} k_B T \cdot n \\
        \Aboxed{ u &= 1.3 \unit{eV}{cm^{-3}} }
    \end{align}
}

{\it 3). [10/26 class]   { {\bf Solar corona and Thomson scattering} (but this problem also uses tools from Ch 1)}\\
\noindent    The corona of the Sun, as seen during a total solar eclipse, consists of very hot, ionized gas, which scatters light from the Sun by means of Thomson scattering ($\sigma_T = 6.65\times10^{-25}$cm$^{2}$).  The annotated photo below is of a total solar eclipse.  The black disk is the Moon just blocking out the Sun.  
\noindent a) For the light from point P, what would the direction of linear polarization relative to the dashed line be? Would you expect to see $100\%$ polarization?  Please explain, but a quantitative answer is not required. }

\answer{
    Due to symmetry, most of the incident radiation will cancel out and the only remaining linear polarisation will be perpendicular to the dashed line. I would expect it to be 100\% polarised from Eq.~3.41 in RL, since $\Pi(\pi / 2) = 1$.
}

{\it \noindent b) The surface brightness of the corona at a few solar radii is roughly $\sim 10^{-9}$ of the surface brightness of the Sun itself, sufficiently dim that it will not blind you.  Assume that Thomson scattering is isotropic and assume the density profile of electrons beyond the solar radius of $R_\odot =7\times10^{10}$cm falls off as $r^{-2}$.   Estimate the electron density in the solar corona at point P.}

\answer{
    The aim of the problem is to find $n_e(r_p)$, the electron density at the radius $r_p$ (where $r_p$ is the distance from the centre of the Sun to the point P). To start, let's write out the radiative transfer equation for pure scattering (RL Eq.~1.86)
    \begin{equation}
        \dv{I}{s} = - \sigma (I - J(r)),
    \end{equation}
    where we can additionally remove the $I$ term as this deals with absorption which is negligible for Thomson scatter (except for huge column densities), as Matt told us in class. Therefore we can simply write that
    \begin{equation}
        \dv{I}{s} = \sigma J(r).
    \end{equation}
    Next we know that in this case, the mean intensity at a radius $r$ is just a function of the emission coefficient from RL Eq.~1.84
    \begin{equation}
        J(r) = \frac{j(r)}{\sigma}.
    \end{equation}
    Now we can apply an earlier equation from RL (Eq.~1.16) to relate the emission coefficient to the power per volume.
    \begin{equation}
        j(r) = \frac{P(r) n_e(r)}{4 \pi},
    \end{equation}
    where $P(r)$ is the power and $n_e(r) = A r^{-2}$ is the electron density. We also know that $P(r) = F(r) \sigma$ and that $F$ is
    \begin{equation}
        F(r) = \pi B_{\odot} \qty(\frac{R_\odot}{r})^2,
    \end{equation}
    there we can combine all of these and write that
    \begin{align}
        \dv{I}{s} &= \sigma J(r) = j(r) \\
                  &= \frac{B_\odot \sigma R_\odot^2}{4 r^2} \cdot \frac{A}{r^2}.
    \end{align}
    We can also note that $r = (r_p^2 + s^2)^{1/2}$ and now perform the integral to find the intensity at the point $P$
    \begin{align}
        I_P &= \frac{A \sigma B_{\odot} R_\odot^2}{4} \int \dd{s} r^{-4}, \\
            &= \frac{A \sigma B_{\odot} R_\odot^2}{4} \int_0^{\infty} \dd{s} (r_p^2 + s^2)^{-2}, \\
            &= \frac{A \pi \sigma B_{\odot} R_\odot^2}{16 r_p^3}
    \end{align}
    Then finally we are given that $I_r = 10^{-9} B_\odot$ and so we can solve for the value of $A$
    \begin{align}
       I_P = 10^{-9} B_\odot &= \frac{A \sigma B_{\odot} R_\odot^2}{4} \cdot \frac{\pi}{4 r_p^3} \\
       A &= \frac{16 \times 10^{-9} \cdot r_p^3}{\pi \sigma R_\odot^2}
    \end{align}
    This gives that the electron density at the point P is (assuming $r_p \approx 2 R_\odot$)
    \begin{align}
        n_e(r_p) &= \frac{16 \times 10^{-9} \cdot r_p^3}{\pi \sigma  R_\odot^2} \cdot \frac{1}{r_p^2} \\
        &= \frac{16 \times 10^{-9} \cdot r_p}{\pi \sigma R_\odot^2} \\
        n_e(r_p) &= \frac{16 \times 10^{-9} \cdot r_p}{\pi \sigma R_\odot^2} \\
        \Aboxed{ n_e(r_p) &= 2.2 \times 10^5 \unit{cm^{-3}} }
    \end{align}
}

{\it \noindent c) Once the eclipse is no longer total, part of the very intense solar photosphere is seen.  On your retina, why are the cells that have the emerging photosphere projected onto them just as damaged as they would be if you were looking at a totally unobstructed sun?}

\answer{
    Intensity is conserved and so the intensity on any cell on the retina that is exposed to the sun will be the same as the intensity on the sun. Therefore, even though every cell on the retina may not have light on it, those cells that \textit{are} exposed will be just as damaged.
}

{\it 4) [10/28 class]  {\bf scattering from a classical bound electron:} This problem follows the end of Ch.~3 in Rybicki and Lightman.

In the case of scattered radiation from a free electron, we considered
the acceleration of the electron due to an incident (plane polarized)
electromagnetic wave:
\begin{equation}
\dot v = \ddot r = \frac{-e E_{\rm inc}}{m_e} ~~~~~{\rm where}~~~~~ E_{\rm inc} = E_0 \sin \omega t.
\end{equation}

In the case of the harmonically bound electron (i.e. sitting in a potential well with $U = m_e \omega_e^2 r^2/2$), the analogous
equation of motion is that of a forced oscillator
\begin{equation}
\ddot r + \omega_e^2 r = \frac{-e E_{\rm inc}}{m_e} .
\end{equation}

\noindent (a) Verify that a solution of the forced oscillator equation
of motion is
\begin{equation}
    r = -\frac{e E_{\rm inc}}{m_e} \frac{1}{\omega_e^2 - \omega^2}.
\end{equation}
(This is not the unique solution and note that the only part that depends on time is $E_{\rm inc}$.)}

\answer{
    First we can differentiate $r$ a bunch of times
    \begin{align}
        r &= -\frac{e E_0}{m_e} \frac{1}{\omega_e^2 - \omega^2} \sin \omega t \\
        \dot{r} &= - \omega \cdot \frac{e E_0}{m_e} \frac{1}{\omega_e^2 - \omega^2} \cos \omega t \\
        \ddot{r} &= \omega^2 \cdot - \frac{e E_0}{m_e} \frac{1}{\omega_e^2 - \omega^2} \sin \omega t \\
        &= - \omega^2 r
    \end{align}
    Now we plug it into the LHS and show that it gives the RHS
    \begin{align}
        \ddot{r} + \omega_e^2 r &= -\omega^2 r + \omega_e^2 r \\
        &= r (\omega_e^2 - \omega^2) \\
        &= \frac{-e E_{\rm inc}}{m_e} \frac{1}{\omega_e^2 - \omega^2} (\omega_e^2 - \omega^2) \\
        &= \frac{-e E_{\rm inc}}{m_e}
    \end{align}
}

{\it \noindent 
(b) Show that the total cross section for scattered radiation
in the forced oscillator case is 
\begin{equation}
\sigma = \sigma_T \left(1-\frac{\omega_e^2}{\omega^2} \right)^{-2}.
\end{equation}}

\answer{
    Here's like 20 lines of algebra to show this is true! :D
    \begin{align*}
        \dv{P}{\Omega} &= \frac{q^2 \ddot{x}^2}{4 \pi c^3} \sin^2 \theta \tag{RL 3.23} \\
        \ddot{x}^2 &= \qty(- \omega_e^2 r - \frac{e E_0 \sin \omega t}{m_e})^2 \tag{square expression from question} \\
        \ddot{x}^2 &= \omega_e^4 r^2 + \frac{e^2 E_0^2 \sin^2 (\omega t)}{m_e^2} - 2 \frac{e E_0 \sin (\omega t) \omega_e^2 r}{m_e} \\
        \dv{P}{\Omega} &= \frac{e^2}{4 \pi c^3} \sin^2 \theta \qty( \omega_e^4 r^2 + \frac{e^2 E_0^2 \sin^2 (\omega t)}{m_e^2} - 2 \frac{e E_0 \sin (\omega t) \omega_e^2 r}{m_e} ) \tag{substitute} \\
        \avg{\dv{P}{\Omega}} &= \frac{e^2}{4 \pi c^3} \sin^2 \theta \qty( \omega_e^4 r^2 + \frac{e^2 E_0^2}{2 m_e^2} ) \tag{average over time} \\
        \avg{S} \dv{\sigma_T}{\Omega} &= \avg{\dv{P}{\Omega}} \tag{RL 3.36} \\
        \dv{\sigma}{\Omega} &= \frac{2 e^2}{E_0^2 c^4} \sin^2 \theta \qty( \omega_e^4 r^2 + \frac{e^2 E_0^2}{2 m_e^2} ) \tag{apply $\avg{S} = c E_0^2 / (8 \pi)$} \\
        &= \sin^2 \theta \qty( \frac{2 e^2 \omega_e^4 r^2}{E_0^2 c^4} + \frac{e^4}{m_e^2 c^4} ) \\
        &= \sin^2 \theta \qty( \frac{2 e^2 \omega_e^4 r^2}{E_0^2 c^4} + r_0^2 ) \tag{RL 3.38} \\
        &= \sin^2 \theta \qty[ \frac{2 e^2 \omega_e^4}{E_0^2 c^4} \qty(\frac{e^2 E_0^2}{2 m_e^2} \frac{1}{(\omega_e^2 - \omega^2)^2}) + r_0^2 ] \tag{definition of $r$ from part a} \\
        &= \sin^2 \theta \qty[ r_0^2 \qty(1 + \frac{\omega_e^4}{(\omega_e^2 - \omega^2)^2}) ] \\
        &= \sin^2 \theta r_0^2 \qty(1 - \frac{\omega_e^2}{\omega^2})^{-2} \tag{reformulate expression} \\
        \sigma &= \int \dd{\Omega} \sin^2 \theta r_0^2 \qty(1 - \frac{\omega_e^2}{\omega^2})^{-2} \\
        \sigma &= \frac{8 \pi}{3} r_0^2 \qty(1 - \frac{\omega_e^2}{\omega^2})^{-2} \\
        \Aboxed{ \sigma &= \sigma_T \qty(1 - \frac{\omega_e^2}{\omega^2})^{-2} }
    \end{align*}
}

\newpage

{\it \noindent (c) Evaluate the limit of $\sigma$ for when $\omega \gg \omega_{e}$. Justify why your result makes intuitive sense. Also evaluate the limit in the alternate limit, when $\omega_{e}\gg\omega$.}

\answer{
    In the limit of $\omega \gg \omega_e$, the cross section is simply the Thomson cross section 
    \begin{equation}
        \boxed{ \sigma = \sigma_T }.
    \end{equation}
    This makes intuitive sense since, if $\omega$ is much larger than $\omega_e$, the electron is barely oscillating within the potential well and so the wave has the same effect as if it wasn't in a well at all.

    In the case of $\omega \ll \omega_e$, we find the following expression
    \begin{align}
        \sigma &= \sigma_T \qty(1 - \frac{\omega_e^2}{\omega^2})^{-2} \\
        \sigma &\approx \sigma_T \qty(-\frac{\omega_e^2}{\omega^2})^{-2} \\
        \Aboxed{ \sigma &\approx \sigma_T \qty(\frac{\omega}{\omega_e})^4 }
    \end{align}
}

{\it \noindent
(d) Write an expression for $F_{\rm rad}$ using this {\it initial} harmonic assumption in terms of $m_e$, $\Gamma\equiv 2 e^2 \omega^2/[3 m_e c^3]$ and \emph{one power of} $\dot r$.
\begin{equation}
F_{\rm rad} = \frac{2e^2}{3 c^3} \frac{d^3 r}{dt^3},
\end{equation}
\begin{equation}
r \propto \cos (\omega t).  {\rm ~~~~~~~~(zeroth ~order~ guess)}
\end{equation}}

\answer{
    First, we can differentiate the expression for $r$ a bunch of times.
    \begin{align*}
        r &\propto \cos (\omega t) \\
        \dot{r} &\propto - \omega \sin (\omega t) \\
        \ddot{r} &\propto - \omega^2 \cos (\omega t) \\
        \dddot{r} &\propto \omega^3 \sin (\omega t) = - \omega^2 \dot{r}
    \end{align*}
    Now we can plug that into the given expression for $F_{\rm rad}$ to find the solution.
    \begin{align*}
        F_{\rm rad} &= \frac{2e^2}{3 c^3} \omega^3 \sin (\omega t) \\
        F_{\rm rad} &= - \frac{2e^2 \omega^2}{3 c^3} \omega \sin (\omega t) \\
        \Aboxed{ F_{\rm rad} &= - \Gamma m_e \dot{r} }
    \end{align*}
}

\newpage

{\it \noindent (e) Imagine now that we have an incident (plane-polarized)
electromagnetic wave of frequency $\omega$, scattering off a ``harmonically
bound" electron, but we now also account
for the damping of the radiation reaction force in the equation of motion
for the electron. This leads to an even more general equation of motion,
that of a forced harmonic oscillator {\it with damping}:
\begin{equation}
\ddot r + \Gamma \dot r + \omega_e^2 r = \frac{-eE_{\rm inc}}{m_e}, {\rm ~~~~~~~~~~ where~~~~} E_{\rm inc} = E_0 e^{-i\omega t}.
\end{equation}

Verifiy that a solution of the above forced oscillator 
with damping equation is
\begin{equation}
r = \frac{-eE_{\rm inc}}{m_e} \left [ \frac{1}{(\omega_e^2 - \omega^2) - i \omega \Gamma} \right].
\end{equation}}

\answer{
    Let's show that the left hand side is the same as the right hand side for this $r$
    \begin{align}
        \dot{r} = - i \omega r \\
        \ddot{r} = - \omega^2 r \\
        \ddot{r} + \Gamma \dot{r} + \omega_e^2 r &= - \omega^2 r - i \omega \Gamma + \omega_e^2 r \\
        &= r (\omega_e^2 - \omega^2 - i \omega \Gamma) \\
        &= \frac{-eE_{\rm inc}}{m_e} \left [ \frac{1}{(\omega_e^2 - \omega^2) - i \omega \Gamma} \right] (\omega_e^2 - \omega^2 - i \omega \Gamma) \\
        &= \frac{-eE_{\rm inc}}{m_e}
    \end{align}
}

{\it \noindent (f) Show that the total cross section for scattered radiation in this case is 
\begin{equation}
\sigma = \frac{4\pi e^2}{m_e c} \left [ \frac{\Gamma \omega^2}{(\omega_e^2-\omega^2)^2 + \omega^2 \Gamma^2} \right].
\end{equation}}

Basically the same as part b but with a different expression:
\begin{align*}
        \dv{P}{\Omega} &= \frac{e^2 \abs{\ddot{r}}^2}{4 \pi c^3} \sin^2 \theta \tag{hint} \\
        \abs{\ddot{r}}^2 &= \abs{-\omega^2 \frac{-eE_{\rm inc}}{m_e} \qty[ \frac{1}{(\omega_e^2 - \omega^2) - i \omega \Gamma}]}^2 \tag{square the expression} \\
        &= \omega^4 \frac{e^2 E_{\rm inc}^2}{m_e^2} \qty[ \frac{1}{(\omega_e^2 - \omega^2) - i \omega \Gamma}] \qty[ \frac{1}{(\omega_e^2 - \omega^2) + i \omega \Gamma}] \\
        &= \frac{e^2 E_{\rm inc}^2 \omega^4}{m_e^2} \qty[ \frac{1}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \\
        \dv{P}{\Omega} &= \frac{e^2}{4 \pi c^3} \sin^2 \theta \frac{e^2 E_{\rm inc}^2 \omega^4}{m_e^2} \qty[ \frac{1}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \tag{substitute} \\
        &= \frac{e^2 E_{\rm inc}^2}{4 \pi m_e} \sin^2 \theta \frac{e^2 \omega^2}{m_e c^3} \omega^2 \qty[ \frac{1}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \tag{collect terms} \\
        &= \frac{3 e^2 E_{\rm inc}^2}{8 \pi m_e} \sin^2 \theta \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \tag{convert to $\Gamma$} \\
        \avg{\dv{P}{\Omega}} &= \frac{3 e^2 E_0^2}{16 \pi m_e} \sin^2 \theta \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \tag{$\avg{E_{\rm inc}^2} = E_0^2$} \\
        \avg{S} \dv{\sigma}{\Omega} &= \avg{\dv{P}{\Omega}} \tag{RL 3.36} \\
        \dv{\sigma}{\Omega} &= \frac{3 e^2}{2 m_e c} \sin^2 \theta \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \tag{apply $\avg{S} = c E_0^2 / (8 \pi)$} \\
        \sigma &= \int \dd{\Omega} \frac{3 e^2}{2 m_e c} \sin^2 \theta \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \\
        \sigma &= \frac{8 \pi}{3} \frac{3 e^2}{2 m_e c} \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \\
        \Aboxed{ \sigma &= \frac{4 \pi e^2}{m_e c} \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] }
\end{align*}

{\it \noindent (g) Now consider the case when the incident radiation has frequency
with $\omega\sim\omega_{e}$ so that $\omega_e^2 -\omega^2 =  (\omega+\omega_e) (\omega-\omega_e)  \approx 2 \omega_e  (\omega-\omega_e)$. Show that your result in this case leads to a cross-section of the form
\begin{equation}
\sigma \approx \left( \frac{\pi e^2}{m_e c}\right) {\cal L},
\end{equation}
\noindent
where ${\cal L}$ is known as the Lorentz profile and it integrates to one ($\int d\nu {\cal L} = 1$) :
\begin{equation}
 {\cal L} = \frac{1}{\pi} \left[ \frac{\Gamma/4\pi}{(\nu - \nu_e)^2 + (\Gamma/4\pi)^2} \right].
 \end{equation}}

 \answer{
    \begin{align*}
        \sigma &= \frac{4 \pi e^2}{m_e c} \qty[ \frac{\Gamma \omega^2}{(\omega_e^2 - \omega^2)^2 + \omega^2 \Gamma^2}] \tag{result from part f} \\
        &= \frac{4 \pi e^2}{m_e c} \qty[ \frac{\Gamma \omega_e^2}{(2 \omega_e (\omega - \omega_e))^2 + \omega_e^2 \Gamma^2}] \tag{apply approximation, set $\omega=\omega_e$} \\
        &= \frac{4 \pi e^2}{m_e c} \qty[ \frac{\Gamma \omega_e^2}{ 4 \omega_e^2 (\omega - \omega_e)^2 + \omega_e^2 \Gamma^2}] \tag{square expression} \\
        &= \frac{4 \pi e^2}{m_e c} \qty[ \frac{\Gamma}{ 4 (\omega - \omega_e)^2 + \Gamma^2}] \tag{cancel $\omega_e$} \\
        &= \qty(\frac{\pi e^2}{m_e c}) \cdot \frac{1}{\pi} \qty[ \frac{\pi \Gamma}{ (\omega - \omega_e)^2 + \frac{\Gamma^2}{4}}] \\
        &= \qty(\frac{\pi e^2}{m_e c}) \cdot \frac{1}{\pi} \qty[ \frac{\pi \Gamma}{ 4 \pi^2 (\nu - \nu_e)^2 + \frac{\Gamma^2}{4}}] \tag{$\omega = 2 \pi \nu$} \\
        \Aboxed{ \sigma &\approx \qty(\frac{\pi e^2}{m_e c}) \cdot \frac{1}{\pi} \qty[ \frac{\Gamma / 4 \pi}{ (\nu - \nu_e)^2 + (\Gamma / 4 \pi)^2}] }  \\
    \end{align*}
}

\end{document}

### 2022-03-15 519_radiative/pset3/solutions.tex
upload radiative stuff too

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 519A; Problem Set 3}
\author{\textbf{Tom Wagg}}

\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}\\#1\\
}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\allowdisplaybreaks

\begin{document}

\maketitle

{\it \noindent 1). {\it [11/2 class]} quick special relativity questions\\
a) At what speed does a clock move if it runs at a rate which is one-half the rate of a clock at rest?}

\answer{
    \begin{align}
        t &= t_0 \gamma \\
        \gamma &= 2 \\
        \Aboxed{ v &= \sqrt{3}/2 c = 0.87 c }
    \end{align}
}

{\it \noindent b) At what speed does a meter stick move if its length is observed to shrink to 0.5 m?}

\answer{
    \begin{align}
        L &= \frac{L_0}{\gamma} \\
        \gamma &= 2 \\
        \Aboxed{ v &= \sqrt{3}/2 c = 0.87 c }
    \end{align}
}

{\it \noindent c) The average lifetime of a $\pi$ meson in its own frame of reference is 26.0 ns. (This is its proper lifetime.) If the $\pi$ meson moves with speed $0.95c$ with respect to the Earth, what is its lifetime as measured by an observer at rest on Earth? What is the average distance it travels before decaying as measured by an observer at rest on Earth?}

\answer{
    \begin{align}
        t &= \gamma t_0 \\
        &= \frac{26 \times 10^{-9} \unit{s}}{\sqrt{1 - 0.95^2}} \\
        \Aboxed{ t &= 8.3 \times 10^{-8} \unit{s} } \\
        d &= v t \\
        &= 0.95 c \cdot 8.3 \times 10^{-8} \unit{s} \\
        \Aboxed{ d &= 24 \unit{m} }
    \end{align}
}

{\it \noindent d) An atomic clock is placed in a jet airplane. The clock measures a time interval of 3600 s when the jet moves with speed 400 m/s. How much longer or shorter a time interval does an identical clock held by an observer on the ground measure?}

\answer{
    This seems like a trick question upon first reading since we are told ``the clock measures'' its time on the plane, not that a stationary observer measures the clock. But assuming it is not a trick question then we find that the time measured on the ground would be longer by a factor
    \begin{equation}
        \boxed{ \gamma = \qty(1 - (400 \unit{m}{s^{-1}}) / c^2)^{-1/2} = 1.0000000000009 }
    \end{equation}
}

{\it 2) {\it [11/4 class]} A massive space fleet of hostile aliens is traveling towards the Earth.

a.) The aliens' ``hyperdrive'' operates at a Lorentz factor of $\gamma=100$.  Still, traveling 100 light years -- the distance from their home planet to the earth in the CMB rest frame -- takes a long time!  How much time passes for the aliens on the ship?  Assume that the time it takes to accelerate to $\gamma=100$ and then back down is small such that most of the trip is spent at $\gamma=100$.}

\answer{
    Since we are in the frame of the aliens on the ship, we just need to find their velocity and divide the distance by it. Since $\gamma$ is so high, the speed of the aliens is going to be extremely close to the speed of light and so we can simply state the time that would pass, given the definition of a light year, would be simply
    \begin{equation}
        t = 100 \unit{years}
    \end{equation}
    And in the alien's frame this would be
    \begin{equation}
        \boxed{ t = 1 \unit{year} }
    \end{equation}
}

{\it b.) In their frame and halfway between the solar system and their home planet, they broadcast isotropically a signal with average bolometric power $P$.  What is the average bolometric flux ($F =\int d\nu F_\nu$) received at both locations?}

\answer{
    We can apply Eq. 4.97b from RL for this problem
    \begin{equation}
        \dv{P_r}{\Omega} = \frac{1}{\gamma^4 (1 - \beta \mu)^4} \dv{P}{\Omega},
    \end{equation}
    where $P_r$ is the received power and $\mu$ is the dependent on the angle such that $\mu = 1$ in the forward direction and $\mu = -1$ in the backward direction. We can write the flux received at both locations as
    \begin{align}
        F &= \dv{P_r}{A} \\
        &= \dv{P_r}{\Omega} \dv{\Omega}{A} \\
        &= \dv{P_r}{\Omega} \frac{1}{(50 \unit{lyr})^2} \\
        F &= \frac{1}{\gamma^4 (1 - \beta \mu)^4} \frac{P}{4 \pi} \frac{1}{(50 \unit{lyr})^2}
    \end{align}
    So now it's just a matter of evaluating! Note that $\beta = \sqrt{1 - 10^{-4}}$ (from rearranging $\gamma = 100$). For the forward direction we have $\mu = 1$ and thus
    \begin{equation}
        \boxed{ F_{\rm sol} = \frac{1}{\gamma^4 (1 - \beta)^4} \frac{P}{4 \pi} \frac{1}{(50 \unit{lyr})^2} = 5.7 P \times 10^{-32} \unit{cm^{-2}} }
    \end{equation}
    and for the backward direction $\mu = -1$ and we get
    \begin{equation}
        \boxed{ F_{\rm alien} = \frac{1}{\gamma^4 (1 + \beta)^4} \frac{P}{4 \pi} \frac{1}{(50 \unit{lyr})^2} = 2.2 P \times 10^{-50} \unit{cm^{-2}} }
    \end{equation}
    Thus we find that the flux in the forward direction is approximately $2.6 \times 10^{18}$ times stronger in the backward direction, oh my!
}

{\it c.)  On the way, their ship hits large dust particles with diameter $D = 1\mu m$ and density $\rho = 1$gr/cm$^3$. What is the energy in each impact?  Compare to the energy of a 10kg bowling ball moving at $36~$km/hr.}

\answer{
    We can take that the relative Lorentz factor between the dust particle and the ship is $\gamma = 100$. Thus the kinetic energy that is absorbed during each impact is
    \begin{equation}
        E = (\gamma - 1) m c^2,
    \end{equation}
    where the minus one is because the rest mass of the particle stays behind (though this factor makes almost no different the final result). Next we just need to solve for the mass using the radius and density
    \begin{equation}
        m = \rho V = \frac{4 \pi r^3 \rho}{3} = 5.24 \times 10^{-12} \unit{kg}
    \end{equation}
    Plugging this all in we find that the energy of each impact is
    \begin{equation}
        \boxed{ E_{\rm grain} = 99 \cdot 5.24 \times 10^{-12} \cdot (3 \times 10^8)^2 \unit{J} = 4.7 \times 10^{10} \unit{erg} }
    \end{equation}
    The kinetic energy of the bowling ball would be
    \begin{equation}
        \boxed{ E_{\rm ball } = \frac{1}{2} m v^2 = 5 \times 10^9 \unit{erg} }
    \end{equation}
    and so the bowling ball has a factor of 10 lower energy than the dust grains.
}

{\it d) The ``reddening'' of these dust grains is such that they absorb $10\%$ of background optical light towards this star. Estimate the average number density of dust grains along the path by assuming that they have a geometric cross section of $\pi (D/2)^2$, valid for $\lambda \lesssim D$.  What is the flux of impacts hitting the ship? (Answer in the Earth's frame.) }

\answer{
    We are told that $10\%$ of light is absorbed by the dust grains and thus we know that
    \begin{equation}
        e^{-\tau} = 0.9,
    \end{equation}
    where $\tau$ is the optical depth. From the definition of the optical depth we can then solve for the for the number density
    \begin{align}
        \tau &= \int n \sigma \dd{s} \\
        \ln \frac{10}{9} &= n \sigma s \\
        \ln \frac{10}{9} &= n \pi \qty(\frac{D}{2})^2 s \\
        n &= \frac{1}{\pi} \ln \frac{10}{9} \qty(\frac{2}{D})^2 \frac{1}{s} \\
        &= \frac{1}{\pi} \ln \frac{10}{9} \qty(\frac{2}{\unit{\mu m}})^2 \frac{2}{100 \unit{lyr}} \\
        \Aboxed{ n &= 2.8 \times 10^{-13} \unit{cm^{-3}} }
    \end{align}
    The flux of impacts hitting the ship can be found as follows
    \begin{align}
        F &= \frac{N}{A t} = \frac{n V}{A t} = \frac{n d}{t} = n v \\
        \Aboxed{ F &= 0.008 \unit{cm^{-2}}{s^{-1}} }
    \end{align}
}

{\it 3) {\it [11/9 class]} free-free. 
Fast radio bursts indicate that the light travels through an electron column that is quite large with $\int n(s) ds \sim 1000$pc~cm$^{-3}$, which is measured by using that the group velocity of radio waves is slower than the speed of light in a plasma with a known frequency dependence. If FRBs were cosmological, which we now know is the case, such an electron column would arise if the come from a redshift of $z\sim 1$.  For this reason, there was intense interest in the FRB phenomena. HII regions can span a huge range of hydrogen densities, from $10-10^6$cm$^{-3}$, and have temperatures of $10^4$K. Another idea was that they were radio bursts from stars, where the electron column comes from within a radius of the star, near the stellar photosphere. The conditions then would be that the column originates from within $\lesssim 1 R_\odot$, i.e. within a solar radius, and the temperature is likely a coronal temperature of $T\sim 10^6$K.\\

a) Assume Solar metallicity. (i) At what fractional accuracy does the metallicity of the star/HII region matter for computing the free-free optical depth?   It's fine to consider just oxygen, which is the most abundant metal, and note that oxygen is a part in a thousand the hydrogen number abundance at Solar metallicity. (ii) How important is the free-free off of helium ions, which comprise 10\% of the total number density of ions?}

\answer{
    From RL Eq. 5.18a we can surmise that
    \begin{equation}
        \alpha^{\rm ff} \propto Z^2 n_i,
    \end{equation}
    and we additionally know that $\tau \propto \alpha$ and so
    \begin{equation}
        \tau^{\rm ff} \propto Z^2 n_i.
    \end{equation}
    For hydrogen, $Z = 1$ and $n_i = n_H$ and we will use this to compare. For the first part of the question, we have that $Z = 8$ and $n_i = n_O = n_H / 1000$. Therefore the fractional accuracy at which metallicity matters is
    \begin{equation}
        \boxed{ {\rm fractional\ accuracy} = \frac{8^2}{1000} = 0.064 }
    \end{equation}
    For the helium we have that $Z = 2$ and $n_i = n_{He} = n_H / 10$. Thus we can write
    \begin{equation}
        \boxed{ {\rm fractional\ accuracy} = \frac{2^2}{10} = 0.4 }
    \end{equation}
}

{\it b) One question the community asked was whether free-free absorption would absorb the FRB in some of these cases, where the observations were all at $\sim 1$GHz. Comment on whether some or all of the HII region and flaring star scenarios would be ruled out by free free absorption from the host nebulae absorbing the flux from the FRB.  Consider the range of densities for the HII region scenario and distances from the stellar surface for the star scenario given above.}

\answer{
    We can answer this question by comparing the optical depth in each case to $\tau = 1$. We can find the optical depth as
    \begin{equation}
        \tau = N \sigma = N \alpha / n,
    \end{equation}
    where $N$ is the column density and the absorption coefficient is found using RL Eq. 5.19b
    \begin{equation}
        \alpha = \frac{4 e^6}{3 m k_B c} \qty(\frac{2 \pi}{3 k_B m})^{1/2} T^{-3/2} Z^2 n_e n_i \nu^{-2} \tilde{g}_{\rm ff}.
    \end{equation}
    So we can combine these to writ ethat the optical depth is
    \begin{equation}
        \tau = \qty(0.018 \unit{cm^5}{K^{3/2}}{s^{-2}}) T^{-3/2} \nu^{-2} N n
    \end{equation}
    Let's first just consider the HII region scenario. In this case, we have that
    \begin{align}
        \tau_{\rm HII} &= \qty(0.018 \unit{cm^5}{K^{3/2}}{s^{-2}}) (10^4 \unit{K})^{-3/2} (1 \unit{GHz})^{-2} (1000 \unit{pc}{cm^{-3}}) \\
        \Aboxed{ \tau_{\rm HII} &= 5.5 n \times 10^{-5} \unit{cm^{3}} }
    \end{align}
    where we have assumed that for an HII region, $Z = 1$ and $n = n_e = n_i$ and that the Gaunt factor is approximately 1. Thus we can see that over the range of densities in an HII region ($n = 10$ to $n = 10^6$), for most densities the optical depth will be much less than 1 but for the higher density regions some and eventually all of the FRBs can be absorbed. In particular, at $n = 18000 \unit{cm^{-3}}$, the optical depth would be equal to 1.

    And also because I really have nothing better to be doing with my time I also made a quick plot of how the fraction absorbed changes as a function of number density. You can see in the plot that once the number density is around $2000 \unit{cm^{-3}}$ then 10\% will get absorbed and once it hits $10^5 \unit{cm^{-3}}$ pretty much everything will get absorbed. Fun :D

    \includegraphics[width=\textwidth]{figures/astro519_ps3_q3b.png}

    Now let's consider the other case. The one small addition is that we need to find the number density by dividing the column density by the distance
    \begin{equation}
        n_{\rm flare} = \frac{N}{s} = \frac{N}{R_\odot}
    \end{equation}
    And with that we can immediately write down the absorption coefficient and optical depth
    \begin{align}
        \tau_{\rm flare} &= \qty(0.018 \unit{cm^5}{K^{3/2}}{s^{-2}}) (10^6 \unit{K})^{-3/2} \qty(\frac{N^2}{R_\odot}) (1 \unit{GHz})^{-2} \\
        \Aboxed{ \tau_{\rm flare} &= 2425 }
    \end{align}
    And thus we see that this optical depth is \textit{wayyy} above one and so everything is absorbed.
}

{\it c) Young neutron stars -- what likely produce FRBs -- often inhabit a bubble of electron-positron pairs.  Could a dense electron-positron pair plasma with temperature $10^{10}$K contribute substantial free-free absorption?}

Yup! Given the dipole approximation, a dense collection of pairs of positive and negative charges will contribute substantial free-free absorption. Assuming that the density is high enough and given that the temperature is particularly high, I imagine this would cause substantial absorption. Some caveats to this are that I'm not sure if the dipole approximation holds because the positive charge will move a lot (unlike in ions) and that I am assuming that there is no annihilation.

{\it 4.) {\it [11/11 class]}  Synchrotron and inverse Compton.\\
Suppose relativistic electrons of fixed energy $\gamma_0$ are being introduced at a constant rate into a medium with (tangled) magnetic field with amplitude $B$. These electrons will emit synchrotron radiation, and, as a result, lose energy. After a sufficiently long time, the electrons will settle into a steady state energy distribution $N(\gamma)$, which we would like to determine. Only consider Lorentz factors where $\gamma \gg 1$.\\
a)  Show that the rate of change of Lorentz factor (i.e. energy divided by $mc^2$) of an individual electron with Lorentz factor $\gamma$ is given by
\begin{equation}
\dot \gamma = \frac{d\gamma}{dt} = -A \gamma^2,   ~~~{\rm where}~~A = \frac{4 e^4 B^2}{9 m^3 c^5}.
\end{equation}}

\answer{
    \begin{align}
        \dot{\gamma} &= \dv{\gamma}{t} \\
                     &= \dv{t} \qty(\frac{E}{m c^2}) \tag{since $E = \gamma m c^2$}\\
                     &= \dv{E}{t} \qty(\frac{1}{m c^2}) \\
                     &= -\frac{P}{m c^2} \\
                     &= -\frac{2 e^2}{3 c^3} \gamma^4 \frac{e^2 B^2}{\gamma^2 m^2} \frac{v_{\perp}^2}{c^2} \frac{1}{m c^2} \tag{RL 6.5a} \\
                     &= -\frac{4 e^2}{9 c^3} \gamma^4 \frac{e^2 B^2}{\gamma^2 m^2} \frac{v^2}{c^2} \frac{1}{m c^2} \tag{RL 6.6} \\
                     &= -\frac{4 e^2}{9 c^3} \gamma^4 \frac{e^2 B^2}{\gamma^2 m^2} \frac{1}{m c^2} \tag{$v \ll c$ since $\gamma \gg 1$} \\
                     &= -\frac{4 e^4 B^2}{9 m^3 c^5} \gamma^2 \\
        \Aboxed{ \dot{\gamma} &= -A \gamma^2 }
    \end{align}
}

{\it b)  Show that $-\dot \gamma N(\gamma)dt$ is equal to the number of electrons whose energies pass through energy $\gamma m c^2$ during time interval $dt$.}

\answer{
    This is easiest to understand with the picture that we drew on the whiteboard. The number of electrons with energies between $\gamma$ and $\gamma + \dd \gamma$ is the same as the flux of electrons over some time interval $\dd{t}$ that pass through an energy $E(\gamma) = \gamma m c^2$. Therefore we can write out the following (adding in a minus since this this is flux down in energy)
    \begin{align}
        F \dd{t} &= - N(\gamma) \dd{\gamma} \\
        F \dd{t} &= - N(\gamma) \dv{\gamma}{t} \dd{t} \\
        \Aboxed{ F \dd{t} &= - \dot{\gamma} N(\gamma) \dd{t} }
    \end{align}
}

{\it c)  Argue that in steady-state $\dot \gamma N(\gamma)$ must be independent of $\gamma$ and thus that $N(\gamma) \propto \gamma^{-2}$ up to a cutoff at $\gamma = \gamma_0$.}

\answer{
    A steady state means that $\dv{t}[N(\gamma)] = 0$, meaning that the number of electrons between $\gamma$ and $\dd{\gamma}$ doesn't change over time. In order to conserve this, it must be the case that the flux over different $\gamma$s cannot be different otherwise you would get a build up. This means that $\dot \gamma N(\gamma)$ must be independent of $\gamma$ in a steady-state.

    \noindent We know that $\dot \gamma \propto \gamma^2$, therefore, if $\dot \gamma N(\gamma)$ has no $\gamma$ dependence, then it must be the case that $N(\gamma) \propto \gamma^{-2}$!
}

{\it d) What is the spectral index of the synchrotron emission from the entire ensemble. Does this spectral index hold at all frequencies?}

\answer{
    From RL 6.20b and 6.22b, we have the the spectral index is defined such that
    \begin{equation}
        s = \frac{p - 1}{2},\qquad \quad {\rm where\ } N(\gamma) \propto \gamma^{-p}
    \end{equation}
    Since we have just found that $N(\gamma) \propto \gamma^{-2}$ in the previous part, this means that $p = 2$. Therefore we have that
    \begin{equation}
        \boxed{ s = 0.5 }
    \end{equation}
    This spectral index does not hold at all frequencies. It is only valid for $\omega < \omega_c$, where the critical frequency is defined as
    \begin{equation}
        \omega_c = \frac{3 \gamma_0 e B}{2 m c} \sin \alpha,
    \end{equation}
    from RL 6.17c.
}

{\it e) Below what value of $B$ is the electron cooling shaped by inverse Compton cooling off the cosmic microwave background. Argue that all the above reasoning still applies except for a different expression for $A$.  Give the new expression.}

\answer{
    Electron cooling will be shaped by inverse Compton cooling off the cosmic microwave background as long as the energy density of the CMB is greater than the magnetic field energy density. From the last problem set (Q2ab), we know that the energy density of the CMB is
    \begin{equation}
        u_{\rm CMB} = 0.25 \unit{eV}{cm^{-3}},
    \end{equation}
    and that the energy density of a magnetic field is
    \begin{equation}
        u_B = \frac{B^2}{8 \pi}.
    \end{equation}
    We can equate these two and solve for the magnetic field strength at which the transition occurs. Therefore, the electron cooling is shaped by the inverse Compton off the CMB as long as $B < B_c$, where
    \begin{equation}
        \boxed{ B_c = \sqrt{2\pi \unit{eV}{cm^{-3}}} = 3.2 \unit{\mu G} }
    \end{equation}
    From RL 7.18, we are given that the following relation holds
    \begin{equation}
        \frac{P_{\rm synch}}{P_{\rm compt}} = \frac{u_{\rm synch}}{u_{\rm compt}},
    \end{equation}
    thus we can follow the same reasoning as the previous parts, except the power will be adjusted by a factor $A_{\rm comp}$, which is the ratio of the energy densities, given by
    \begin{equation}
        A_{\rm comp} = \frac{2 \pi}{B^2} \unit{eV}{cm^{-3}}
    \end{equation}
    This means that the final $A$ value is
    \begin{equation}
        \boxed{ A = \frac{8 \pi e^4}{9 m^3 c^5} \unit{eV}{cm^{-3}} }
    \end{equation}
}

\end{document}

### 2022-03-15 519_radiative/pset4/solutions.tex
upload radiative stuff too

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 519A; Problem Set 4, some of it at least}
\author{\textbf{Tom Wagg}}

\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}\\#1\\
}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\allowdisplaybreaks

\begin{document}

\maketitle

{\it 1.  The temperature in a star varies with depth.   Assume that we can approximate the source function $S_\nu$ of a star as $S_\nu = a_\nu +b_\nu \tau_\nu$, where $\tau_\nu$ is defined as the optical depth from the surface to some depth (i.e. along a vector {\bf normal} to the surface...not necessarily along the path you are considering later on). We will show in class that this $S_\nu$ is actually a reasonable model: even more amazingly we will find that  for a grey atmosphere in the Eddington approximation not only does this form apply but with $a_\nu =b_\nu$.\\ 
a) What is the intensity at the surface and how does it relate to the source function evaluated at $\tau_\nu = 1$?  (Assume a ray perpendicular to the surface.)}

\answer{
    \begin{align}
        \dv{I_\nu}{\tau_\nu} &= I_\nu - S_\nu \\
        \dv{I_\nu}{\tau_\nu} &= I_\nu - (a_\nu + b_\nu \tau_\nu) \\
        I_v(\tau_\nu) &= a_\nu + b_\nu (1 + \tau_\nu) + C e^{\tau_\nu} \tag{use Wolfram Alpha to solve} \\
        I_v(\tau_\nu) &= a_\nu + b_\nu (1 + \tau_\nu) \tag{apply boundary condition} \\
        \Aboxed{ I_v(0) &= a_\nu + b_\nu }
    \end{align}
    From comparison, we have that $S_{\nu}(\tau_\nu = 1) = a_\nu + b_\nu$ and thus $S_{\nu}(\tau_\nu = 1) = I_v(\tau_\nu = 0)$.
}

{\it b) Apply the result in part (a) to stellar lines versus the stellar continuum.  How does the intensity at the center of a line differ from the continuum intensity? Please discuss both the case where the temperature decreases with increasing radial coordinate and the case where there is an inversion such that it increases with increasing radial coordinate.  (See problem 3 where we work out a somewhat more specific example.)}

\answer{
    First let's consider the case in which the temperature decreases with increasing radial coordinate. In this case, the continuum source function is greater than the line source function. Applying our result from part a we can therefore say that the intensity of the continuum is greater than the intensity of the line centre.

    For the case in which the temperature increases with increasing radial coordinate the continuum source function is less than the line source function. Applying our result from part a we can therefore say that the opposite relation and so the intensity of the continuum is less than the intensity of the line centre.
}

{\it c) The source function we have assumed does not apply in the chromosphere of the Sun, which has a temperature increasing with  increasing radius (whereas in the photosphere temperature is decreasing with radius). Is it lines that have larger or that have smaller optical depths that are more likely to be seen in emission? Justify your answer.}

\answer{
    The \textbf{larger} optical depths! If the optical depth was too small (specifically, smaller than the continuum optical depth) then you would not be able to distinguish the line emission from the continuum as the optical depth increases inwards for the photosphere but outwards for the chromosphere.
}

{\it d) Now we will generalize what we did in part a) to sum rays from different angles.  Argue that the emergent specific intensity for a ray with an angle $\theta$ normal to the (planar) stellar surface is given by $I_\nu = S_\nu(\tau_\nu = \cos(\theta))$.  Remember that $\tau_\nu$ in this expression is the optical depth normal to the surface.  This is known as the Eddington-Barbier approximation, where the `approximation' is in the form of the source function (which we will motivate in class).}

\answer{
    If the ray come at an angle $\theta$ normal to the plane this means that the \textit{effective} optical depth is $\tau_\nu / \cos \theta$. Therefore, since we know that the intensity at the surface is $S_\nu(\tau_\nu = 1)$, in order to get an effective optical depth of 1, the optical depth normal to the plane would need to be $\cos \theta$ instead of $1$. 
}

{\it e) From $I_\nu = S_\nu(\tau_\nu = \cos(\theta))$, calculate the flux at the surface and find $F_\nu =  \pi (a_\nu + b_\nu \times 2 /3)$.  Thus, $F_\nu = \pi S_\nu(\tau_\nu = 2/3)$!  The vaunted two-thirds law!  Of course, the flux at the surface is easy to relate to the flux at any distance from the star by $(R_*/r)^2$ dilution.}

\answer{
    \begin{align}
        F_\nu &= \int \dd{\Omega} I_\nu \cos \theta \\
              &= \int \dd{\Omega} \cos \theta \qty(a_\nu + b_\nu \cos \theta) \\
              &= 2 \pi \int_0^1 \dd{\mu} \mu (a_\nu + b_\nu \mu) \\
        \Aboxed {F_\nu &= \pi \qty(a_\nu + \frac{2}{3}b_\nu) }
    \end{align}
}

{\it f) Reason (qualitatively) how the above calculations relate to stellar limb darkening.  Assume that the temperature decreases with increasing radius.}

\answer{
    Stellar limb darkening is where the edges of a star appear darker than the centre. When viewing the limbs we are looking at a region of lower temperature as the question states. Since the temperature is lower, this mean that the source function will be lower. In the previous part we found that the flux at the surface is proportional to the source function and thus this means the flux in the limbs is lower, leading to stellar limb darkening.
}

\end{document}

### 2022-03-28 531_stars/pset1/solutions.tex
first two questions

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\title{Astro 531; Problem Set 1}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\maketitle

\question{\textbf{2.1 - Stellar Properties}}

\question{Part a - Radius}
\answer{
    \begin{align}
        L &= 4 \pi R^2 \sigma T_{\rm eff}^4 \\
        R &= \sqrt{\frac{L}{4 \pi \sigma T_{\rm eff}^4}} \\
        \Aboxed{ R &= 2570 \unit{R_{\odot}} }
    \end{align}
}

\question{Part b - Density}
\answer{
    If we assume that the star is a constant density throughout then we can use the definition of density and the radius we calculated above.
    \begin{align}
        \rho &= \frac{M}{\frac{4}{3} \pi R^3} \\
        \Aboxed{ \rho &= 1.66 \times 10^{-6} \unit{kg}{m^{-3}} }
    \end{align}
}

\question{Part c - Escape velocity}
\answer{
    \begin{equation}
        \boxed{ v_{\rm esc} = \sqrt{ \frac{2 G M}{R} } = 54500 \unit{m}{s^{-1}} }
    \end{equation}
}

\question{Part d - Comparisons to Sun}
\answer{
    This star is clearly much larger than the sun (by a factor of 2570) but is \textbf{much} less dense (since the mean density of the sun is approximately $1400 \unit{kg}{m^{-3}}$, so the density is $10^{9}$ times lower than the sun). Finally, the escape velocity is lower than the sun (around 0.088 times lower). 
}

\pagebreak

\question{\textbf{3.5 - Kepler's 3rd Law from Virial Theorem}}
\answer{
    The virial theorem gives the following relations between energies
    \begin{equation}
        E_{\rm kin} = - \frac{1}{2} E_{\rm pot}
    \end{equation}
    For the orbits of planets around the sun, their potential energy will come from the gravitational potential and so
    \begin{equation}
        E_{\rm pot} = - \frac{G m_1 m_2}{a},
    \end{equation}
    and the kinetic energy is simply
    \begin{equation}
        E_{\rm kin} = \frac{1}{2} m_1 v_1^2 + \frac{1}{2} m_2 v_2^2.
    \end{equation}
    For the next part, it will be useful to define the distance from each body to the centre of mass, $a_i$, in terms of the masses
    \begin{equation}
        a_i = \frac{a \cdot m_{1 - i}}{m_1 + m_2} \\
    \end{equation}
    and additionally, it's useful to write the velocity in terms of the period and masses
    \begin{align}
        v_i &= \frac{2 \pi a_i}{P} \\
        v_i &= \frac{2 \pi}{P} \frac{a \cdot m_{1 - i}}{m_1 + m_2}
    \end{align}
    Now we can put this all together in the expression for the virial theorem.
    \begin{align}
        \frac{1}{2} m_1 v_1^2 + \frac{1}{2} m_2 v_2^2 &= \frac{1}{2} \cdot \frac{G m_1 m_2}{a}, \\
        \frac{1}{2} m_1 \qty(\frac{2 \pi}{P} \frac{a \cdot m_2}{m_1 + m_2})^2 + \frac{1}{2} m_2  \qty(\frac{2 \pi}{P} \frac{a \cdot m_1}{m_1 + m_2})^2 &= \frac{G m_1 m_2}{2 a} \\
        \frac{2 \pi^2 a^2}{P^2 M^2} \qty( m_1 m_2^2 + m_2 m_1^2 ) &= \frac{G m_1 m_2}{2 a} \\
        \frac{a^3}{P^2} &= \frac{G}{4 \pi^2} \frac{M^2 m_1 m_2}{\qty( m_1 m_2^2 + m_2 m_1^2 )} \\
        \frac{a^3}{P^2} &= \frac{G}{4 \pi^2} \frac{M^2}{\qty( m_1 + m_2 )} \\
        \Aboxed{ \frac{a^3}{P^2} &= \frac{G M}{4 \pi^2} }
    \end{align}
}

\end{document}

### 2022-04-01 531_stars/pset1/solutions.tex
most of first stars problem set

@@
 
-\question{Part a - Radius}
+\question{Part a - Radi  us}
 \answer{
@@
 
+\pagebreak
+
+\question{\textbf{4.4 - Central Temperature-Density}}
+\answer{
+    Here's a plot of the three different stars on (pretty much) the same figure as the textbook (I made the boundaries using Onno Pols' lecture notes but shaded them the same way as the textbook).
+I've also indicated the main sequence turnoff and the exhaustion of central helium as dots in the tracks (where the tracks evolve from the bottom left to top right).
+    \begin{center}
+        \includegraphics[width=0.8\textwidth]{figures/central_T_rho.pdf}
+    \end{center}
+    From these tracks we can conclude that over time each of the stars is increasing in temperature and density within its core.
+For the $1 \unit{M_{\odot}}$ star, it moves from the ideal gas regime to the degenerate (but non-relativistic) regime during core helium burning.
+This makes sense as the core of the star at this stage would be a white dwarf, despite it still being ``living'' in the outer later.
+
+    The $12 \unit{M_{\odot}}$ star remains in the ideal gas regime all the way up until central helium exhaustion.
+Once helium is exhausted in the core, the star drifts into the extremely relativistic degenerate regime.
+This makes sense as we would expect this star to end its life as a neutron star.
+
+    Finally, the $120 \unit{M_{\odot}}$ star very closely follows the radiation-ideal gas boundary, particularly during its main sequence.
+After this it moves more clearly into the ideal gas regime and, though it appears to end its life here, the models only evolve through carbon burning.
+In reality, we would expect more complex evolution beyond this point and that the star would end as a black hole (given that its final mass is still around $30 \unit{M_{\odot}}$ even with all of the mass loss).
+}
+
+\pagebreak
+
+\question{\textbf{5.1 - Contribution to Opacity}}
+
+\question{Part a - Identifying regimes}
+\answer{
+    Electron scattering essentially dominates at any point at which the star is fully ionised.
+In terms of the plot this means that the electron scattering dominates in the far right at temperatures approximately above $10^{7.5} \unit{K}$.
+In this region the opacity is almost constant with increasing pressure.
+
+    For bound-free and free-free absorption, they dominate the opacity within temperatures of around $10^{5.5} - 10^{7.5} \unit{K}$.
+In this region we can see that the slopes are all approximately following $T^{-7/2}$ as we could expect.
+}
+
+\question{Part b - Check dependence}
+\answer{
+    For electron scattering the slope is 0 (there is no temperature dependence), as we would expect.\\\\
+    To check the temperature and density dependence for opacity with bound-free and free-free absorption we need to check that the slope is around $-7/2$.
+So just as a test, for the density of $10^{2} \unit{g}{cm^{-3}}$, the slope is
+    \begin{equation}
+        \mathrm{slope} = \frac{2 - 0.2}{6.5 - 7} = -3.6
+    \end{equation}
+    which is approximately $-7/2$, especially given that I pulled these numbers off the plot by eye!
+
+    We can also check the density dependence by checking the values at a fixed temperature.
+At a value of $\log T = 6.2$, the opacities of the densities with $\log \rho \in [0, -2, -4]$ are approximately $\log \kappa = [2, 0.9, -0.1]$.
+So the change in opacity is about the same for the same change in the density, thus showing a linear dependence as we would expect.
+}
+
+% \clearpage
+
+\question{\textbf{6.1 - Mass Luminosity Relation}}
+
+\question{Part a - Check relation for Appendix D}
+\answer{
+    I have plotted the masses and luminosities of the stars at different metallicities in the top panel of the figure below in order to compare against the mass-luminosity relation (using the same exponents as Figure 2.3 of the textbook), which I show as a black dotted line.
+    We can see that the mass-luminosity relation agrees well with the data for most stars.
+It is only for stars above $30 \unit{M_{\odot}}$ where the relation seems to no longer apply.
+This coincides with the mass where stars start to be radiation dominated as we talked about in class so perhaps this is the reason for the deviation.
+}
+
+\question{Part b - Luminosities}
+\answer{
+    As the question states, the luminosity of massive stars is about the same for both compositions, but the luminosity of the metal poor low-mass stars is \textit{higher} than those of $Z = 0.014$ stars.
+We can see this exactly in the figure above.
+Specifically, in the bottom panel I show the ratio of the two luminosities and we can see that for low mass stars, the luminosity of metal poor ones is higher.
+
+    The short answer to this trend is that \textbf{low-mass stars do not have electron scattering as a significant source of opacity}.
+To elaborate: low-mass stars tend to be cooler than higher-mass stars, and so they are less ionised and have less electron scattering.
+This means that the opacity of low-mass stars comes almost entirely from bound-free absorption (free-free absorption wouldn't be applicable since $\kappa_{bf} \gg \kappa_{ff}$ for a metal abundance of $Z > 0.001$).
+Now bound-free absorption is directly proportional to the metallicity and thus metal-poor low-mass stars have lower opacity (and so higher luminosity) than metal-rich low-mass stars.
+The same is not true for high-mass stars since electron scattering is independent of metallicity and so when the star is more metal-poor the electron scattering still limits the luminosity even when the bound-free absorption is low.
+}
+
+\begin{center}
+    \includegraphics[width=\textwidth]{figures/ml_relation.pdf}
+\end{center}
+
 \end{document}

### 2022-04-05 531_stars/pset1/solutions.tex
create new title style, add last question

@@
 
-\title{Astro 531; Problem Set 1}
+\newcommand{\psetnum}{1}
+\newcommand{\class}{ASTR 531 - Stellar Interiors}
+
+\newcommand{\tomtitle}{
+    \noindent {\huge \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
+    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
+    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
+}
+
+\title{\class : Problem Set \psetnum}
 \author{\textbf{Tom Wagg}}
@@
 
-\maketitle
+\tomtitle
 
@@
 
-\question{Part a - Radi  us}
+\question{Part a - Radius}
 \answer{
@@
 
-\pagebreak
-
 \question{\textbf{5.1 - Contribution to Opacity}}
@@
 
-% \clearpage
-
 \question{\textbf{6.1 - Mass Luminosity Relation}}
@@
 }
+
+\begin{center}
+    \includegraphics[width=\textwidth]{figures/ml_relation.pdf}
+\end{center}
 
@@
 
+\question{\textbf{7.2 - Convection Criterions}}
+
+\question{Part a and Part b - Convection zones with different criterions}
+\answer{
+    I combined these two parts since I put all of the information on the same plot:
     \begin{center}
-    \includegraphics[width=\textwidth]{figures/ml_relation.pdf}
+        \includegraphics[width=0.5\textwidth]{figures/criterion_shift_zones.jpg}
     \end{center}
+    The $\mu$ profile decreases with increasing radius since fusion is creating heavier elements in the core.
+As the question asks I adopted some arbitrary limits (red lines) for the convection zone when using the Schwarzschild criterion.
+
+    When instead adopting the Ledoux criterion, the convection zone will shrink (as shown by the green lines).
+I have two qualitative arguments for this.
+First, since the Ledoux criterion is identical to the Schwarzschild criterion except with an extra term for the $\mu$ gradient, this criterion is \textit{harder} to fulfill.
+This means that the fraction of the star that is convective should decrease and thus the zone should shrinks.
+
+    Additionally, I also considered the `blob' of gas.
+When switching from Schwarzschild to Ledoux, a `blob' travelling upwards now is not only at a different temperature than the surrounding regions, but also probably heavier due to the increased $\mu$ relative to the surroundings.
+This would decrease the upper limit of the convection zone since a `blob' could not travel to the same height.
+The same argument can be applied in the opposite direction and so a `blob' can't sink as far into the star, increasing the lower limit.
+}
+
+\question{Part c - Convection effects on $\mu$ gradient}
+\answer{
+    The $\mu$ gradient flattens out within the convection zone.
+This is because all of the material is mixed and thus the $\mu$ at all radii within the zone because the average of the $\mu$ values that were in the zone initially.
+See the image below for my sketch of this.
+    \begin{center}
+        \includegraphics[width=0.7\textwidth]{figures/mixing.jpg}
+    \end{center}
+}
 

### 2022-04-05 531_stars/pset1/solutions.tex
minor typos fixed

@@
 \usepackage{empheq}
+\usepackage{wrapfig}
 
@@
     \begin{center}
-        \includegraphics[width=0.5\textwidth]{figures/criterion_shift_zones.jpg}
+        \includegraphics[width=0.45\textwidth]{figures/criterion_shift_zones.jpg}
     \end{center}
+    Yes yes, I'm sure you're pondering why I never pursued a career in art despite my excessive talents, but please let's focus on the question at hand!
 The $\mu$ profile decreases with increasing radius since fusion is creating heavier elements in the core.
@@
 First, since the Ledoux criterion is identical to the Schwarzschild criterion except with an extra term for the $\mu$ gradient, this criterion is \textit{harder} to fulfill.
-This means that the fraction of the star that is convective should decrease and thus the zone should shrinks.
+This means that the fraction of the star that is convective should decrease and thus the zone should shrink.
 
-    Additionally, I also considered the `blob' of gas.
-When switching from Schwarzschild to Ledoux, a `blob' travelling upwards now is not only at a different temperature than the surrounding regions, but also probably heavier due to the increased $\mu$ relative to the surroundings.
+    Additionally, I considered the `blob' of gas.
+When switching from Schwarzschild to Ledoux, a `blob' travelling upwards now is not only at a different temperature than the surrounding regions, but also \textit{heavier} due to the increased $\mu$ relative to the surroundings.
 This would decrease the upper limit of the convection zone since a `blob' could not travel to the same height.
@@
     The $\mu$ gradient flattens out within the convection zone.
-This is because all of the material is mixed and thus the $\mu$ at all radii within the zone because the average of the $\mu$ values that were in the zone initially.
+This is because all of the material is mixed and thus the $\mu$ at all radii within the zone are equal to the average of the $\mu$ values that were in the zone initially.
 See the image below for my sketch of this.

### 2022-04-05 558_exoplanets/pset1/kepler_solver.py
exoplanets hw1 q1

@@ -1,0 +1,110 @@
+import numpy as np
+
+# ANSI colour codes
+RED = "\033[0;31m"
+GREEN = "\033[0;32m"
+UNDERLINE = "\033[4m"
+END = "\033[0m"
+
+
+def g(E, e, M):
+    """ Kepler's equation in functional form (g(E, e, M) = 0) """
+    return E - e * np.sin(E) - M
+
+
+def g_prime(E, e):
+    """ Derivative of g(E, e, M) """
+    return 1 - e * np.cos(E)
+
+
+def solve_kepler_equation(e, M, tol=1e-10, max_steps=10000, verbose=False):
+    """Solve Kepler's equation using Newton's method
+
+    Parameters
+    ----------
+    e : `float`
+        Eccentricity
+    M : `float`
+        Mean Anomaly
+    tol : `float`, optional
+        Tolerance, meaning the difference between E across steps at which to declare we've converged to a
+        solution (too big and the value will be wrong, too small and you may not converge), by default 1e-10
+    max_steps : `int`, optional
+        Maximum number of steps to take, by default 10000
+    verbose : `bool`, optional
+        Whether to print out detailed information, by default False
+
+    Returns
+    -------
+    E : `float`
+        Eccentric anomaly
+    """
+    # make an initial guess of the eccentric anomaly
+    E = M + 0.85 * e * np.sign(np.sin(M))
+    steps = 0
+    tol_to_dp = int(abs(np.log10(tol)))
+
+    if verbose:
+        print("Initial guess:", np.round(E, tol_to_dp))
+
+    # loop until you reach the maximum number of steps
+    while steps < max_steps:
+        # calculate the next value of E
+        new_E = E - g(E, e, M) / g_prime(E, e)
+        if verbose:
+            print(f"Step {steps} guess: {np.round(new_E, tol_to_dp)}")
+
+        # check whether the value of E has change significantly (if not then return)
+        if np.abs(new_E - E) < tol:
+            return E
+
+        # update E and the steps
+        E = new_E
+        steps += 1
+
+    raise ValueError("Did not converge")
+
+
+def test_kepler_solver(n_tests=10000, **kwargs):
+    """Test whether the solver for Kepler's equation is working
+
+    Parameters
+    ----------
+    n_tests : `int`, optional
+        Number of random tests to do, by default 10000
+
+    Returns
+    -------
[…]

### 2022-04-05 558_exoplanets/pset1/kepler_solver.py
add some text to make the submission better

@@ -3,6 +3,7 @@ import numpy as np
 # ANSI colour codes
 RED = "\033[0;31m"
 GREEN = "\033[0;32m"
+BOLD = "\033[1m"
 UNDERLINE = "\033[4m"
 END = "\033[0m"
 
@@ -78,7 +79,7 @@ def test_kepler_solver(n_tests=10000, **kwargs):
     e, M, E : `tuple of array of floats` :p
         Eccentricity and anomalies, returned only when tests fail, otherwise None returned
     """
-    print(f"Running {UNDERLINE}{n_tests}{END} tests of the solver")
+    print(f"Running {n_tests} tests of the Kepler equation solver")
 
     # create random e in [0, 1) and random M in [0, 2 pi)
     e = np.random.uniform(0, 1, size=n_tests)
@@ -103,6 +104,8 @@ def test_kepler_solver(n_tests=10000, **kwargs):
 
 
 def main():
+    print()
+    print(f"{BOLD}{UNDERLINE}Problem 1 - Create Kepler equation solver and show it works{END}")
     test_kepler_solver(tol=1e-9)
 
 

### 2022-04-08 558_exoplanets/pset1/kepler_solver.py
progress on pset1

@@ -1,4 +1,5 @@
 import numpy as np
+import astropy.units as u
 
 # ANSI colour codes
 RED = "\033[0;31m"
@@ -64,6 +65,14 @@ def solve_kepler_equation(e, M, tol=1e-10, max_steps=10000, verbose=False):
         steps += 1
 
     raise ValueError("Did not converge")
+
+
+def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
+    M = 2 * np.pi / P * (t - t_p)
+    E = solve_kepler_equation(e, M)
+    f = 2 * np.arctan(((1 + e) / (1 - e))**(0.5) * np.tan(E / 2))
+
+    return k_star * (np.cos(omega + f) + e * np.cos(omega)) + gamma
 
 
 def test_kepler_solver(n_tests=10000, **kwargs):

### 2022-04-09 558_exoplanets/pset1/solutions.tex
reorganise 558 pset1 folder

--- 2022-04-01 531_stars/pset1/solutions.tex
@@
 
-\title{Astro 531; Problem Set 1}
+\newcommand{\psetnum}{1}
+\newcommand{\class}{ASTR 558 - Exoplanets}
+
+\newcommand{\tomtitle}{
+    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
+    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
+    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
+}
+
+\title{\class : Problem Set \psetnum}
 \author{\textbf{Tom Wagg}}
@@
 
-\maketitle
+\tomtitle
 
-\question{\textbf{2.1 - Stellar Properties}}
+\question{\textbf{Kepler Solver}}
+\answer{
 
-\question{Part a - Radi  us}
-\answer{
-    \begin{align}
-        L &= 4 \pi R^2 \sigma T_{\rm eff}^4 \\
-        R &= \sqrt{\frac{L}{4 \pi \sigma T_{\rm eff}^4}} \\
-        \Aboxed{ R &= 2570 \unit{R_{\odot}} }
-    \end{align}
 }
 
-\question{Part b - Density}
-\answer{
-    If we assume that the star is a constant density throughout then we can use the definition of density and the radius we calculated above.
-    \begin{align}
-        \rho &= \frac{M}{\frac{4}{3} \pi R^3} \\
-        \Aboxed{ \rho &= 1.66 \times 10^{-6} \unit{kg}{m^{-3}} }
-    \end{align}
-}
-
-\question{Part c - Escape velocity}
-\answer{
-    \begin{equation}
-        \boxed{ v_{\rm esc} = \sqrt{ \frac{2 G M}{R} } = 54500 \unit{m}{s^{-1}} }
-    \end{equation}
-}
-
-\question{Part d - Comparisons to Sun}
-\answer{
-    This star is clearly much larger than the sun (by a factor of 2570) but is \textbf{much} less dense (since the mean density of the sun is approximately $1400 \unit{kg}{m^{-3}}$, so the density is $10^{9}$ times lower than the sun).
-Finally, the escape velocity is lower than the sun (around 0.088 times lower).
-}
-
-\pagebreak
-
-\question{\textbf{3.5 - Kepler's 3rd Law from Virial Theorem}}
-\answer{
-    The virial theorem gives the following relations between energies
-    \begin{equation}
-        E_{\rm kin} = - \frac{1}{2} E_{\rm pot}
-    \end{equation}
-    For the orbits of planets around the sun, their potential energy will come from the gravitational potential and so
-    \begin{equation}
-        E_{\rm pot} = - \frac{G m_1 m_2}{a},
-    \end{equation}
-    and the kinetic energy is simply
-    \begin{equation}
-        E_{\rm kin} = \frac{1}{2} m_1 v_1^2 + \frac{1}{2} m_2 v_2^2.
-    \end{equation}
-    For the next part, it will be useful to define the distance from each body to the centre of mass, $a_i$, in terms of the masses
-    \begin{equation}
-        a_i = \frac{a \cdot m_{1 - i}}{m_1 + m_2} \\
-    \end{equation}
-    and additionally, it's useful to write the velocity in terms of the period and masses
-    \begin{align}
-        v_i &= \frac{2 \pi a_i}{P} \\
-        v_i &= \frac{2 \pi}{P} \frac{a \cdot m_{1 - i}}{m_1 + m_2}
-    \end{align}
-    Now we can put this all together in the expression for the virial theorem.
-    \begin{align}
-        \frac{1}{2} m_1 v_1^2 + \frac{1}{2} m_2 v_2^2 &= \frac{1}{2} \cdot \frac{G m_1 m_2}{a}, \\
-        \frac{1}{2} m_1 \qty(\frac{2 \pi}{P} \frac{a \cdot m_2}{m_1 + m_2})^2 + \frac{1}{2} m_2  \qty(\frac{2 \pi}{P} \frac{a \cdot m_1}{m_1 + m_2})^2 &= \frac{G m_1 m_2}{2 a} \\
-        \frac{2 \pi^2 a^2}{P^2 M^2} \qty( m_1 m_2^2 + m_2 m_1^2 ) &= \frac{G m_1 m_2}{2 a} \\
-        \frac{a^3}{P^2} &= \frac{G}{4 \pi^2} \frac{M^2 m_1 m_2}{\qty( m_1 m_2^2 + m_2 m_1^2 )} \\
-        \frac{a^3}{P^2} &= \frac{G}{4 \pi^2} \frac{M^2}{\qty( m_1 + m_2 )} \\
-        \Aboxed{ \frac{a^3}{P^2} &= \frac{G M}{4 \pi^2} }
-    \end{align}
-}
-
-\pagebreak
-
-\question{\textbf{4.4 - Central Temperature-Density}}
-\answer{
-    Here's a plot of the three different stars on (pretty much) the same figure as the textbook (I made the boundaries using Onno Pols' lecture notes but shaded them the same way as the textbook).
-I've also indicated the main sequence turnoff and the exhaustion of central helium as dots in the tracks (where the tracks evolve from the bottom left to top right).
-    \begin{center}
-        \includegraphics[width=0.8\textwidth]{figures/central_T_rho.pdf}
-    \end{center}
-    From these tracks we can conclude that over time each of the stars is increasing in temperature and density within its core.
-For the $1 \unit{M_{\odot}}$ star, it moves from the ideal gas regime to the degenerate (but non-relativistic) regime during core helium burning.
-This makes sense as the core of the star at this stage would be a white dwarf, despite it still being ``living'' in the outer later.
-
-    The $12 \unit{M_{\odot}}$ star remains in the ideal gas regime all the way up until central helium exhaustion.
-Once helium is exhausted in the core, the star drifts into the extremely relativistic degenerate regime.
-This makes sense as we would expect this star to end its life as a neutron star.
-
-    Finally, the $120 \unit{M_{\odot}}$ star very closely follows the radiation-ideal gas boundary, particularly during its main sequence.
-After this it moves more clearly into the ideal gas regime and, though it appears to end its life here, the models only evolve through carbon burning.
-In reality, we would expect more complex evolution beyond this point and that the star would end as a black hole (given that its final mass is still around $30 \unit{M_{\odot}}$ even with all of the mass loss).
-}
-
-\pagebreak
-
-\question{\textbf{5.1 - Contribution to Opacity}}
-
-\question{Part a - Identifying regimes}
-\answer{
-    Electron scattering essentially dominates at any point at which the star is fully ionised.
-In terms of the plot this means that the electron scattering dominates in the far right at temperatures approximately above $10^{7.5} \unit{K}$.
-In this region the opacity is almost constant with increasing pressure.
-
-    For bound-free and free-free absorption, they dominate the opacity within temperatures of around $10^{5.5} - 10^{7.5} \unit{K}$.
-In this region we can see that the slopes are all approximately following $T^{-7/2}$ as we could expect.
-}
-
-\question{Part b - Check dependence}
-\answer{
-    For electron scattering the slope is 0 (there is no temperature dependence), as we would expect.\\\\
-    To check the temperature and density dependence for opacity with bound-free and free-free absorption we need to check that the slope is around $-7/2$.
-So just as a test, for the density of $10^{2} \unit{g}{cm^{-3}}$, the slope is
-    \begin{equation}
-        \mathrm{slope} = \frac{2 - 0.2}{6.5 - 7} = -3.6
-    \end{equation}
-    which is approximately $-7/2$, especially given that I pulled these numbers off the plot by eye!
-
-    We can also check the density dependence by checking the values at a fixed temperature.
-At a value of $\log T = 6.2$, the opacities of the densities with $\log \rho \in [0, -2, -4]$ are approximately $\log \kappa = [2, 0.9, -0.1]$.
-So the change in opacity is about the same for the same change in the density, thus showing a linear dependence as we would expect.
-}
-
-% \clearpage
-
-\question{\textbf{6.1 - Mass Luminosity Relation}}
-
-\question{Part a - Check relation for Appendix D}
-\answer{
-    I have plotted the masses and luminosities of the stars at different metallicities in the top panel of the figure below in order to compare against the mass-luminosity relation (using the same exponents as Figure 2.3 of the textbook), which I show as a black dotted line.
-    We can see that the mass-luminosity relation agrees well with the data for most stars.
-It is only for stars above $30 \unit{M_{\odot}}$ where the relation seems to no longer apply.
-This coincides with the mass where stars start to be radiation dominated as we talked about in class so perhaps this is the reason for the deviation.
-}
-
-\question{Part b - Luminosities}
-\answer{
-    As the question states, the luminosity of massive stars is about the same for both compositions, but the luminosity of the metal poor low-mass stars is \textit{higher} than those of $Z = 0.014$ stars.
-We can see this exactly in the figure above.
-Specifically, in the bottom panel I show the ratio of the two luminosities and we can see that for low mass stars, the luminosity of metal poor ones is higher.
-
-    The short answer to this trend is that \textbf{low-mass stars do not have electron scattering as a significant source of opacity}.
-To elaborate: low-mass stars tend to be cooler than higher-mass stars, and so they are less ionised and have less electron scattering.
-This means that the opacity of low-mass stars comes almost entirely from bound-free absorption (free-free absorption wouldn't be applicable since $\kappa_{bf} \gg \kappa_{ff}$ for a metal abundance of $Z > 0.001$).
-Now bound-free absorption is directly proportional to the metallicity and thus metal-poor low-mass stars have lower opacity (and so higher luminosity) than metal-rich low-mass stars.
-The same is not true for high-mass stars since electron scattering is independent of metallicity and so when the star is more metal-poor the electron scattering still limits the luminosity even when the bound-free absorption is low.
-}
-
-\begin{center}
-    \includegraphics[width=\textwidth]{figures/ml_relation.pdf}
-\end{center}
-
 \end{document}

### 2022-04-09 558_exoplanets/pset1/planet_finder.py

@@ -67,14 +67,6 @@ def solve_kepler_equation(e, M, tol=1e-10, max_steps=10000, verbose=False):
     raise ValueError("Did not converge")
 
 
-def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
-    M = 2 * np.pi / P * (t - t_p)
-    E = solve_kepler_equation(e, M)
-    f = 2 * np.arctan(((1 + e) / (1 - e))**(0.5) * np.tan(E / 2))
-
-    return k_star * (np.cos(omega + f) + e * np.cos(omega)) + gamma
-
-
 def test_kepler_solver(n_tests=10000, **kwargs):
     """Test whether the solver for Kepler's equation is working
 
@@ -110,6 +102,24 @@ def test_kepler_solver(n_tests=10000, **kwargs):
         fail_frac = len(should_be_zero[np.isclose(should_be_zero, np.zeros(n_tests))]) / len(should_be_zero)
         print(f"{RED}Ah sad, you failed, looks like {np.round(fail_frac * 100, 1)}% of tests failed{END}")
         return e, M, E
+
+
+def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
+    M = 2 * np.pi / P * (t - t_p)
+    E = solve_kepler_equation(e, M)
+    f = 2 * np.arctan(((1 + e) / (1 - e))**(0.5) * np.tan(E / 2))
+
+    return k_star * (np.cos(omega + f) + e * np.cos(omega)) + gamma
+
+
+class PlanetFinder():
+    def __init__(file):
+        self.file = file
+    
+        planet = pd.read_csv("data/mystery_planet01.txt", sep="\t", names=["time", "rv", "rv_err"])
+        self.data_time = planet["time"] - planet["time"].min()
+        self.data_rv = planet["rv"]
+        self.data_rv_err = planet["rv_err"]
 
 
 def main():

### 2022-04-09 558_exoplanets/pset1/planet_finder.py
add least squares stuff

@@ -1,5 +1,10 @@
 import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from matplotlib.colors import LogNorm
 import astropy.units as u
+import astropy.constants as const
+import emcee
 
 # ANSI colour codes
 RED = "\033[0;31m"
@@ -113,13 +118,28 @@ def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
 
 
 class PlanetFinder():
-    def __init__(file):
+    def __init__(self, file):
         self.file = file
-    
+
         planet = pd.read_csv("data/mystery_planet01.txt", sep="\t", names=["time", "rv", "rv_err"])
-        self.data_time = planet["time"] - planet["time"].min()
-        self.data_rv = planet["rv"]
-        self.data_rv_err = planet["rv_err"]
+        self.time = planet["time"] - planet["time"].min()
+        self.rv = planet["rv"]
+        self.rv_err = planet["rv_err"]
+        self.period_lsq = None
+
+    def least_squares_period(self, period_range):
+
+        least_squares = np.zeros_like(period_range)
+        for i, period in enumerate(period_range):
+            folded_period = self.time % period
+            order = np.argsort(folded_period)
+            folded_period = folded_period[order]
+            rvs = self.rv[order].values
+
+            least_squares[i] = np.sum(((rvs[1:] - rvs[:-1]))**2)
+
+        self.period_lsq = period_range[least_squares.argmin()]
+        return self.period_lsq
 
 
 def main():

### 2022-04-09 558_exoplanets/pset1/planet_finder.py
transfer everything into planet_finder

@@ -13,6 +13,23 @@ BOLD = "\033[1m"
 UNDERLINE = "\033[4m"
 END = "\033[0m"
 
+plt.rc('font', family='serif')
+plt.rcParams['text.usetex'] = False
+fs = 24
+
+# update various fontsizes to match
+params = {'figure.figsize': (12, 8),
+          'legend.fontsize': fs,
+          'axes.labelsize': fs,
+          'xtick.labelsize': 0.6 * fs,
+          'ytick.labelsize': 0.6 * fs,
+          'axes.linewidth': 1.1,
+          'xtick.major.size': 7,
+          'xtick.minor.size': 4,
+          'ytick.major.size': 7,
+          'ytick.minor.size': 4}
+plt.rcParams.update(params)
+
 
 def g(E, e, M):
     """ Kepler's equation in functional form (g(E, e, M) = 0) """
@@ -117,11 +134,45 @@ def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
     return k_star * (np.cos(omega + f) + e * np.cos(omega)) + gamma
 
 
+def get_rvs(times, P, k_star, t_p, gamma, omega, e):
+    return np.array([radial_velocity(P=P, k_star=k_star, t=t, t_p=t_p,
+                                     gamma=gamma, omega=omega, e=e) for t in times])
+
+
+def log_prior(theta):
+    bounds = [(110, 112), (0.2, 0.7), (10.2, 11.2), (-1, 1), (0.01, 2 * np.pi), (0, 1)]
+    for i in range(len(bounds)):
+        if theta[i] < bounds[i][0] or theta[i] >= bounds[i][1]:
+            return -np.inf
+    return 0.0
+
+
+def log_likelihood(theta, data_times, data_rvs, data_errors):
+    P, k_star, t_p, gamma, omega, e = theta
+    model_rvs = get_rvs(data_times, P=P, k_star=k_star, t_p=t_p,
+                        gamma=gamma, omega=omega, e=e)
+
+    sigma2 = data_errors**2
+    return -0.5 * np.sum((data_rvs - model_rvs)**2 / sigma2 + np.log(sigma2))
+
+
+def log_prob(theta, data_times, data_rvs, data_errors):
+    prior = log_prior(theta)
+    if not np.isfinite(prior):
+        return prior
+    else:
+        return log_likelihood(theta, data_times, data_rvs, data_errors) + prior
+
+
+def planet_mass(P, k_star, e, m_star, i):
[…]

### 2022-04-09 558_exoplanets/pset1/planet_finder.py
add a lot of comments, remove old files

@@ -5,6 +5,7 @@ from matplotlib.colors import LogNorm
 import astropy.units as u
 import astropy.constants as const
 import emcee
+import corner
 
 # ANSI colour codes
 RED = "\033[0;31m"
@@ -127,6 +128,30 @@ def test_kepler_solver(n_tests=10000, **kwargs):
 
 
 def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
+    """Find the radial velocity of a star
+
+    Parameters
+    ----------
+    k_star : `float`
+        Velocity semi-amplitude
+    omega : `float`
+        Pericenter angle
+    e : `float`
+        Eccentricity
+    t : `float`
+        Time
+    t_p : `float`
+        Time of Perihelion
+    P : `float`
+        Period
+    gamma : `float`
+        CoM radial velocity
+
+    Returns
+    -------
+    v_rv : `float`
+        Radial velocity
+    """
     M = 2 * np.pi / P * (t - t_p)
     E = solve_kepler_equation(e, M)
     f = 2 * np.arctan(((1 + e) / (1 - e))**(0.5) * np.tan(E / 2))
@@ -135,12 +160,50 @@ def radial_velocity(k_star, omega, e, t, t_p, P, gamma):
 
 
 def get_rvs(times, P, k_star, t_p, gamma, omega, e):
+    """Get radial velocities over time for a system
+
+    Parameters
+    ----------
+    times : `list`
+        Times at which to evaluate
+    k_star : `float`
+        Velocity semi-amplitude
+    omega : `float`
+        Pericenter angle
+    e : `float`
+        Eccentricity
+    t_p : `float`
+        Time of Perihelion
+    P : `float`
+        Period
+    gamma : `float`
+        CoM radial velocity
+
+    Returns
+    -------
+    v_rv : `np.array`
+        Radial velocities
+    """
     return np.array([radial_velocity(P=P, k_star=k_star, t=t, t_p=t_p,
                                      gamma=gamma, omega=omega, e=e) for t in times])
 
 
-def log_prior(theta):
-    bounds = [(110, 112), (0.2, 0.7), (10.2, 11.2), (-1, 1), (0.01, 2 * np.pi), (0, 1)]
+def log_prior(theta, bounds):
+    """Find the log of the prior using some bounds. Basically don't let the MCMC choose values out of these
+    bounds
+
+    Parameters
+    ----------
+    theta : `list`
+        List of the 6 parameters (P, k_star etc.)
[…]

### 2022-04-09 558_exoplanets/pset1/planet_finder.py
format the fit better and add instuctions

@@ -485,7 +485,9 @@ def main():
     initial_guesses = [finder.period_lsq, 0.5 * (finder.rv.max() - finder.rv.min()), 10.7, -0.1, 5, 0.92]
     bounds = [(110, 112), (0.2, 0.7), (10.2, 11.2), (-1, 1), (0.01, 2 * np.pi), (0, 1)]
     finder.run_mcmc(initial_guesses=initial_guesses, bounds=bounds)
-    print(f"Looks like the best fits are {GREEN}{BOLD}{finder.best_fit}{END}")
+
+    formatted_fit = ', '.join(['{:.2e}']*len(finder.best_fit)).format(*finder.best_fit)
+    print(f"Looks like the best fits are {BOLD}{GREEN}[" + formatted_fit + f"]{END}")
     finder.mcmc_corner_plot()
     finder.plot_best_fit()
 

### 2022-04-09 558_exoplanets/pset1/solutions.tex
add periodogram

@@
 
-\question{\textbf{Kepler Solver}}
+\noindent All code that I used for this homework can be found in \href{https://www.github.com/TomWagg/uw-grad-classes/blob/main/558_exoplanets/pset1/planet_finder.py}{\texttt{planet\_finder.py}}.
+The various functions handle the different questions and if you want to re-run it then you can just run \texttt{python planet\_finder.py} and it'll take about 10 minutes.\\
+
+\question{\textbf{Q1. Kepler Solver}}
 \answer{
+    I wrote a solver for Kepler's equation in the function \texttt{solve\_kepler\_equation()} and validated it by running 10,000 checks by putting the resulting eccentric anomaly back into the Kepler equation (this is done in \texttt{test\_kepler\_solver()}).
+}
 
+\question{\textbf{Q2. Radial Velocity Equation}}
+\answer{
+    This is implemented in \texttt{radial\_velocity()}. It is formulated slightly different from class because I figured I could simplify some of the trig by turning
+    \begin{equation}
+        v_{rv} = k_* \qty[ \cos f(t) \cos \omega - \sin f(t) \sin \omega + e \cos \omega ] + \gamma
+    \end{equation}
+    into the following by using the addition identities.
+    \begin{equation}
+        v_{rv} = k_* \qty[ \cos (f(t) + \omega) + e \cos \omega ] + \gamma
+    \end{equation}
+}
+
+\question{\textbf{Q3. Find planet period}}
+\answer{
+    I followed the method that Eric demonstrated in class in order to find the period.
+I chose a grid of periods over which to search and, for each of them, folded the data based on this period, sorted it by phase and then calculate the sum of the square of the difference between adjacent points.
+I plot this in a periodogram in Figure 1.
+    \begin{figure}
+        \centering
+        % \includegraphics[width=\textwidth]{figures/periodogram_FAKE.pdf}
+        \caption{Simple periodogram for the mysterious planet. I've highlighted the minimum with the grey bar which occurs around 111.47 days.}
+    \end{figure}
+    The minimum of this plot should give the period and so I find that the period of the planet is
+    \begin{equation}
+        \boxed{ P = 111.47 \unit{days} }
+    \end{equation}
+}
+
+\question{\textbf{4. Fit planet parameters}}
+\answer{
+    \begin{figure}
+        \centering
+        \includegraphics[width=\textwidth]{figures/mcmc_corner.pdf}
+        \caption{Corner plot of the various planet parameters that I fit using \texttt{emcee}}
+    \end{figure}
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.7\textwidth]{figures/best_fit.pdf}
+        \caption{Best fit to the data overlaid on top of the data folded using the best fitted period.}
+    \end{figure}
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.7\textwidth]{figures/planet_mass.pdf}
+        \caption{Contour plot of possible planet masses given the best fit parameters. Contours of shown for masses of Earth and Neptune.}
+    \end{figure}
 }

### 2022-04-09 558_exoplanets/pset1/planet_finder.py

@@ -335,6 +335,19 @@ class PlanetFinder():
             # calculate the sum of the squares for this period
             least_squares[i] = np.sum(((rvs[1:] - rvs[:-1]))**2)
 
+        fig, ax = plt.subplots()
+
+        ax.scatter(period_range, least_squares, s=0.5)
+
+        ax.axvline(period_range[least_squares.argmin()], color="grey", alpha=0.2, zorder=-1, lw=10)
+
+        ax.set_xlabel(r"Period, $P \, [{\rm days}]$")
+        ax.set_ylabel("Sum of squares of difference")
+
+        plt.savefig("figures/periodogram.pdf", format="pdf", bbox_inches="tight")
+
+        plt.show()
+
         # pick the minimum and return it
         self.period_lsq = period_range[least_squares.argmin()]
         return self.period_lsq
@@ -475,7 +488,7 @@ def main():
     print()
     print(f"{BOLD}{UNDERLINE}Problem 3 - Find period of planet{END}")
     finder = PlanetFinder(file="data/mystery_planet01.txt")
-    finder.least_squares_period(period_range=np.linspace(100, 120, 25000))
+    finder.least_squares_period(period_range=np.linspace(10, 500, 25000))
     print(f"Looks like the period of this planet is about {GREEN}{BOLD}{finder.period_lsq:1.4f}{END} days")
 
     print()

### 2022-04-10 558_exoplanets/pset1/solutions.tex
pset1 finished for exoplanets

--- 2022-04-09 558_exoplanets/pset1/solutions.tex
@@
 
-\question{\textbf{Kepler Solver}}
+\noindent All code that I used for this homework can be found in \href{https://www.github.com/TomWagg/uw-grad-classes/blob/main/558_exoplanets/pset1/planet_finder.py}{\texttt{planet\_finder.py}}.
+Just run \texttt{python planet\_finder.py} to reproduce all results.
+\\
+
+\question{\textbf{Q1. Kepler Solver}}
 \answer{
-
+    I wrote a solver for Kepler's equation in the function \texttt{solve\_kepler\_equation()} and validated it in \texttt{test\_kepler\_solver()} by putting the resulting eccentric anomalies back into the Kepler equation and checking that the equation still holds.
 }
 
+\question{\textbf{Q2. Radial Velocity Equation}}
+\answer{
+    This is implemented in \texttt{radial\_velocity()}.
+It is formulated slightly differently from class because I figured I could simplify some of the trig and write the formula instead as the following by using the addition identities.
+    \begin{equation}
+        v_{rv} = k_* \qty[ \cos (f(t) + \omega) + e \cos \omega ] + \gamma
+    \end{equation}
+    \vspace{-1cm}
+}
+
+\question{\textbf{Q3. Find planet period}}
+\answer{
+    I followed the method that Eric demonstrated in class in order to find the period.
+I chose a grid of periods over which to search and, for each of them, folded the data based on this period, sorted it by phase and then calculate the sum of the square of the difference between adjacent points.
+I plot this in a periodogram in the figure below.
+
+    \begin{center}
+        \includegraphics[width=0.55\textwidth]{figures/periodogram.pdf}
+    \end{center}
+    The minimum of this plot should give the period (which I highlight with a grey bar) and so I find that the period of the planet is
+    \begin{equation}
+        \boxed{ P = 111.47 \unit{days} }
+    \end{equation}
+}
+
+\question{\textbf{4. Fit planet parameters}}
+\answer{
+    Now for the main event!
+I fitted for the planet parameters $(P, k_*, t_p, \gamma, \omega, e)$ using \texttt{emcee}.
+I set the initial guess for $P$ using the answer from Q3.
+For $k_*$ I used $0.5(v_{rv, {\rm max}} - v_{rv, {\rm min}})$ from the definition.
+I guessed values for the other four just by playing around with them until the fit looked sort of close.
+This gave initial guesses as
+    \begin{equation}
+        \mathrm{initial\ guesses} = [111.47 \unit{days}, 0.4973 \unit{m}{s^{-1}}, 10.7 \unit{days}, -0.1 \unit{m}{s^{-1}}, 5 \unit{rad}, 0.92]
+    \end{equation}
+    for $P, k_*, t_p, \gamma, \omega, e$ respectively. I then applied the following bounds as a prior
+    \begin{equation}
+        \begin{split}
+            \mathrm{bounds} =
+                [&(110, 112) \unit{days}, (0.2, 0.7) \unit{m}{s^{-1}}, (10.2, 11.2) \unit{days},\\
+                &(-1, 1) \unit{m}{s^{-1}}, (0.01, 2 \pi) \unit{rad}, (0, 1)]
+        \end{split}
+    \end{equation}
+    Next I ran \texttt{emcee} using 16 walkers, with 5000 steps each and a 1000 step burn in.
+I've plotted the parameter distribution as a corner plot and \textbf{I put it at the end of this document}.
+
+    There are some interesting correlations.
+The one between $P$ and $t_p$ makes a lot of sense, if the period is longer then the sharp transition is also going to move.
+I'm less confident about the reasoning behind the correlations between $e$ and $k_*$ plus $\gamma$ but I \textit{think} this is because all of these parameters each tend to `stretch' the model in radial velocity and so they tend to have interdependencies?
+
+    To get the best fit for the parameters, I take the median values from each of these distributions. The best fit values are:
+    \begin{equation}
+        \boxed{ P, k_*, t_p, \gamma, \omega, e = 111.31 \unit{days}, 0.471 \unit{m}{s^{-1}}, 10.92 \unit{days}, -0.0019 \unit{m}{s^{-1}}, 5.251 \unit{rad}, 0.932}
+    \end{equation}
+    I plot this with the folded data (on the same period as the fit) in the plot below - and it shows that we have good agreement, fun!
+    \begin{center}
+        \includegraphics[width=0.8\textwidth]{figures/best_fit.pdf}
+    \end{center}
+    So now I got intrigued and wondered what this meant for the mass of the planet. Since we know that $k_*$ is defined as
+    \begin{equation}
+        k_* = \qty(\frac{2 \pi G}{P (1 - e^2)})^{1/3} \frac{M_p \sin \iota}{(M_* + M_p)^{2/3}},
+    \end{equation}
+    where $\iota$ is the inclination and $M_*$ is the star mass.
+It should be fairly reasonable to rearrange this to find the planet mass and make the assumption that $M_* \gg M_p$ (this worries me slightly but I figure it'll be fine except in extreme cases).
+This gives that the planet mass is
+    \begin{equation}
+        M_p = \frac{k_* M_*^{2/3}}{\sin \iota} \qty(\frac{2 \pi G}{P (1 - e^2)})^{-1/3},
+    \end{equation}
+    So now we know everything in this equation from the best fit except $\iota$ and $M_*$.
+I made a grid of these values and plotted the planet mass as a contour plot below along with some lines indicating Earth and Neptune.
+
+    \begin{center}
+        \includegraphics[width=\textwidth]{figures/planet_mass.pdf}
+    \end{center}
+
+    Eric hasn't given us any constraints on either of these parameters so it's hard to say more but if we could constrain the inclination to be larger than a small angle and had a guess for the stellar type then we'd quickly get a good idea for the planet mass from this plot!
+:D
+}
+
+\clearpage
+
+\question{\textbf{6. Gauss Functions}}
+
+\question{Part a - Conservation of Angular Momentum}
+\answer{
+    We are given that the Gauss functions are defined such that
+    \begin{align}
+        \va{x} &= f \va{x}_0 + g \va{v}_0 \\
+        \va{v} &= \dot{f} \va{x}_0 + \dot{g} \va{v}_0
+    \end{align}
+    Let's use these to show that when angular momentum is conserved then the relation given holds. If angular momentum is conserved then it must be the case that
+    \begin{equation}
+        \va{x}_0 \cross \va{v}_0 = \va{x} \cross \va{v}
+    \end{equation}
+    We can now apply the fact that the cross product is distributive and the following identities
+    \begin{align}
+        \va{a} \cross \va{a} &= 0, \\
+        \va{a} \cross \va{b} &= - \va{b} \cross \va{a},
+    \end{align}
+    to the conservation equation to show the required relation.
+    \begin{align}
+        \va{x}_0 \cross \va{v}_0 &= \va{x} \cross \va{v} \\
+                                 &= \qty(f \va{x}_0 + g \va{v}_0) \cross (\dot{f} \va{x}_0 + \dot{g} \va{v}_0) \\
+                                 &= (f \dot{f}) (\va{x}_0 \cross \va{x}_0) + (f \dot{g}) (\va{x}_0 \cross \va{v}_0) + (g \dot{f}) (\va{v}_0 \cross \va{x}_0) (g \dot{g}) (\va{v}_0 \cross \va{v}_0) \\
+                                 &= (f \dot{g}) (\va{x}_0 \cross \va{v}_0) - (g \dot{f}) (\va{x}_0 \cross \va{v}_0) \\
+        \va{x}_0 \cross \va{v}_0 &= (f \dot{g} - g \dot{f}) (\va{x}_0 \cross \va{v}_0) \\
+                               \Aboxed{ \implies 1 &= f \dot{g} - g \dot{f} }
+    \end{align}
+}
+
+\question{Part b - Kepler's Equation Satisfied}
+\answer{
+    Right then, let's get into the weeds! We are given that the Gauss functions are
+    \begin{align}
+        f &= \frac{a}{r_0} \qty(\cos (E - E_0) - 1) + 1 \\
+        g &= (t - t_0) + \frac{1}{n} \qty(\sin(E - E_0) - (E - E_0)) \\
+        \dot{f} &= -\frac{a^2}{r r_0} n \sin (E - E_0) \\
+        \dot{g} &= \frac{a}{r} \qty(\cos (E - E_0) - 1) + 1
+    \end{align}
+    We can write these in a nicer form by assuming that $E_0 = M_0 = 0$ and $M = n (t - t_0)$.
+    \begin{align}
+        f &= \frac{a}{r_0} \qty(\cos E - 1) + 1 \\
+        g &= \frac{1}{n} \qty(M + \sin E - E) \\
+        \dot{f} &= -\frac{a^2}{r r_0} n \sin E \\
+        \dot{g} &= \frac{a}{r} \qty(\cos E - 1) + 1
+    \end{align}
+    It will also be useful to apply
+    \begin{align}
+        M &= E - e \sin E \\
+        r &= a (1 - e \cos E) \\
+        r_0 &= a (1 - e)
+    \end{align}
+    And now we just plug the whole gaggle of equations into our relation from the previous question
+    \begin{align}
+        f \dot{g} - g \dot{f} &= \qty[\frac{a}{r_0} \qty(\cos E - 1) + 1] \qty[\frac{a}{r} \qty(\cos E - 1) + 1] - \qty[\frac{1}{n} \qty(M + \sin E - E)] \qty[-\frac{a^2}{r r_0} n \sin E] \\
+        &= \frac{a^2}{r r_0} \qty[ \qty(\qty(\cos E - 1) + \frac{r_0}{a}) \qty(\qty(\cos E - 1) + \frac{r}{a}) + \sin E \qty(M + \sin E - E)] \\
+        &= \frac{a^2}{r r_0} \qty[ \qty(\cos E - 1 + 1 - e) \qty(\cos E - 1 + 1 - e \cos E) + \sin E \qty(M + \sin E - E)] \\
+        &= \frac{a^2}{r r_0} \qty[ \qty(\cos E - e) \qty(\cos E - e \cos E) + \sin E \qty(\sin E - e \sin E)] \\
+        &= \frac{a^2}{r r_0} (1 - e) \qty[ \qty(\cos E - e) \cos E + \sin^2 E] \\
+        &= \frac{a^2}{r r_0} (1 - e) \qty[ (1 - e \cos E) (\cos^2 E + \sin^2 E)] \\
+        &= \frac{a^2}{r r_0} (1 - e) (1 - e \cos E) \\
+        &= \frac{a^2}{r r_0} \frac{r_0}{a} \frac{r}{a} \\
+        \Aboxed{ f \dot{g} - g \dot{f}&= 1 }
+    \end{align}
+}
+
+\clearpage
+
+\begin{center}
+    \section*{MCMC Corner Plot of Planet Parameters}
+    \includegraphics[width=\textwidth]{figures/mcmc_corner.pdf}
+\end{center}
+
 \end{document}

### 2022-04-10 558_exoplanets/pset1/planet_finder.py

@@ -499,7 +499,7 @@ def main():
     bounds = [(110, 112), (0.2, 0.7), (10.2, 11.2), (-1, 1), (0.01, 2 * np.pi), (0, 1)]
     finder.run_mcmc(initial_guesses=initial_guesses, bounds=bounds)
 
-    formatted_fit = ', '.join(['{:.2e}']*len(finder.best_fit)).format(*finder.best_fit)
+    formatted_fit = ', '.join(['{:.4e}']*len(finder.best_fit)).format(*finder.best_fit)
     print(f"Looks like the best fits are {BOLD}{GREEN}[" + formatted_fit + f"]{END}")
     finder.mcmc_corner_plot()
     finder.plot_best_fit()

### 2022-04-11 558_exoplanets/pset1/planet_finder.py
km/s not m/s, add star data

@@ -395,8 +395,8 @@ class PlanetFinder():
         plt.rcParams.update(params)
 
         corner.corner(
-            self.samples, labels=[r"$P\, [{\rm days}]$", r"$k_{*}\, [{\rm m/s}]$", r"$t_p\, [{\rm days}]$",
-                                  r"$\gamma\, [{\rm m / s}]$", r"$\omega \, [{\rm rad}]$", r"$e$"]
+            self.samples, labels=[r"$P\, [{\rm days}]$", r"$k_{*}\, [{\rm km / s}]$", r"$t_p\, [{\rm days}]$",
+                                  r"$\gamma\, [{\rm km / s}]$", r"$\omega \, [{\rm rad}]$", r"$e$"]
         )
 
         plt.savefig("figures/mcmc_corner.pdf", format="pdf", bbox_inches="tight")
@@ -436,7 +436,7 @@ class PlanetFinder():
         ax.axhline(0, color="grey", linestyle="dotted")
 
         ax.set_xlabel("Phase [days]")
-        ax.set_ylabel(r"$v_{rv} \, [{\rm m /s}]$")
+        ax.set_ylabel(r"$v_{rv} \, [{\rm km / s}]$")
         ax.legend()
 
         plt.savefig("figures/best_fit.pdf", format="pdf", bbox_inches="tight")
@@ -452,28 +452,39 @@ class PlanetFinder():
         MS, I = np.meshgrid(m_star_range, i_range)
 
         m_planets = planet_mass(self.best_fit[0] * u.day,
-                                self.best_fit[1] * u.m / u.s,
+                                self.best_fit[1] * u.km / u.s,
                                 self.best_fit[-1], i=I, m_star=MS).to(u.kg) / const.M_earth
 
-        cont = ax.contourf(i_range, m_star_range, m_planets.T, norm=LogNorm(),
-                           levels=np.logspace(np.log10(0.08), 3, 25), cmap="magma")
+        cont = ax.contourf(i_range.to(u.deg), m_star_range, m_planets.T, norm=LogNorm(),
+                   levels=np.logspace(np.log10(0.1), np.log10(20), 15), cmap="magma")
         cbar = fig.colorbar(cont)
+        cbar.set_ticks([0.1, 0.3, 1, 5, 20])
 
         ax.set_yscale("log")
-        ax.set_xlim(0, np.pi/2)
+        ax.set_xlim(0, 90)
[…]

### 2022-04-12 558_exoplanets/pset1/planet_finder.py
dividing by the wrong mass

@@ -453,7 +453,7 @@ class PlanetFinder():
 
         m_planets = planet_mass(self.best_fit[0] * u.day,
                                 self.best_fit[1] * u.km / u.s,
-                                self.best_fit[-1], i=I, m_star=MS).to(u.kg) / const.M_earth
+                                self.best_fit[-1], i=I, m_star=MS).to(u.kg) / const.M_jup
 
         cont = ax.contourf(i_range.to(u.deg), m_star_range, m_planets.T, norm=LogNorm(),
                    levels=np.logspace(np.log10(0.1), np.log10(20), 15), cmap="magma")
@@ -470,9 +470,6 @@ class PlanetFinder():
 
         ax.errorbar(x=89.6, y=0.98, xerr=0.4, yerr=0.1, color="black", lw=2)
 
-        big_names = ax.contour(i_range.to(u.deg), m_star_range, m_planets.T, norm=LogNorm(),
-                               levels=[0.3, 1, 12.7], colors="white", linewidths=2)
-
         def planet_labels(x):
             if x == 0.3:
                 return "Saturn"
@@ -482,6 +479,9 @@ class PlanetFinder():
                 return "Kepler-25b"
             else:
                 return "ERROR"
+
+        big_names = ax.contour(i_range.to(u.deg), m_star_range, m_planets.T, norm=LogNorm(),
+                               levels=[0.3, 1, 12.7], colors="white", linewidths=2)
 
         ax.clabel(big_names, [0.3, 1, 12.7], fmt=planet_labels,
                   use_clabeltext=True, fontsize=0.7 * fs, manual=[(30, 1), (60, 5e-2), (80, 1.2e-2)])

### 2022-04-12 558_exoplanets/pset1/solutions.tex
fix units and whatnot

@@
     \begin{equation}
-        \mathrm{initial\ guesses} = [111.47 \unit{days}, 0.4973 \unit{m}{s^{-1}}, 10.7 \unit{days}, -0.1 \unit{m}{s^{-1}}, 5 \unit{rad}, 0.92]
+        \mathrm{initial\ guesses} = [111.47 \unit{days}, 0.4973 \unit{km}{s^{-1}}, 10.7 \unit{days}, -0.1 \unit{km}{s^{-1}}, 5 \unit{rad}, 0.92]
     \end{equation}
@@
             \mathrm{bounds} =
-                [&(110, 112) \unit{days}, (0.2, 0.7) \unit{m}{s^{-1}}, (10.2, 11.2) \unit{days},\\
-                &(-1, 1) \unit{m}{s^{-1}}, (0.01, 2 \pi) \unit{rad}, (0, 1)]
+                [&(110, 112) \unit{days}, (0.2, 0.7) \unit{km}{s^{-1}}, (10.2, 11.2) \unit{days},\\
+                &(-1, 1) \unit{km}{s^{-1}}, (0.01, 2 \pi) \unit{rad}, (0, 1)]
         \end{split}
@@
     \begin{equation}
-        \boxed{ P, k_*, t_p, \gamma, \omega, e = 111.31 \unit{days}, 0.471 \unit{m}{s^{-1}}, 10.92 \unit{days}, -0.0019 \unit{m}{s^{-1}}, 5.251 \unit{rad}, 0.932}
+        \boxed{ P, k_*, t_p, \gamma, \omega, e = 111.31 \unit{days}, 0.471 \unit{km}{s^{-1}}, 10.92 \unit{days}, -0.0019 \unit{km}{s^{-1}}, 5.251 \unit{rad}, 0.932}
     \end{equation}
@@
 
-    Eric hasn't given us any constraints on either of these parameters so it's hard to say more but if we could constrain the inclination to be larger than a small angle and had a guess for the stellar type then we'd quickly get a good idea for the planet mass from this plot!
-:D
+    After some quick googling of the parameters that I found, it looks like our mystery planet is none other than HD 80606b!
+In this case, I can use the values of $M_* = 0.98 \pm 0.1 \unit{M_\odot}$ and $\iota = 89.6 \pm 0.4$ from \href{https://arxiv.org/abs/0902.4457}{Table 1 of this paper}.
+Plugging these into the equation above give that the mass of the planet should be
+    \begin{equation}
+        M_p \approx 5.59 \unit{M_{\rm Jup}}
+    \end{equation}
+    The real answer is $4.0 \pm 0.3 \unit{M_{\rm Jup}}$...but that's not far off given this is the most basic fit I could come up with in just a couple of hours, very fun! :D
 }

### 2022-04-12 558_exoplanets/pset1/solutions.tex
update text given plot changes

@@
     So now we know everything in this equation from the best fit except $\iota$ and $M_*$.
-I made a grid of these values and plotted the planet mass as a contour plot below along with some lines indicating Earth and Neptune.
+I made a grid of these values and plotted the planet mass as a contour plot below along with some lines indicating Saturn, Jupiter and Kepler-25b (one of the most massive planets found).
 
@@
     After some quick googling of the parameters that I found, it looks like our mystery planet is none other than HD 80606b!
-In this case, I can use the values of $M_* = 0.98 \pm 0.1 \unit{M_\odot}$ and $\iota = 89.6 \pm 0.4$ from \href{https://arxiv.org/abs/0902.4457}{Table 1 of this paper}.
+In this case, I can use the values of $M_* = 0.98 \pm 0.1 \unit{M_\odot}$ and $\iota = 89.6 \pm 0.4$ from \href{https://arxiv.org/abs/0902.4457}{Table 1 of this paper} (shown as black point with errorbars on plot).
 Plugging these into the equation above give that the mass of the planet should be

### 2022-04-12 531_stars/pset2/solutions.tex
first two questions of pset2 stars

--- 2022-04-01 531_stars/pset1/solutions.tex
@@
 \usepackage{xcolor}
+\usepackage{mhchem}
 \hypersetup{
@@
 
-\title{Astro 531; Problem Set 1}
+\newcommand{\psetnum}{2}
+\newcommand{\class}{ASTR 531 - Stellar Interiors and Evolution}
+
+\newcommand{\tomtitle}{
+    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
+    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
+    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
+}
+
+\title{\class : Problem Set \psetnum}
 \author{\textbf{Tom Wagg}}
@@
 
-\maketitle
+\tomtitle
 
-\question{\textbf{2.1 - Stellar Properties}}
-
-\question{Part a - Radi  us}
+\question{\textbf{8.1 - Mass Defect Fraction}}
 \answer{
-    \begin{align}
-        L &= 4 \pi R^2 \sigma T_{\rm eff}^4 \\
-        R &= \sqrt{\frac{L}{4 \pi \sigma T_{\rm eff}^4}} \\
-        \Aboxed{ R &= 2570 \unit{R_{\odot}} }
-    \end{align}
+    The mass defects of the respective equations are as follows
+    \begin{center}
+        \begin{tabular}{c|c|c}
+            Phase & Equation & $\Delta m / m$\\
+            \hline
+            Hydrogen & $4 \ce{^{1}H} \to \ce{^{4}He}$ & 0.007145 \\
+            Helium & $3 \ce{^{4}He} \to \ce{^{12}C}$ & 0.000650 \\
+            Carbon & $\ce{^{12}C} + \ce{^{4}He} \to \ce{^{16}O}$ & 0.000481 \\
+            Oxygen & $2 \ce{^{16}O} \to \ce{^{28}Si} + \ce{^{4}He}$ & 0.000322 \\
+            Silicon & $2 \ce{^{28}Si} \to \ce{^{56}Fe}$ & 0.000338 \\
+        \end{tabular}
+    \end{center}
+    You can see that the trend is that subsequent reactions have lower mass defect fractions and produce less energy per reaction (with the exception of Silicon vs.
+Oxygen fusion.) Since less energy is available to counter gravity in each phase, this implies that stars will spend the majority of their time in the hydrogen fusion phase and progressively less time in each fusion phase.
 }
 
-\question{Part b - Density}
+\question{\textbf{8.4 - Minimum Core Masses}}
 \answer{
-    If we assume that the star is a constant density throughout then we can use the definition of density and the radius we calculated above.
-    \begin{align}
-        \rho &= \frac{M}{\frac{4}{3} \pi R^3} \\
-        \Aboxed{ \rho &= 1.66 \times 10^{-6} \unit{kg}{m^{-3}} }
-    \end{align}
+    Equation 8.24 in the textbook gives that the minimum core mass required to initiate a fusion phase with ignition temperature $T_{\rm ign}$ is
+    \begin{equation}
+        M_{\rm crit} \approx \qty[ \frac{\mathcal{R} K_1}{\mu_c \mu_e^{5/3} G^2} \cdot T_{\rm ign}]^{3/4}
+    \end{equation}
+    The only variables here are $\mu_c, \mu_e$ and $T_{\rm ign}$.
+The textbook gives us that $\mu_e = \mu_c = 2$ for all fusion phases after Helium.
+Thus, given that the minimum core mass for Helium fusion is $0.3 \unit{M_{\odot}}$ (and that we know the ignition temperature is $10^8 \unit{K}$ from Table 8.4), the core masses required for the subsequent phases will be
+    \begin{equation}
+        M_{\rm crit} \approx 0.3 \unit{M_{\odot}} \cdot \qty(\frac{T_{\rm ign}}{10^{8} \unit{K}})^{3/4}
+    \end{equation}
+    So we can tabulate all of this for brevity
+    \begin{center}
+        \begin{tabular}{c|c|c}
+            Phase & $T_{\rm ign} / \unit{K}$ & $M_{\rm crit} / \unit{M_{\odot}} $\\
+            \hline
+            Helium & $10^8$ & 0.3 \\
+            Carbon & $6 \times 10^8$ & 1.15 \\
+            Neon & $9 \times 10^8$ & 1.56 \\
+            Oxygen & $10^9$ & 1.69 \\
+            Silicon & $3 \times 10^9 $ & 3.85 \\
+        \end{tabular}
+    \end{center}
 }
 
-\question{Part c - Escape velocity}
-\answer{
-    \begin{equation}
-        \boxed{ v_{\rm esc} = \sqrt{ \frac{2 G M}{R} } = 54500 \unit{m}{s^{-1}} }
-    \end{equation}
-}
-
-\question{Part d - Comparisons to Sun}
-\answer{
-    This star is clearly much larger than the sun (by a factor of 2570) but is \textbf{much} less dense (since the mean density of the sun is approximately $1400 \unit{kg}{m^{-3}}$, so the density is $10^{9}$ times lower than the sun).
-Finally, the escape velocity is lower than the sun (around 0.088 times lower).
-}
-
-\pagebreak
-
-\question{\textbf{3.5 - Kepler's 3rd Law from Virial Theorem}}
-\answer{
-    The virial theorem gives the following relations between energies
-    \begin{equation}
-        E_{\rm kin} = - \frac{1}{2} E_{\rm pot}
-    \end{equation}
-    For the orbits of planets around the sun, their potential energy will come from the gravitational potential and so
-    \begin{equation}
-        E_{\rm pot} = - \frac{G m_1 m_2}{a},
-    \end{equation}
-    and the kinetic energy is simply
-    \begin{equation}
-        E_{\rm kin} = \frac{1}{2} m_1 v_1^2 + \frac{1}{2} m_2 v_2^2.
-    \end{equation}
-    For the next part, it will be useful to define the distance from each body to the centre of mass, $a_i$, in terms of the masses
-    \begin{equation}
-        a_i = \frac{a \cdot m_{1 - i}}{m_1 + m_2} \\
-    \end{equation}
-    and additionally, it's useful to write the velocity in terms of the period and masses
-    \begin{align}
-        v_i &= \frac{2 \pi a_i}{P} \\
-        v_i &= \frac{2 \pi}{P} \frac{a \cdot m_{1 - i}}{m_1 + m_2}
-    \end{align}
-    Now we can put this all together in the expression for the virial theorem.
-    \begin{align}
-        \frac{1}{2} m_1 v_1^2 + \frac{1}{2} m_2 v_2^2 &= \frac{1}{2} \cdot \frac{G m_1 m_2}{a}, \\
-        \frac{1}{2} m_1 \qty(\frac{2 \pi}{P} \frac{a \cdot m_2}{m_1 + m_2})^2 + \frac{1}{2} m_2  \qty(\frac{2 \pi}{P} \frac{a \cdot m_1}{m_1 + m_2})^2 &= \frac{G m_1 m_2}{2 a} \\
-        \frac{2 \pi^2 a^2}{P^2 M^2} \qty( m_1 m_2^2 + m_2 m_1^2 ) &= \frac{G m_1 m_2}{2 a} \\
-        \frac{a^3}{P^2} &= \frac{G}{4 \pi^2} \frac{M^2 m_1 m_2}{\qty( m_1 m_2^2 + m_2 m_1^2 )} \\
-        \frac{a^3}{P^2} &= \frac{G}{4 \pi^2} \frac{M^2}{\qty( m_1 + m_2 )} \\
-        \Aboxed{ \frac{a^3}{P^2} &= \frac{G M}{4 \pi^2} }
-    \end{align}
-}
-
-\pagebreak
-
-\question{\textbf{4.4 - Central Temperature-Density}}
-\answer{
-    Here's a plot of the three different stars on (pretty much) the same figure as the textbook (I made the boundaries using Onno Pols' lecture notes but shaded them the same way as the textbook).
-I've also indicated the main sequence turnoff and the exhaustion of central helium as dots in the tracks (where the tracks evolve from the bottom left to top right).
-    \begin{center}
-        \includegraphics[width=0.8\textwidth]{figures/central_T_rho.pdf}
-    \end{center}
-    From these tracks we can conclude that over time each of the stars is increasing in temperature and density within its core.
-For the $1 \unit{M_{\odot}}$ star, it moves from the ideal gas regime to the degenerate (but non-relativistic) regime during core helium burning.
-This makes sense as the core of the star at this stage would be a white dwarf, despite it still being ``living'' in the outer later.
-
-    The $12 \unit{M_{\odot}}$ star remains in the ideal gas regime all the way up until central helium exhaustion.
-Once helium is exhausted in the core, the star drifts into the extremely relativistic degenerate regime.
-This makes sense as we would expect this star to end its life as a neutron star.
-
-    Finally, the $120 \unit{M_{\odot}}$ star very closely follows the radiation-ideal gas boundary, particularly during its main sequence.
-After this it moves more clearly into the ideal gas regime and, though it appears to end its life here, the models only evolve through carbon burning.
-In reality, we would expect more complex evolution beyond this point and that the star would end as a black hole (given that its final mass is still around $30 \unit{M_{\odot}}$ even with all of the mass loss).
-}
-
-\pagebreak
-
-\question{\textbf{5.1 - Contribution to Opacity}}
-
-\question{Part a - Identifying regimes}
-\answer{
-    Electron scattering essentially dominates at any point at which the star is fully ionised.
-In terms of the plot this means that the electron scattering dominates in the far right at temperatures approximately above $10^{7.5} \unit{K}$.
-In this region the opacity is almost constant with increasing pressure.
-
-    For bound-free and free-free absorption, they dominate the opacity within temperatures of around $10^{5.5} - 10^{7.5} \unit{K}$.
-In this region we can see that the slopes are all approximately following $T^{-7/2}$ as we could expect.
-}
-
-\question{Part b - Check dependence}
-\answer{
-    For electron scattering the slope is 0 (there is no temperature dependence), as we would expect.\\\\
-    To check the temperature and density dependence for opacity with bound-free and free-free absorption we need to check that the slope is around $-7/2$.
-So just as a test, for the density of $10^{2} \unit{g}{cm^{-3}}$, the slope is
-    \begin{equation}
-        \mathrm{slope} = \frac{2 - 0.2}{6.5 - 7} = -3.6
-    \end{equation}
-    which is approximately $-7/2$, especially given that I pulled these numbers off the plot by eye!
-
-    We can also check the density dependence by checking the values at a fixed temperature.
-At a value of $\log T = 6.2$, the opacities of the densities with $\log \rho \in [0, -2, -4]$ are approximately $\log \kappa = [2, 0.9, -0.1]$.
-So the change in opacity is about the same for the same change in the density, thus showing a linear dependence as we would expect.
-}
-
-% \clearpage
-
-\question{\textbf{6.1 - Mass Luminosity Relation}}
-
-\question{Part a - Check relation for Appendix D}
-\answer{
-    I have plotted the masses and luminosities of the stars at different metallicities in the top panel of the figure below in order to compare against the mass-luminosity relation (using the same exponents as Figure 2.3 of the textbook), which I show as a black dotted line.
-    We can see that the mass-luminosity relation agrees well with the data for most stars.
-It is only for stars above $30 \unit{M_{\odot}}$ where the relation seems to no longer apply.
-This coincides with the mass where stars start to be radiation dominated as we talked about in class so perhaps this is the reason for the deviation.
-}
-
-\question{Part b - Luminosities}
-\answer{
-    As the question states, the luminosity of massive stars is about the same for both compositions, but the luminosity of the metal poor low-mass stars is \textit{higher} than those of $Z = 0.014$ stars.
-We can see this exactly in the figure above.
-Specifically, in the bottom panel I show the ratio of the two luminosities and we can see that for low mass stars, the luminosity of metal poor ones is higher.
-
-    The short answer to this trend is that \textbf{low-mass stars do not have electron scattering as a significant source of opacity}.
-To elaborate: low-mass stars tend to be cooler than higher-mass stars, and so they are less ionised and have less electron scattering.
-This means that the opacity of low-mass stars comes almost entirely from bound-free absorption (free-free absorption wouldn't be applicable since $\kappa_{bf} \gg \kappa_{ff}$ for a metal abundance of $Z > 0.001$).
-Now bound-free absorption is directly proportional to the metallicity and thus metal-poor low-mass stars have lower opacity (and so higher luminosity) than metal-rich low-mass stars.
-The same is not true for high-mass stars since electron scattering is independent of metallicity and so when the star is more metal-poor the electron scattering still limits the luminosity even when the bound-free absorption is low.
-}
-
-\begin{center}
-    \includegraphics[width=\textwidth]{figures/ml_relation.pdf}
-\end{center}
-
 \end{document}

### 2022-04-12 531_stars/pset2/solutions.tex
stars pset2 q3

@@
 
+\question{\textbf{9.1 - Typical Timescales}}
+\answer{
+    For calculating the timescales based on $M, R, L$, I use the following equations from the textbook (specifically, 9.3, 9.4, 9.5d)
+    \begin{equation}
+        \tau_{\rm dyn} = \sqrt{\frac{1}{G \bar{\rho}}}, \qquad \tau_{\rm KH} = \frac{G M^2}{R L}, \qquad \tau_{\rm nuc} \approx \frac{M / \unit{M_{\odot}}}{L / \unit{L_{\odot}}} 10^{10} \unit{yr}
+    \end{equation}
+    Specifically, this gives the following values
+    \begin{center}
+        \begin{tabular}{c||c|c|c||c}
+            Star & $\tau_{\rm dyn} / \unit{s}$ & $\tau_{\rm KH} / \unit{s}$ & $\tau_{\rm nuc} / \unit{s}$ & Ratios \\
+            \hline
+            1$\unit{M_{\odot}}$ MS & $3.3 \times 10^{03}$ & $9.9 \times 10^{14}$ & $3.2 \times 10^{17}$ & $1 : 3.0 \times 10^{11} : 9.7 \times 10^{13}$ \\
+            60$\unit{M_{\odot}}$ MS & $2.4 \times 10^{04}$ & $3.0 \times 10^{11}$ & $2.4 \times 10^{13}$ & $1 : 1.2 \times 10^{07} : 9.7 \times 10^{08}$\\
+            15$\unit{M_{\odot}}$ RSG & $1.6 \times 10^{08}$ & $1.5 \times 10^{08}$ & $1.1 \times 10^{13}$ & $1.1 : 1 : 7.0 \times 10^{04}$ \\
+            0.6$\unit{M_{\odot}}$ WD & $5.5 \times 10^{00}$ & $3.0 \times 10^{19}$ & N/A & $1 : 5.4 \times 10^{18}$ \\
+        \end{tabular}
+    \end{center}
+    There's a couple of things to note here.
+First, in the general case of a lower mass main sequence star, the dynamical timescale is far shorter than both other timescales meaning that these stars are always in quasi-hydrostatic equilibrium.
+The thermal timescale is also shorter than the nuclear timescale and so the star is in thermal equilibrium for most of its life.
+
+    For higher mass main sequence stars, the nuclear timescale is significantly reduced (by a factor of 10,000), thus higher mass stars will run out of fuel much more quickly.
+The thermal timescale is also relatively closer to this nuclear timescale so we'd expect this star to spend more of its time out of thermal equilibrium.
+
+    For RSGs, an interesting thing to note if that the dynamical timescale is actually slightly longer than the thermal timescale!
+For WDs the dynamical timescale is incredibly short (5 seconds!), whilst the thermal timescale is much longer than for any of the other stars.
+}
+
 \end{document}

### 2022-04-12 531_stars/pset2/solutions.tex
stars pset2 q4

@@
 
+\question{\textbf{9.2 - Hey, where'd the sun go!?}}
+\answer{
+    If the sun were to suddenly stop nuclear fusion then how it would take us to realise would depend on what we were measuring.
+If we happen to have a fancy neutrino detector then the answer would be the photon travel time since the neutrinos would immediately stop appearing - I feel as though people would probably assume instrumental errors rather than the sun giving up on fusion though ;)
+
+    Assuming we're just using good old EM detectors, then I think the answer is instead the \textbf{dynamical timescale}.
+We can immediately rule out the nuclear timescale since that's only relevant with nuclear fuel running out (rather than being arbitrarily cut off as in this case).
+One may argue for the thermal timescale, since this is how long it would actually take for the Sun's luminosity to reflect its lack of internal energy production.
+However, upon the removal of the central energy source, the sun would start to contract on the dynamical timescale to remain in hydrostatic equilibrium and thus we could measure this contraction and notice the sun has broken in this time (about an hour for the sun).
+}
+
 \end{document}

### 2022-04-13 531_stars/pset2/solutions.tex
start of stars pset2 q5
adjust todo
finished with stars pset2

--- 2022-04-12 531_stars/pset2/solutions.tex
@@
     The only variables here are $\mu_c, \mu_e$ and $T_{\rm ign}$.
-The textbook gives us that $\mu_e = \mu_c = 2$ for all fusion phases after Helium.
-Thus, given that the minimum core mass for Helium fusion is $0.3 \unit{M_{\odot}}$ (and that we know the ignition temperature is $10^8 \unit{K}$ from Table 8.4), the core masses required for the subsequent phases will be
+The textbook gives us that $\mu_e = \mu_c = 2$ for all fusion phases after helium.
+Thus, given that the minimum core mass for helium fusion is $0.3 \unit{M_{\odot}}$ (and that we know the ignition temperature is $10^8 \unit{K}$ from Table 8.4), the core masses required for the subsequent phases will be
     \begin{equation}
@@
     There's a couple of things to note here.
-First, in the general case of a lower mass main sequence star, the dynamical timescale is far shorter than both other timescales meaning that these stars are always in quasi-hydrostatic equilibrium.
+First, in the general case of a lower mass main sequence star, the dynamical timescale is far shorter than both other timescales, meaning that these stars are always in quasi-hydrostatic equilibrium.
 The thermal timescale is also shorter than the nuclear timescale and so the star is in thermal equilibrium for most of its life.
@@
 
+\question{\textbf{9.2 - Hey, where'd the sun go!?}}
+\answer{
+    Well it depends what we are measuring.
+If we happen to have a fancy neutrino detector then the answer would be the photon travel time since the neutrinos would immediately stop appearing - though I think people would probably assume some sort of instrument error at first, rather than the sun kicking the hypothetical bucket ;)
+
+    Assuming we're just using good old EM detectors, then I think the answer is instead the \textbf{dynamical timescale}.
+We can immediately rule out the nuclear timescale since that's only relevant with nuclear fuel running out (rather than being arbitrarily cut off).
+One may argue for the thermal timescale, since this is how long it would actually take for the Sun's luminosity to reflect its lack of nuclear energy production.
+However, upon the removal of the central energy source, the sun would start to contract on the dynamical timescale and thus we could measure this contraction and notice the sun has broken in this time (about an hour for the sun).
+}
+
+\question{\textbf{11.1 - Density structure}}
+\answer{
+    A polytrope is defined to obey the relation $P \sim \rho^{\gamma}$, or equivalently $\rho \sim P^{1/\gamma}$.
+Therefore, a star with $\gamma =4/3$ has a more concentrated density structure than a star with $\gamma =5/3$.
+We can see this is true because $\rho$ follows a steeper power law with pressure when $\gamma = 4/3$ and thus more of the mass will be concentrated towards the centre and decrease quickly towards the surface when pressure declines.
+}
+
 \end{document}

### 2022-04-25 531_stars/exam1/exam.tex
start writing the exam 1 for stars

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{mhchem}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}
\usepackage[shortlabels]{enumitem}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\psetnum}{1}
\newcommand{\class}{ASTR 531 - Stellar Interiors and Evolution}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Exam \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Problem Set \psetnum}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\tomtitle

\vspace{-1.5cm}

\section*{Thriving under high pressure}
\question{Dearest Grad Student,\\

\noindent My name is Professor Daniel Eyemund and I'm reaching out to you with an exciting business idea - diamonds! I've realised that I could get around applying for grants by generating a large number of diamonds and discretely selling them! They might be rare on Earth but with the right conditions perhaps we could create a \textbf{bunch} of them out in space. The plan is as follows:
\begin{enumerate}[nosep]
    \item Collect a big cloud of carbon
    \item Squish it down and inject it into a white dwarf
    \item Create diamonds
    \item \dots
    \item Profit?
\end{enumerate}
Your mission, should you choose to accept it (\dots which you have to because this is an exam), is to find out whether a white dwarf could produce diamonds, if so, what kind of white dwarf we need! I feel your experiences with stellar interiors, not to mention high pressure environments, would make you ideally suited to this task. Please note your timely response would be appreciated as I will be meeting with Elon to pitch the idea in 30 minutes.\\\\
Yours sincerely,

\noindent Prof. D. Eyemund}
\answer{}

\question{\textbf{Part a - Complete degeneracy pressure for electrons}}

\question{Given that, for complete degeneracy, the $n_e(p)$ distribution is rectangular for $p < p_{\rm F}$ such that $n_e(p) = 2 / h^3$, show that the pressure scales as $P_e \propto n_e^{5/3}$.}

\answer{}

\question{\textbf{Part b - Polytrope central pressure}}

\question{By showing that part a is true, you've proven that we can use an $n = 3/2$ polytrope! It'll now be useful to know an expression for the central pressure of a polytrope star.\\

\noindent Use the fact that $P_c = K \rho_c^{1 + 1/n}$ to show that $P_c = A \frac{G M^2}{R^4}$ (and find the constant $A$ in terms of $D_n, \pi, n, M_n$). Hint: You can find $K$ by comparing an expression for $\alpha$ and an expression for $M_n$. Additionally note that $\rho_c = \bar{\rho} D_n$.}

\answer{
    To start, we need to find an expression for $K$. We have the definition of $\alpha$ from the Lane-Emden equation (see Eq. 11.5).
    \begin{equation}
        \alpha = \qty[\frac{(n + 1) K}{4 \pi G \rho_c^{1 - 1/n}}]^{1/2}
    \end{equation}
    Next we can use Equation 11.10a to find another expression for $\alpha$ from the definition of $M_n$
    \begin{align}
        M &= 4 \pi a^3 \rho_c M_n \\
        \alpha &= \qty(\frac{M}{M_n 4 \pi \rho_c})^{1/3}
    \end{align}
    Combining these yields an expression for $K$
    \begin{align}
        \qty[\frac{(n + 1) K}{4 \pi G \rho_c^{1 - 1/n}}]^{1/2} &= \qty(\frac{M}{M_n 4 \pi \rho_c})^{1/3} \\
        K &= \qty(\frac{M}{M_n 4 \pi \rho_c})^{2/3} \frac{4 \pi G \rho_c^{1 - 1 / n}}{(n + 1)} \\
        K &= \qty(\frac{M}{M_n})^{2/3} \frac{(4 \pi)^{1/3} G \rho_c^{1/3 - 1 / n}}{(n + 1)}
    \end{align}
    Now, given that $P_c = K \rho_c^{1 + 1/n}$, this means that we can write the central pressure as
    \begin{align}
        P_c &= \qty(\frac{M}{M_n})^{2/3} \frac{(4 \pi)^{1/3} G \rho_c^{1/3 - 1 / n}}{(n + 1)} \rho_c^{1 + 1 / n} \\
            &= \qty(\frac{M}{M_n})^{2/3} \frac{(4 \pi)^{1/3} G \rho_c^{4/3}}{(n + 1)} \\
            &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] M^{2/3} \rho_c^{4/3}
    \end{align}
    We also know that the average density is given by $\bar{\rho} = \frac{M}{\frac{4}{3} \pi R^3}$ and it is related to the central density as $\rho_c = \bar{\rho} D_n$. Therefore we can express the central pressure
    \begin{align}
        P_c &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] M^{2/3} \qty(\frac{M D_n}{\frac{4}{3} \pi R^3})^{4/3} \\
            &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] \qty(\frac{D_n}{\frac{4}{3} \pi})^{4/3} \frac{M^2}{R^4} \\
        P_c &= \underbrace{\qty[\frac{(3 D_n)^{4/3}}{4 \pi (n + 1) M_n^{2/3}}]}_{=A} \frac{G M^2}{R^4}
    \end{align}
}

\question{\textbf{Part c - White dwarf $\mathbf{M}$-$\mathbf{P_c}$ relation}}

\question{Now let's take that and create a relation for the central pressure of a white dwarf as only a function of mass. Apply the $M$-$R$ relation for WDs (assuming $\mu_e = 2$) and that $n = 3/2$ to show that $P_c = \frac{A G}{B^4} M^{C}$ where $B$ and $C$ are constants you should find.}

\answer{
    The WD $M$-$R$ relation with $\mu_e = 2$ gives that
    \begin{equation}
        R = 0.012 \unit{R_{\odot}} \qty(\frac{M}{\unit{M_{\odot}}})^{-1/3}
    \end{equation}
    This gives the final result of the central pressure as a function of WD mass
    \begin{equation}
        \boxed{ P_{c, {\rm WD}} = \frac{A G}{B^4} M^{C} }
    \end{equation}
    where $B = 0.012 \unit{R_{\odot}} \unit{M_{\odot}}^{1/3}$ and $C = 10/3$.
}

\end{document}

### 2022-04-25 531_stars/exam1/exam.tex
finish up first draft of exam

@@
 
+\question{\textbf{Part d - Minimum white dwarf mass to produce diamonds}}
+
+\question{Carbon requires a pressure of at least $\sim 10^{10} \unit{Pa}$ to form diamonds.
+Find the minimum mass a white dwarf would need to convert carbon to diamonds in its core (assuming $A = 0.77$, $B=1.05 \unit{m}{kg^{1/3}}$, $C=10/3$ and $G = 6.67 \times 10^{-11} \unit{m^3}{kg^{-1}}{s^2}$).\\
+
+\noindent What does this imply for which sorts of white dwarfs could work for Prof.\ Eyemund?
+How would you answer change qualitatively if the professor could only inject carbon near the surface rather than at the centre?}
+\answer{}
+
+\question{\textbf{Extra Credit - The Verdict}}
+
+\question{Craft a response to Prof.\ Eyemund summarising the situation. Bonus points will be awarded for each dreadful pun.}
+
+\answer{
+    The solution is left as an exercise for the reader.
+}
+
 \end{document}

### 2022-04-26 531_stars/exam1/exam.tex
finished first exam

@@
     \item Collect a big cloud of carbon
-    \item Squish it down and inject it into a white dwarf
-    \item Create diamonds
+    \item Squish it down and inject it into the centre a white dwarf
+    \item Create diamonds (and worry about extracting them later)
     \item \dots
@@
 \end{enumerate}
-Your mission, should you choose to accept it (\dots which you have to because this is an exam), is to find out whether a white dwarf could produce diamonds, if so, what kind of white dwarf we need!
-I feel your experiences with stellar interiors, not to mention high pressure environments, would make you ideally suited to this task.
+Your mission, should you choose to accept it (\dots which you have to because this is an exam), is to find out whether a white dwarf has enough pressure at the centre to produce diamonds, and if so, what mass of white dwarf we need!
+I feel your experiences with stellar interiors, not to mention the high pressure environment known as grad school, would make you ideally suited to this task.
 Please note your timely response would be appreciated as I will be meeting with Elon to pitch the idea in 30 minutes.\\\\
@@
 
-\answer{}
+\answer{
+    First, we should solve for the Fermi momentum in terms of the number density from the fact that the integral should give the number density.
+    \begin{align}
+        n_e &= \int_0^{p_{\rm F}} n_e \dd[3]{p} \\
+            &= \frac{2}{h^3} \int_0^{p_{\rm F}} 4 \pi p^2 \dd{p} \\
+        n_e &= \frac{8 \pi p_{\rm F}^3}{3 h^3}
+    \end{align}
+    And so we can turn this around to get an expression for the Fermi momentum in terms of $n_e$
+    \begin{equation}
+        p_{\rm F} = \qty(\frac{3 n_e h^3}{8 \pi})^{1/3}
+    \end{equation}
+    Now, we can use this expression for the pressure integral (from Eq. 4.7) to find that the desired relation
+    \begin{align}
+        P_e &= \frac{1}{3} \int p \cdot v(p) \cdot n_e(p) \dd[3]{p} \\
+            &= \frac{1}{3} \int_0^{p_{\rm F}} p \cdot \frac{p}{m_e} \cdot \frac{2}{h^3} \dd[3]{p} \\
+            &= \frac{2}{3 m_e h^3} \int_0^{p_{\rm F}} p^2 \cdot 4 \pi p^2 \dd{p} \\
+            &= \frac{8 \pi}{3 m_e h^3} \cdot p_{\rm F}^5 \\
+            &= \frac{8 \pi}{3 m_e h^3} \cdot \qty(\frac{3 n_e h^3}{8 \pi})^{5/3} \\
+            &= \qty(\frac{3 h^3}{8 \pi})^{2/3} \cdot \frac{n_e^{5/3}}{m_e} \\
+        \Aboxed{ P_e &\propto n_e^{5/3} }
+    \end{align}
+}
 
@@
             &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] \qty(\frac{D_n}{\frac{4}{3} \pi})^{4/3} \frac{M^2}{R^4} \\
-        P_c &= \underbrace{\qty[\frac{(3 D_n)^{4/3}}{4 \pi (n + 1) M_n^{2/3}}]}_{=A} \frac{G M^2}{R^4}
+        \Aboxed{ P_c &= \underbrace{\qty[\frac{(3 D_n)^{4/3}}{4 \pi (n + 1) M_n^{2/3}}]}_{=A} \frac{G M^2}{R^4} }
     \end{align}
@@
 
+\clearpage
+
 \question{\textbf{Part d - Minimum white dwarf mass to produce diamonds}}
@@
 \question{Carbon requires a pressure of at least $\sim 10^{10} \unit{Pa}$ to form diamonds.
-Find the minimum mass a white dwarf would need to convert carbon to diamonds in its core (assuming $A = 0.77$, $B=1.05 \unit{m}{kg^{1/3}}$, $C=10/3$ and $G = 6.67 \times 10^{-11} \unit{m^3}{kg^{-1}}{s^2}$).\\
+Find the minimum mass a white dwarf would need to convert carbon to diamonds in its core (assuming $A = 0.77$, $B = 1.05 \times 10^{17} \unit{m}{kg^{1/3}}$, $C=10/3$ and $G = 6.67 \times 10^{-11} \unit{m^3}{kg^{-1}}{s^2}$ and recalling that $\unit{M_{\odot}} = 2 \times 10^{30} \unit{kg}$).\\
 
@@
 How would you answer change qualitatively if the professor could only inject carbon near the surface rather than at the centre?}
-\answer{}
+
+\answer{
+    First we can solve for the minimum mass as
+    \begin{align}
+        M_{\rm min} &= \qty(\frac{10^{10} 10^{17 \cdot 4}}{0.77 \cdot 6.67 \times 10^{-11}})^{3/10} \unit{kg} \\
+                    &= 3.25 \times 10^{26} \unit{kg} \\
+        \Aboxed{ M_{\rm min} &= 1.6 \times 10^{-4} \unit{M_\odot} }
+    \end{align}
+    This answer implies that \textbf{all} white dwarfs could create diamonds at their cores!
+If the professor could only inject carbon closer to the surface then the minimum mass would probably increase since the pressure declines sharply as you near the surface.
+}
 
@@
 
-\question{Craft a response to Prof.\ Eyemund summarising the situation. Bonus points will be awarded for each dreadful pun.}
+\question{Craft a pithy yet informative response to Prof.\ Eyemund summarising the situation. Bonus points will be awarded for each dreadful pun.}
 
 \answer{
-    The solution is left as an exercise for the reader.
+    The solution is left as an exercise for the reader :D
 }

### 2022-05-07 531_stars/pset3/solutions.tex
set up stars pset3

--- 2022-04-12 531_stars/pset2/solutions.tex
@@
 
-\newcommand{\psetnum}{2}
+\newcommand{\psetnum}{3}
 \newcommand{\class}{ASTR 531 - Stellar Interiors and Evolution}
@@
 
-\question{\textbf{8.1 - Mass Defect Fraction}}
-\answer{
-    The mass defects of the respective equations are as follows
-    \begin{center}
-        \begin{tabular}{c|c|c}
-            Phase & Equation & $\Delta m / m$\\
-            \hline
-            Hydrogen & $4 \ce{^{1}H} \to \ce{^{4}He}$ & 0.007145 \\
-            Helium & $3 \ce{^{4}He} \to \ce{^{12}C}$ & 0.000650 \\
-            Carbon & $\ce{^{12}C} + \ce{^{4}He} \to \ce{^{16}O}$ & 0.000481 \\
-            Oxygen & $2 \ce{^{16}O} \to \ce{^{28}Si} + \ce{^{4}He}$ & 0.000322 \\
-            Silicon & $2 \ce{^{28}Si} \to \ce{^{56}Fe}$ & 0.000338 \\
-        \end{tabular}
-    \end{center}
-    You can see that the trend is that subsequent reactions have lower mass defect fractions and produce less energy per reaction (with the exception of Silicon vs.
-Oxygen fusion.) Since less energy is available to counter gravity in each phase, this implies that stars will spend the majority of their time in the hydrogen fusion phase and progressively less time in each fusion phase.
-}
+\question{\textbf{12.2 - Pre-MS Radii and Timescales}}
+\answer{}
 
-\question{\textbf{8.4 - Minimum Core Masses}}
-\answer{
-    Equation 8.24 in the textbook gives that the minimum core mass required to initiate a fusion phase with ignition temperature $T_{\rm ign}$ is
-    \begin{equation}
-        M_{\rm crit} \approx \qty[ \frac{\mathcal{R} K_1}{\mu_c \mu_e^{5/3} G^2} \cdot T_{\rm ign}]^{3/4}
-    \end{equation}
-    The only variables here are $\mu_c, \mu_e$ and $T_{\rm ign}$.
-The textbook gives us that $\mu_e = \mu_c = 2$ for all fusion phases after Helium.
-Thus, given that the minimum core mass for Helium fusion is $0.3 \unit{M_{\odot}}$ (and that we know the ignition temperature is $10^8 \unit{K}$ from Table 8.4), the core masses required for the subsequent phases will be
-    \begin{equation}
-        M_{\rm crit} \approx 0.3 \unit{M_{\odot}} \cdot \qty(\frac{T_{\rm ign}}{10^{8} \unit{K}})^{3/4}
-    \end{equation}
-    So we can tabulate all of this for brevity
-    \begin{center}
-        \begin{tabular}{c|c|c}
-            Phase & $T_{\rm ign} / \unit{K}$ & $M_{\rm crit} / \unit{M_{\odot}} $\\
-            \hline
-            Helium & $10^8$ & 0.3 \\
-            Carbon & $6 \times 10^8$ & 1.15 \\
-            Neon & $9 \times 10^8$ & 1.56 \\
-            Oxygen & $10^9$ & 1.69 \\
-            Silicon & $3 \times 10^9 $ & 3.85 \\
-        \end{tabular}
-    \end{center}
-}
+\question{\textbf{15.4 - Metallicity and Mass Loss Rates}}
+\answer{}
 
-\question{\textbf{9.1 - Typical Timescales}}
-\answer{
-    For calculating the timescales based on $M, R, L$, I use the following equations from the textbook (specifically, 9.3, 9.4, 9.5d)
-    \begin{equation}
-        \tau_{\rm dyn} = \sqrt{\frac{1}{G \bar{\rho}}}, \qquad \tau_{\rm KH} = \frac{G M^2}{R L}, \qquad \tau_{\rm nuc} \approx \frac{M / \unit{M_{\odot}}}{L / \unit{L_{\odot}}} 10^{10} \unit{yr}
-    \end{equation}
-    Specifically, this gives the following values
-    \begin{center}
-        \begin{tabular}{c||c|c|c||c}
-            Star & $\tau_{\rm dyn} / \unit{s}$ & $\tau_{\rm KH} / \unit{s}$ & $\tau_{\rm nuc} / \unit{s}$ & Ratios \\
-            \hline
-            1$\unit{M_{\odot}}$ MS & $3.3 \times 10^{03}$ & $9.9 \times 10^{14}$ & $3.2 \times 10^{17}$ & $1 : 3.0 \times 10^{11} : 9.7 \times 10^{13}$ \\
-            60$\unit{M_{\odot}}$ MS & $2.4 \times 10^{04}$ & $3.0 \times 10^{11}$ & $2.4 \times 10^{13}$ & $1 : 1.2 \times 10^{07} : 9.7 \times 10^{08}$\\
-            15$\unit{M_{\odot}}$ RSG & $1.6 \times 10^{08}$ & $1.5 \times 10^{08}$ & $1.1 \times 10^{13}$ & $1.1 : 1 : 7.0 \times 10^{04}$ \\
-            0.6$\unit{M_{\odot}}$ WD & $5.5 \times 10^{00}$ & $3.0 \times 10^{19}$ & N/A & $1 : 5.4 \times 10^{18}$ \\
-        \end{tabular}
-    \end{center}
-    There's a couple of things to note here.
-First, in the general case of a lower mass main sequence star, the dynamical timescale is far shorter than both other timescales meaning that these stars are always in quasi-hydrostatic equilibrium.
-The thermal timescale is also shorter than the nuclear timescale and so the star is in thermal equilibrium for most of its life.
+\question{\textbf{16.1 - RGB Radii}}
+\answer{}
 
-    For higher mass main sequence stars, the nuclear timescale is significantly reduced (by a factor of 10,000), thus higher mass stars will run out of fuel much more quickly.
-The thermal timescale is also relatively closer to this nuclear timescale so we'd expect this star to spend more of its time out of thermal equilibrium.
-
-    For RSGs, an interesting thing to note if that the dynamical timescale is actually slightly longer than the thermal timescale!
-For WDs the dynamical timescale is incredibly short (5 seconds!), whilst the thermal timescale is much longer than for any of the other stars.
-}
+\question{\textbf{17.1 - Helium Flash Duration}}
+\answer{}
 

### 2022-05-07 531_stars/pset3/solutions.tex
stars q1a

@@
 
-\question{\textbf{12.2 - Pre-MS Radii and Timescales}}
-\answer{}
+\question{\textbf{12.2 - Early Radii and Timescales}}
+
+\question{Part a - Radii Estimations}
+\answer{
+    Let's use a couple of different relations from the textbook to get the radii at different times.
+A protostar becomes ionised and stars the Hayashi concentration phase when it's radius is on the order of
+    \begin{equation}
+        R_{\rm Hayashi, start} \approx 100 \unit{R_{\odot}} \qty(\frac{M}{\unit{M_\odot}})
+    \end{equation}
+    We find the radius of the protostar once the Hayashi concentration phase comes to an end is approximately a factor of 50 lower (based on assumptions of the temperature and opacity) such that
+    \begin{equation}
+        R_{\rm Hayashi, end} \approx 2 \unit{R_{\odot}} \qty(\frac{M}{\unit{M_\odot}})
+    \end{equation}
+    The radius at the start of the PMS phase will be the same as the end of the Hayashi concentration phase.
+    \begin{equation}
+        R_{\rm PMS, start} = R_{\rm Hayashi, end}
+    \end{equation}
+    Finally, the radius at the end of the PMS phase is the same as the radius at ZAMS and so we can write that
+    \begin{equation}
+        R_{\rm PMS, end} = R_{\rm ZAMS} = \unit{R_\odot} \qty(\frac{M}{\unit{M_\odot}})^{0.7}
+    \end{equation}
+    So now we can plug in numbers for the different masses of stars that we considered
+    \begin{center}
+        \begin{tabular}{c|cccc}
+            $M / \unit{M_\odot}$ & $R_{\rm Hayashi, start} / \unit{R_\odot}$ & $R_{\rm Hayashi, end} / \unit{R_\odot}$ & $R_{\rm PMS, start} / \unit{R_\odot}$ & $R_{\rm PMS, end} / \unit{R_\odot}$ \\
+            \hline
+            0.3 & 30 & 0.6 & 0.6 & 0.43 \\
+            3 & 300 & 6 & 6 & 2.16 \\
+            30 & 3000 & 60 & 60 & 10.8 \\
+        \end{tabular}
+    \end{center}
+}
 

### 2022-05-08 531_stars/pset3/solutions.tex
stars q1b

@@
     Let's use a couple of different relations from the textbook to get the radii at different times.
-A protostar becomes ionised and stars the Hayashi concentration phase when it's radius is on the order of
+A protostar becomes ionised and stars the Hayashi concentration phase when it's radius is on the order of (Eq.
+12.13)
     \begin{equation}
@@
     \end{equation}
-    We find the radius of the protostar once the Hayashi concentration phase comes to an end is approximately a factor of 50 lower (based on assumptions of the temperature and opacity) such that
+    We find the radius of the protostar once the Hayashi concentration phase comes to an end is approximately a factor of 50 lower (based on assumptions of the temperature and opacity) such that (page 12-8)
     \begin{equation}
@@
     \end{equation}
-    Finally, the radius at the end of the PMS phase is the same as the radius at ZAMS and so we can write that
+    Finally, the radius at the end of the PMS phase is the same as the radius at ZAMS and so we can write that (Eq. 12.16)
     \begin{equation}
@@
 
+\question{Part b - Timescale estimations}
+\answer{
+    The duration of the Hayashi concentration phase is given in Eq. 12.15 but we also showed that the timescale scales as $1 / M$ such that
+    \begin{equation}
+        \tau_{\rm Hayashi} \approx 10^{6} \unit{yr} \qty(\frac{M_{\odot}}{M})
+    \end{equation}
+    The duration of the PMS phase is given by Eq. 12.17 so we have that
+    \begin{equation}
+        \tau_{\rm PMS} \approx 6 \times 10^7 \unit{yr} \qty(\frac{M}{\unit{M_\odot}})^{-2.5}
+    \end{equation}
+    So now we can plug in numbers for the different masses of stars that we considered
+    \begin{center}
+        \begin{tabular}{c|cc}
+            $M / \unit{M_\odot}$ & $\tau_{\rm Hayashi} / \unit{yr}$ & $\tau_{\rm PMS} / \unit{yr}$ \\
+            \hline
+            0.3 & $3.33 \times 10^6$ & $1.22 \times 10^9$ \\
+            3 & $3.33 \times 10^5$ & $3.85 \times 10^6$ \\
+            30 & $3.33 \times 10^4$ & $1.22 \times 10^4$ \\
+        \end{tabular}
+    \end{center}
+}
+
 \question{\textbf{15.4 - Metallicity and Mass Loss Rates}}

### 2022-05-08 531_stars/pset3/solutions.tex
pset3 q3 stars

@@
 \question{\textbf{15.4 - Metallicity and Mass Loss Rates}}
-\answer{}
+\answer{
+    \todo{Need to check code/method with Emily}
+}
 
 \question{\textbf{16.1 - RGB Radii}}
-\answer{}
+\answer{
+    From inspection of Figure 16.1 we can find values for $L$ and $T_{\rm eff}$ at the start and end of the RGB phase. This phase starts at C and ends at F. We can then use the fact that
+    \begin{equation}
+        R = \sqrt{\frac{L}{4 \pi \sigma T_{\rm eff}^4}}
+    \end{equation}
+    to get the radii. Since I'm in a mood for tables today, let's make another!
+    \begin{center}
+        \begin{tabular}{c|ccc}
+            Stage & $\log (L / \unit{L_\odot})$ & $\log (T_{\rm eff} / \unit{K})$ & $R / \unit{R_{\odot}}$ \\
+            \hline
+            C (Start of RGB) & 0.4 & 3.7 & 2.1 \\
+            F (End of RGB) & 3.4 & 3.48 & 183.1
+        \end{tabular}
+    \end{center}
+}
 

### 2022-05-08 531_stars/pset3/solutions.tex
stars pset3 q4

@@
 
+\clearpage
+
 \question{\textbf{17.1 - Helium Flash Duration}}
-\answer{}
+\answer{
+    First we need to calculate the potential energy of a uniform sphere. We know that gravitational potential is given by
+    \begin{equation}
+        \Phi = - \frac{G M}{r}
+    \end{equation}
+    So we just need to integrate this over a series of teeny tiny shells over the core of the star. Note we can assume a uniform density so $\rho(r) = \rho$. Let's do it!
+    \begin{align}
+        U &= - \int_0^R \frac{G M(r)}{r} \rho(r) \dd[3]{r} \\
+          &= - \int_0^R \frac{G (\rho(r) \frac{4}{3} \pi r^3)}{r} \rho(r) 4 \pi r^2 \dd{r} \\
+          &= - \frac{16 G \pi^2 \rho^2}{3}\int_0^R r^4 \dd{r} \\
+        U &= - \frac{16 G \pi^2 \rho^2 R^5}{15}
+    \end{align}
+    We can rearrange this to get rid of the radius and just write it as a function of density and mass.
+    \begin{align}
+        R &= \qty(\frac{3 M}{4 \pi \rho})^{1/3} \\
+        U &= - \frac{16 G \pi^2 \rho^2}{15} \qty(\frac{3 M}{4 \pi \rho})^{5/3} \\
+        U &= - \frac{G \qty(36 \pi M^5 \rho)^{1/3}}{5}
+    \end{align}
+    Nice, what a...clean expression?\footnote{To be fair, it looks much better if we did it in terms of $M$ and $R$: $U = -\frac{3}{5} G M^2 / R$}.
+Now let's use this to find the difference in potential energy for a $0.5 \unit{M_\odot}$ core expanding from a density of $\rho = 10^{6} \unit{g}{cm^{-3}}$ to $\rho = 10^{4} \unit{g}{cm^{-3}}$.
+    \begin{equation}
+        \Delta U = - \frac{G \qty(36 \pi (0.5 \unit{M_\odot})^5)^{1/3}}{5} \qty[(10^{4} \unit{g}{cm^{-3}})^{1/3} - (10^{6} \unit{g}{cm^{-3}})^{1/3}] = 5 \times 10^{42} \unit{J}
+    \end{equation}
+    Now we can calculate the duration of the Helium flash given the energy production and efficiency
+    \begin{equation}
+        \tau_{\rm flash} = \frac{\Delta U}{L \cdot \phi} = \frac{5 \times 10^{49} \unit{J}}{10^{10} \unit{L_\odot} \cdot 0.2} = 2.5 \times 10^{40} \unit{erg}{L_\odot^{-1}}
+    \end{equation}
+    Now because I'm an ``astronomer'' I can 100\% (totally) understand ergs and $L_\odot$ intuitively, but just in case some readers don't live and breathe cgs in the same way this translates to
+    \begin{equation}
+        \boxed{ \tau_{\rm flash} = 0.21 \unit{yr} }
+    \end{equation}
+
+}
 

### 2022-05-10 531_stars/pset3/solutions.tex
add some todos
still trying to get mass loss to work
confused on vink
finished stars pset3

@@
     \end{equation}
-    The radius at the start of the PMS phase will be the same as the end of the Hayashi concentration phase.
+    The radius at the start of the PMS phase is the same as the end of the Hayashi concentration phase.
     \begin{equation}
@@
 \answer{
-    \todo{Need to check code/method with Emily}
+    I found it easier to grab the files from online rather than copy them from the textbook so it was easy enough to just do this for all of the masses rather than only $20 \unit{M_\odot}$ and $60 \unit{M_\odot}$.
+
+    You can see how I calculated the mass loss rates in \href{https://github.com/TomWagg/uw-grad-classes/blob/main/531_stars/pset3/code.ipynb}{this Jupyter notebook} using Vink (2001) mass loss (Eq.\ 15.14).
+The exact values at ZAMS and TAMS are under ``Part a Answers'' if you want to see them.
+I'm going to brush over those because they're just numbers and instead focus on the later parts which are easier to qualitatively discuss.
+
+    In the plots below I show the average MS mass loss rate (averaged between ZAMS and TAMS) on the left, and I then multiply this by the length of the main sequence to get the fraction of mass that is lost, shown on the right.
+The dotted lines show the `real' mass loss rate which was contained in the files.
+We can see that the rates I calculated match up rather well with these!
+    \begin{center}
+        \includegraphics[width=0.495\textwidth]{figures/mean_MS_mlr.pdf}
+        \includegraphics[width=0.495\textwidth]{figures/mass_fraction_lost.pdf}
+    \end{center}
+    \textbf{Metallicity Effects:} First, let's talk about what happens for the different metallicities.
+You can immediately see that the higher metallicity line (blue) is much higher than the low metallicity line (orange) - note these are log scales!
+The lower metallicity stars have fewer metals and thus fewer lines that can be used to drive winds.
+Therefore mass loss is lower for lower metallicity stars.
+
+    \noindent We also see that my calculated rates don't quite match up with those in the files so let's just address that quickly.
+First the overall rates seem to be lower pretty much everywhere.
+This makes sense because we are only calculating one type of winds but the files include lots of different winds.
+
+    \textbf{Small mass differences:} For masses below $20 \unit{M_\odot}$ we seem to be under-predicting the mass loss rates and this makes sense because Vink mass loss only applies to hot and luminous stars and are zero for these stars (specifically for anything with $T_{\rm eff} < 12500 \unit{K}$ or $L < 3 \times 10^4 \unit{L_\odot}$).
+If we added the Nieuwenhujzen \& de Jager fit then those winds would probably explains most of the difference (since these are just main sequence stars.)
+
+    \textbf{Large mass differences:} For masses above $80 \unit{M_\odot}$ things start to go wrong as well.
+For an $85 \unit{M_\odot}$ star, we are under-predicting the mass loss by nearly an order of magnitude.
+I \textit{think} this might be LBV winds in the real model.
+I'm honestly unsure why we over-predict for a $120 \unit{M_\odot}$ but the Vink fits aren't supposed to be used above 50,000 K anyway so we're in dodgy regions anyway.
+If there's an obvious reason for this I'd love to know!
+:)
 }
@@
     \end{center}
+    So the radius is increasing by nearly two orders of magnitude!
 }
-
-\clearpage
 
@@
     \begin{equation}
-        \boxed{ \tau_{\rm flash} = 0.21 \unit{yr} }
+        \boxed{ \tau_{\rm flash} = 0.21 \unit{yr} \approx 76 \unit{days} }
     \end{equation}
-
+    Quite the `flash' indeed!
 }

### 2022-05-11 531_stars/pset4/solutions.tex
set up hw4 for stars

--- 2022-05-08 531_stars/pset3/solutions.tex
@@
 
-\newcommand{\psetnum}{3}
+\newcommand{\psetnum}{4}
 \newcommand{\class}{ASTR 531 - Stellar Interiors and Evolution}
@@
 
-\question{\textbf{12.2 - Early Radii and Timescales}}
+\question{\textbf{20.2 - White Dwarf Luminosity}}
+\answer{}
 
-\question{Part a - Radii Estimations}
-\answer{
-    Let's use a couple of different relations from the textbook to get the radii at different times.
-A protostar becomes ionised and stars the Hayashi concentration phase when it's radius is on the order of (Eq.
-12.13)
-    \begin{equation}
-        R_{\rm Hayashi, start} \approx 100 \unit{R_{\odot}} \qty(\frac{M}{\unit{M_\odot}})
-    \end{equation}
-    We find the radius of the protostar once the Hayashi concentration phase comes to an end is approximately a factor of 50 lower (based on assumptions of the temperature and opacity) such that (page 12-8)
-    \begin{equation}
-        R_{\rm Hayashi, end} \approx 2 \unit{R_{\odot}} \qty(\frac{M}{\unit{M_\odot}})
-    \end{equation}
-    The radius at the start of the PMS phase will be the same as the end of the Hayashi concentration phase.
-    \begin{equation}
-        R_{\rm PMS, start} = R_{\rm Hayashi, end}
-    \end{equation}
-    Finally, the radius at the end of the PMS phase is the same as the radius at ZAMS and so we can write that (Eq. 12.16)
-    \begin{equation}
-        R_{\rm PMS, end} = R_{\rm ZAMS} = \unit{R_\odot} \qty(\frac{M}{\unit{M_\odot}})^{0.7}
-    \end{equation}
-    So now we can plug in numbers for the different masses of stars that we considered
-    \begin{center}
-        \begin{tabular}{c|cccc}
-            $M / \unit{M_\odot}$ & $R_{\rm Hayashi, start} / \unit{R_\odot}$ & $R_{\rm Hayashi, end} / \unit{R_\odot}$ & $R_{\rm PMS, start} / \unit{R_\odot}$ & $R_{\rm PMS, end} / \unit{R_\odot}$ \\
-            \hline
-            0.3 & 30 & 0.6 & 0.6 & 0.43 \\
-            3 & 300 & 6 & 6 & 2.16 \\
-            30 & 3000 & 60 & 60 & 10.8 \\
-        \end{tabular}
-    \end{center}
-}
-
-\question{Part b - Timescale estimations}
-\answer{
-    The duration of the Hayashi concentration phase is given in Eq. 12.15 but we also showed that the timescale scales as $1 / M$ such that
-    \begin{equation}
-        \tau_{\rm Hayashi} \approx 10^{6} \unit{yr} \qty(\frac{M_{\odot}}{M})
-    \end{equation}
-    The duration of the PMS phase is given by Eq. 12.17 so we have that
-    \begin{equation}
-        \tau_{\rm PMS} \approx 6 \times 10^7 \unit{yr} \qty(\frac{M}{\unit{M_\odot}})^{-2.5}
-    \end{equation}
-    So now we can plug in numbers for the different masses of stars that we considered
-    \begin{center}
-        \begin{tabular}{c|cc}
-            $M / \unit{M_\odot}$ & $\tau_{\rm Hayashi} / \unit{yr}$ & $\tau_{\rm PMS} / \unit{yr}$ \\
-            \hline
-            0.3 & $3.33 \times 10^6$ & $1.22 \times 10^9$ \\
-            3 & $3.33 \times 10^5$ & $3.85 \times 10^6$ \\
-            30 & $3.33 \times 10^4$ & $1.22 \times 10^4$ \\
-        \end{tabular}
-    \end{center}
-}
-
-\question{\textbf{15.4 - Metallicity and Mass Loss Rates}}
-\answer{
-    \todo{Need to check code/method with Emily}
-}
-
-\question{\textbf{16.1 - RGB Radii}}
-\answer{
-    From inspection of Figure 16.1 we can find values for $L$ and $T_{\rm eff}$ at the start and end of the RGB phase. This phase starts at C and ends at F. We can then use the fact that
-    \begin{equation}
-        R = \sqrt{\frac{L}{4 \pi \sigma T_{\rm eff}^4}}
-    \end{equation}
-    to get the radii. Since I'm in a mood for tables today, let's make another!
-    \begin{center}
-        \begin{tabular}{c|ccc}
-            Stage & $\log (L / \unit{L_\odot})$ & $\log (T_{\rm eff} / \unit{K})$ & $R / \unit{R_{\odot}}$ \\
-            \hline
-            C (Start of RGB) & 0.4 & 3.7 & 2.1 \\
-            F (End of RGB) & 3.4 & 3.48 & 183.1
-        \end{tabular}
-    \end{center}
-}
-
-\question{\textbf{17.1 - Helium Flash Duration}}
+\question{\textbf{23.2 - Central Temperature-Density Gradient}}
 \answer{}

### 2022-05-11 531_stars/pset4/solutions.tex
started first question HW4 stars

@@
 \question{\textbf{20.2 - White Dwarf Luminosity}}
-\answer{}
+
+\question{Part a}
+\answer{
+    The luminosity of a white dwarf in the slow cooling phase is given by Eq.\ 20.10 in the textbook
+    \begin{equation}
+        \frac{L}{\unit{L_\odot}} \approx 5.2 \times 10^{10} \frac{M}{\unit{M_\odot}} \mu_{\rm ion}^{-7/5} \qty(\frac{t}{\unit{yr}})^{-7.5}
+    \end{equation}
+    Since we are comparing white dwarfs with the same cooling age, the only relevant factors are the mass and $\mu_{\rm ion}$ when comparing a H-rich WD to He-rich and C-rich.
+
+    The values of $\mu_{\rm ion}$ for these WDs are $1, 4, 12$ respectively. This means that the relative luminosity of the WDs is
+    \begin{equation}
+        \frac{L}{L_{\rm H-rich}} = 1 : 0.14 : 0.03
+    \end{equation}
+    respectively. This shows that H-rich WDs are the brightest for a given cooling age, following my He-rich and then C-rich WDs.
+
+    \todo{Can we assume that the mass is constant?}
+}
+
+\question{Part b}
+\answer{
+    \todo{Unsure, maybe larger ions make cooling happen faster? Why?}
+}
 

### 2022-05-11 531_stars/pset4/solutions.tex
not sure on Q2 stars HW4

@@
 \question{\textbf{23.2 - Central Temperature-Density Gradient}}
-\answer{}
+\answer{
+    The evolution of a star in the $T_c$--$\rho_c$ diagram becomes less steep at late evolution phases. This means that for a given increase in density, the increase in temperature is not as strong.
+    \todo{I think this is probably something to do with the mass defect being lower and so the energy/temperature production is lower?}
+}
 

### 2022-05-14 531_stars/pset4/solutions.tex
stars hw4 q3

@@
 
-\question{Part a}
+\question{Part a - Difference in luminosity}
 \answer{
@@
 
-\question{Part b}
+\question{Part b - Reason for differences}
 \answer{
@@
 
+\clearpage
+
+\question{\textbf{25.1 - Non-spherical mass loss in rotating stars}}
+
+\question{Part a - Mass loss latitude trends}
+\answer{
+    The star in Figure 25.3 is very massive, for these stars the winds are driven mainly by radiation pressure.
+The Von Zeipel effect means that the effective temperature of the pole is higher than at the equator.
+This means that for most cases (the left panel) mass loss is much higher at the poles than at the equator.
+
+    In certain cases (the right panel), the decreased $T_{\rm eff}$ at the equator actually leads to a higher mass loss rate due to the formation of a bi-stability disk.
+This disk has higher mass loss because its lower effective temperature puts it on the other side of a bi-stability jump from the pole and so its winds are driven by different ions that more effectively remove mass from the star.
+}
+
+\question{Part b - Terminal wind velocity latitude trends}
+\answer{
+    In rapidly rotating stars, the effective surface gravity at a given radius becomes a function of latitude, with stronger gravity at the poles than the equation.
+The escape velocity is a direct function of this surface gravity.
+We have previously shown that the terminal wind velocity is approximately equal to the escape velocity (see Section 15.2.2 of the textbook).
+For these reasons the terminal wind velocity is higher at the poles than at the equator.
+}
+
+\question{Part c - Density of bi-stability disk}
+\answer{
+    A rotation-induced bi-stability disk must by definition have a lower effective temperature than the wind in the poles and it also has a lower terminal wind velocity.
+The lower temperature naturally leads to higher densities and the lower velocity allows more gas to remain in the disk.
+    \todo{check this part, particularly the velocity bit}
+}
+
+\question{\textbf{27.1 - Limit for NS vs.\ BH for rotating stars}}
+\answer{}
+
 \end{document}

### 2022-05-14 531_stars/pset4/solutions.tex
hw4 q4 stars

@@
 \question{\textbf{27.1 - Limit for NS vs.\ BH for rotating stars}}
-\answer{}
+\answer{
+    The limit of stars with initial masses resulting in NSs or BHs could change due to rotation in the following ways.
+
+    \textit{Higher mass limit:} The mass needed to create a BH would need to increase due to the increased mass loss due to rotation.
+This means that less mass is available to go into the creation of a remnant and so stars that would have resulted in BHs instead result in NSs.
+
+    \textit{Lower mass limit:} However in addition to the overall mass loss, the mass of the helium core instead increases for rapidly-rotating stars due to the induced mixing.
+A larger core mass can also lead to the larger remnant mass and thus more easily create a BH with a given initial mass.
+
+    Overall, I think that\dots
+    \todo{Check with Emily what's more important}
+}
 

### 2022-05-16 531_stars/pset4/solutions.tex
q1 and q2 done stars hw4

@@
     respectively. This shows that H-rich WDs are the brightest for a given cooling age, following my He-rich and then C-rich WDs.
-
-    \todo{Can we assume that the mass is constant?}
 }
@@
 \answer{
-    \todo{Unsure, maybe larger ions make cooling happen faster? Why?}
+    The cooling of a WD is due to the loss of the thermal energy of the ions.
+The specific heat of a substance gives a measure for how much energy is required to change its temperature by a given amount.
+Therefore, a smaller specific heat means that cooling an object is easier and will happen more quickly.
+The specific heat of the ion is given by
+    \begin{equation}
+        c_V = \frac{3 k}{2 m_{\rm ion}}
+    \end{equation}
+    This means that for helium and carbon, the increased $m_{\rm ion}$ results in a smaller specific heat and thus faster cooling rate.
 }
@@
 \answer{
-    The evolution of a star in the $T_c$--$\rho_c$ diagram becomes less steep at late evolution phases. This means that for a given increase in density, the increase in temperature is not as strong.
-    \todo{I think this is probably something to do with the mass defect being lower and so the energy/temperature production is lower?}
+    For most of its life the star follows
+    \begin{equation}
+        T_c \sim \rho_c^{1/3},
+    \end{equation}
+    however this trend becomes less steep over time. This is because the full expression is actually
+    \begin{equation}
+        T_c \sim M_c^{2/3} \rho_c^{1/3},
+    \end{equation}
+    For the later fusion phases, the core mass becomes much lower (see Eq.\ 14.10) and so we can't assume it to be constant.
+Therefore, we see that for late fusion phases the increase in temperature for a given density is lower, thus the evolution in the plot is less steep.
 }
-
-\clearpage
 

### 2022-05-16 531_stars/exam2/exam.tex
start exam 2 stars

--- 2022-04-25 531_stars/exam1/exam.tex
@@
 
-\newcommand{\psetnum}{1}
+\newcommand{\psetnum}{2}
 \newcommand{\class}{ASTR 531 - Stellar Interiors and Evolution}
@@
 
-\title{\class : Problem Set \psetnum}
+\title{\class : Exam \psetnum}
 \author{\textbf{Tom Wagg}}
@@
 
-\section*{Thriving under high pressure}
-\question{Dearest Grad Student,\\
+\section*{Binary no more: A breakup story}
 
-\noindent My name is Professor Daniel Eyemund and I'm reaching out to you with an exciting business idea - diamonds!
-I've realised that I could get around applying for grants by generating a large number of diamonds and discretely selling them!
-They might be rare on Earth but with the right conditions perhaps we could create a \textbf{bunch} of them out in space.
-The plan is as follows:
-\begin{enumerate}[nosep]
-    \item Collect a big cloud of carbon
-    \item Squish it down and inject it into a white dwarf
-    \item Create diamonds
-    \item \dots
-    \item Profit?
-\end{enumerate}
-Your mission, should you choose to accept it (\dots which you have to because this is an exam), is to find out whether a white dwarf could produce diamonds, if so, what kind of white dwarf we need!
-I feel your experiences with stellar interiors, not to mention high pressure environments, would make you ideally suited to this task.
-Please note your timely response would be appreciated as I will be meeting with Elon to pitch the idea in 30 minutes.\\\\
-Yours sincerely,
-
-\noindent Prof. D. Eyemund}
-\answer{}
-
-\question{\textbf{Part a - Complete degeneracy pressure for electrons}}
-
-\question{Given that, for complete degeneracy, the $n_e(p)$ distribution is rectangular for $p < p_{\rm F}$ such that $n_e(p) = 2 / h^3$, show that the pressure scales as $P_e \propto n_e^{5/3}$.}
-
-\answer{}
-
-\question{\textbf{Part b - Polytrope central pressure}}
-
-\question{By showing that part a is true, you've proven that we can use an $n = 3/2$ polytrope! It'll now be useful to know an expression for the central pressure of a polytrope star.\\
-
-\noindent Use the fact that $P_c = K \rho_c^{1 + 1/n}$ to show that $P_c = A \frac{G M^2}{R^4}$ (and find the constant $A$ in terms of $D_n, \pi, n, M_n$).
-Hint: You can find $K$ by comparing an expression for $\alpha$ and an expression for $M_n$.
-Additionally note that $\rho_c = \bar{\rho} D_n$.}
-
-\answer{
-    To start, we need to find an expression for $K$. We have the definition of $\alpha$ from the Lane-Emden equation (see Eq. 11.5).
-    \begin{equation}
-        \alpha = \qty[\frac{(n + 1) K}{4 \pi G \rho_c^{1 - 1/n}}]^{1/2}
-    \end{equation}
-    Next we can use Equation 11.10a to find another expression for $\alpha$ from the definition of $M_n$
 \begin{align}
-        M &= 4 \pi a^3 \rho_c M_n \\
-        \alpha &= \qty(\frac{M}{M_n 4 \pi \rho_c})^{1/3}
+    v = \sqrt{G (M_1 + M_2) \qty(\frac{2}{r} - \frac{1}{a})}
 \end{align}
-    Combining these yields an expression for $K$
-    \begin{align}
-        \qty[\frac{(n + 1) K}{4 \pi G \rho_c^{1 - 1/n}}]^{1/2} &= \qty(\frac{M}{M_n 4 \pi \rho_c})^{1/3} \\
-        K &= \qty(\frac{M}{M_n 4 \pi \rho_c})^{2/3} \frac{4 \pi G \rho_c^{1 - 1 / n}}{(n + 1)} \\
-        K &= \qty(\frac{M}{M_n})^{2/3} \frac{(4 \pi)^{1/3} G \rho_c^{1/3 - 1 / n}}{(n + 1)}
-    \end{align}
-    Now, given that $P_c = K \rho_c^{1 + 1/n}$, this means that we can write the central pressure as
-    \begin{align}
-        P_c &= \qty(\frac{M}{M_n})^{2/3} \frac{(4 \pi)^{1/3} G \rho_c^{1/3 - 1 / n}}{(n + 1)} \rho_c^{1 + 1 / n} \\
-            &= \qty(\frac{M}{M_n})^{2/3} \frac{(4 \pi)^{1/3} G \rho_c^{4/3}}{(n + 1)} \\
-            &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] M^{2/3} \rho_c^{4/3}
-    \end{align}
-    We also know that the average density is given by $\bar{\rho} = \frac{M}{\frac{4}{3} \pi R^3}$ and it is related to the central density as $\rho_c = \bar{\rho} D_n$.
-Therefore we can express the central pressure
-    \begin{align}
-        P_c &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] M^{2/3} \qty(\frac{M D_n}{\frac{4}{3} \pi R^3})^{4/3} \\
-            &= \qty[\frac{(4 \pi)^{1/3} G}{(n + 1) M_n^{2/3}}] \qty(\frac{D_n}{\frac{4}{3} \pi})^{4/3} \frac{M^2}{R^4} \\
-        P_c &= \underbrace{\qty[\frac{(3 D_n)^{4/3}}{4 \pi (n + 1) M_n^{2/3}}]}_{=A} \frac{G M^2}{R^4}
-    \end{align}
-}
-
-\question{\textbf{Part c - White dwarf $\mathbf{M}$-$\mathbf{P_c}$ relation}}
-
-\question{Now let's take that and create a relation for the central pressure of a white dwarf as only a function of mass.
-Apply the $M$-$R$ relation for WDs (assuming $\mu_e = 2$) and that $n = 3/2$ to show that $P_c = \frac{A G}{B^4} M^{C}$ where $B$ and $C$ are constants you should find.}
-
-\answer{
-    The WD $M$-$R$ relation with $\mu_e = 2$ gives that
-    \begin{equation}
-        R = 0.012 \unit{R_{\odot}} \qty(\frac{M}{\unit{M_{\odot}}})^{-1/3}
-    \end{equation}
-    This gives the final result of the central pressure as a function of WD mass
-    \begin{equation}
-        \boxed{ P_{c, {\rm WD}} = \frac{A G}{B^4} M^{C} }
-    \end{equation}
-    where $B = 0.012 \unit{R_{\odot}} \unit{M_{\odot}}^{1/3}$ and $C = 10/3$.
-}
 

### 2022-05-16 531_stars/pset4/solutions.tex
finished HW4 stars

@@
 \answer{
-    A rotation-induced bi-stability disk must by definition have a lower effective temperature than the wind in the poles and it also has a lower terminal wind velocity.
-The lower temperature naturally leads to higher densities and the lower velocity allows more gas to remain in the disk.
-    \todo{check this part, particularly the velocity bit}
+    The density of the gas scales as
+    \begin{equation}
+        \rho \sim \frac{\dot{M}}{v_{\infty}}
+    \end{equation}
+    For a rotation-induced bi-stability disk, the mass loss is higher than at the poles at its slightly lower $T_{\rm eff}$ places it just below a bi-stability jump (see Section 15.3).
+Moreover, the terminal wind velocity ($v_\infty$) is higher on the lower side of a bi-stability jump (see Section 15.2.2).
+The combination of these effects leads to a significant increase in density in the disk relative to the poles.
 }
@@
 
-    \textit{Higher mass limit:} The mass needed to create a BH would need to increase due to the increased mass loss due to rotation.
-This means that less mass is available to go into the creation of a remnant and so stars that would have resulted in BHs instead result in NSs.
-
-    \textit{Lower mass limit:} However in addition to the overall mass loss, the mass of the helium core instead increases for rapidly-rotating stars due to the induced mixing.
-A larger core mass can also lead to the larger remnant mass and thus more easily create a BH with a given initial mass.
-
-    Overall, I think that\dots
-    \todo{Check with Emily what's more important}
+    \begin{itemize}
+        \item \textit{Case for higher mass limit:} The mass needed to create a BH would need to increase to compensate for the increased mass loss due to rotation.
+Otherwise, less mass is available to go into the creation of a remnant and so stars that would have resulted in BHs instead result in NSs.
+        \item \textit{Case for lower mass limit:} Though the overall mass is decreased for a rapidly-rotating star, the mass of the helium core \textit{increases} due to the induced mixing.
+A larger core mass leads to a larger remnant mass and thus more easily create a BH with a given initial mass.
+    \end{itemize}
+    Overall, the winner here is the latter reasoning.
+Core mass is the most important parameter for determining the destiny of any massive star and so if all stars were rapidly-rotating the mass limit between NSs and BHs would decrease.
 }

### 2022-05-16 531_stars/pset4/solutions.tex
fix some typos

@@
 \answer{
-    In rapidly rotating stars, the effective surface gravity at a given radius becomes a function of latitude, with stronger gravity at the poles than the equation.
+    In rapidly rotating stars, the effective surface gravity at a given radius becomes a function of latitude, with stronger gravity at the poles than the equator.
 The escape velocity is a direct function of this surface gravity.
@@
     For a rotation-induced bi-stability disk, the mass loss is higher than at the poles at its slightly lower $T_{\rm eff}$ places it just below a bi-stability jump (see Section 15.3).
-Moreover, the terminal wind velocity ($v_\infty$) is higher on the lower side of a bi-stability jump (see Section 15.2.2).
+Moreover, the terminal wind velocity ($v_\infty$) is lower on the lower side of a bi-stability jump (see Section 15.2.2).
 The combination of these effects leads to a significant increase in density in the disk relative to the poles.
@@
 \answer{
-    The limit of stars with initial masses resulting in NSs or BHs could change due to rotation in the following ways.
-
+    The limit of stars with initial masses resulting in NSs or BHs could change due to rotation in the following ways:
     \begin{itemize}

### 2022-05-17 531_stars/exam2/exam.tex
set up question for exam2

@@
 \usepackage{mhchem}
+\usepackage[normalem]{ulem}
 \hypersetup{
@@
 
-\section*{Binary no more: A breakup story}
+\section*{Getting a white dwarf to chill out}
+\vspace{-0.35cm}
+\subsubsection*{Prof.\ Eyemund Returns}
+\vspace{0.25cm}
+\question{Dearest Grad Student,\\
 
-\begin{align}
-    v = \sqrt{G (M_1 + M_2) \qty(\frac{2}{r} - \frac{1}{a})}
-\end{align}
+    \noindent I have good news and bad news.
+The bad news is that it seems that our venture into the diamond creation business has gone slightly awry.
+The other faculty became suspicious after they noticed I was using that goblet of diamonds as a paperweight and after some lengthy arguments I have been legally forbidden from ever using white dwarfs for diamond production!
+The good news?
+Well they didn't say anything about neutron stars...\\
+
+    \vspace{-0.3cm}
+    \noindent Now the problem of course is that we've invested quite a lot of \sout{your PhD funding} money into finding that sample of white dwarfs.
+But I realised we don't need to \emph{find} a neutron star, we just need to \emph{make} a neutron star!!\\
+
+    \vspace{-0.3cm}
+    \noindent So I've done some experiments with adding mass to existing white dwarfs, but for some reason none of them reach the Chandrasekhar limit before exploding dramatically.
+I'm not entirely sure why this happens but it seems to be temperature dependent, therefore I've acquired a Cooling Ray to stop the photosphere from getting too hot!
+It can cool at any rate we want, but it costs more to do so.
+So my question for you is \textbf{What's the lowest rate we could cool the outer layers of a white dwarf whilst it accretes material and still create a neutron star?}\\\\
+    Yours sincerely,
+
+    \noindent Prof. D. Eyemund
+    \answer{}
+}
+
+\question{\textbf{Part a - What's with those explosions?}}
+
+\question{Explain what is physically occuring at the surface of the white dwarf to cause exoplosions as the temperature increases during accretion}
+
+\answer{}
+
+\question{\textbf{Part b - How quickly do we need to cool it?}}
+
+\question{}
+
+\answer{}
 

### 2022-05-17 531_stars/exam2/exam.tex
part a+b exam2 stars

@@
     \noindent So I've done some experiments with adding mass to existing white dwarfs, but for some reason none of them reach the Chandrasekhar limit before exploding dramatically.
-I'm not entirely sure why this happens but it seems to be temperature dependent, therefore I've acquired a Cooling Ray to stop the photosphere from getting too hot!
+I'm not entirely sure why this happens but it seems to be temperature dependent, therefore I've acquired a Cooling Ray to stop the envelope from getting too hot!
 It can cool at any rate we want, but it costs more to do so.
@@
 
-\question{Explain what is physically occuring at the surface of the white dwarf to cause exoplosions as the temperature increases during accretion}
+\question{Explain what is physically occuring at the surface of the white dwarf that leads to exoplosions during accretion.}
 
-\answer{}
+\answer{
+    It's the ignition of thermonuclear fusion!
+As the white dwarf accretes material, the temperature of the envelope starts to heat up.
+Eventually, the temperature reaches the ignition temperature of H-fusion and thus fusion is rapidly ignited throughout the envelope, leading to an exoplosive ejection of the accreted shell of material.
+}
 
-\question{\textbf{Part b - How quickly do we need to cool it?}}
+\clearpage
 
-\question{}
+\question{\textbf{Part b - How quickly does an accreting WD heat up?}}
 
-\answer{}
+\question{Derive the rate at which an accreting WD heats:
+\begin{equation}
+    \dot{T} = f(M_{\rm WD}, R_{\rm WD}, M_{\rm env}, \mu, \dot{M}_a)
+\end{equation}
+It should be a function of the mass of the white dwarf, radius of the white dwarf, mass of the surrounding envelope, mean molecular weight of envelope and the accretion rate.
+
+You should assume that all gravitational energy of the accreted material is converted to thermal energy in the WD envelope during accretion and that the white dwarf is already very close to the Chandrasekhar limit.
+Additionally, you can assume the envelope is an ideal gas.
+and its mass remains approximately constant.}
+
+\answer{
+    The gravitational energy for the accreted mass on the white dwarf is just a function of the masses and the separation.
+    \begin{equation}
+        E_{\rm grav} = - \frac{G M_{\rm WD} M_a}{R_{\rm WD}}
+    \end{equation}
+    Since the envelope of the WD is an ideal gas, the average thermal energy is
+    \begin{equation}
+        \avg{E_{\rm thermal}} = \frac{3}{2} k_B T
+    \end{equation}
+    We can convert this to the total thermal energy of the envelope by multiplying by the total number of particles in the envelope
+    \begin{equation}
+        E_{\rm thermal} = \avg{E_{\rm thermal}} \cdot N = \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H}
+    \end{equation}
+    Now, since we are assuming that all gravitational energy is converted to thermal energy then the time derivatives of these expressions should be equal and we can solve for the change in temperature
+    \begin{align}
+        - \dv{E_{\rm grav}}{t} &= \dv{E_{\rm thermal}}{t} \\
+        \dv{t} \frac{G M_{\rm WD} M_a}{R_{\rm WD}} &= \dv{t} \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H} \\
+        \frac{G M_{\rm WD} \dot{M}_a}{R_{\rm WD}} &= \frac{3}{2} k_B \dot{T} \cdot \frac{M_{\rm env}}{\mu m_H} \\
+        \Aboxed{ \dot{T} &= \frac{2}{3 k_B} \frac{\mu m_H}{M_{\rm env}} \frac{G M_{\rm WD} \dot{M}_a}{R_{\rm WD}} }
+    \end{align}
+}
+
+\question{\textbf{Part c - How fast does the laser need to cool the WD?}}
+
+\question{Calculate the minimum cooling rate required for the WD to reach the Chandrasekhar limit before any explosions occur.}
 

### 2022-05-17 531_stars/exam2/exam.tex
exam 2 stars: part c

@@
 
-\question{Calculate the minimum cooling rate required for the WD to reach the Chandrasekhar limit before any explosions occur.}
+\question{Calculate the minimum cooling rate required for the He-WD to reach the Chandrasekhar limit before any explosions occur.}
+
+\answer{
+    First, let's calculate the time until the WD reaches the Chandrasekhar limit
+    \begin{equation}
+        t_{\rm ch} = \frac{M_{\rm ch} - M_{\rm WD}}{\dot{M}_a}
+    \end{equation}
+    Now we can solve for the minimum cooling rate of the laser that keeps the temperature under the H-fusion ignition temperature in this amount of time.
+    \begin{align}
+        T_{\rm ign} &> (\dot{T}_{\rm accretion} - \dot{T}_{\rm laser}) t_{\rm Ch} \\
+        \dot{T}_{\rm laser} &> \dot{T}_{\rm accretion} - \frac{T_{\rm ign}}{t_{\rm Ch}} \\
+        \Aboxed{ \dot{T}_{\rm laser}&> \dot{M}_a \qty[\frac{2}{3 k_B} \frac{\mu m_H}{M_{\rm env}} \frac{G M_{\rm WD}}{R_{\rm WD}} - \frac{T_{\rm ign}}{M_{\rm ch} - M_{\rm WD}}] }
+    \end{align}
+}
 

### 2022-05-17 531_stars/exam2/exam.tex
stars exam 2: part c numbers

@@
 
-\question{Calculate the minimum cooling rate required for the He-WD to reach the Chandrasekhar limit before any explosions occur.}
+\question{Calculate the minimum cooling rate required for the WD to reach the Chandrasekhar limit before any explosions occur.
+
+Plug in numbers for a He-WD with $M_{\rm WD} = 1.45 \unit{M_\odot}$, $M_{\rm env} = 10^{-5} \unit{M_\odot}$ and $\dot{M}_a = 10^{-8} \unit{M_\odot}{yr^{-1}}$}
 
@@
     \end{align}
+    Finally, we can start plugging numbers into this.
+Assuming the white dwarf follows the mass-radius relation then the radius is $R_{\rm WD} = 0.012 M_{\rm WD}^{-1/3}$.
+We additionally know that $T_{\rm ign} = 4 \times 10^6$ for H-fusion and $M_{\rm ch} = 1.46 \unit{M_\odot}, \mu = 2$ for a He-WD.
+Putting all of this into the equation above gives
+    \begin{equation}
+        \boxed{ \dot{T}_{\rm laser} \approx 10^6 \unit{K}{yr^{-1}}}
+    \end{equation}
 }

### 2022-05-18 531_stars/exam2/exam.tex
stars exam 2: done pending Emily approval

@@
 
-\vspace{-1.5cm}
+\vspace{-0.5cm}
+
+\noindent \textbf{[Exam format]}
+
+\noindent This exam is open note but closed book/internet. You have access to a list of constants and maybe use a tool for converting units (e.g.\ an iPython instance with astropy installed).
 
@@
 The bad news is that it seems that our venture into the diamond creation business has gone slightly awry.
-The other faculty became suspicious after they noticed I was using that goblet of diamonds as a paperweight and after some lengthy arguments I have been legally forbidden from ever using white dwarfs for diamond production!
+The other faculty members became suspicious after they noticed I was using that goblet of diamonds as a paperweight and after some lengthy arguments I have been legally forbidden from ever using white dwarfs for diamond production!
 The good news?
@@
     \vspace{-0.3cm}
-    \noindent Now the problem of course is that we've invested quite a lot of \sout{your PhD funding} money into finding that sample of white dwarfs.
+    \noindent Now the problem of course is that we've invested rather a lot of \sout{your PhD funding} money into finding that sample of white dwarfs.
 But I realised we don't need to \emph{find} a neutron star, we just need to \emph{make} a neutron star!!\\
@@
     \noindent So I've done some experiments with adding mass to existing white dwarfs, but for some reason none of them reach the Chandrasekhar limit before exploding dramatically.
-I'm not entirely sure why this happens but it seems to be temperature dependent, therefore I've acquired a Cooling Ray to stop the envelope from getting too hot!
+I'm not entirely sure why this happens but it seems to be temperature dependent, therefore I've acquired a Cooling Ray$^{\rm TM}$ to stop the envelope from getting too hot!
 It can cool at any rate we want, but it costs more to do so.
-So my question for you is \textbf{What's the lowest rate we could cool the outer layers of a white dwarf whilst it accretes material and still create a neutron star?}\\\\
+So my question for you is \textbf{What is the lowest rate we could cool the outer layers of a white dwarf as it accretes material and still create a neutron star?}\\\\
     Yours sincerely,
@@
 
+\clearpage
+
 \question{\textbf{Part a - What's with those explosions?}}
 
-\question{Explain what is physically occuring at the surface of the white dwarf that leads to exoplosions during accretion.}
+\question{Explain the physical mechanism that leads to the explosions at the surface of a white dwarf as it accretes material.}
 
@@
 
-\clearpage
-
 \question{\textbf{Part b - How quickly does an accreting WD heat up?}}
 
-\question{Derive the rate at which an accreting WD heats:
+\question{Derive the rate at which a WD heats up as it accretes material:
 \begin{equation}
@@
 \end{equation}
-It should be a function of the mass of the white dwarf, radius of the white dwarf, mass of the surrounding envelope, mean molecular weight of envelope and the accretion rate.
+Express your answer in terms of the mass of the white dwarf, radius of the white dwarf, mass of the surrounding envelope, mean molecular weight of envelope and the accretion rate.
 
 You should assume that all gravitational energy of the accreted material is converted to thermal energy in the WD envelope during accretion and that the white dwarf is already very close to the Chandrasekhar limit.
-Additionally, you can assume the envelope is an ideal gas.
-and its mass remains approximately constant.}
+Additionally, you can assume the envelope is an ideal gas and its mass remains approximately constant.}
 
@@
     \begin{equation}
-        E_{\rm thermal} = \avg{E_{\rm thermal}} \cdot N = \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H}
+        E_{\rm thermal} = \avg{E_{\rm thermal}} \cdot N_{\rm env} = \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H}
     \end{equation}
@@
         - \dv{E_{\rm grav}}{t} &= \dv{E_{\rm thermal}}{t} \\
-        \dv{t} \frac{G M_{\rm WD} M_a}{R_{\rm WD}} &= \dv{t} \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H} \\
+        \dv{t} \qty[\frac{G M_{\rm WD} M_a}{R_{\rm WD}}] &= \dv{t} \qty[\frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H}] \\
         \frac{G M_{\rm WD} \dot{M}_a}{R_{\rm WD}} &= \frac{3}{2} k_B \dot{T} \cdot \frac{M_{\rm env}}{\mu m_H} \\
@@
 
-Plug in numbers for a He-WD with $M_{\rm WD} = 1.45 \unit{M_\odot}$, $M_{\rm env} = 10^{-5} \unit{M_\odot}$ and $\dot{M}_a = 10^{-8} \unit{M_\odot}{yr^{-1}}$}
+Plug in numbers for a He-WD with $M_{\rm WD} = 1.45 \unit{M_\odot}$, $M_{\rm env} = 10^{-5} \unit{M_\odot}$ and $\dot{M}_a = 10^{-8} \unit{M_\odot}{yr^{-1}}$.
+If you are unsure of the temperature at which an explosion occurs then you can use $T = 10^7 \unit{K}$.}
 
@@
         \dot{T}_{\rm laser} &> \dot{T}_{\rm accretion} - \frac{T_{\rm ign}}{t_{\rm Ch}} \\
-        \Aboxed{ \dot{T}_{\rm laser}&> \dot{M}_a \qty[\frac{2}{3 k_B} \frac{\mu m_H}{M_{\rm env}} \frac{G M_{\rm WD}}{R_{\rm WD}} - \frac{T_{\rm ign}}{M_{\rm ch} - M_{\rm WD}}] }
+        \dot{T}_{\rm laser}&> \dot{M}_a \qty[\frac{2}{3 k_B} \frac{\mu m_H}{M_{\rm env}} \frac{G M_{\rm WD}}{R_{\rm WD}} - \frac{T_{\rm ign}}{M_{\rm ch} - M_{\rm WD}}]
     \end{align}
@@
     \end{equation}
+    Fun fact: to cool the envelope of this white dwarf, the laser would need around $10^{20}$ times the power consumption of the US, so it seems Professor Eyemund's quest for diamonds must go on!
 }

### 2022-05-19 531_stars/exam2/exam.tex
new attempt at stars exam!

--- 2022-05-17 531_stars/exam2/exam.tex
@@
 
-\vspace{-1.5cm}
+\vspace{-0.5cm}
 
-\section*{Getting a white dwarf to chill out}
-\vspace{-0.35cm}
-\subsubsection*{Prof.\ Eyemund Returns}
+\noindent \textbf{[Exam format]}
+
+\noindent This exam is open note but closed book/internet. You have access to a list of constants and maybe use a tool for converting units (e.g.\ an iPython instance with astropy installed).
+
+\section*{Binary Limbo: How can you(r separation) go?!}
 \vspace{0.25cm}
-\question{Dearest Grad Student,\\
-
-    \noindent I have good news and bad news.
-The bad news is that it seems that our venture into the diamond creation business has gone slightly awry.
-The other faculty became suspicious after they noticed I was using that goblet of diamonds as a paperweight and after some lengthy arguments I have been legally forbidden from ever using white dwarfs for diamond production!
-The good news?
-Well they didn't say anything about neutron stars...\\
-
-    \vspace{-0.3cm}
-    \noindent Now the problem of course is that we've invested quite a lot of \sout{your PhD funding} money into finding that sample of white dwarfs.
-But I realised we don't need to \emph{find} a neutron star, we just need to \emph{make} a neutron star!!\\
-
-    \vspace{-0.3cm}
-    \noindent So I've done some experiments with adding mass to existing white dwarfs, but for some reason none of them reach the Chandrasekhar limit before exploding dramatically.
-I'm not entirely sure why this happens but it seems to be temperature dependent, therefore I've acquired a Cooling Ray to stop the envelope from getting too hot!
-It can cool at any rate we want, but it costs more to do so.
-So my question for you is \textbf{What's the lowest rate we could cool the outer layers of a white dwarf whilst it accretes material and still create a neutron star?}\\\\
-    Yours sincerely,
-
-    \noindent Prof. D. Eyemund
-    \answer{}
-}
-
-\question{\textbf{Part a - What's with those explosions?}}
-
-\question{Explain what is physically occuring at the surface of the white dwarf that leads to exoplosions during accretion.}
-
-\answer{
-    It's the ignition of thermonuclear fusion!
-As the white dwarf accretes material, the temperature of the envelope starts to heat up.
-Eventually, the temperature reaches the ignition temperature of H-fusion and thus fusion is rapidly ignited throughout the envelope, leading to an exoplosive ejection of the accreted shell of material.
-}
-
-\clearpage
-
-\question{\textbf{Part b - How quickly does an accreting WD heat up?}}
-
-\question{Derive the rate at which an accreting WD heats:
-\begin{equation}
-    \dot{T} = f(M_{\rm WD}, R_{\rm WD}, M_{\rm env}, \mu, \dot{M}_a)
-\end{equation}
-It should be a function of the mass of the white dwarf, radius of the white dwarf, mass of the surrounding envelope, mean molecular weight of envelope and the accretion rate.
-
-You should assume that all gravitational energy of the accreted material is converted to thermal energy in the WD envelope during accretion and that the white dwarf is already very close to the Chandrasekhar limit.
-Additionally, you can assume the envelope is an ideal gas.
-and its mass remains approximately constant.}
-
-\answer{
-    The gravitational energy for the accreted mass on the white dwarf is just a function of the masses and the separation.
-    \begin{equation}
-        E_{\rm grav} = - \frac{G M_{\rm WD} M_a}{R_{\rm WD}}
-    \end{equation}
-    Since the envelope of the WD is an ideal gas, the average thermal energy is
-    \begin{equation}
-        \avg{E_{\rm thermal}} = \frac{3}{2} k_B T
-    \end{equation}
-    We can convert this to the total thermal energy of the envelope by multiplying by the total number of particles in the envelope
-    \begin{equation}
-        E_{\rm thermal} = \avg{E_{\rm thermal}} \cdot N = \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H}
-    \end{equation}
-    Now, since we are assuming that all gravitational energy is converted to thermal energy then the time derivatives of these expressions should be equal and we can solve for the change in temperature
-    \begin{align}
-        - \dv{E_{\rm grav}}{t} &= \dv{E_{\rm thermal}}{t} \\
-        \dv{t} \frac{G M_{\rm WD} M_a}{R_{\rm WD}} &= \dv{t} \frac{3}{2} k_B T \cdot \frac{M_{\rm env}}{\mu m_H} \\
-        \frac{G M_{\rm WD} \dot{M}_a}{R_{\rm WD}} &= \frac{3}{2} k_B \dot{T} \cdot \frac{M_{\rm env}}{\mu m_H} \\
-        \Aboxed{ \dot{T} &= \frac{2}{3 k_B} \frac{\mu m_H}{M_{\rm env}} \frac{G M_{\rm WD} \dot{M}_a}{R_{\rm WD}} }
-    \end{align}
-}
-
-\question{\textbf{Part c - How fast does the laser need to cool the WD?}}
-
-\question{Calculate the minimum cooling rate required for the He-WD to reach the Chandrasekhar limit before any explosions occur.}
-
-\answer{
-    First, let's calculate the time until the WD reaches the Chandrasekhar limit
-    \begin{equation}
-        t_{\rm ch} = \frac{M_{\rm ch} - M_{\rm WD}}{\dot{M}_a}
-    \end{equation}
-    Now we can solve for the minimum cooling rate of the laser that keeps the temperature under the H-fusion ignition temperature in this amount of time.
-    \begin{align}
-        T_{\rm ign} &> (\dot{T}_{\rm accretion} - \dot{T}_{\rm laser}) t_{\rm Ch} \\
-        \dot{T}_{\rm laser} &> \dot{T}_{\rm accretion} - \frac{T_{\rm ign}}{t_{\rm Ch}} \\
-        \Aboxed{ \dot{T}_{\rm laser}&> \dot{M}_a \qty[\frac{2}{3 k_B} \frac{\mu m_H}{M_{\rm env}} \frac{G M_{\rm WD}}{R_{\rm WD}} - \frac{T_{\rm ign}}{M_{\rm ch} - M_{\rm WD}}] }
-    \end{align}
-}
 

### 2022-05-19 531_stars/exam2/exam.tex
set up question

@@
 
-\section*{Binary Limbo: How can you(r separation) go?!}
-\vspace{0.25cm}
+\section*{Binary Limbo: How low can you(r separation) go?!}
+
+\question{For each type of mass transfer listed below state whether there is a minimum separation that a binary can achieve during mass transfer.\\
+If you answer yes
+\begin{itemize}
+    \item Explain what physically occurs in the system when it reaches this minimum separation
+    \item Find an expression for the minimum separation in terms of the initial separation and initial mass ratio $q = M_{2, i} / M_{1, i}$ (assuming that the rate of mass transfer is constant)
+\end{itemize}
+If you answer no
+\begin{itemize}
+    \item Explain why a minimum separation does not exist
+\end{itemize}
+}
+
+\question{\textbf{Part a - Conservative and Stable Mass Transfer}}
+\answer{}
+
+\question{\textbf{Part b - Unstable Mass Transfer}}
+\answer{}
 

### 2022-05-19 531_stars/exam2/exam.tex
new version of stars exam 2

@@
 
-\noindent This exam is open note but closed book/internet. You have access to a list of constants and maybe use a tool for converting units (e.g.\ an iPython instance with astropy installed).
+\noindent This exam is open note but closed book/internet.
+\textit{[My purpose with this is that students will have written down Eq.\ 28.15 somewhere so can use that but they don't have the full explanations from the book about mass transfer and so will need to explain that themselves]}
 
@@
 \question{\textbf{Part a - Conservative and Stable Mass Transfer}}
-\answer{}
+\answer{
+    \textbf{Yes}, a minimum separation exists.
+This minimum occurs when the mass ratio of the binary flips, meaning that the initially more massive star becomes less massive than the secondary star due to the transfer of mass.
+At this point, the binary stops shrinking and the separation increases instead until mass transfer ends.
 
-\question{\textbf{Part b - Unstable Mass Transfer}}
-\answer{}
+    We can find an expression for this point using Eq.\ 28.15 from the textbook. This gives that the change in separation during mass transfer is
+    \begin{equation}
+        \frac{\dot{a}}{a} = 2 \frac{\dot{M}_1}{M_1} \qty(\frac{M_1}{M_2} - 1)
+    \end{equation}
+    Therefore, since we are assuming that the mass transfer rate is constant, we can integrate this to find the minimum separation.
+    \begin{align}
+        \int_{a_i}^{a_{\rm min}} \frac{1}{a} \dd{a} &= 2 \frac{\dot{M}_1}{M_1} \qty(\frac{M_1}{M_2} - 1) \int_0^{t_{\rm min}} \dd{t} \\
+        \ln \frac{a_{\rm min}}{a_i} &= 2 \frac{\dot{M}_1}{M_1} \qty(\frac{M_1}{M_2} - 1) \cdot t_{\rm min}
+    \end{align}
+    The time at which the minimum occurs is the time at which the masses are equal. Therefore, this time is when half the difference in their initial masses has been transferred and so
+    \begin{equation}
+        t_{\rm min} = -\frac{M_1 - M_2}{2 \dot{M}_1}
+    \end{equation}
+    where the negative sign accounts for the fact that $\dot{M}_1$ is negative. Putting this all together and substituting $q$ we find that
+    \begin{align}
+        \ln \frac{a_{\rm min}}{a_i} &= 2 - q - \frac{1}{q} \\
+        \Aboxed{ a_{\rm min} &= a_i \, e^{2 - q - \frac{1}{q}} }
+    \end{align}
+    This function is plotted in Figure~\ref{fig:min_sep} for students reading the solutions who are interested but this is of course not necessary in their answer.
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.8\textwidth]{relative_min_separation.pdf}
+        \caption{Minimum separation relative to initial separation for different initial mass ratios (for conservative, stable mass transfer at a constant rate).}
+        \label{fig:min_sep}
+    \end{figure}
+}
+
+\question{\textbf{Part b - Dynamically Unstable Mass Transfer}}
+\answer{
+    \textbf{No}, there is no minimum separation for dynamically unstable mass transfer.
+This is because this is a runaway mass transfer process.
+The donor star does not contract sufficiently (or even expands further) and so does not balance with its shrinking Roche Lobe and thus the mass transfer becomes so fast that the donor star is no longer in hydrostatic equilibrium.
+
+    This leads to an extremely high mass transfer rate such that soon \textit{both} stars overflow their Roche Lobes and this initiates a common envelope event.
+This common envelope is the reason that no minimum separation exists.
+The stars orbit through the gas of the common envelope which results in friction and dramatically decreases the orbital separation.
+There is no limit to how close this effect may drive the stars, indeed in many cases a common envelope leads to a stellar merger in which the stars merge entirely and thus have zero orbital separation.
+}
 

### 2022-08-20 581_observing/observing_planning/solutions.tex
observing planning activity

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\psetnum}{1}
\newcommand{\class}{ASTR 581 - Observing}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Observing Planning \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Observing Planning}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\tomtitle

\noindent I use astroplan/astropy throughout. You can see my code here: \url{https://github.com/TomWagg/uw-grad-classes/tree/main/581_observing/observing_planning/code.ipynb}\\

\question{1. APO Location}
\answer{
    Latitude: 32.7803 degrees North\\
    Longitude: 105.8203 degrees West
}

\question{2. APO Sunset}
\answer{
    Sunset is at 01:01 UTC and 19:01 MST.
}

\question{3. Sidereal Time}
\answer{
    Honestly a little confused on this because astropy gave me an hour angle but I \textit{think} the time is 23:23:16.
}

\question{4. Moon Phase}
\answer{
    Waning Crescent
}

\question{5. Moon time}
\answer{
    06:49 UTC
}

\question{6. Moon inteference}
\answer{
    I think it depends. The moon is waning crescent so it's not too bright, and it doesn't rise until the B half so A half observations won't be affected. For the B half it will depend on the separation of the target from the moon and how bright it is.
}

\clearpage

\question{7. RA/Dec of objects}
\answer{
    M91: (88.859260, +14.4958754)\\
    NGC 5866: (26.6231708, +55.7633927)
}

\question{8. Observable from MRO}
\answer{
    Only NGC 5866 is observable at this time of year
}

\question{9. Airmass conditions}
\answer{
    NGC 5866 will only have an airmass less than 1.5 for about 30 minutes after sunset (before twilight) so it may be difficult to see.
}

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.7\textwidth]{mro_galaxies.pdf}
\end{figure}

\question{10. Morning twilight}
\answer{
    Astronomial morning twilight occurs at 12:02:20 UTC so any time after that you'll start struggling to see fainter sources.
}

\question{11. Best seasons to observe}
\answer{
    March is best for M91, and May is best for NGC 5866. These months maximise the time this object spends with airmass $< 1.5$ at MRO.
}

\question{12. Dust?}
\answer{
    No need to worry. These objects have galactic latitudes of 76.82963912 and 52.5908177 degrees so they are far from the galactic plane.
}

\end{document}

### 2022-10-08 541_ism/week1/solutions.tex
start astr541, add week1 q1a

--- 2022-02-24 pset3/solutions.tex
@@
 
-\title{Astro 507; Problem Set 3}
+\title{ASTR 541; Week 1}
 \author{\textbf{Tom Wagg}}
@@
 
-\question{1. \textbf{FIRAS}}
-
-\answer{}
-
-\question{2a. \textbf{Entropy}}
+\question{1a. \textbf{Hydrogen density}}
 \answer{
-    Let's derive the entropy for an ideal, non-relativistic Fermi gas (in terms of $V$, $z$ and $T$). For this I will follow Lecture 10. The entropy is defined as
+    We can first calculate the density of the disc as
+    \begin{align}
+        \rho &= \frac{M}{\pi R_{\rm disc}^2 H} \\
+             &= 1.9 \times 10^{-24} \unit{g}{cm^{-3}}
+    \end{align}
+    Now we need to convert the density to a number density as follows
+    \begin{align}
+        \rho &= m_{\rm H} n_{\rm H} + m_{\rm He} n_{\rm He} \\
+             &= n_{\rm H} \qty(m_{\rm H} + m_{\rm He} \frac{n_{\rm He}}{n_{\rm H}}) \\
+             &= n_{\rm H} \qty(m_{\rm H} + 0.1 m_{\rm He}) \\
+             &= 1.4 n_{\rm H} m_p \\
+        n_{\rm H} &= \frac{\rho}{1.4 m_p}
+    \end{align}
+    This gives that the average number density of hydrogen is
     \begin{equation}
-        S = \frac{U + PV - N \mu}{T},
-    \end{equation}
-    so we just need to use definitions of $U$, $P$, $N$ and $\mu$. The definitions of $\mu$ is trivial when using the fugacity
-    \begin{equation}
-        \mu = k_B T \ln z
-    \end{equation}
-    For the others, I refer to lecture 10 where we can find the definition of $n$ on slide 5
-    \begin{equation}
-        N = n V = \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V F_{1/2}(z)
-    \end{equation}
-    and the definition of $P$ on slide 7
-    \begin{equation}
-        P = \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z)
-    \end{equation}
-    For the total energy, we don't have it from the slides but we can derive it in the same way using Fermi-Dirac integrals.
-It starts in the same way as $n$ but with an extra factor of $\epsilon$ in the numerator, then we use the substitution $w = \epsilon / k_B T$
-    \begin{align}
-        U &= (2s + 1) \frac{V}{h^3} \int \frac{\epsilon}{e^{\frac{\epsilon - \mu}{k_B T}} + 1} \dd{\va{p}} \\
-        U &= \frac{4\pi (2s + 1)}{2 h^3} (2 m k_B T)^{3/2} k_B T \int_0^\infty \dd{w} \frac{w^{3/2}}{e^w z^{-1} + 1} \\
-        U &= \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T F_{3/2}(z)
-    \end{align}
-    Now it's simply a matter of plugging all of these into the entropy expression.
-    \begin{align}
-        S &= \frac{U + PV - N \mu}{T} \\
-          &= \frac{1}{T} \qty[ \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T F_{3/2}(z) + \frac{4(2s + 1)}{3 \pi^{1/2}} V k_B T \lambda^{-3} F_{3/2}(z) + \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} V k_B T \ln z F_{1/2}(z)] \\
-          &= \frac{2(2s + 1) V k_B T}{\pi^{1/2} \lambda^3} \qty[ F_{3/2}(z) + \frac{2}{3} F_{3/2}(z) + \ln z F_{1/2}(z)] \\
-        \Aboxed{ S &= \frac{2(2s + 1) V k_B T}{\pi^{1/2} \lambda^3} \qty[ \frac{5}{3} F_{3/2}(z) + F_{1/2}(z) \ln z] }
-    \end{align}
-}
-
-\question{2b. \textbf{Expanding pressure}}
-\answer{
-    \begin{align}
-        \frac{P}{n k_B T} &= \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z) \qty[\frac{2(2s + 1)}{\pi^{1/2} \lambda^3} k_B T F_{1/2}(z)]^{-1} \\
-                          &= \frac{2 F_{3/2}(z)}{3 F_{1/2}(z)}
-    \end{align}
-    \todo{come back to this on Thursday}
-}
-
-\question{\textbf{3. Brown Dwarf}}
-
-\question{3a. \textbf{Fugacity}}
-\answer{
-    The general goal here is to find the number density based on the central temperature and density and then solve for the Fermi-Dirac integral and we can then invert that to find the fugacity.
-To start, we know that
-    \begin{equation}
-        \rho = m_e n_e + m_{\rm H} n_{\rm H} + m_{\rm He} n_{\rm He}
-    \end{equation}
-    We can rewrite this (and neglect the electron term) as
-    \begin{equation}
-        \rho = n_{\rm H} \qty(m_{\rm H} + m_{\rm He} \frac{n_{\rm He}}{n_{\rm H}})
-    \end{equation}
-    Now we can plug in the masses of hydrogen and helium as well as their relative abundance (given as 0.1 in the problem description).
-    \begin{equation}
-        n_{\rm H} = \frac{\rho_c}{1.4 m_p}
-    \end{equation}
-    Now we just relate this to the electron number density simply as
-    \begin{equation}
-        n = \qty(2 \frac{n_{\rm He}}{n_{\rm H}} + 1) n_{\rm H}
-    \end{equation}
-    And combining the two gives an expression for the number density in terms of the central density
-    \begin{equation}
-        n = \frac{1.2 \rho_c}{1.4 m_p}
-    \end{equation}
-    The other part that we need is the de-Broglie wavelength which depends on the central temperature as
-    \begin{equation}
-        \lambda = \frac{h}{\sqrt{2 \pi m_e k_B T_c}}
-    \end{equation}
-    Now let's use this with the equation for the number density from Lecture 10 and solve for the Fermi-Dirac integral (plugging in values in the final equation).
-    \begin{align}
-        n &= \frac{2(2s + 1)}{\pi^{1/2} \lambda^3} F_{1/2}(z) \\
-        F_{1/2}(z) &= \frac{n \pi^{1/2} \lambda^3}{2(2s + 1)} \\
-        F_{1/2}(z) &= \frac{\pi^{1/2}}{2(2s + 1)} \frac{1.2 \rho_c}{1.4 m_p} \qty(\frac{h}{\sqrt{2 \pi m_e k_B T_c}})^3 \\
-        F_{1/2}(z) &= 2.08
-    \end{align}
-    I took this value and inverted it using the approximate analytic expressions from the paper (Aymerich-Humet+1981) to find the fugacity.
-    \begin{equation}
-        \boxed{ z = 1.65 }
+        \boxed{ n_{\rm H} = 0.8 \unit{cm^{-3}} }
     \end{equation}
@@
 
-\question{3b. \textbf{Pressure}}
-\answer{
-    In order to calculate the pressure we need to apply the same pressure equation as 2a
-    \begin{equation}
-        P = \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z)
-    \end{equation}
-    In this specific case, $s = 1/2$, $T = 6 \times 10^6 \unit{K}$, $\lambda$ is calculated in the same way as part a and we found that $z = 1.65$ in part a. This gives the pressure as
-    \begin{equation}
-        \boxed{ P = 1.92 \times 10^{16} \unit{Pa} }
-    \end{equation}
-}
-
-\question{3c. \textbf{Relativistic?}}
-\answer{
-    In lecture we found that electrons are relativistic when their fermi momentum is such that
-    \begin{equation}
-        p_{\rm F} \sim m_e c
-    \end{equation}
-    We showed that in terms of density this was equivalent to
-    \begin{equation}
-        \rho_{\rm rel} \sim 2 \times 10^6 \unit{g}{cm^{-3}} \cdot \frac{\mu_e}{2 m_p}
-    \end{equation}
-    The last term with $\mu_e$ is going to be approximately of order unity and so, since the central density is only $325 \unit{g}{cm^{-3}}$ we have that the electrons \textbf{are} relativistic
-    \begin{equation}
-        \boxed{ \rho_c \ll \rho_{\rm rel} \implies \text{non-relativistic} }
-    \end{equation}
-}
-
-\question{3d. \textbf{Ideal Gas}}
-\answer{
-    For this part we simply need to calculate the pressure assuming that it was an ideal gas. In this case
-    \begin{equation}
-        P_{\rm ideal} = n k_B T
-    \end{equation}
-    We found $n$ in part a and $T = 6 \times 10^6 \unit{K}$ and thus we can quickly plug these in to find that
-    \begin{equation}
-        \boxed{P_{\rm ideal} = 1.38 \times 10^{16} \unit{Pa}}
-    \end{equation}
-    We can therefore see that the degeneracy contributes significantly since the pressure is about 40\% larger when including degeneracy effects.
-}
-
-\question{3e. \textbf{Polytropes}}
-\answer{
-    We are given the two following equations
-    \begin{align}
-        \rho_c &= 8.44 \unit{g}{cm^{-3}} \frac{M}{R^3} \\
-        R M^{1/3} &= \frac{0.13}{(\mu_e F_{1/2}(z))^{2/3}}
-    \end{align}
-    First, let's quickly find the mass per electron in units of the proton mass, $\mu_e$, knowing that the gas is 90\% hydrogen (1 nucleon, 1 electron) and 10\% helium (4 nucleons, 2 electrons).
-    \begin{equation}
-        \mu_e = 0.9 \cdot 1 + 0.1 \cdot 2 = 1.1
-    \end{equation}
-    Now we can combine these equations to instead write that
-    \begin{align}
-        M &= \frac{\rho_c R^3}{8.44 \unit{g}{cm^{-3}}} \\
-        R &= \qty[\frac{0.13}{(\mu_e F_{1/2}(z))^{2/3}} \qty(\frac{\rho_c}{8.44 \unit{g}{cm^{-3}}})^{-1/3}]^{1/2}
-    \end{align}
-    Plugging in our numbers $\mu_e = 1.1$, $z = 1.65$ and $\rho_c = 325 \unit{g}{cm^{-3}}$ to get R (and then M) gives
-    \begin{empheq}[box=\fbox]{align}
-        R &= 0.15 \unit{R_{\odot}} \\
-        M &= 0.13 \unit{M_{\odot}}
-    \end{empheq}
-}
-
 \end{document}

### 2022-10-08 541_ism/week1/problem_1.py

@@ -1,0 +1,13 @@
+import astropy.units as u
+import astropy.constants as const
+import numpy as np
+
+Mdisc = 4e9 * u.Msun
+Rdisc = 15 * u.kpc
+H = 200 * u.pc
+
+density = (Mdisc / (np.pi * Rdisc**2 * H)).to(u.g / u.cm**3)
+print(density)
+
+n_H = (density / (1.4 * const.m_p)).to(u.cm**(-3))
+print(n_H)

### 2022-10-08 541_ism/week1/problem_1.py
541, week1 1c

@@ -5,9 +5,29 @@ import numpy as np
 Mdisc = 4e9 * u.Msun
 Rdisc = 15 * u.kpc
 H = 200 * u.pc
+Vdisc = (np.pi * Rdisc**2 * H).to(u.cm**3)
 
-density = (Mdisc / (np.pi * Rdisc**2 * H)).to(u.g / u.cm**3)
-print(density)
+disc_density = (Mdisc / Vdisc).to(u.g / u.cm**3)
+print(f"Disc density {disc_density:1.1e}")
 
-n_H = (density / (1.4 * const.m_p)).to(u.cm**(-3))
-print(n_H)
+n_H = (disc_density / (1.4 * const.m_p)).to(u.cm**(-3))
+print(f"n_H {n_H:1.1e}")
+print()
+
+
+mc_vol = 4/3 * np.pi * (15 * u.pc)**3
+h2_mass = 2 * const.m_p
+mc_density = h2_mass * 100 * u.cm**(-3)
+mc_mass_single = (mc_vol * mc_density).to(u.Msun)
+print(f"Mass of a MC: {mc_mass_single:1.1e}")
+
+mc_count = 0.3 * Mdisc / (mc_mass_single)
+print(f"Number of MCs: {mc_count:1.1e}")
+print()
+
+# print(f"Fraction of volume occupied {(mc_vol / Vdisc).decompose():1.1e}")
+
+v_los = (np.pi * (15 * u.pc)**2 * (8.5 * u.kpc)).to(u.cm**3)
+print("Volume along line of sight")
+
+print((v_los / Vdisc * mc_count).decompose())

### 2022-10-08 541_ism/week1/solutions.tex
541, week1 1c
541 week1 1d attempt

@@
 
+\question{1b. \textbf{Dust grain density}}
+\answer{
+    ???
+}
+
+\question{1c. \textbf{Molecular clouds - frequency and mass}}
+\answer{
+    The typical mass can be found by multiplying the number density, mass of particles and volume.
+    \begin{align}
+        M_{\rm MC} &= m_{\rm H_2} n_{H_2} V_{\rm MC} \\
+                   &= 2 m_{p} n_{H_2} \qty(\frac{4}{3} \pi R_{\rm MC}^3) \\
+        \Aboxed{ M_{\rm MC} &= 10^{5} \unit{M_\odot} }
+    \end{align}
+    The approximate number of clouds in the galaxy is then just
+    \begin{align}
+        N_{\rm MC} &= \frac{0.3 M_{\rm disc}}{M_{\rm MC}} \\
+        \Aboxed{ N_{\rm MC} &= 1.1 \times 10^{4} }
+    \end{align}
+}
+
+\question{1d. \textbf{Molecular clouds getting in the way}}
+\answer{
+    If the centre of a molecular cloud passes within a radius of the line of sight then it will obscure our view. This defines the volume within which a molecular cloud could block our sight as
+    \begin{equation}
+        V_{\rm LoS} = \pi R_{\rm MC}^2 \cdot 8.5 \unit{kpc}
+    \end{equation}
+    Given the total volume of the disc and the approximate number of molecular clouds in the disc, this implies that the expectation value of the number of clouds intersecting our line of sight is
+    \begin{equation}
+        \boxed{ N = \frac{V_{\rm LoS}}{V_{\rm disc}} N_{\rm MC} = 0.73 }
+    \end{equation}
+}
+
 \end{document}

### 2022-10-08 541_ism/week1/solutions.tex
541 - finish week 1 problems

@@
 
-\title{ASTR 541; Week 1}
+\newcommand{\psetnum}{1}
+\newcommand{\class}{ASTR 541 - Interstellar Medium}
+
+\newcommand{\tomtitle}{
+    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
+    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
+    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
+}
+
+\title{\class : Week \psetnum}
 \author{\textbf{Tom Wagg}}
@@
 
-\maketitle
+\tomtitle{}
 
@@
 \answer{
-    ???
+    The mass of each dust grain is given by
+    \begin{align}
+        m_{\rm DG} &= \frac{4}{3} \pi R_{\rm RG}^3 \cdot \rho_{\rm RG} \\
+                   &= 8.4 \times 10^{-15} \unit{g}
+    \end{align}
+    We are given the total dust mass and we already calculated the volume of the disc in part a so now the number density is just
+    \begin{align}
+        n_{\rm DG} &= \frac{M_{\rm dust}}{m_{\rm DG}} \cdot \frac{1}{V_{\rm disc}} \\
+        \Aboxed{ n_{\rm DG} &= 1.6 \times 10^{-12} \unit{cm^{-3}} }
+    \end{align}
 }
@@
         M_{\rm MC} &= m_{\rm H_2} n_{H_2} V_{\rm MC} \\
-                   &= 2 m_{p} n_{H_2} \qty(\frac{4}{3} \pi R_{\rm MC}^3) \\
-        \Aboxed{ M_{\rm MC} &= 10^{5} \unit{M_\odot} }
+                   &= (2 m_{p}) \cdot n_{H_2} \cdot \qty(\frac{4}{3} \pi R_{\rm MC}^3) \\
+        \Aboxed{ M_{\rm MC} &= 7 \times 10^{4} \unit{M_\odot} }
     \end{align}
@@
         N_{\rm MC} &= \frac{0.3 M_{\rm disc}}{M_{\rm MC}} \\
-        \Aboxed{ N_{\rm MC} &= 1.1 \times 10^{4} }
+        \Aboxed{ N_{\rm MC} &= 1.7 \times 10^{4} }
     \end{align}
@@
     \begin{equation}
-        V_{\rm LoS} = \pi R_{\rm MC}^2 \cdot 8.5 \unit{kpc}
+        V_{\rm LoS} = \pi R_{\rm MC}^2 \cdot 8.5 \unit{kpc} = 6 \times 10^{6} \unit{pc^3}
     \end{equation}

### 2022-10-08 541_ism/week1/problem_1.py

@@ -14,6 +14,17 @@ n_H = (disc_density / (1.4 * const.m_p)).to(u.cm**(-3))
 print(f"n_H {n_H:1.1e}")
 print()
 
+v_dg = 4/3 * np.pi * (1000 * u.Angstrom)**3
+rho_dg = 2 * u.g / u.cm**3
+m_dg = (v_dg * rho_dg).to(u.g)
+print(f"Mass of a dust grain: {m_dg:1.1e}")
+
+m_dust = 0.7 / 100 * Mdisc
+N_dg = m_dust / m_dg
+n_dg = (N_dg / Vdisc).to(u.cm**(-3))
+print(f"Number density of dust grains {n_dg: 1.1e}")
+print()
+
 
 mc_vol = 4/3 * np.pi * (15 * u.pc)**3
 h2_mass = 2 * const.m_p
@@ -27,7 +38,7 @@ print()
 
 # print(f"Fraction of volume occupied {(mc_vol / Vdisc).decompose():1.1e}")
 
-v_los = (np.pi * (15 * u.pc)**2 * (8.5 * u.kpc)).to(u.cm**3)
-print("Volume along line of sight")
+v_los = (np.pi * (15 * u.pc)**2 * (8.5 * u.kpc)).to(u.pc**3)
+print(f"Volume along line of sight: {v_los:1.1e}")
 
 print((v_los / Vdisc * mc_count).decompose())

### 2022-10-10 541_ism/week1/problem_1.py
fix part c density (thanks Andy!)

@@ -27,8 +27,7 @@ print()
 
 
 mc_vol = 4/3 * np.pi * (15 * u.pc)**3
-h2_mass = 2 * const.m_p
-mc_density = h2_mass * 100 * u.cm**(-3)
+mc_density = 1.4 * 2 * const.m_p * 100 * u.cm**(-3)
 mc_mass_single = (mc_vol * mc_density).to(u.Msun)
 print(f"Mass of a MC: {mc_mass_single:1.1e}")
 

### 2022-10-10 541_ism/week1/solutions.tex
update solutions with part c changes

@@
 \answer{
-    The typical mass can be found by multiplying the number density, mass of particles and volume.
+    First we can find the density of a cloud using the number density that we've been given and the conversion that we found in part a to account for the helium fraction (I'm ignoring dust here because it is such a small fraction).
+    \begin{equation}
+        \rho_{\rm MC} = 1.4 \cdot 2 m_p \cdot n(H_2)
+    \end{equation}
+    The typical mass can be found by multiplying the density and volume.
     \begin{align}
-        M_{\rm MC} &= m_{\rm H_2} n_{H_2} V_{\rm MC} \\
-                   &= (2 m_{p}) \cdot n_{H_2} \cdot \qty(\frac{4}{3} \pi R_{\rm MC}^3) \\
-        \Aboxed{ M_{\rm MC} &= 7 \times 10^{4} \unit{M_\odot} }
+        M_{\rm MC} &= \rho_{\rm MC} V_{\rm MC} \\
+                   &= 1.4 \cdot (2 m_{p}) \cdot n_{H_2} \cdot \qty(\frac{4}{3} \pi R_{\rm MC}^3) \\
+        \Aboxed{ M_{\rm MC} &= 9.8 \times 10^{4} \unit{M_\odot} }
     \end{align}
@@
         N_{\rm MC} &= \frac{0.3 M_{\rm disc}}{M_{\rm MC}} \\
-        \Aboxed{ N_{\rm MC} &= 1.7 \times 10^{4} }
+        \Aboxed{ N_{\rm MC} &= 1.2 \times 10^{4} }
     \end{align}
@@
     \begin{equation}
-        \boxed{ N = \frac{V_{\rm LoS}}{V_{\rm disc}} N_{\rm MC} = 0.73 }
+        \boxed{ \mathbb{E}[N_{\rm LoS}] = \frac{V_{\rm LoS}}{V_{\rm disc}} N_{\rm MC} = 0.52}
     \end{equation}

### 2022-10-10 541_ism/week2/solutions.tex
setup 541 week 2 problem set

--- 2022-10-08 541_ism/week1/solutions.tex
@@
 
-\title{ASTR 541; Week 1}
+\newcommand{\psetnum}{2}
+\newcommand{\class}{ASTR 541 - Interstellar Medium}
+
+\newcommand{\tomtitle}{
+    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
+    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
+    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
+}
+
+\title{\class : Week \psetnum}
 \author{\textbf{Tom Wagg}}
@@
 
-\maketitle
+\tomtitle{}
 
-\question{1a. \textbf{Hydrogen density}}
+\question{\textbf{1. Maxwellians}}
+
 \answer{
-    We can first calculate the density of the disc as
-    \begin{align}
-        \rho &= \frac{M}{\pi R_{\rm disc}^2 H} \\
-             &= 1.9 \times 10^{-24} \unit{g}{cm^{-3}}
-    \end{align}
-    Now we need to convert the density to a number density as follows
-    \begin{align}
-        \rho &= m_{\rm H} n_{\rm H} + m_{\rm He} n_{\rm He} \\
-             &= n_{\rm H} \qty(m_{\rm H} + m_{\rm He} \frac{n_{\rm He}}{n_{\rm H}}) \\
-             &= n_{\rm H} \qty(m_{\rm H} + 0.1 m_{\rm He}) \\
-             &= 1.4 n_{\rm H} m_p \\
-        n_{\rm H} &= \frac{\rho}{1.4 m_p}
-    \end{align}
-    This gives that the average number density of hydrogen is
-    \begin{equation}
-        \boxed{ n_{\rm H} = 0.8 \unit{cm^{-3}} }
-    \end{equation}
+    .
 }
 
-\question{1b. \textbf{Dust grain density}}
+\question{\textbf{2. The Dust Collector}}
+
 \answer{
-    ???
-}
-
-\question{1c. \textbf{Molecular clouds - frequency and mass}}
-\answer{
-    The typical mass can be found by multiplying the number density, mass of particles and volume.
-    \begin{align}
-        M_{\rm MC} &= m_{\rm H_2} n_{H_2} V_{\rm MC} \\
-                   &= 2 m_{p} n_{H_2} \qty(\frac{4}{3} \pi R_{\rm MC}^3) \\
-        \Aboxed{ M_{\rm MC} &= 10^{5} \unit{M_\odot} }
-    \end{align}
-    The approximate number of clouds in the galaxy is then just
-    \begin{align}
-        N_{\rm MC} &= \frac{0.3 M_{\rm disc}}{M_{\rm MC}} \\
-        \Aboxed{ N_{\rm MC} &= 1.1 \times 10^{4} }
-    \end{align}
-}
-
-\question{1d. \textbf{Molecular clouds getting in the way}}
-\answer{
-    If the centre of a molecular cloud passes within a radius of the line of sight then it will obscure our view. This defines the volume within which a molecular cloud could block our sight as
-    \begin{equation}
-        V_{\rm LoS} = \pi R_{\rm MC}^2 \cdot 8.5 \unit{kpc}
-    \end{equation}
-    Given the total volume of the disc and the approximate number of molecular clouds in the disc, this implies that the expectation value of the number of clouds intersecting our line of sight is
-    \begin{equation}
-        \boxed{ N = \frac{V_{\rm LoS}}{V_{\rm disc}} N_{\rm MC} = 0.73 }
-    \end{equation}
+    .
 }

### 2022-10-10 541_ism/week2/problem_1.py

@@ -1,0 +1,5 @@
+import astropy.units as u
+import astropy.constants as const
+import numpy as np
+from scipy.integrate import quad
+

### 2022-10-10 541_ism/week2/solutions.tex
week 2 problem 1 astr541

@@
 \answer{
-    .
+    The minimum velocity needed to ionise a hydrogen atom can be calculated by setting the kinetic energy equal to the ionisation energy
+    \begin{align}
+        \frac{1}{2} m_{e} v_{\rm ionise}^2 &= E_{\rm ionise} \\
+        v_{\rm ionise} &= \sqrt{\frac{2 E_{\rm ionise}}{m_e}} \\
+        \Aboxed{ v_{\rm ionise} &= 2187 \unit{km}{s^{-1}}}
+    \end{align}
+    We now need to integrate the Maxwellian (yay?).
+Now I don't intend to do that analytically because I am not sadistic and therefore will be numerically integrating it will \texttt{scipy}'s quadrature function.
+Additionally, it doesn't play nicely with infinity so we're going to be sneaky and flip it around
+    \begin{align}
+        f_{\rm ionise} &= \int_{v_{\rm ionise}}^{\infty} 4 \pi \qty(\frac{m_e}{2 \pi k_B T})^{\frac{3}{2}} \exp\qty(-\frac{m v^2}{2 k_B T}) v^2 \dd{v} \\
+        f_{\rm ionise} &= 1 - \int_0^{v_{\rm ionise}} 4 \pi \qty(\frac{m_e}{2 \pi k_B T})^{\frac{3}{2}} \exp\qty(-\frac{m v^2}{2 k_B T}) v^2 \dd{v}
+    \end{align}
+    Now we can evaluate this for both temperatures given
+    \begin{equation}
+        \boxed{ f_{\rm ionise, HII} = 6.5 \times 10^{-7},\quad f_{\rm ionise, halo} = 0.96 }
+    \end{equation}
+    From this we can conclude that collisional ionisation is extremely unlikely to occur in an HII region, but it will be dominant for hot halo gas.
 }
@@
 
-\answer{
-    .
-}
+\answer{}
 

### 2022-10-10 541_ism/week2/problem_1.py

@@ -3,3 +3,15 @@ import astropy.constants as const
 import numpy as np
 from scipy.integrate import quad
 
+
+def maxwellian(v, m, T):
+    exp_term = np.exp(- m * v**2 / (2 * const.k_B.si.value * T))
+    return (4 * np.pi * (m / (2 * np.pi * const.k_B.si.value * T))**(3/2) * v**2 * exp_term)
+
+E_ionise = 13.6 * u.eV
+v_ionise = np.sqrt(2 * E_ionise / const.m_e).to(u.km / u.s)
+print(v_ionise)
+
+for T in [1e4, 1e6]:
+    prob = 1 - quad(maxwellian, 0, v_ionise.si.value, args=(const.m_e.si.value, T))[0]
+    print(f"At T={T * u.K:1.1e}, fraction {prob:1.1e} of electrons are able to ionise")

### 2022-10-12 541_ism/week2/solutions.tex
541 - finish week 2 problems

@@
 
-\answer{}
+\answer{
+    The rate at which we collect dust grains will be given by the usual form of
+    \begin{equation}
+        \mathcal{R} = n \sigma v
+    \end{equation}
+    We are given a collecting area, $A$, which corresponds to the cross section and velocity, $v_{\rm W}$, so those parts are easy.
+The more complicated part is that we need to know the number density of the ISM dust particles.\\
+
+    \noindent Luckily for us, the problem from last week is going to be rather relevant! So we can apply what we did in the first parts of last week's problem set
+    \begin{align}
+        \rho_{\rm ISM} &= 1.4 m_p \cdot n_{\rm H} \\
+        \rho_{\rm dust} &= 0.005 \rho_{\rm ISM} \\
+        m_{\rm dust, particle} &= \frac{4}{3} \pi a^3 \cdot \rho_{\rm dust, particle} \\
+    \end{align}
+    Then we can use these things to get the number density of dust
+    \begin{align}
+        n_{\rm dust} &= \frac{\rho_{\rm dust}}{m_{\rm dust, particle}} \\
+                     &= \frac{0.005 \cdot 1.4 m_p \cdot n_{\rm H}}{\frac{4}{3} \pi a^3 \cdot \rho_{\rm dust, particle}} \\
+        n_{\rm dust} &= 3.1 \times 10^{-13} \unit{cm^{-3}}
+    \end{align}
+    Now we can use this to find the required area to get a rate of 1 particle per hour
+    \begin{equation}
+        \boxed{ A = \frac{\mathcal{R}}{n_{\rm dust} v_W} = \frac{1 \unit{hr^{-1}}}{3 \times 10^{-13} \unit{cm^{-3}} \cdot 26 \unit{km}{s^{-1}}} = 350 \unit{cm^2} }
+    \end{equation}
+
+}
 

### 2022-10-12 541_ism/week2/problem_2.py

@@ -1,0 +1,16 @@
+import astropy.units as u
+import astropy.constants as const
+import numpy as np
+
+n_H = 0.22 * u.cm**(-3)
+rho = 1.4 * const.m_p * n_H
+rho_dust = 0.5 / 100 * rho
+m_dust_particle = 4/3 * np.pi * (0.1 * u.micrometer)**3 * (2 * u.g * u.cm**(-3))
+n_dust = rho_dust / m_dust_particle
+
+print(f"Number density of dust {n_dust.to(u.cm**(-3)): 1.1e}")
+
+v = 26 * u.km / u.s
+
+collecting_area = ((1 / u.hr) / (n_dust * v)).to(u.cm**2)
+print(f"Collecting area required: {collecting_area:1.1f}")

### 2022-10-17 541_ism/week3/solutions.tex
astr541 - week 3 problem 1

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{mhchem}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\psetnum}{3}
\newcommand{\class}{ASTR 541 - Interstellar Medium}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Week \psetnum}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\tomtitle{}

\noindent For reference, if you'd ever like to see the code that I've used to get my answers to these, \href{https://github.com/TomWagg/uw-grad-classes/tree/main/541_ism}{here's a link to my GitHub repo}! (\#astropy.units for life)\\

\question{\textbf{1. Stromgren Spheres}}

\question{1a. What matters about the star?}

\answer{
    When finding the Stromgren radius for a region we need to know the ionisation rate. We derived in class that this radius is given as
    \begin{equation}
        Q_{\rm ionise} = \int_{\nu_{\rm ionise}}^{\infty} \frac{L_\nu}{h \nu} \dd{\nu}
    \end{equation}
    Therefore, we need to know the \emph{luminosity} of an O7V star and then we can integrate that to get the ionisation coefficient. But goodness me that would be far too \textit{Draine}ing so we can look it up in the textbook instead! From Table 15.1 we find that the rate is
    \begin{equation}
        \boxed{ Q_{\rm ionise, H} = 10^{48.75} \unit{s^{-1}} }
    \end{equation}
}

\question{1b. A hydrogen sphere}
\answer{
    We derived in class that the radius of a Stromgren sphere is given by the following equation
    \begin{equation}
        R_{\rm S} = \qty[ \frac{3}{4 \pi} \frac{Q_{\rm ionise}}{n_e n_{\rm ion} \alpha(T)} ]^{1/3}
    \end{equation}
    So to find the radius all we need to do is get values for each of the quantities above. For this part of the question we're just focussing on the hydrogen sphere and so the values are fairly simple
    \begin{equation}
        n_e = n_{\rm H},\quad n_{\rm ion} = n_{\rm H},\quad \alpha(T) = 4 \times 10^{-13} \unit{cm^{3}}{s^{-1}},\quad Q_{\rm ionise} = Q_{\rm ionise, H}
    \end{equation}
    where $n_{\rm H} = 5000 \unit{cm^{-3}}$ was given in the problem. Plugging these numbers in gives the radius as
    \begin{equation}
        \boxed{ R_{\rm H} = 0.17 \unit{pc} }
    \end{equation}
    This may at first glance cause you undue stress - ``this violates our understanding that HII regions are (several) tens of pc in size, I just did a quiz on this'' you say! Worry not. You can note that the density used in this question is \emph{extremely} high and so we are dealing with an ``ultra-compact'' HII region in this case and hence the radius is smaller than you may predict.
}

\clearpage

\question{1c. A helium sphere}
\answer{
    For the helium sphere it is a little different since we need to account for the hydrogen as well. As a simple first step, we are given the values of $\alpha$ and $Q$ (relative to hydrogen) in the question
    \begin{equation}
        \alpha = 8 \times 10^{-13} \unit{cm^3}{s^{-1}},\quad Q_{\rm ionise} = 0.135 \cdot Q_{\rm ionise, H}
    \end{equation}
    The difference lies in the number of electrons per ion. In this case we need to include the electrons from both the hydrogen and the helium. Each supplies only one electron since the star is not hot enough to doubly ionise helium consistently.
    \begin{equation}
        n_{\rm e} = n_{\rm H} + n_{\rm He} = 1.1 n_{\rm H},\quad n_{\rm ion} = n_{\rm He} = 0.1 n_{\rm H}
    \end{equation}
    We can plug these numbers back into the equation above to get
    \begin{equation}
        \boxed{ R_{\rm He} = 0.14 \unit{pc} }
    \end{equation}
    And so we find that the helium sphere is smaller than the hydrogen sphere. So we'd expect (based on our very approximate assumptions!) to find an inner region where H is fully ionise and He is singly ionised and a smaller outer region where only H is ionised.
}

\question{1d. Timescales}
\answer{
    So how long does this all take? We know that the recombination timescale is much longer than the photoionisation timescale and so this is what sets the steady-state timescale. The recombination timescale can be written as
    \begin{equation}
        \tau_{\rm recomb} = \frac{1}{\alpha n_{\rm ion}}
    \end{equation}
    Since $\alpha$ is the recombination rate for a given volume and $n_{\rm ion}$ is the number of ions in a given volume just waiting for a chance to recombine. If we use the values of $\alpha$ and $n_{\rm ion}$ from parts b and c to get the timescales for H and He we find that
    \begin{equation}
        \tau_{\rm recomb, H} = 15.8 \unit{yr},\quad \tau_{\rm recomb, He} = 79.2 \unit{yr}
    \end{equation}
    Given these values, the overall time it takes for the cloud to reach a steady-state is simply the large of the two and therefore
    \begin{equation}
        \boxed{ \tau_{\rm steady} = 79.2 \unit{yr} }
    \end{equation}
}

\question{1e. Formation}
\answer{
    Given our values in the previous question, it is clear that the recombination timescale is shorter for the hydrogen sphere and so \textbf{the H sphere forms first}. This is because the number density of H is 10 times higher (whilst the $\alpha$ is only two times smaller) and therefore the recombination happens at a faster rate.
}

\end{document}

### 2022-10-17 541_ism/week3/problem_1.py

@@ -1,0 +1,31 @@
+import astropy.units as u
+import numpy as np
+
+
+def stromgren_radius(Q, n_e, n_ion, alpha):
+    return (3 / (4 * np.pi) * Q / (n_e * n_ion * alpha))**(1/3)
+
+
+# Q_H and ratio from Draine Table 15.1
+Q_H = 10**(48.75) * u.s**(-1)
+Q_He = 0.135 * Q_H
+
+# test for an O3V star
+# Q_H = 10**(49.64) * u.s**(-1)
+# Q_He = 0.251 * Q_H
+
+# values given in the question
+n_H = 5000 * u.cm**(-3)
+alpha_H = 4e-13 * u.cm**3 * u.s**(-1)
+n_He = 0.1 * n_H
+alpha_He = 2 * alpha_H
+
+# calculate radii
+r_H = stromgren_radius(Q_H, n_H, n_H, alpha_H).to(u.pc)
+r_He = stromgren_radius(Q_He, n_H + n_He, n_He, alpha_He).to(u.pc)
+
+print(f"Radius for H is {r_H:1.2f}")
+print(f"Radius for He is {r_He:1.2f}")
+
+print(f"Timescale for H {(1 / (alpha_H * n_H)).to(u.yr):1.1f}")
+print(f"Timescale for He {(1 / (alpha_He * n_He)).to(u.yr):1.1f}")

### 2022-10-23 541_ism/week3/solutions.tex
add write up for w3p2 541

@@
 
+\clearpage
+
+\question{\textbf{2. Spectroscopic Notation}}
+
+\answer{
+    I'll first describe the algorithm for how one can determine the terms and then just show the answer that my python program gives :D\\
+
+    \noindent The inputs for this algorithm are: the quantum numbers of the subshell being filled $n$, $l$ and the total number of electrons to add $N_e$. The format of each term will be
+    \begin{equation}
+        \mathrm{term} = \,^{2 S + 1}\mathcal{L}_{J}^p
+    \end{equation}
+    \begin{enumerate}
+        \item Work the possible states that an electron can take (all pairings of $m_l$ and $m_s$ where $m_l \in [-l, l]$, $m_s \in [-\frac{1}{2}, \frac{1}{2}]$)
+        \item List \textit{every} possible combination of size $N_e$ of these states that satisfies the Pauli exclusion principle (each electron has a unique pairing of $m_l$ and $m_s$).
+This is rather a hassle by hand, but is trivial in Python!
+        \item Given a combination $\{(m_l, m_s)_i \ \forall i \in [1, N_e]\}$, let $M_l = \sum_{i = 1}^{N_e} m_{l, i}$ and $M_s = \sum_{i = 1}^{N_e} m_{l, i}$.
+Compute $M_l$ and $M_s$ for each combination and populate a matrix indexed on $M_l$ (rows) and $M_s$ (columns), where each entry is the total number of combinations resulting in that $(M_l, M_s)$ pairing.
+        \item While the matrix still has a nonzero entry
+        \begin{enumerate}
+            \item Find the first (in terms of decreasing $M_l$) row that has a nonzero entry - let the index of this row be $L$
+            \item Find the last column (in terms of increasing $M_s$) with a nonzero entry - let the index of this column be $S$
+            \item Store an entry of $(L, S)$ for this element/ion
+            \item Subtract a unitary matrix that ranges from the $(-L, L)$ rows and from the $(-S, S)$ columns from the overall matrix
+        \end{enumerate}
+        \item For each ($L, S$) entry, output a term for each $J$ in the range $[|L - S|, |L + S|]$
+        \item Go get yourself a treat of some sort, you deserve it after all of that!
+    \end{enumerate}
+    Additionally, if we want to sort these in terms of energy then we can apply Hund's rules:
+    \begin{enumerate}
+        \item First sort descending by multiplicity, $(2 S + 1)$
+        \item Next sort descending by angular momentum, $L$
+        \item Next sort by angular momentum, $J$. If the shell is half-filled then sort ascending, otherwise descending
+    \end{enumerate}
+    Now we want to get the terms for O\,II.
+In this case, the electronic configuration is $1 s^2, 2s^2, 2p^3$ and so the non-filled level is $2p$ and it has $3$ electrons.
+If we input $n = 2$, $l = p = 1$, $N_e = 3$ into my program\footnote{\url{https://www.github.com/TomWagg/uw-grad-classes/541_ism/week3/problem2.py}}, then we find that the terms are (in order of increasing energy)
+    \begin{equation}
+        \boxed{ {\rm O II} = {}^{4} S_{\frac{3}{2}}, {}^{2}D_{\frac{3}{2}}, {}^{2} D_{\frac{5}{2}}, {}^{2} P_{\frac{1}{2}}, {}^{2} P_{\frac{3}{2}} }
+    \end{equation} 
+}
+
 \end{document}

### 2022-10-23 541_ism/week3/problem_2.py

@@ -1,0 +1,88 @@
+import numpy as np
+import itertools
+import collections
+
+L_lookup = ["S", "P", "D", "F", "G", "H", "I", "J", "K"]
+
+
+def format_terms(*terms):
+    strings = []
+    for S, L, J in terms:
+        term_string = None
+        if not J.is_integer():
+            term_string = f"{S}{L}({int(J * 2)}/2)"
+        else:
+            term_string = f"{S}{L}{int(J)}"
+        strings.append(term_string)
+    return strings
+
+
+def get_spectroscopic_terms(n, l, n_electron):
+    assert l < n
+    m_l_range = list(range(-l, l + 1))
+    m_s_range = [-1/2, 1/2]
+
+    # these states are available to the electrons in this shell
+    electron_states = [(m_l, m_s) for m_l in m_l_range for m_s in m_s_range]
+
+    # find all combination of these states
+    combs = itertools.combinations(electron_states, n_electron)
+
+    # create a dict to keep track of the L, S combos (for the matrix)
+    LS_combos = collections.defaultdict(int)
+
+    # two sets for tracking the total L and S values in each combo
+    M_l_vals = set()
+    M_s_vals = set()
+
+    # get the information about each possible combination
+    for comb in combs:
+        M_l, M_s = np.asarray(comb).sum(axis=0)
+        LS_combos[(M_l, M_s)] += 1
+        M_l_vals.add(M_l)
+        M_s_vals.add(M_s)
+
+    # turn the sets of values into ranges and start an empty matrix
+    M_s_range = np.arange(min(M_s_vals), max(M_s_vals) + 1)
+    M_l_range = np.arange(min(M_l_vals), max(M_l_vals) + 1)
+    matrix = np.zeros((len(M_l_range), len(M_s_range))).astype(int)
+
+    # populate the matrix
+    for i, M_l in enumerate(M_l_range):
+        for j, M_s in enumerate(M_s_range):
+            matrix[i, j] = LS_combos[(M_l, M_s)]
+
+    # keep track of the terms
+    terms = []
+
+    # while there is still a combination remaining in the matrix
+    while matrix.sum() > 0:
+        # go through each row
+        for i in range(len(matrix)):
+            # check if any of the columns are nonzero
+            columns = matrix[i] > 0
+            if columns.any():
+                # if yes, then build a unitary matrix with the same width and reflected height
+                unitary = np.array([np.repeat(False, len(columns)) if abs(j) > abs(M_l_range[i])
[…]

### 2022-10-23 541_ism/week3/solutions.tex
finish week 3 problems - 541

--- 2022-10-17 541_ism/week3/solutions.tex
@@
 
+\clearpage
+
+\question{\textbf{2. Spectroscopic Notation}}
+
+\answer{
+    I'll first describe the algorithm for how one can determine the terms and then, given that I'm lazy and I did not do it by hand, I'll just show the answer that my python program gives :D\\
+
+    \noindent The inputs for this algorithm are: the quantum numbers of the subshell being filled, $n$ and $l$, and the total number of electrons in the partially filled shell, $N_e$.
+The format of each term will be
+    \begin{equation}
+        \mathrm{term} = \,^{2 S + 1}\mathcal{L}_{J}^p
+    \end{equation}
+    \begin{enumerate}
+        \item \textbf{States:} Work the possible states that an electron can take (all pairings of $m_l$ and $m_s$ where $m_l \in [-l, l]$, $m_s \in [-\frac{1}{2}, \frac{1}{2}]$)
+        \item \textbf{Combinations:} List \textit{every} possible combination of size $N_e$ of these states that satisfies the Pauli exclusion principle (each electron has a unique pairing of $m_l$ and $m_s$).
+Yes, this is indeed rather a hassle by hand!
+BUT is trivial in Python!
+:D
+        \item \textbf{Build matrix:} Given a combination $\{(m_l, m_s)_i \ \forall i \in [1, N_e]\}$, let $M_l = \sum_{i = 1}^{N_e} m_{l, i}$ and $M_s = \sum_{i = 1}^{N_e} m_{l, i}$.
+Compute $M_l$ and $M_s$ for each combination and populate a matrix indexed on $M_l$ (rows) and $M_s$ (columns), where each entry is the total number of combinations resulting in that $(M_l, M_s)$ pairing.
+        \item \textbf{Reduce matrix:} While the matrix still has a nonzero entry
+        \begin{enumerate}
+            \item Find the row with the largest $M_l$ value with a nonzero entry - let the index of this row be $L$
+            \item Find the column \textit{within this row} with the largest $M_s$ value with a nonzero entry - let the index of this column be $S$
+            \item Store the pair $(L, S)$ as a possible term
+            \item Subtract a unitary matrix - ranging from the $(-L, L)$ rows and from the $(-S, S)$ columns - from the overall matrix
+        \end{enumerate}
+        \item \textbf{Expand terms:} For each ($L, S$) pair, output a term for each $J$ in the range $[|L - S|, |L + S|]$
+        \item \textbf{Prevent mental breakdown:} Go get yourself a cookie, you deserve it after all of that!
+    \end{enumerate}
+    Additionally, if we want to sort these in terms of energy then we can apply Hund's rules:
+    \begin{enumerate}
+        \item First sort descending by multiplicity, $(2 S + 1)$
+        \item Next sort descending by angular momentum, $L$
+        \item Next sort by angular momentum, $J$. If the shell is half-filled then sort ascending, otherwise descending
+    \end{enumerate}
+    Now we want to get the terms for O\,II.
+In this case, the electronic configuration is $1 s^2, 2s^2, 2p^3$ and so the non-filled level is $2p$ and it has $3$ electrons.
+If we input $n = 2$, $l = p = 1$, $N_e = 3$ into my program\footnote{\url{https://www.github.com/TomWagg/uw-grad-classes/541_ism/week3/problem2.py}}, then we find that the terms are (in order of \emph{increasing} energy)
+    \begin{equation}
+        \boxed{ {\rm O II} = {}^{4} S_{\frac{3}{2}}, {}^{2}D_{\frac{3}{2}}, {}^{2} D_{\frac{5}{2}}, {}^{2} P_{\frac{1}{2}}, {}^{2} P_{\frac{3}{2}} }
+    \end{equation} 
+}
+
 \end{document}

### 2022-10-23 541_ism/week3/problem_2.py

@@ -3,6 +3,7 @@ import itertools
 import collections
 
 L_lookup = ["S", "P", "D", "F", "G", "H", "I", "J", "K"]
+level_sizes = [2, 6, 10, 14]
 
 
 def format_terms(*terms):
@@ -10,9 +11,9 @@ def format_terms(*terms):
     for S, L, J in terms:
         term_string = None
         if not J.is_integer():
-            term_string = f"{S}{L}({int(J * 2)}/2)"
+            term_string = f"{S}{L_lookup[L]}({int(J * 2)}/2)"
         else:
-            term_string = f"{S}{L}{int(J)}"
+            term_string = f"{S}{L_lookup[L]}{int(J)}"
         strings.append(term_string)
     return strings
 
@@ -72,16 +73,21 @@ def get_spectroscopic_terms(n, l, n_electron):
                 # work out the value of L and S for this term(s) based on the matrix
                 L, S = int(abs(M_l_range[i])), M_s_range[columns].max()
 
-                # add a term for each possible J value
-                for J in np.arange(np.abs(L - S), np.abs(L + S) + 1):
-                    terms.append((int(2 * S + 1), L_lookup[L], J))
+                # add a term for each possible J valuev in the correct order based on how filled the level is
+                J_range = np.arange(abs(L - S), abs(L + S) + 1)
+                if n_electron > level_sizes[l] / 2:
+                    J_range = reversed(J_range)
+                for J in J_range:
+                    terms.append((int(2 * S + 1), L, J))
 
                 # go start back at the first row now that the matrix has changed
                 break
 
-    return terms
+    # sort based on Hund's 1st and 2nd rules
+    return sorted(terms, key=lambda x: (x[0], x[1]), reverse=True)
 
 
+# get terms for OII
 terms = get_spectroscopic_terms(2, 1, 3)
 
 # print out the formatted terms

### 2022-10-23 541_ism/final_project/configurations.py
start playing around with project code

@@ -1,0 +1,69 @@
+import numpy as np
+from constants import L_lookup, level_sizes
+
+
+def read_elements():
+    element_list = []
+    with open("elements.csv") as elements:
+        element_list = [line.rstrip("\n").split("\t") for line in elements]
+    return np.transpose(element_list)
+
+
+def electrons_from_element(el_string):
+    n_elec, symbols, names = read_elements()
+    electrons = []
+    if len(el_string) <= 2:
+        electrons = n_elec[np.char.lower(symbols) == el_string.lower()]
+    else:
+        electrons = n_elec[np.char.lower(names) == el_string.lower()]
+
+    if len(electrons) == 0:
+        raise ValueError(f"Cannot recognise this element: `{el_string}`")
+
+    return int(electrons[0])
+
+
+def parse_electrons(source):
+    split_source = source.split(" ")
+    if len(split_source) > 1:
+        element, ionisation = split_source
+        n_I = sum([int(ionisation[i] == "I") for i in range(len(ionisation))])
+    else:
+        element = split_source[0]
+        n_I = 1
+    return electrons_from_element(element) - n_I + 1
+
+
+def electronic_configuration(n_electron):
+    order = [(1, 0), (2, 0), (2, 1), (3, 0), (3, 1),
+             (4, 0), (3, 2), (4, 1), (4, 2), (5, 0), (5, 1)]
+
+    exceptions = {
+        24: [(1, 0, 2), (2, 0, 2), (2, 1, 6), (3, 0, 2), (3, 1, 6), (4, 0, 1), (3, 2, 5)],
+        29: [(1, 0, 2), (2, 0, 2), (2, 1, 6), (3, 0, 2), (3, 1, 6), (4, 0, 1), (3, 2, 10)]
+    }
+
+    if n_electron in exceptions:
+        return exceptions[n_electron]
+
+    configuration = []
+    for n, l in order:
+        n_filled = level_sizes[l] if level_sizes[l] <= n_electron else n_electron
+        n_electron -= n_filled
+        configuration.append((n, l, n_filled))
+
+        if n_electron <= 0:
+            break
+
+    return configuration
+
+
+def format_configuration(configuration):
+    config_string = ""
+    for n, l, n_e in configuration:
+        config_string += f"{n}{L_lookup[l].lower()}{n_e} "
+    return config_string.rstrip()
+
+
+for i in range(1, 37):
+    print(i, ":", format_configuration(electronic_configuration(i)))

### 2022-10-23 541_ism/final_project/constants.py

@@ -1,0 +1,5 @@
+# convert a value of L to the letter representation
+L_lookup = ["S", "P", "D", "F", "G", "H", "I", "J", "K"]
+
+# get the size of a level for each value of l
+level_sizes = [2, 6, 10, 14]

### 2022-10-23 541_ism/final_project/terms.py

@@ -1,0 +1,109 @@
+import numpy as np
+import itertools
+import collections
+from constants import L_lookup, level_sizes
+
+
+def format_terms(*terms, use_latex=False):
+    strings = []
+    for S, L, J in terms:
+        term_string = None
+        if not J.is_integer():
+            if use_latex:
+                term_string = rf"$^{{{S}}} {L_lookup[L]}_{{\frac{{{int(J * 2)}}}{{2}}}}$"
+            else:
+                term_string = f"{S}{L_lookup[L]}({int(J * 2)}/2)"
+        else:
+            if use_latex:
+                term_string = rf"$^{{{S}}} {L_lookup[L]}_{{{int(J)}}}$"
+            else:
+                term_string = f"{S}{L_lookup[L]}{int(J)}"
+        strings.append(term_string)
+    return strings
+
+
+def get_spectroscopic_terms(n, l, n_electron):
+    assert l < n
+    assert n_electron <= level_sizes[l]
+    m_l_range = list(range(-l, l + 1))
+    m_s_range = [-1/2, 1/2]
+
+    # these states are available to the electrons in this shell
+    electron_states = [(m_l, m_s) for m_l in m_l_range for m_s in m_s_range]
+
+    # find all combination of these states
+    combs = itertools.combinations(electron_states, n_electron)
+
+    # create a dict to keep track of the L, S combos (for the matrix)
+    LS_combos = collections.defaultdict(int)
+
+    # two sets for tracking the total L and S values in each combo
+    M_l_vals = set()
+    M_s_vals = set()
+
+    # get the information about each possible combination
+    for comb in combs:
+        M_l, M_s = np.asarray(comb).sum(axis=0)
+        LS_combos[(M_l, M_s)] += 1
+        M_l_vals.add(M_l)
+        M_s_vals.add(M_s)
+
+    # turn the sets of values into ranges and start an empty matrix
+    M_s_range = np.arange(min(M_s_vals), max(M_s_vals) + 1)
+    M_l_range = np.arange(min(M_l_vals), max(M_l_vals) + 1)
+    matrix = np.zeros((len(M_l_range), len(M_s_range))).astype(int)
+
+    # populate the matrix
+    for i, M_l in enumerate(M_l_range):
+        for j, M_s in enumerate(M_s_range):
+            matrix[i, j] = LS_combos[(M_l, M_s)]
+
+    # keep track of the terms
+    terms = []
+
+    # while there is still a combination remaining in the matrix
+    while matrix.sum() > 0:
+        # go through each row
[…]

### 2022-10-26 541_ism/week4/solutions.tex
ISM week 4 problems

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{mhchem}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\psetnum}{3}
\newcommand{\class}{ASTR 541 - Interstellar Medium}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Week \psetnum}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\tomtitle{}

\noindent For reference, if you'd ever like to see the code that I've used to get my answers to these, \href{https://github.com/TomWagg/uw-grad-classes/tree/main/541_ism}{here's a link to my GitHub repo}! (\#astropy.units for life)\\

\question{\textbf{1. Collisions}}

\question{Part I: Diagram}

\answer{
    Here's a diagram of a collisional excitation (complete with a Jess level pun)
    \begin{center}
        \includegraphics[width=\textwidth]{~/Downloads/diagram for ISM(1).png}
    \end{center}
}

\question{Part II: Energy Conservation}
\answer{
    We can write down the energy conservation by considering the kinetic energy of the system and the internal energy of species A.
    \begin{equation}
        \boxed{ \frac{1}{2} \mu v_1^2 = h \nu_{21} + \frac{1}{2} \mu v_2^2 }
    \end{equation}
    where $\mu$ is the reduced mass of the two species.
}

\question{Part III: Steady-State Equation}
\answer{
    Assuming a steady-state means that the rate of excitations must equal the rate of \textit{de}-excitations. We can write this in terms of number densities and collisional rate coefficients as
    \begin{equation}
        \boxed{ n_A n_B \gamma_{12} = n_{A*} n_B \gamma_{21} }
    \end{equation}
    where $n_B$ is the number density of species B and $n_A$, $n_{A*}$ are the number densities of species A in the ground and excited state respectively. Each $\gamma$ is the collisional rate coefficient for an excitation or de-excitation respectively.
}

\question{Part IV: Cross Sections}
\answer{
    Now's it time to expand things out a little. We know from lectures that the collisional rate coefficient can be written as
    \begin{equation}
        \gamma_{xy} = \int f(v_x) \sigma_{xy}(v_x) v_x \dd{v_x}
    \end{equation}
    Using this definition, we can transform our steady-state equation into the following
    \begin{equation}
        \boxed{ n_A n_B \int f(v_1) \sigma_{12}(v_1) v_1 \dd{v_1} = n_{A*} n_B \int f(v_2) \sigma_{21}(v_2) v_2 \dd{v_2} }
    \end{equation}
}

\question{Part V: Detailed balance (this is where the magic happens)}
\answer{
    Due to the elegance of detailed balance we can know make all of our lives rather simpler. Detailed balance tells us that we don't need to consider \emph{every} velocity, we can just consider only a single velocity that would excite species A, $v_1$ and resulting velocity from that collision, $v_2$ (and the rest will all cancel out). This means that we can then remove the integral and be left with a rather more pleasant relation.
    \begin{equation}
        \boxed{ n_A n_B f(v_1) \sigma_{12}(v_1) v_1 \dd{v_1} = n_{A*} n_B f(v_2) \sigma_{21}(v_2) v_2 \dd{v_2} }
    \end{equation}
}

\question{Part VI: The Algebra Bit$^{\rm TM}$}
\answer{
    Okie dokie folks, strap in, here we go! First, let's cancel out the $n_B$ (note we could have done this earlier, I didn't do that until now because I wanted to, totally not because I didn't notice).
    \begin{equation}
        n_A f(v_1) \sigma_{12}(v_1) v_1 \dd{v_1} = n_{A*} f(v_2) \sigma_{21}(v_2) v_2 \dd{v_2}
    \end{equation}
    Next it would be good to have a relation between $\dd{v_1}$ and $\dd{v_2}$. Let's go back to our conservation of energy equation (our old friend from part II). Taking the total derivative of each side and setting them equal (since we are in a steady-state), we find that
    \begin{equation}
        v_1 \dd{v_1} = v_2 \dd{v_2}
    \end{equation}
    where the $\mu$ has cancelled. Well that's convenient, now we can just cross out those terms on each side of our equation to get
    \begin{equation}
        f(v_1) \sigma_{12}(v_1) = \frac{n_{A*}}{n_A} f(v_2) \sigma_{21}(v_2)
    \end{equation}
    where I've moved the $n_A$ over to the other side for reasons that will soon become clear (or perhaps just to give myself an air of mystery, take your pick).

    \noindent Next let's go through the rather arduous task of expanding the maxwellian terms. Note here that since $v_1$ and $v_2$ are \emph{relative} velocities, so the mass terms are the reduced mass $\mu$ in both cases. This means that the form of the maxwellian we shall looks like this
    \begin{equation}
        f(v) = 4 \pi \qty(\frac{\mu}{2 \pi k_B T})^{\frac{3}{2}} \exp\qty(-\frac{\mu v^2}{2 k_B T}) v^2
    \end{equation}
    Let's plug that into both sides and cancel down
    \begin{align}
        \sigma_{12}(v_1) \qty[4 \pi \qty(\frac{\mu}{2 \pi k_B T})^{\frac{3}{2}} \exp\qty(-\frac{\mu v^2}{2 k_B T}) v^2] &= \frac{n_{A*}}{n_A} \sigma_{21}(v_2) \qty[4 \pi \qty(\frac{\mu}{2 \pi k_B T})^{\frac{3}{2}} \exp\qty(-\frac{\mu v^2}{2 k_B T}) v^2] \\
        \sigma_{12}(v_1) \qty[\exp\qty(-\frac{\mu v_1^2}{2 k_B T}) v_1^2] &= \frac{n_{A*}}{n_A} \sigma_{21}(v_2) \qty[\exp\qty(-\frac{\mu v_2^2}{2 k_B T}) v_2^2] \\
        \sigma_{12}(v_1) v_1^2 &= \frac{n_{A*}}{n_A} \sigma_{21}(v_2) \qty[\exp\qty(-\frac{\mu (v_2^2 - v_1^2)}{2 k_B T}) v_2^2]
    \end{align}
    Jolly good! Now for my big reveal of why I moved the $n_A$ over. Well, since collisions are occurring and changing the excitation levels of atoms in a steady-state we must have an equilibrium between kinetic temperature and excitation temperature. Stretching our memories back to the LTE lecture we can remember that \textbf{this means the Boltzmann equation holds}, which gives
    \begin{equation}
        \frac{n_{A*}}{n_A} = \frac{g_2}{g_1} \exp \qty(- \frac{h\nu}{k_B T})
    \end{equation}
    Splendid! Now let's plug \emph{that} into our equation from before to find
    \begin{align}
        \sigma_{12}(v_1) v_1^2 &= \sigma_{21}(v_2) \frac{g_2}{g_1} \exp\qty(-\frac{h \nu}{k_B T}) \qty[\exp\qty(-\frac{\mu (v_2^2 - v_1^2)}{2 k_B T}) v_2^2] \\
        g_1 \sigma_{12}(v_1) v_1^2 &= \sigma_{21}(v_2) g_2 \qty[\exp\qty(-\frac{\mu (v_2^2 - v_1^2 + \frac{2}{\mu} h \nu)}{2 k_B T}) v_2^2]
    \end{align}
    Now get this, that term in the exponential $(v_2^2 - v_1^2 + \frac{2}{\mu} h \nu)$ is just 0 (from our conservation of energy equation from earlier!). That means we can scrap the exponential and leave ourselves with a tidy little relation between the degeneracies, cross sections and velocities.
    \begin{equation}
        \boxed{ g_1 \sigma_{12}(v_1) v_1^2 = \sigma_{21}(v_2) g_2 v_2^2 }
    \end{equation}
    We made it! Now don't forget to submit your offerings to Detailed Balance to ensure future expressions can be simplified in the same way.
}

\end{document}

### 2022-10-31 541_ism/week0/solutions.tex
forgot about week0 problem for ISM

--- 2022-10-17 541_ism/week3/solutions.tex
@@
 
-\newcommand{\psetnum}{3}
+\newcommand{\psetnum}{0}
 \newcommand{\class}{ASTR 541 - Interstellar Medium}
@@
 
-\question{\textbf{1. Stromgren Spheres}}
+\question{\textbf{1. Introduction to (Baryonic) Accounting}}
 
-\question{1a. What matters about the star?}
+\noindent\emph{The general aim of this problem is to prove that the majority of baryonic matter does not lie in stars in galaxy or galaxies in clusters in a back-of-the-envelope calculation}
+\answer{
+    This answer heavily follows the solution laid out by \href{https://arxiv.org/pdf/astro-ph/9712020.pdf}{the paper that is the linked} in the question.
+To begin, we are given that $\Omega_B = 0.0455$ at $z = 0$.
+We can multiply this by the critical density,
+    \begin{equation}
+        \rho_c = \frac{3 H_0^2}{8 \pi G} \approx 9 \times 10^{-27} \unit{kg}{m^{-3}},
+    \end{equation}
+    to find that the mass density of baryonic matter is
+    \begin{equation}
+        \rho_B \approx 4 \times 10^{-28} \unit{kg}{m^{-3}}.
+    \end{equation}
 
-\answer{
-    When finding the Stromgren radius for a region we need to know the ionisation rate. We derived in class that this radius is given as
+    \noindent Next we can calculate the mass density of stars and galaxies and compare it to this value.
+We can consider that most luminosity is coming from either a spheroid or a disk component and write that the density for each is
     \begin{equation}
-        Q_{\rm ionise} = \int_{\nu_{\rm ionise}}^{\infty} \frac{L_\nu}{h \nu} \dd{\nu}
+        \rho = \mathcal{L}_B f_B \avg{M / L_B},
     \end{equation}
-    Therefore, we need to know the \emph{luminosity} of an O7V star and then we can integrate that to get the ionisation coefficient.
-But goodness me that would be far too \textit{Draine}ing so we can look it up in the textbook instead!
-From Table 15.1 we find that the rate is
+    where $\mathcal{L}_B$ is the mean luminosity density, $f_B$ is the fraction of light produced by this component and $\avg{M / L_B}$ is the mean mass-to-light ratio for this component.
+
+    Let's assume that the only two components are spheroids and discs.
+The paper gives us that the fractions and mass-to-light ratios (in $\unit{M_{\odot}} / \unit{L_\odot}$) for each of these are $(0.385, 6.5)$ and $(0.582, 1.5)$ respectively.
+Summing these gives that the total mass density is
     \begin{equation}
-        \boxed{ Q_{\rm ionise, H} = 10^{48.75} \unit{s^{-1}} }
+        \rho = \mathcal{L}_B (0.385 \cdot 6.5 + 0.582 \cdot 1.5) \unit{M_{\odot}} / \unit{L_\odot} \approx 3.4 \mathcal{L}_B \unit{M_{\odot}} / \unit{L_\odot},
     \end{equation}
-}
+    From the paper we have that the mean luminosity density is approximately (assuming $h \approx 1$)
+    \begin{equation}
+        \mathcal{L}_B \approx 2 \times 10^{8} \unit{L_\odot}{Mpc^{-3}}
+    \end{equation}
+    Plugging this back in we get that the mass density of stars is
+    \begin{equation}
+        \rho_{\rm stars} \approx 6.8 \times 10^8 \unit{M_\odot}{Mpc^{-3}}
+    \end{equation}\\
 
-\question{1b. A hydrogen sphere}
-\answer{
-    We derived in class that the radius of a Stromgren sphere is given by the following equation
+    \noindent We can now compare $\rho_B$ and $\rho_{\rm stars}$ and note that
     \begin{equation}
-        R_{\rm S} = \qty[ \frac{3}{4 \pi} \frac{Q_{\rm ionise}}{n_e n_{\rm ion} \alpha(T)} ]^{1/3}
+        \boxed{ \frac{\rho_{\rm stars}}{\rho_B} \approx 11\% }
     \end{equation}
-    So to find the radius all we need to do is get values for each of the quantities above.
-For this part of the question we're just focussing on the hydrogen sphere and so the values are fairly simple
-    \begin{equation}
-        n_e = n_{\rm H},\quad n_{\rm ion} = n_{\rm H},\quad \alpha(T) = 4 \times 10^{-13} \unit{cm^{3}}{s^{-1}},\quad Q_{\rm ionise} = Q_{\rm ionise, H}
-    \end{equation}
-    where $n_{\rm H} = 5000 \unit{cm^{-3}}$ was given in the problem. Plugging these numbers in gives the radius as
-    \begin{equation}
-        \boxed{ R_{\rm H} = 0.17 \unit{pc} }
-    \end{equation}
-    This may at first glance cause you undue stress - ``this violates our understanding that HII regions are (several) tens of pc in size, I just did a quiz on this'' you say!
-Worry not.
-You can note that the density used in this question is \emph{extremely} high and so we are dealing with an ``ultra-compact'' HII region in this case and hence the radius is smaller than you may predict.
-}
-
-\clearpage
-
-\question{1c. A helium sphere}
-\answer{
-    For the helium sphere it is a little different since we need to account for the hydrogen as well.
-As a simple first step, we are given the values of $\alpha$ and $Q$ (relative to hydrogen) in the question
-    \begin{equation}
-        \alpha = 8 \times 10^{-13} \unit{cm^3}{s^{-1}},\quad Q_{\rm ionise} = 0.135 \cdot Q_{\rm ionise, H}
-    \end{equation}
-    The difference lies in the number of electrons per ion.
-In this case we need to include the electrons from both the hydrogen and the helium.
-Each supplies only one electron since the star is not hot enough to doubly ionise helium consistently.
-    \begin{equation}
-        n_{\rm e} = n_{\rm H} + n_{\rm He} = 1.1 n_{\rm H},\quad n_{\rm ion} = n_{\rm He} = 0.1 n_{\rm H}
-    \end{equation}
-    We can plug these numbers back into the equation above to get
-    \begin{equation}
-        \boxed{ R_{\rm He} = 0.14 \unit{pc} }
-    \end{equation}
-    And so we find that the helium sphere is smaller than the hydrogen sphere.
-So we'd expect (based on our very approximate assumptions!) to find an inner region where H is fully ionise and He is singly ionised and a smaller outer region where only H is ionised.
-}
-
-\question{1d. Timescales}
-\answer{
-    So how long does this all take?
-We know that the recombination timescale is much longer than the photoionisation timescale and so this is what sets the steady-state timescale.
-The recombination timescale can be written as
-    \begin{equation}
-        \tau_{\rm recomb} = \frac{1}{\alpha n_{\rm ion}}
-    \end{equation}
-    Since $\alpha$ is the recombination rate for a given volume and $n_{\rm ion}$ is the number of ions in a given volume just waiting for a chance to recombine.
-If we use the values of $\alpha$ and $n_{\rm ion}$ from parts b and c to get the timescales for H and He we find that
-    \begin{equation}
-        \tau_{\rm recomb, H} = 15.8 \unit{yr},\quad \tau_{\rm recomb, He} = 79.2 \unit{yr}
-    \end{equation}
-    Given these values, the overall time it takes for the cloud to reach a steady-state is simply the large of the two and therefore
-    \begin{equation}
-        \boxed{ \tau_{\rm steady} = 79.2 \unit{yr} }
-    \end{equation}
-}
-
-\question{1e. Formation}
-\answer{
-    Given our values in the previous question, it is clear that the recombination timescale is shorter for the hydrogen sphere and so \textbf{the H sphere forms first}.
-This is because the number density of H is 10 times higher (whilst the $\alpha$ is only two times smaller) and therefore the recombination happens at a faster rate.
+    So from our back-of-the-envelope estimate we find that stars represent a small fraction of the baryonic matter within the Universe!!
 }

### 2022-10-31 541_ism/week0/problem_1.py

@@ -1,0 +1,14 @@
+import astropy.units as u
+import astropy.constants as const
+import numpy as np
+
+
+omega_B = 0.0455
+rho_crit = 9e-27 * u.kg * u.m**(-3)
+rho_B = omega_B * rho_crit
+
+print(f"{rho_B.to(u.Msun * u.Mpc**(-3)):1.1e}")
+
+mean_lum = 2e8 * u.Lsun * u.Mpc**(-3) * 3.4 * (u.Msun / u.Lsun)
+
+print((mean_lum / rho_B).decompose() * 100)

### 2022-10-31 541_ism/final_project/configurations.py
add `formatted` options to funcs

@@ -34,7 +34,7 @@ def parse_electrons(source):
     return electrons_from_element(element) - n_I + 1
 
 
-def electronic_configuration(n_electron):
+def electronic_configuration(n_electron, formatted=False):
     order = [(1, 0), (2, 0), (2, 1), (3, 0), (3, 1),
              (4, 0), (3, 2), (4, 1), (4, 2), (5, 0), (5, 1)]
 
@@ -55,7 +55,10 @@ def electronic_configuration(n_electron):
         if n_electron <= 0:
             break
 
-    return configuration
+    if formatted:
+        return format_configuration(configuration)
+    else:
+        return configuration
 
 
 def format_configuration(configuration):
@@ -65,5 +68,5 @@ def format_configuration(configuration):
     return config_string.rstrip()
 
 
-for i in range(1, 37):
-    print(i, ":", format_configuration(electronic_configuration(i)))
+# for i in range(1, 37):
+#     print(i, ":", format_configuration(electronic_configuration(i)))

### 2022-10-31 541_ism/final_project/run.py

@@ -1,0 +1,15 @@
+from linetools.lists.linelist import LineList
+import terms
+import configurations as con
+
+ism = LineList("ISM")
+
+sulfur_II = ism._data[(ism._data["ion"] == 6) & (ism._data["Z"] == 8)]# & (ism._data["A"] > 0.0)]
+
+for line in sulfur_II:
+    print(line["name"], line["wrest"], line["Ek"])
+    # print(line["nj"], line["Jj"], line["gj"], "-", line["nk"], line["Jk"], line["gk"])
+
+print(con.format_configuration(con.electronic_configuration(8)))
+print(con.format_configuration(con.electronic_configuration(3)))
+print(terms.format_terms(*terms.get_spectroscopic_terms(2, 0, 1)))

### 2022-10-31 541_ism/final_project/terms.py

@@ -22,7 +22,7 @@ def format_terms(*terms, use_latex=False):
     return strings
 
 
-def get_spectroscopic_terms(n, l, n_electron):
+def get_spectroscopic_terms(n, l, n_electron, formatted=False):
     assert l < n
     assert n_electron <= level_sizes[l]
     m_l_range = list(range(-l, l + 1))
@@ -89,21 +89,16 @@ def get_spectroscopic_terms(n, l, n_electron):
                 break
 
     # sort based on Hund's 1st and 2nd rules
-    return sorted(terms, key=lambda x: (x[0], x[1]), reverse=True)
+    terms = sorted(terms, key=lambda x: (x[0], x[1]), reverse=True)
+
+    if formatted:
+        return format_terms(*terms)
+    else:
+        return terms
 
 
 # DRAINE Table 4.1
-for ne in [1, 2]:
-    print(*format_terms(*get_spectroscopic_terms(1, 0, ne)))
-for ne in range(1, 7):
-    print(*format_terms(*get_spectroscopic_terms(2, 1, ne)))
-
-
-# get_spectroscopic_terms(3, 0, 1)
-# get_spectroscopic_terms(3, 0, 2)
-# get_spectroscopic_terms(3, 1, 1)
-# get_spectroscopic_terms(3, 1, 2)
-# get_spectroscopic_terms(3, 1, 3)
-# get_spectroscopic_terms(3, 1, 4)
-# get_spectroscopic_terms(3, 1, 5)
-# get_spectroscopic_terms(3, 1, 6)
+# for ne in [1, 2]:
+#     print(*format_terms(*get_spectroscopic_terms(1, 0, ne)))
+# for ne in range(1, 7):
+#     print(*format_terms(*get_spectroscopic_terms(2, 1, ne)))

### 2022-10-31 541_ism/week5/solutions.tex
start ISM week 5

--- 2022-10-31 541_ism/week0/solutions.tex
@@
 
-\newcommand{\psetnum}{0}
+\newcommand{\psetnum}{5}
 \newcommand{\class}{ASTR 541 - Interstellar Medium}
@@
 
-\question{\textbf{1. Introduction to (Baryonic) Accounting}}
-
-\noindent\emph{The general aim of this problem is to prove that the majority of baryonic matter does not lie in stars in galaxy or galaxies in clusters in a back-of-the-envelope calculation}
-\answer{
-    This answer heavily follows the solution laid out by \href{https://arxiv.org/pdf/astro-ph/9712020.pdf}{the paper that is the linked} in the question.
-To begin, we are given that $\Omega_B = 0.0455$ at $z = 0$.
-We can multiply this by the critical density,
-    \begin{equation}
-        \rho_c = \frac{3 H_0^2}{8 \pi G} \approx 9 \times 10^{-27} \unit{kg}{m^{-3}},
-    \end{equation}
-    to find that the mass density of baryonic matter is
-    \begin{equation}
-        \rho_B \approx 4 \times 10^{-28} \unit{kg}{m^{-3}}.
-    \end{equation}
-
-    \noindent Next we can calculate the mass density of stars and galaxies and compare it to this value.
-We can consider that most luminosity is coming from either a spheroid or a disk component and write that the density for each is
-    \begin{equation}
-        \rho = \mathcal{L}_B f_B \avg{M / L_B},
-    \end{equation}
-    where $\mathcal{L}_B$ is the mean luminosity density, $f_B$ is the fraction of light produced by this component and $\avg{M / L_B}$ is the mean mass-to-light ratio for this component.
-
-    Let's assume that the only two components are spheroids and discs.
-The paper gives us that the fractions and mass-to-light ratios (in $\unit{M_{\odot}} / \unit{L_\odot}$) for each of these are $(0.385, 6.5)$ and $(0.582, 1.5)$ respectively.
-Summing these gives that the total mass density is
-    \begin{equation}
-        \rho = \mathcal{L}_B (0.385 \cdot 6.5 + 0.582 \cdot 1.5) \unit{M_{\odot}} / \unit{L_\odot} \approx 3.4 \mathcal{L}_B \unit{M_{\odot}} / \unit{L_\odot},
-    \end{equation}
-    From the paper we have that the mean luminosity density is approximately (assuming $h \approx 1$)
-    \begin{equation}
-        \mathcal{L}_B \approx 2 \times 10^{8} \unit{L_\odot}{Mpc^{-3}}
-    \end{equation}
-    Plugging this back in we get that the mass density of stars is
-    \begin{equation}
-        \rho_{\rm stars} \approx 6.8 \times 10^8 \unit{M_\odot}{Mpc^{-3}}
-    \end{equation}\\
-
-    \noindent We can now compare $\rho_B$ and $\rho_{\rm stars}$ and note that
-    \begin{equation}
-        \boxed{ \frac{\rho_{\rm stars}}{\rho_B} \approx 11\% }
-    \end{equation}
-    So from our back-of-the-envelope estimate we find that stars represent a small fraction of the baryonic matter within the Universe!!
-}
+\question{\textbf{1. }}
 

### 2022-10-31 541_ism/week5/problem_1.py

@@ -1,0 +1,8 @@
+import astropy.units as u
+import numpy as np
+
+import sys
+sys.path.append("../final_project")
+
+import terms
+import configurations as con

### 2022-11-01 541_ism/week5/solutions.tex
week5 p1 ISM

@@
 \usepackage{mhchem}
+\usepackage{array}
 \hypersetup{
@@
 
+\newcolumntype{C}{>{$}c<{$}}
+
 \begin{document}
@@
 
-\question{\textbf{1. }}
+\question{\textbf{1. SII Transitions}}
+
+\begin{center}
+    \includegraphics[width=0.5\textwidth]{energy_levels.pdf}
+\end{center}
+
+\question{1a. Electronic Configuration}
+\answer{
+    I used my Python program (and the Aufbau Principle) to find that the electronic configuration of SII is
+    \begin{equation}
+        \boxed{ \mathrm{SII} = 1s^2 2s^2 2p^6 3s^2 \mathbf{3p^3} }
+    \end{equation}
+    where the last subshell is only half-filled and so this is where we get our terms.
+}
+
+\question{1b. Spectroscopic Terms}
+\answer{
+    Again I put my feet up and let Python do the heavy lifting to find (and sort!) the spectroscopic terms as
+    \begin{equation}
+        \boxed{ \mathrm{SII} = {}^4\mathrm{S}_{3/2}, {}^2\mathrm{D}_{3/2,\, 5/2}, {}^2\mathrm{P}_{1/2,\, 3/2}}
+    \end{equation}
+    and additionally plug those into a plotting function to get the diagram above.
+}
+
+\question{1c. Cooling Rate}
+\answer{
+    As noted in the question, the cooling rate per SII ion is calculated as the sum of line intensities. We can therefore write the cooling rate per ion as a product of sums.
+    \begin{equation}
+        I = \sum_{i = 2}^{3} \sum_{j = 1}^{i - 1} \frac{n_i}{n_{\rm SII}} A_{ij} \Delta E_{ij}
+    \end{equation}
+    where $n_i / n_{\rm SII}$ is the fraction of ions that are in state $i$ (table 2 from the question), $A_{ij}$ is the Einstein coefficient for the transition from $i$ to $j$ (table 1 from the question) and $\Delta E_{ij}$ is the difference in energy between states $i$ and $j$ (table 1 from the question).
+Then we plug, we chug, we put it all in a table.
+    \begin{center}
+        \boxed{\begin{tabular}{C|C}
+            n_e [\unit{cm^{-3}}] & I [\unit{erg}{s^{-1}}{ion^{-1}}]\\
+            \hline
+            10   & 5.15 \times 10^{-19} \\
+            10^3 & 4.36 \times 10^{-17} \\
+            10^5 & 3.17 \times 10^{-16}
+        \end{tabular}}
+    \end{center}
+}
+
+\question{1d. Excitation Temperature}
+\answer{
+    The excitation temperature for a transition from $u$ to $l$ can be written as
+    \begin{equation}
+        e^{-E_{ul} / k_B T_{\rm ex}} = \frac{n_u}{n_l} \frac{g_l}{g_u}
+    \end{equation}
+    Or if we rearrange it
+    \begin{equation}
+        T_{\rm ex} = - \frac{E_{ul}}{k_B} \qty(\ln \frac{n_u}{n_l} \frac{g_l}{g_u})^{-1}
+    \end{equation}
+    The transition from ${}^2\mathrm{D}_{3/2} \to {}^4\mathrm{S}_{3/2}$ corresponds to a transition from $n=2$ to $n=1$ in our labelling scheme.
+The degeneracies of these two states are identical so we can cancel the $g$ factors.
+It is then simply a matter of plugging in the $\Delta E$ and $n$ values from the tables in the question to get values for the temperature.
+    \begin{center}
+        \boxed{\begin{tabular}{C|C}
+            n_e [\unit{cm^{-3}}] & T_{\rm ex} [\unit{K}]\\
+            \hline
+            10   & 2.3 \times 10^{3} \\
+            10^3 & 4.6 \times 10^{3} \\
+            10^5 & 9.6 \times 10^{3}
+        \end{tabular}}
+    \end{center}
+}
+
+\question{1e. Trends}
+\answer{
+    We can summarise the trends in a table by looking at the ratio of the cooling rates and temperatures to the initial electron density values.
+We see that the cooling rates and excitation temperatures both increase with increasing electron densities.
+    \begin{center}
+        \begin{tabular}{C|C|C}
+            n_e [\unit{cm^{-3}}] & I / I(n_e = 10 \unit{cm^{-3}}) & T_{\rm ex} / T_{\rm ex}(n_e = 10 \unit{cm^{-3}})\\
+            \hline
+            10   & 1 & 1 \\
+            10^3 & 85 & 2.01 \\
+            10^5 & 620 & 4.22
+        \end{tabular}
+    \end{center}
+    The cooling rate increases since a higher electron density results in more collisions and more ions in excited states (collisions are exciting!) - this results in more transitions and more emissions and thus more cooling.
+The excitation temperature increases since it corresponds to the internal energy of the ions and so more collisions and more excited ions implies a higher excitation temperature.
+
+    If we were to increase the density even higher the excitation temperature would eventually level off at it reaches equilibrium with the kinetic temperature of the gas (${\sim}10000\unit{K}$).
+}
+
+\question{1f. Diagnostics}
+\answer{
+    Given that the change in energy between the $n=2$ and $n=3$ states is small, the emission line intensity ratio would be a useful diagnostic for \textbf{density}.
+}
+
+\question{1g. Low Density Ratios}
+\answer{
+    The ratio of the intensities in a low density regime is given by
+    \begin{equation}
+        \frac{I_{31}}{I_{21}}_{n \ll n_{\rm crit}} = \frac{\Omega_{31}}{\Omega_{21}}
+    \end{equation}
+    I \emph{could} look these values up and plug them in, \emph{or} I could be lazy and just take a look at Figure 18.4 from Draine and read off the ratio.
+I will leave it as an exercise to the reader to guess which method I used.
+    \begin{equation}
+        \boxed{ \frac{I_{31}}{I_{21}}_{n \ll n_{\rm crit}} \approx 1.44 }
+    \end{equation}
+}
+
+\question{1h. High Density Ratios}
+\answer{
+    The ratio of the intensities in a high density regime is given by
+    \begin{equation}
+        \frac{I_{31}}{I_{21}}_{n \gg n_{\rm crit}} = \frac{g_3}{g_2} \frac{A_{31}}{A_{21}}
+    \end{equation}
+    We know all of these values (recalling $g = 2 J + 1$), and so plugging them in we find
+    \begin{equation}
+        \boxed{ \frac{I_{31}}{I_{21}}_{n \gg n_{\rm crit}} \approx 0.44 }
+    \end{equation}
+}
 

### 2022-11-01 541_ism/week5/problem_1.py

@@ -1,4 +1,5 @@
 import astropy.units as u
+import astropy.constants as const
 import numpy as np
 
 import sys
@@ -6,3 +7,40 @@ sys.path.append("../final_project")
 
 import terms
 import configurations as con
+
+print("Part a")
+print(f"Electronic configuration of SII is: {con.electronic_configuration(15, formatted=True)}")
+
+print("\nPart b")
+print("Spectroscopic terms of SII in order are:", *terms.get_spectroscopic_terms(3, 1, 3, formatted=True))
+
+print("\nPart c")
+ns = np.array([[0.999, 0.00008, 0.0004],
+               [0.964, 0.0088, 0.027],
+               [0.787, 0.084, 0.129]])
+A_31 = 2.6e-4 / u.s
+A_21 = 8.8e-4 / u.s
+lambda_31 = 6716.5 * u.Angstrom
+lambda_21 = 6730.8 * u.Angstrom
+base_rate = None
+for i, n_e in enumerate([10, 1e3, 1e5]):
+    n_1, n_2, n_3 = ns[i]
+
+    I_31 = n_3 * A_31 * const.h * const.c / lambda_31
+    I_21 = n_2 * A_21 * const.h * const.c / lambda_21
+
+    I = I_31 + I_21
+    if i == 0:
+        base_rate = I
+    print(f"Cooling rate for log(n_e)={np.log10(n_e)}: {I.to(u.erg / u.s):1.2e}, {(I / base_rate).decompose():1.1f}")
+
+
+print("\nPart d")
+base_T = None
+for i, n_e in enumerate([10, 1e3, 1e5]):
+    n_1, n_2, n_3 = ns[i]
+    factor = (-const.h * const.c / lambda_31) / const.k_B
+    t_ex_21 = factor * np.log(n_2 / n_1)**(-1)
+    if i == 0:
+        base_T = t_ex_21
+    print(f"Excitation temperature: {t_ex_21.to(u.K): 1.1e}, {(t_ex_21 / base_T).decompose():1.2f}")

### 2022-11-01 541_ism/final_project/levels.py
start making energy level diagrams

@@ -1,0 +1,72 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import astropy.units as u
+
+import terms
+
+plt.rc('font', family='serif')
+plt.rcParams['text.usetex'] = False
+fs = 24
+
+# update various fontsizes to match
+params = {'figure.figsize': (12, 8),
+          'legend.fontsize': 0.7 * fs,
+          'axes.labelsize': fs,
+          'xtick.labelsize': 0.7 * fs,
+          'ytick.labelsize': 0.7 * fs,
+          'axes.linewidth': 1.1,
+          'xtick.major.size': 7,
+          'xtick.minor.size': 4,
+          'ytick.major.size': 7,
+          'ytick.minor.size': 4}
+plt.rcParams.update(params)
+
+
+def plot_energy_levels(spec_terms, transitions, title=None):
+    x_vals = np.arange(len(transitions) + 2)
+
+    fig, ax = plt.subplots(figsize=(6, 10))
+
+    previous_L = spec_terms[0][1]
+    colour_ind = 0
+    height = 1
+    heights = [1]
+    for n, term in enumerate(spec_terms):
+        L = term[1]
+        height += 1
+        if L != previous_L:
+            height += 2
+            colour_ind += 1
+        ax.plot(x_vals, np.ones_like(x_vals) * height, color=f"C{colour_ind}", lw=2)
+        ax.annotate(*terms.format_terms(term, use_latex=True), xy=(-0.1, height),
+                    ha="right", va="center", fontsize=fs)
+        ax.annotate(f"n = {n + 1}", xy=(x_vals[-1] + 0.1, height),
+                    ha="left", va="center", fontsize=0.8*fs)
+        previous_L = L
+        heights.append(height)
+
+    for x, transition in enumerate(transitions):
+        upper_height = heights[transition[0]]
+        lower_height = heights[transition[1]]
+        midpoint = lower_height + ((upper_height - lower_height) / 2)
+        ax.annotate("", xytext=(x + 1, upper_height), xy=(x + 1, lower_height),
+                    arrowprops=dict(arrowstyle="-|>", linewidth=2, color="grey"))
+
+        wavelength = transition[2].to(u.Angstrom).value
+        t_string = f"{wavelength:1.1f}" if wavelength < 1e5 else f"{wavelength:1.1e}"
[…]

### 2022-11-01 541_ism/final_project/configurations.py
add latex formatting for configurations

@@ -4,7 +4,7 @@ from constants import L_lookup, level_sizes
 
 def read_elements():
     element_list = []
-    with open("elements.csv") as elements:
+    with open("data/elements.csv") as elements:
         element_list = [line.rstrip("\n").split("\t") for line in elements]
     return np.transpose(element_list)
 
@@ -34,7 +34,7 @@ def parse_electrons(source):
     return electrons_from_element(element) - n_I + 1
 
 
-def electronic_configuration(n_electron, formatted=False):
+def get_configuration(n_electron, formatted=False, use_latex=False):
     order = [(1, 0), (2, 0), (2, 1), (3, 0), (3, 1),
              (4, 0), (3, 2), (4, 1), (4, 2), (5, 0), (5, 1)]
 
@@ -56,15 +56,18 @@ def electronic_configuration(n_electron, formatted=False):
             break
 
     if formatted:
-        return format_configuration(configuration)
+        return format_configuration(configuration, use_latex=use_latex)
     else:
         return configuration
 
 
-def format_configuration(configuration):
+def format_configuration(configuration, use_latex=False):
     config_string = ""
     for n, l, n_e in configuration:
-        config_string += f"{n}{L_lookup[l].lower()}{n_e} "
+        if use_latex:
+            config_string += rf"${{{n}}}{{{L_lookup[l].lower()}}}^{{{n_e}}}$ "
+        else:
+            config_string += f"{n}{L_lookup[l].lower()}{n_e} "
     return config_string.rstrip()
 
 

### 2022-11-01 541_ism/final_project/configurations.py
add __all__'s

@@ -1,6 +1,8 @@
 import numpy as np
 from constants import L_lookup, level_sizes
 
+__all__ = ["parse_electrons", "get_configuration", "format_configuration"]
+
 
 def read_elements():
     element_list = []

### 2022-11-01 541_ism/final_project/constants.py

@@ -1,3 +1,5 @@
+__all__ = ["L_lookup", "level_sizes"]
+
 # convert a value of L to the letter representation
 L_lookup = ["S", "P", "D", "F", "G", "H", "I", "J", "K"]
 

### 2022-11-01 541_ism/final_project/levels.py

@@ -4,6 +4,8 @@ import astropy.units as u
 
 import terms
 
+__all__ = ["plot_energy_levels"]
+
 plt.rc('font', family='serif')
 plt.rcParams['text.usetex'] = False
 fs = 24
@@ -22,16 +24,19 @@ params = {'figure.figsize': (12, 8),
 plt.rcParams.update(params)
 
 
-def plot_energy_levels(spec_terms, transitions, title=None):
+def plot_energy_levels(spec_terms, transitions, title=None, fig=None, ax=None, show=True):
     x_vals = np.arange(len(transitions) + 2)
 
-    fig, ax = plt.subplots(figsize=(6, 10))
+    if fig is None or ax is None:
+        fig, ax = plt.subplots(figsize=(6, 10))
+
+    max_n = max(transitions)[0]
 
     previous_L = spec_terms[0][1]
     colour_ind = 0
     height = 1
     heights = [1]
-    for n, term in enumerate(spec_terms):
+    for n, term in enumerate(spec_terms[:max_n]):
         L = term[1]
         height += 1
         if L != previous_L:
@@ -62,7 +67,9 @@ def plot_energy_levels(spec_terms, transitions, title=None):
         ax.set_title(title, fontsize=fs, y=-0.05)
     ax.axis("off")
 
-    plt.show()
+    if show:
+        plt.show()
+    return fig, ax
 
 
 # plot_energy_levels(terms.get_spectroscopic_terms(3, 1, 3)[:3],

### 2022-11-01 541_ism/final_project/run.py

@@ -1,15 +1,0 @@
-from linetools.lists.linelist import LineList
-import terms
-import configurations as con
-
-ism = LineList("ISM")
-
-sulfur_II = ism._data[(ism._data["ion"] == 6) & (ism._data["Z"] == 8)]# & (ism._data["A"] > 0.0)]
-
-for line in sulfur_II:
-    print(line["name"], line["wrest"], line["Ek"])
-    # print(line["nj"], line["Jj"], line["gj"], "-", line["nk"], line["Jk"], line["gk"])
-
-print(con.format_configuration(con.electronic_configuration(8)))
-print(con.format_configuration(con.electronic_configuration(3)))
-print(terms.format_terms(*terms.get_spectroscopic_terms(2, 0, 1)))

### 2022-11-01 541_ism/final_project/terms.py

@@ -3,6 +3,8 @@ import itertools
 import collections
 from constants import L_lookup, level_sizes
 
+__all__ = ["format_terms", "get_spectroscopic_terms"]
+
 
 def format_terms(*terms, use_latex=False):
     strings = []
@@ -22,9 +24,10 @@ def format_terms(*terms, use_latex=False):
     return strings
 
 
-def get_spectroscopic_terms(n, l, n_electron, formatted=False):
-    assert l < n
-    assert n_electron <= level_sizes[l]
+def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False):
+    assert l < n, "`l` must be less than `n`"
+    assert n_electron <= level_sizes[l], "Number electrons must be no more than subshell capacity"
+    assert n_electron > 0, "Number electrons must be positive"
     m_l_range = list(range(-l, l + 1))
     m_s_range = [-1/2, 1/2]
 
@@ -92,7 +95,7 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False):
     terms = sorted(terms, key=lambda x: (x[0], x[1]), reverse=True)
 
     if formatted:
-        return format_terms(*terms)
+        return format_terms(*terms, use_latex=use_latex)
     else:
         return terms
 

### 2022-11-01 541_ism/final_project/atom.py
create atom class

@@ -1,0 +1,46 @@
+import configurations as con
+import terms
+import levels
+
+
+class Atom():
+    def __init__(self, name=None, n_electron=None, formatted=True, use_latex=False, no_cache=False):
+        if n_electron is None and name is None:
+            raise ValueError("One of `n_electron` or `name` must be supplied")
+        elif n_electron is None:
+            n_electron = con.parse_electrons(name)
+
+        self.name = name
+        self._n_electron = n_electron
+        self.formatted = formatted
+        self.use_latex = use_latex
+        self.no_cache = no_cache
+
+    def __repr__(self):
+        return f"<Atom: {self.name}, {self.configuration}, {self.terms[-1]}>"
+
+    @property
+    def _configuration(self):
+        return con.get_configuration(self._n_electron)
+
+    @property
+    def configuration(self):
+        if self.formatted:
+            return con.format_configuration(self._configuration, use_latex=self.use_latex)
+        else:
+            return self._configuration
+
+    @property
+    def _terms(self):
+        return terms.get_spectroscopic_terms(*self._configuration[-1])
+
+    @property
+    def terms(self):
+        if self.formatted:
+            return terms.format_terms(*self._terms, use_latex=self.use_latex)
+        else:
+            return self._terms
+
+    def plot_energy_levels(self, transitions, **kwargs):
+        return levels.plot_energy_levels(spec_terms=self._terms, transitions=transitions,
+                                         title=self.name, **kwargs)

### 2022-11-01 541_ism/final_project/terms.py
add stepbystep option for terms

@@ -6,6 +6,11 @@ from constants import L_lookup, level_sizes
 __all__ = ["format_terms", "get_spectroscopic_terms"]
 
 
+def _underline_print(str):
+    print(str)
+    print("".join(["-" for _ in range(len(str))]))
+
+
 def format_terms(*terms, use_latex=False):
     strings = []
     for S, L, J in terms:
@@ -24,7 +29,7 @@ def format_terms(*terms, use_latex=False):
     return strings
 
 
-def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False):
+def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, stepbystep=False):
     assert l < n, "`l` must be less than `n`"
     assert n_electron <= level_sizes[l], "Number electrons must be no more than subshell capacity"
     assert n_electron > 0, "Number electrons must be positive"
@@ -33,9 +38,19 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False):
 
     # these states are available to the electrons in this shell
     electron_states = [(m_l, m_s) for m_l in m_l_range for m_s in m_s_range]
+    if stepbystep:
+        _underline_print("Step 1: Possible electron states (m_l, m_s)")
+        # print("-------------------------------------------")
+        print(electron_states)
+        print()
 
     # find all combination of these states
     combs = itertools.combinations(electron_states, n_electron)
+    if stepbystep:
+        _underline_print("Step 2: All possible combinations of states")
+        print(list(combs))
+        print()
+        combs = itertools.combinations(electron_states, n_electron)
 
     # create a dict to keep track of the L, S combos (for the matrix)
     LS_combos = collections.defaultdict(int)
@@ -61,6 +76,13 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False):
         for j, M_s in enumerate(M_s_range):
             matrix[i, j] = LS_combos[(M_l, M_s)]
 
+    if stepbystep:
+        _underline_print("Step 3: Initial matrix")
+        print(matrix)
+        print()
+
+        _underline_print("Step 4: Matrix reduction")
+
     # keep track of the terms
     terms = []
 
@@ -81,15 +103,25 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False):
                 # work out the value of L and S for this term(s) based on the matrix
                 L, S = int(abs(M_l_range[i])), M_s_range[columns].max()
 
[…]

### 2022-11-01 541_ism/final_project/levels.py
sort transitions, add label options

@@ -24,13 +24,16 @@ params = {'figure.figsize': (12, 8),
 plt.rcParams.update(params)
 
 
-def plot_energy_levels(spec_terms, transitions, title=None, fig=None, ax=None, show=True):
+def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=True, show_n_labels=True,
+                       fig=None, ax=None, show=True, figsize=(6, 10)):
     x_vals = np.arange(len(transitions) + 2)
 
     if fig is None or ax is None:
-        fig, ax = plt.subplots(figsize=(6, 10))
+        fig, ax = plt.subplots(figsize=figsize)
 
     max_n = max(transitions)[0]
+
+    transitions = sorted(transitions, key=lambda x: (x[1], x[0]), reverse=False)
 
     previous_L = spec_terms[0][1]
     colour_ind = 0
@@ -43,10 +46,13 @@ def plot_energy_levels(spec_terms, transitions, title=None, fig=None, ax=None, s
             height += 2
             colour_ind += 1
         ax.plot(x_vals, np.ones_like(x_vals) * height, color=f"C{colour_ind}", lw=2)
-        ax.annotate(*terms.format_terms(term, use_latex=True), xy=(-0.1, height),
-                    ha="right", va="center", fontsize=fs)
-        ax.annotate(f"n = {n + 1}", xy=(x_vals[-1] + 0.1, height),
-                    ha="left", va="center", fontsize=0.8*fs)
+
+        if show_term_labels:
+            ax.annotate(*terms.format_terms(term, use_latex=True), xy=(-0.1, height),
+                        ha="right", va="center", fontsize=fs)
+        if show_n_labels:
+            ax.annotate(f"n = {n + 1}", xy=(x_vals[-1] + 0.1, height),
+                        ha="left", va="center", fontsize=0.8*fs)
         previous_L = L
         heights.append(height)
 
@@ -59,8 +65,9 @@ def plot_energy_levels(spec_terms, transitions, title=None, fig=None, ax=None, s
 
         wavelength = transition[2].to(u.Angstrom).value
         t_string = f"{wavelength:1.1f}" if wavelength < 1e5 else f"{wavelength:1.1e}"
-        ax.annotate(t_string + r"$\rm \AA$", xy=((x + 1) * 1.01, midpoint), rotation=90,
-                    va="center", fontsize=0.6*fs, color="grey")
[…]

### 2022-11-01 541_ism/final_project/atom.py
create a name is one isn't given

@@ -10,7 +10,7 @@ class Atom():
         elif n_electron is None:
             n_electron = con.parse_electrons(name)
 
-        self.name = name
+        self.name = name if name is not None else f"{n_electron} electrons"
         self._n_electron = n_electron
         self.formatted = formatted
         self.use_latex = use_latex

### 2022-11-01 541_ism/final_project/atom.py
add docstrings to atom and configurations

@@ -5,6 +5,25 @@ import levels
 
 class Atom():
     def __init__(self, name=None, n_electron=None, formatted=True, use_latex=False, no_cache=False):
+        """A class for accessing various electronic configuration/spectroscopic term/energy level diagram
+        functions.
+
+        Parameters
+        ----------
+        name : `str`, optional
+            The name of an atom or ion, by default None
+        n_electron : `int`, optional
+            A number of electrons, by default None
+        formatted : `bool`, optional
+            Whether to format variable into strings, by default True
+        use_latex : `bool`, optional
+            Whether to turn string variables into LaTeX, by default False
+
+        Raises
+        ------
+        ValueError
+            If neither `name` or `n_electron` are supplied
+        """
         if n_electron is None and name is None:
             raise ValueError("One of `n_electron` or `name` must be supplied")
         elif n_electron is None:

### 2022-11-01 541_ism/final_project/configurations.py

@@ -5,6 +5,13 @@ __all__ = ["parse_electrons", "get_configuration", "format_configuration"]
 
 
 def read_elements():
+    """Read the list of elements from our tiny csv file
+
+    Returns
+    -------
+    element_list : :class:`~np.ndarray`
+        List of elements, each has (n_electron, symbol, name)
+    """
     element_list = []
     with open("data/elements.csv") as elements:
         element_list = [line.rstrip("\n").split("\t") for line in elements]
@@ -12,8 +19,27 @@ def read_elements():
 
 
 def electrons_from_element(el_string):
+    """Convert a string representing an element to a number of electrons
+
+    Parameters
+    ----------
+    el_string : `str`
+        String that gives an element
+
+    Returns
+    -------
+    n_electron : `int`
+        Number of electrons corresponding to the string
+
+    Raises
+    ------
+    ValueError
+        If the string supplied doesn't match any elements
+    """
     n_elec, symbols, names = read_elements()
     electrons = []
+
+    # if they give a short string then match to a symbol, otherwise name to name
     if len(el_string) <= 2:
         electrons = n_elec[np.char.lower(symbols) == el_string.lower()]
     else:
@@ -26,6 +52,7 @@ def electrons_from_element(el_string):
 
 
 def parse_electrons(source):
+    # TODO: handle ionisation better
     split_source = source.split(" ")
     if len(split_source) > 1:
         element, ionisation = split_source
@@ -37,26 +64,49 @@ def parse_electrons(source):
 
 
 def get_configuration(n_electron, formatted=False, use_latex=False):
+    """Get the electronic configuration of an atom/ion given a number of electrons
+
+    Parameters
+    ----------
+    n_electron : `int`
+        Number of electrons
+    formatted : `bool`, optional
+        Whether to format the configuration into a string, by default False
+    use_latex : `bool`, optional
+        Whether to use LaTeX in the formatted string, by default False
+
+    Returns
+    -------
+    configuration : `various`
+        The electronic configuration
+    """
+    # the order in which to fill shells following the Aufbau principle
+    # each tuple is (n, l)
     order = [(1, 0), (2, 0), (2, 1), (3, 0), (3, 1),
              (4, 0), (3, 2), (4, 1), (4, 2), (5, 0), (5, 1)]
 
+    # some exceptions to the principle which we handle manually
     exceptions = {
[…]

### 2022-11-01 541_ism/final_project/terms.py
docstrings for terms

@@ -6,12 +6,33 @@ from constants import L_lookup, level_sizes
 __all__ = ["format_terms", "get_spectroscopic_terms"]
 
 
-def _underline_print(str):
-    print(str)
-    print("".join(["-" for _ in range(len(str))]))
+def _underline_print(string):
+    """Same as a regular print call but underlined with hyphens
+
+    Parameters
+    ----------
+    string : `str`
+        Some string to print out
+    """
+    print(string)
+    print("".join(["-" for _ in range(len(string))]))
 
 
 def format_terms(*terms, use_latex=False):
+    """Format a list of spectroscopic terms
+
+    Parameters
+    ----------
+    *terms : `tuple`
+        Each `term` should be a tuple of `2S+1, L, J`
+    use_latex : `bool`, optional
+        Whether to use LaTeX in the formatted string, by default False
+
+    Returns
+    -------
+    term_strings : `list` of `str`s
+        A list of strings representing each term
+    """
     strings = []
     for S, L, J in terms:
         term_string = None
@@ -30,6 +51,28 @@ def format_terms(*terms, use_latex=False):
 
 
 def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, stepbystep=False):
+    """Compute the spectroscopic terms for a given level (n, l) that contains n_electron electrons
+
+    Parameters
+    ----------
+    n : `int`
+        Quantum number of state
+    l : `int`
+        Angular momentum of state
+    n_electron : `int`
+        Number of electrons in the state
+    formatted : `bool`, optional
+        Whether to format the result as a string, by default False
+    use_latex : `bool`, optional
+        Whether to use LaTeX for the formatted string, by default False
+    stepbystep : `bool`, optional
+        Whether to print out the step-by-step algorithmic approach to getting the terms, by default False
+
+    Returns
+    -------
+    terms : `various`
+        The spectroscopic terms
+    """
     assert l < n, "`l` must be less than `n`"
     assert n_electron <= level_sizes[l], "Number electrons must be no more than subshell capacity"
     assert n_electron > 0, "Number electrons must be positive"
@@ -121,6 +164,8 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, 
         _underline_print("Step 5: Expand terms")
         print(*format_terms(*terms))
         print()
+
+        # it's crazy that I can add a cookie to my code
         print("Step 6: Go eat this 🍪")
 
     # sort based on Hund's 1st and 2nd rules

### 2022-11-01 541_ism/final_project/levels.py
docstring and comments for levels

@@ -26,61 +26,105 @@ plt.rcParams.update(params)
 
 def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=True, show_n_labels=True,
                        fig=None, ax=None, show=True, figsize=(6, 10)):
-    x_vals = np.arange(len(transitions) + 2)
+    """Plot an energy level diagram
 
+    Parameters
+    ----------
+    spec_terms : `list` of `tuples`
+        Spectroscopic terms for each energy level, the first is n = 1
+    transitions : `list` of `tuples`
+        A list of transitions, where each tuple is (n_u, n_l, lambda), the number of the upper state, the
+        number of the lower state and the wavelength of the transition
+    title : `str`, optional
+        A title for the plot, by default None
+    show_term_labels : `bool`, optional
+        Whether to label each energy level with its term, by default True
+    show_n_labels : `bool`, optional
+        Whether to label each energy level with its n, by default True
+    fig : `Figure`, optional
+        A matplotlib Figure on which to plot, by default it will be automatically create
+    ax : Axis, optional
+        A matplotlib Axis on which to plot, by default it will be automatically create
+    show : `bool`, optional
+        Whether to show the matplotlib figure immediately, by default True
+    figsize : `tuple`, optional
+        Size of the figure, by default (6, 10)
+
+    Returns
+    -------
+    fig, ax
+        The figure and axis with the plot on it
+    """
+    # create a figure if necessary
     if fig is None or ax is None:
         fig, ax = plt.subplots(figsize=figsize)
 
+    # work out the max n used in the transitions and sort the transitions
     max_n = max(transitions)[0]
-
     transitions = sorted(transitions, key=lambda x: (x[1], x[0]), reverse=False)
 
+    # create some dummy x values (1 for each transition plus some padding on the sides)
+    x_vals = np.arange(len(transitions) + 2)
+
+    # track the previous L, colour indicator, current height and all previous heights
     previous_L = spec_terms[0][1]
     colour_ind = 0
     height = 1
     heights = [1]
+
+    # loop over the terms (ignoring any above transitions)
     for n, term in enumerate(spec_terms[:max_n]):
-        L = term[1]
+        # add to the height
         height += 1
+
+        # if the L has changed then add extra to the height and change colour
+        L = term[1]
         if L != previous_L:
[…]

### 2022-11-05 541_ism/week6/problem_1.py
move problem to week 6

@@ -1,0 +1,10 @@
+from linetools.spectra.xspectrum1d import XSpectrum1D
+
+from linetools.guis import xspecgui as ltxsg
+
+# fname = '~/Downloads/J0929+4644_172_157_spec.fits'
+fname = '~/Downloads/J0943+0531_106_34_redspec.fits'
+
+sp = XSpectrum1D.from_file(fname)
+
+ltxsg.main(sp)

### 2022-11-06 541_ism/final_project/atom.py
add some ionisation functions

@@ -4,7 +4,8 @@ import levels
 
 
 class Atom():
-    def __init__(self, name=None, n_electron=None, formatted=True, use_latex=False, no_cache=False):
+    def __init__(self, name=None, n_electron=None, n_ion=0,
+                 formatted=True, use_latex=False, no_cache=False):
         """A class for accessing various electronic configuration/spectroscopic term/energy level diagram
         functions.
 
@@ -14,6 +15,8 @@ class Atom():
             The name of an atom or ion, by default None
         n_electron : `int`, optional
             A number of electrons, by default None
+        n_ion : `int`, optional
+            Number of times it has been ionised, by default None
         formatted : `bool`, optional
             Whether to format variable into strings, by default True
         use_latex : `bool`, optional
@@ -27,10 +30,11 @@ class Atom():
         if n_electron is None and name is None:
             raise ValueError("One of `n_electron` or `name` must be supplied")
         elif n_electron is None:
-            n_electron = con.parse_electrons(name)
+            n_electron, n_ion = con.parse_electrons(name)
 
         self.name = name if name is not None else f"{n_electron} electrons"
         self._n_electron = n_electron
+        self._n_ion = n_ion
         self.formatted = formatted
         self.use_latex = use_latex
         self.no_cache = no_cache
@@ -40,7 +44,7 @@ class Atom():
 
     @property
     def _configuration(self):
-        return con.get_configuration(self._n_electron)
+        return con.get_configuration(self._n_electron - self._n_ion)
 
     @property
     def configuration(self):

### 2022-11-06 541_ism/final_project/configurations.py

@@ -1,5 +1,6 @@
+from copy import copy
 import numpy as np
-from constants import L_lookup, level_sizes
+from constants import L_lookup, level_sizes, aufbau_order
 
 __all__ = ["parse_electrons", "get_configuration", "format_configuration"]
 
@@ -51,25 +52,33 @@ def electrons_from_element(el_string):
     return int(electrons[0])
 
 
-def parse_electrons(source):
-    # TODO: handle ionisation better
-    split_source = source.split(" ")
-    if len(split_source) > 1:
-        element, ionisation = split_source
-        n_I = sum([int(ionisation[i] == "I") for i in range(len(ionisation))])
+def parse_electrons(input_string):
+    final_char = input_string[-1]
+    if final_char in ["+", "-"]:
+        n_ion = int(input_string[-2]) * (-1 if final_char == "-" else 1)
+        n_elec = electrons_from_element(input_string[:-2])
+    elif final_char == "I":
+        n_ion = 0
+        i = -2
+        while input_string[i] == "I":
+            n_ion += 1
+            i -= 1
+        n_elec = electrons_from_element(input_string[:i + 1])
     else:
-        element = split_source[0]
-        n_I = 1
-    return electrons_from_element(element) - n_I + 1
+        n_elec = electrons_from_element(input_string)
+        n_ion = 0
+    return n_elec, n_ion
 
 
-def get_configuration(n_electron, formatted=False, use_latex=False):
+def get_configuration(n_electron, n_ion=0, formatted=False, use_latex=False):
     """Get the electronic configuration of an atom/ion given a number of electrons
 
     Parameters
     ----------
     n_electron : `int`
         Number of electrons
+    n_ion : `int`
+        Number of times to ionise
     formatted : `bool`, optional
         Whether to format the configuration into a string, by default False
     use_latex : `bool`, optional
@@ -80,10 +89,10 @@ def get_configuration(n_electron, formatted=False, use_latex=False):
     configuration : `various`
         The electronic configuration
     """
-    # the order in which to fill shells following the Aufbau principle
-    # each tuple is (n, l)
-    order = [(1, 0), (2, 0), (2, 1), (3, 0), (3, 1),
-             (4, 0), (3, 2), (4, 1), (4, 2), (5, 0), (5, 1)]
+    # if we are gaining electrons then just add them directly
+    if n_ion < 0:
+        n_electron += abs(n_ion)
+        n_ion = 0
 
     # some exceptions to the principle which we handle manually
[…]

### 2022-11-06 541_ism/final_project/constants.py

@@ -5,3 +5,7 @@ L_lookup = ["S", "P", "D", "F", "G", "H", "I", "J", "K"]
 
 # get the size of a level for each value of l
 level_sizes = [2, 6, 10, 14]
+
+# the order in which to fill shells following the Aufbau principle each tuple is (n, l)
+aufbau_order = [(1, 0), (2, 0), (2, 1), (3, 0), (3, 1),
+                (4, 0), (3, 2), (4, 1), (4, 2), (5, 0), (5, 1)]

### 2022-11-06 541_ism/final_project/atom.py
handle single + - signs

@@ -32,7 +32,7 @@ class Atom():
         elif n_electron is None:
             n_electron, n_ion = con.parse_electrons(name)
 
-        self.name = name if name is not None else f"{n_electron} electrons"
+        self.name = name if name is not None else f"({n_electron} - {n_ion}) electrons"
         self._n_electron = n_electron
         self._n_ion = n_ion
         self.formatted = formatted

### 2022-11-06 541_ism/final_project/configurations.py

@@ -53,10 +53,28 @@ def electrons_from_element(el_string):
 
 
 def parse_electrons(input_string):
+    """Parse the number of electrons and ionisation from an input string.
+
+    Parameters
+    ----------
+    input_string : `str`
+        An input string, something like "Carbon", "HII", "O3+"
+
+    Returns
+    -------
+    n_elec : `int`
+        Number of electrons
+    n_ion : `int`
+        Number of times to ionise
+    """
     final_char = input_string[-1]
     if final_char in ["+", "-"]:
-        n_ion = int(input_string[-2]) * (-1 if final_char == "-" else 1)
-        n_elec = electrons_from_element(input_string[:-2])
+        if input_string[-2].isdigit():
+            n_ion = int(input_string[-2]) * (-1 if final_char == "-" else 1)
+            n_elec = electrons_from_element(input_string[:-2])
+        else:
+            n_ion = -1 if final_char == "-" else 1
+            n_elec = electrons_from_element(input_string[:-1])
     elif final_char == "I":
         n_ion = 0
         i = -2
@@ -116,7 +134,7 @@ def get_configuration(n_electron, n_ion=0, formatted=False, use_latex=False):
                 break
 
     if n_ion > 0:
-        ionise_configuration(configuration=configuration, n_ion=n_ion)
+        configuration = ionise_configuration(configuration=configuration, n_ion=n_ion)
 
     # format the result if the user wants to
     if formatted:

### 2022-11-06 541_ism/final_project/atom.py
add warning if user asks terms when >1 half-filled

@@ -1,11 +1,12 @@
 import configurations as con
 import terms
 import levels
+import warnings
 
 
 class Atom():
     def __init__(self, name=None, n_electron=None, n_ion=0,
-                 formatted=True, use_latex=False, no_cache=False):
+                 formatted=True, use_latex=False, quiet=False):
         """A class for accessing various electronic configuration/spectroscopic term/energy level diagram
         functions.
 
@@ -21,6 +22,8 @@ class Atom():
             Whether to format variable into strings, by default True
         use_latex : `bool`, optional
             Whether to turn string variables into LaTeX, by default False
+        quiet : `bool`, optional
+            Whether to silence warnings, by default False
 
         Raises
         ------
@@ -37,14 +40,14 @@ class Atom():
         self._n_ion = n_ion
         self.formatted = formatted
         self.use_latex = use_latex
-        self.no_cache = no_cache
+        self.quiet = quiet
 
     def __repr__(self):
         return f"<Atom: {self.name}, {self.configuration}, {self.terms[-1]}>"
 
     @property
     def _configuration(self):
-        return con.get_configuration(self._n_electron - self._n_ion)
+        return con.get_configuration(n_electron=self._n_electron, n_ion=self._n_ion)
 
     @property
     def configuration(self):
@@ -55,6 +58,9 @@ class Atom():
 
     @property
     def _terms(self):
+        if not self.quiet and con.has_many_half_filled_shells(self._configuration):
+            warnings.warn(("This atom/ion has more than one half-filled subshell! The terms returned by this "
+                           "function only correspond to the *outermost* half-filled subshell."))
         return terms.get_spectroscopic_terms(*self._configuration[-1])
 
     @property

### 2022-11-06 541_ism/final_project/configurations.py

@@ -179,3 +179,11 @@ def ionise_configuration(configuration, n_ion):
             break
     configuration = [(n, l, n_filled) for n, l, n_filled in sorted(ionised_configuration) if n_filled > 0]
     return configuration
+
+
+def has_many_half_filled_shells(configuration):
+    n_half_filled = 0
+    for _, l, n_filled in configuration:
+        if n_filled < level_sizes[l]:
+            n_half_filled += 1
+    return n_half_filled > 1

### 2022-11-06 541_ism/final_project/configurations.py
add docstrings to new configuration functions

@@ -109,7 +109,7 @@ def get_configuration(n_electron, n_ion=0, formatted=False, use_latex=False):
     """
     # if we are gaining electrons then just add them directly
     if n_ion < 0:
-        n_electron += abs(n_ion)
+        n_electron += (-n_ion)
         n_ion = 0
 
     # some exceptions to the principle which we handle manually
@@ -133,6 +133,7 @@ def get_configuration(n_electron, n_ion=0, formatted=False, use_latex=False):
             if n_electron <= 0:
                 break
 
+    # ionise the configuration if necessary
     if n_ion > 0:
         configuration = ionise_configuration(configuration=configuration, n_ion=n_ion)
 
@@ -168,21 +169,59 @@ def format_configuration(configuration, use_latex=False):
 
 
 def ionise_configuration(configuration, n_ion):
+    """Ionise a configuration a given number of times
+
+    Parameters
+    ----------
+    configuration : `list` of `tuples`
+        Electronic configuration
+    n_ion : `int`
+        Number of times to ionise, must be positive
+
+    Returns
+    -------
+    ionised_configuration : `list` of `tuples`
+        Configuration after ionisation
+    """
+    assert n_ion > 0, "Number of ionisations must be positive"
+
+    # sort in descending n, l
     ionised_configuration = sorted(copy(configuration), reverse=True)
+
+    # step through each shell
     for i, conf in enumerate(ionised_configuration):
         n, l, n_filled = conf
+
+        # remove electrons from this shell
         n_left = n_filled - n_ion
         ionised_configuration[i] = (n, l, n_left)
+
+        # if there are still electrons left to remove then change n_ion
         if n_left < 0:
             n_ion -= n_filled
         else:
             break
+
+    # add the configuration back in order for any shells that have a positive n_filled value
     configuration = [(n, l, n_filled) for n, l, n_filled in sorted(ionised_configuration) if n_filled > 0]
     return configuration
 
 
 def has_many_half_filled_shells(configuration):
+    """Check whether a configuration has more than 1 half-filled subshell
+
+    Parameters
+    ----------
+    configuration : `list` of `tuples`
+        Electronic configuration
+
+    Returns
+    -------
+    flag : `bool`
+        Returns true when more than 1 are half-filled
+    """
     n_half_filled = 0
+    # check each subshell and whether it has reached the level size
     for _, l, n_filled in configuration:
         if n_filled < level_sizes[l]:
             n_half_filled += 1

### 2022-11-06 541_ism/final_project/atom.py
convert electrons into a string

@@ -35,7 +35,10 @@ class Atom():
         elif n_electron is None:
             n_electron, n_ion = con.parse_electrons(name)
 
-        self.name = name if name is not None else f"({n_electron} - {n_ion}) electrons"
+        assert n_ion <= n_electron, "Can't ionise more electrons than you have!"
+
+        self.auto_name = name is None
+        self.name = name
         self._n_electron = n_electron
         self._n_ion = n_ion
         self.formatted = formatted
@@ -43,7 +46,11 @@ class Atom():
         self.quiet = quiet
 
     def __repr__(self):
-        return f"<Atom: {self.name}, {self.configuration}, {self.terms[-1]}>"
+        if self.auto_name:
+            self.name = con.electrons_to_string(n_electron=self._n_electron, n_ion=self._n_ion)
+        lowest_term = self.terms[-1] if len(self.terms) > 0 else "No terms"
+        config = self.configuration if self.configuration != "" else "No electrons"
+        return f"<Atom: {self.name}, {config}, {lowest_term}>"
 
     @property
     def _configuration(self):
@@ -61,7 +68,10 @@ class Atom():
         if not self.quiet and con.has_many_half_filled_shells(self._configuration):
             warnings.warn(("This atom/ion has more than one half-filled subshell! The terms returned by this "
                            "function only correspond to the *outermost* half-filled subshell."))
-        return terms.get_spectroscopic_terms(*self._configuration[-1])
+        if len(self._configuration) > 0:
+            return terms.get_spectroscopic_terms(*self._configuration[-1])
+        else:
+            return []
 
     @property
     def terms(self):
@@ -70,6 +80,34 @@ class Atom():
         else:
             return self._terms
 
+    def ionise(self, delta_n_ion=1):
+        """Ionise the Atom a number of times
+
+        Parameters
+        ----------
+        delta_n_ion : `int`, optional
+            Number of time to ionise, can be negative to add electrons, by default 1
+        """
+        if self._n_ion + delta_n_ion > self._n_electron:
+            raise ValueError("Can't ionise more electrons than you have!")
+        self._n_ion += delta_n_ion
+        self.auto_name = True
+
     def plot_energy_levels(self, transitions, **kwargs):
+        """Plot energy level diagram for the Atom
+
+        Parameters
+        ----------
+        transitions : `list` of `tuples`
+            A list of transitions, where each tuple is (n_u, n_l, lambda), the number of the upper state, the
+            number of the lower state and the wavelength of the transition
[…]

### 2022-11-06 541_ism/final_project/configurations.py

@@ -88,6 +88,37 @@ def parse_electrons(input_string):
     return n_elec, n_ion
 
 
+def electrons_to_string(n_electron, n_ion):
+    if n_ion < 0:
+        ion_string = f"{-n_ion}-"
+    else:
+        # add 1 since n_ion=1 corresponds to II not I
+        n_ion += 1
+
+        # convert n_ion to a roman numeral
+        roman_num = [1, 4, 5, 9, 10, 40, 50, 90, 100, 400, 500, 900, 1000]
+        roman_sym = ["I", "IV", "V", "IX", "X", "XL", "L", "XC", "C", "CD", "D", "CM", "M"]
+        i = 12
+
+        ion_string = ""
+        while n_ion:
+            div = n_ion // roman_num[i]
+            n_ion %= roman_num[i]
+
+            while div:
+                ion_string += roman_sym[i]
+                div -= 1
+            i -= 1
+    n_elec, symbols, _ = read_elements()
+
+    matching_symbol = symbols[n_elec.astype(int) == n_electron]
+    if len(matching_symbol) == 0:
+        raise ValueError(f"Can't find an element to match your `n_electron` value - `{n_electron}`")
+
+    # print(symbols, n_electron)
+    return f"{matching_symbol[0]}{ion_string}"
+
+
 def get_configuration(n_electron, n_ion=0, formatted=False, use_latex=False):
     """Get the electronic configuration of an atom/ion given a number of electrons
 

### 2022-11-06 541_ism/final_project/configurations.py
docstring/comments for electrons_to_string

@@ -89,6 +89,27 @@ def parse_electrons(input_string):
 
 
 def electrons_to_string(n_electron, n_ion):
+    """Convert a number of electrons and ionisations into a string.
+
+    Ionisations will be represented with roman numerals when non-negative.
+
+    Parameters
+    ----------
+    n_electron : `int`
+        Number of electrons
+    n_ion : `int`
+        Number of ionisations (may be negative)
+
+    Returns
+    -------
+    el_string : `str`
+        String representing the input
+
+    Raises
+    ------
+    ValueError
+        If the number of electrons is not in the table of elements
+    """
     if n_ion < 0:
         ion_string = f"{-n_ion}-"
     else:
@@ -109,13 +130,15 @@ def electrons_to_string(n_electron, n_ion):
                 ion_string += roman_sym[i]
                 div -= 1
             i -= 1
-    n_elec, symbols, _ = read_elements()
 
+    # find the matching symbol in the elements table
+    n_elec, symbols, _ = read_elements()
     matching_symbol = symbols[n_elec.astype(int) == n_electron]
+
+    # raise an error if it isn't there
     if len(matching_symbol) == 0:
         raise ValueError(f"Can't find an element to match your `n_electron` value - `{n_electron}`")
 
-    # print(symbols, n_electron)
     return f"{matching_symbol[0]}{ion_string}"
 
 

### 2022-11-06 541_ism/final_project/configurations.py
update configurations __all__

@@ -2,7 +2,8 @@ from copy import copy
 import numpy as np
 from constants import L_lookup, level_sizes, aufbau_order
 
-__all__ = ["parse_electrons", "get_configuration", "format_configuration"]
+__all__ = ["parse_electrons", "get_configuration", "format_configuration",  "electrons_from_element",
+           "electrons_to_string", "ionise_configuration", "has_many_half_filled_shells"]
 
 
 def read_elements():

### 2022-11-07 541_ism/final_project/configurations.py
handle roman input above III

@@ -68,21 +68,33 @@ def parse_electrons(input_string):
     n_ion : `int`
         Number of times to ionise
     """
-    final_char = input_string[-1]
+    # get the final character for strings of more than 1 character
+    final_char = input_string[-1] if len(input_string) > 1 else ""
+
+    # check for O3+, H- style
     if final_char in ["+", "-"]:
+        # handle the case where it is [number][+/-]
         if input_string[-2].isdigit():
             n_ion = int(input_string[-2]) * (-1 if final_char == "-" else 1)
             n_elec = electrons_from_element(input_string[:-2])
+        # handle the case where it is just a +/-
         else:
             n_ion = -1 if final_char == "-" else 1
             n_elec = electrons_from_element(input_string[:-1])
-    elif final_char == "I":
+
+    # check for OVI, SII style
+    elif final_char in ["I", "V"]:
         n_ion = 0
+
+        # loop backwards through the string until the character isn't a 'I' or 'V'
+        # second condition is to ignore the first character (isn't relevant and making Vanadium work)
         i = -2
-        while input_string[i] == "I":
-            n_ion += 1
+        while input_string[i] in ["I", "V"] and (-i) < len(input_string):
             i -= 1
         n_elec = electrons_from_element(input_string[:i + 1])
+        n_ion = _roman_to_int(input_string[i + 1:])
+
+    # otherwise just lookup the element
     else:
         n_elec = electrons_from_element(input_string)
         n_ion = 0
@@ -281,3 +293,31 @@ def has_many_half_filled_shells(configuration):
         if n_filled < level_sizes[l]:
             n_half_filled += 1
     return n_half_filled > 1
+
+
+def _roman_to_int(s):
+    """Convert roman numerals to an integer
+
+    I just copied this: `https://www.tutorialspoint.com/roman-to-integer-in-python`_
+
+    Parameters
+    ----------
+    s : `str`
+        A string of roman numerals (only works up to 10)
+
+    Returns
+    -------
+    num : `int`
+        Number corresponding to roman numerals
+    """
+    roman = {'I': 1, 'IV': 4, 'V': 5, 'IX': 9, 'X': 10}
+    i = 0
+    num = 0
+    while i < len(s):
+        if i + 1 < len(s) and s[i:i + 2] in roman:
+            num += roman[s[i:i + 2]]
+            i += 2
+        else:
+            num += roman[s[i]]
+            i += 1
+    return num

### 2022-11-07 541_ism/final_project/proposal.tex
add final project proposal for ISM

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\class}{ASTR 541 - Interstellar Medium}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Final Project Proposal \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Final Project Proposal}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\tomtitle{}

\vspace{-1.5cm}

\section{General Idea}
I enjoyed the lectures on electronic configuration and spectroscopic terms and I was intrigued by the concept of implementing an algorithm to solve the problems that we did in class by hand. Therefore, for my project I intend to write some code to work out electronic configurations, spectroscopic terms and also create energy level diagrams.

\section{Intended Products}
\subsection{Code}
I'd like to design a small bit of code to do the following. Note in several places I have to say \emph{``(only guaranteed to work for atoms)''} because it seems that Hund's rules and the Aufbau Principle do not apply to ions in the same way - there don't seem to be rules for ions as far as I can tell.\\

\noindent \textbf{Electronic Configurations}
\begin{itemize}
    \item \textbf{Input:} either a number of electrons and ionisations or a string like ``HII'' or ``O3+''
    \item \textbf{Output:} the electronic configuration in a list of $[n, l, n_e)$ tuples - \emph{(only guaranteed to work for atoms)}
    \item Add options to return in ASCII or LaTeX format
\end{itemize}

\noindent \textbf{Spectroscopic Terms}
\begin{itemize}
    \item \textbf{Input:} Subshell parameters, $n, l$ and number of electrons for the shell, $n_e$
    \item \textbf{Output:} the spectroscopic terms in the a list of $(2S+1, L, J)$ tuples - \emph{(only guaranteed to work for atoms)}
    \item Add options to return in ASCII or LaTeX format
    \item Sort terms by energy based on Hund's rules
    \item Work out the ground state spectroscopic terms given an electronic configuration
\end{itemize}

\noindent \textbf{Energy Level diagrams}
\begin{itemize}
    \item \textbf{Input:} a list of spectroscopic terms (or a string to be parsed and it'll calculate them) for a series of levels, a list of transitions between levels as tuples $(n_up, n_low, \lambda)$ - these transitions could be pulled directly from \texttt{linetools}
    \item \textbf{Output:} an energy level diagram with levels labelled with their terms, colour coded and annotated with transition wavelengths
    \item I'm also interesting in adding conditions for whether a transition is forbidden and changing the style of the arrow if it is
    \item So big picture I could get to a point where I write something like \texttt{plot\_energy\_levels("SII", transitions)} and it'll work out the electronic configuration, then the spectroscopic terms, plot up the levels and then add the transitions that you specify.
\end{itemize}

\subsection{Discussion}

I also think I could write/talk about a couple of different things that I found whilst reading:
\begin{itemize}
    \item My implementation of the algorithms for working out the configurations/terms \emph{(writing only, no one wants to hear this in the presentation haha)}
    \item There is a reason for the periodic table's weird shape, we can talk about it!
    \item Symmetry of spectroscopic terms within subshells
    \item The problems with the Aufbau Principle and how electron shielding plays a role in messing it up
\end{itemize}

\section{Format}
My intended format will be a link to the GitHub repository containing the code (which will be fully comments and docstringed), as well as a Jupyter Notebook that demonstrates its use and discusses the points above.

\end{document}

### 2022-11-07 541_ism/final_project/atom.py
format terms more intelligently

@@ -76,7 +76,7 @@ class Atom():
     @property
     def terms(self):
         if self.formatted:
-            return terms.format_terms(*self._terms, use_latex=self.use_latex)
+            return terms.format_terms(self._terms, use_latex=self.use_latex)
         else:
             return self._terms
 

### 2022-11-07 541_ism/final_project/levels.py

@@ -88,7 +88,7 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
 
         # add term and n labels as desired
         if show_term_labels:
-            ax.annotate(*terms.format_terms(term, use_latex=True), xy=(-0.1, height),
+            ax.annotate(terms.format_terms(term, use_latex=True), xy=(-0.1, height),
                         ha="right", va="center", fontsize=fs)
         if show_n_labels:
             ax.annotate(f"n = {n + 1}", xy=(x_vals[-1] + 0.1, height),

### 2022-11-07 541_ism/final_project/terms.py

@@ -18,12 +18,12 @@ def _underline_print(string):
     print("".join(["-" for _ in range(len(string))]))
 
 
-def format_terms(*terms, use_latex=False):
+def format_terms(terms, use_latex=False):
     """Format a list of spectroscopic terms
 
     Parameters
     ----------
-    *terms : `tuple`
+    terms : `list` of `tuple`s or `tuple`
         Each `term` should be a tuple of `2S+1, L, J`
     use_latex : `bool`, optional
         Whether to use LaTeX in the formatted string, by default False
@@ -33,21 +33,24 @@ def format_terms(*terms, use_latex=False):
     term_strings : `list` of `str`s
         A list of strings representing each term
     """
+    if isinstance(terms, tuple):
+        terms = [terms]
     strings = []
     for S, L, J in terms:
         term_string = None
         if not J.is_integer():
             if use_latex:
-                term_string = rf"$^{{{S}}} {L_lookup[L]}_{{\frac{{{int(J * 2)}}}{{2}}}}$"
+                term_string = rf"${{}}^{{{S}}} {L_lookup[L]}_{{\frac{{{int(J * 2)}}}{{2}}}}$"
             else:
                 term_string = f"{S}{L_lookup[L]}({int(J * 2)}/2)"
         else:
             if use_latex:
-                term_string = rf"$^{{{S}}} {L_lookup[L]}_{{{int(J)}}}$"
+                term_string = rf"${{}}^{{{S}}} {L_lookup[L]}_{{{int(J)}}}$"
             else:
                 term_string = f"{S}{L_lookup[L]}{int(J)}"
         strings.append(term_string)
-    return strings
+    
+    return ", ".join(strings)
 
 
 def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, stepbystep=False):
@@ -162,7 +165,7 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, 
     if stepbystep:
         print()
         _underline_print("Step 5: Expand terms")
-        print(*format_terms(*terms))
+        print(format_terms(terms))
         print()
 
         # it's crazy that I can add a cookie to my code
@@ -172,6 +175,6 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, 
     terms = sorted(terms, key=lambda x: (x[0], x[1]), reverse=True)
 
     if formatted:
-        return format_terms(*terms, use_latex=use_latex)
[…]

### 2022-11-07 541_ism/final_project/atom.py
add parity handling for spec_terms

@@ -48,7 +48,7 @@ class Atom():
     def __repr__(self):
         if self.auto_name:
             self.name = con.electrons_to_string(n_electron=self._n_electron, n_ion=self._n_ion)
-        lowest_term = self.terms[-1] if len(self.terms) > 0 else "No terms"
+        lowest_term = self.terms.split(", ")[-1] if len(self.terms) > 0 else "No terms"
         config = self.configuration if self.configuration != "" else "No electrons"
         return f"<Atom: {self.name}, {config}, {lowest_term}>"
 

### 2022-11-07 541_ism/final_project/terms.py

@@ -36,20 +36,21 @@ def format_terms(terms, use_latex=False):
     if isinstance(terms, tuple):
         terms = [terms]
     strings = []
-    for S, L, J in terms:
+    for S, L, J, parity in terms:
         term_string = None
+        parity_string = "o" if parity == 0 else ""
         if not J.is_integer():
             if use_latex:
-                term_string = rf"${{}}^{{{S}}} {L_lookup[L]}_{{\frac{{{int(J * 2)}}}{{2}}}}$"
+                term_string = rf"${{}}^{{{S}}} {L_lookup[L]}_{{\frac{{{int(J * 2)}}}{{2}}}}^{{\rm {parity_string}}}$"
             else:
-                term_string = f"{S}{L_lookup[L]}({int(J * 2)}/2)"
+                term_string = f"{S}{L_lookup[L]}({int(J * 2)}/2){parity_string}"
         else:
             if use_latex:
-                term_string = rf"${{}}^{{{S}}} {L_lookup[L]}_{{{int(J)}}}$"
+                term_string = rf"${{}}^{{{S}}} {L_lookup[L]}_{{{int(J)}}}^{{\rm {parity_string}}}$"
             else:
-                term_string = f"{S}{L_lookup[L]}{int(J)}"
+                term_string = f"{S}{L_lookup[L]}{int(J)}{parity_string}"
         strings.append(term_string)
-    
+
     return ", ".join(strings)
 
 
@@ -81,6 +82,11 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, 
     assert n_electron > 0, "Number electrons must be positive"
     m_l_range = list(range(-l, l + 1))
     m_s_range = [-1/2, 1/2]
+
+    if (l % 2) == 0:
+        parity = 1
+    else:
+        parity = (n_electron % 2) == 0
 
     # these states are available to the electrons in this shell
     electron_states = [(m_l, m_s) for m_l in m_l_range for m_s in m_s_range]
@@ -154,7 +160,7 @@ def get_spectroscopic_terms(n, l, n_electron, formatted=False, use_latex=False, 
                 if n_electron > level_sizes[l] / 2:
                     J_range = reversed(J_range)
                 for J in J_range:
-                    terms.append((int(2 * S + 1), L, J))
+                    terms.append((int(2 * S + 1), L, J, parity))
 
                 if stepbystep:
                     print(unitary, f"{int(2 * S + 1)}{L_lookup[L]}")

### 2022-11-07 541_ism/final_project/levels.py
implement different style lines for forbidden transitions

@@ -4,7 +4,7 @@ import astropy.units as u
 
 import terms
 
-__all__ = ["plot_energy_levels"]
+__all__ = ["plot_energy_levels", "is_forbidden"]
 
 plt.rc('font', family='serif')
 plt.rcParams['text.usetex'] = False
@@ -100,6 +100,9 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
 
     # loop over transitions
     for x, transition in enumerate(transitions):
+        t_flag = apply_selection_rules(spec_terms[transition[0] - 1], spec_terms[transition[1] - 1])
+        linestyle = "-" if t_flag == 0 else ("dotted" if t_flag == 1 else "dashed")
+
         # work out the height of the upper and lower levels (plus a midpoint)
         upper_height = heights[transition[0]]
         lower_height = heights[transition[1]]
@@ -107,7 +110,7 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
 
         # plot an arrow from the upper to the lower
         ax.annotate("", xytext=(x + 1, upper_height), xy=(x + 1, lower_height),
-                    arrowprops=dict(arrowstyle="-|>", linewidth=2, color="grey"))
+                    arrowprops=dict(arrowstyle="-|>", linewidth=2, linestyle=linestyle, color="grey"))
 
         # add an annotation of the wavelength at the midpoint of the arrow
         wavelength = transition[2].to(u.Angstrom).value
@@ -128,3 +131,37 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
     if show:
         plt.show()
     return fig, ax
+
+
+def apply_selection_rules(term_low, term_up):
+    """Apply electric dipole transition selection rules to a pair of spectroscopic terms.
+
+    This follows the rules from Draine Section 6.7
+
+    Parameters
+    ----------
+    term_low : `tuple`
+        Spectroscopic term of lower level in form (2S+1, L, J, parity)
+    term_up : `tuple`
+        Spectroscopic term of lower level in form (2S+1, L, J, parity)
+
+    Returns
+    -------
+    flag : `int
+        A flag describing the type of transition (0=permitted, 1=semiforbidden, 2=forbidden)
+    """
+    mult_low, L_low, J_low, parity_low = term_low
[…]

### 2022-11-07 541_ism/final_project/levels.py
add reasons for forbidden flag!

@@ -100,8 +100,8 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
 
     # loop over transitions
     for x, transition in enumerate(transitions):
-        t_flag = apply_selection_rules(spec_terms[transition[0] - 1], spec_terms[transition[1] - 1])
-        linestyle = "-" if t_flag == 0 else ("dotted" if t_flag == 1 else "dashed")
+        t_flag, _ = apply_selection_rules(spec_terms[transition[0] - 1], spec_terms[transition[1] - 1])
+        linestyle = "-" if t_flag == 0 else ("dashed" if t_flag == 1 else "dotted")
 
         # work out the height of the upper and lower levels (plus a midpoint)
         upper_height = heights[transition[0]]
@@ -110,7 +110,9 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
 
         # plot an arrow from the upper to the lower
         ax.annotate("", xytext=(x + 1, upper_height), xy=(x + 1, lower_height),
-                    arrowprops=dict(arrowstyle="-|>", linewidth=2, linestyle=linestyle, color="grey"))
+                    arrowprops=dict(arrowstyle="-", linewidth=2, shrinkB=5, linestyle=linestyle, color="grey"))
+        ax.annotate("", xytext=(x + 1, upper_height), xy=(x + 1, lower_height),
+                    arrowprops=dict(arrowstyle="-|>", linewidth=0, mutation_scale=20, color="grey"))
 
         # add an annotation of the wavelength at the midpoint of the arrow
         wavelength = transition[2].to(u.Angstrom).value
@@ -147,8 +149,10 @@ def apply_selection_rules(term_low, term_up):
 
     Returns
     -------
-    flag : `int
+    flag : `int`
         A flag describing the type of transition (0=permitted, 1=semiforbidden, 2=forbidden)
+    reason : `str`
+        A list of reasons for why a flag was chosen
     """
     mult_low, L_low, J_low, parity_low = term_low
     mult_up, L_up, J_up, parity_up = term_up
@@ -164,4 +168,18 @@ def apply_selection_rules(term_low, term_up):
     semiforbidden = parity_changed & delta_J_allowed & delta_L_allowed
     permitted = semiforbidden & spin_unchanged
 
-    return 0 if permitted else (1 if semiforbidden else 2)
+    reason_strings = [
+        "The parity changed",
[…]

### 2022-11-07 541_ism/week6/solutions.tex
move problem to week 6
remove mhchems
the spectra are done, and my braincells are gone

--- 2022-10-31 541_ism/week5/solutions.tex
@@
 \usepackage{xcolor}
-\usepackage{mhchem}
+\usepackage{array}
 \hypersetup{
@@
 
-\newcommand{\psetnum}{5}
+\newcommand{\psetnum}{6}
 \newcommand{\class}{ASTR 541 - Interstellar Medium}
@@
 
-\question{\textbf{1. }}
+\question{\textbf{1. Spectroscopy}}
+\answer{
+    I'm going to summarise my results in a single table to make it a bit easier to parse.
+    \begin{center}
+        \boxed{
+        \begin{tabular}{l|c|c|c|c|c|c}
+            Galaxy & Redshift & SII Ratio & O3HB & N2 & O3N2 & Oxygen Abundance \\
+            \hline
+            J0929+4644\_172\_157 & 0.017 & 1.27 & 5.45 & 0.05 & 2.06 & 8.17 \\
+            J0943+0531\_106\_34 & 0.228 & 1.30 & 0.43 & 0.55 & -0.11 & 8.77
+        \end{tabular}}
+    \end{center}
+    \textbf{Density regimes:} Both of these galaxies are in a low density regime since the SII ratios are both close to 1.44 (which we calculated in the last homework).
+
+    \noindent \textbf{Spectrum Noise:} The second galaxy is noisier because it is at a higher redshift and therefore all fluxes are lower.
+
+    \noindent \textbf{Galaxy Masses:} The second galaxy has a higher oxygen abundance.
+Therefore I think this means that \textbf{the second galaxy must be more massive}.
+The reason for this is that massive stars enrich the gas and increase the oxygen abundance.
+So a higher stellar mass would result in a larger oxygen abundance.
+}
 

### 2022-11-07 541_ism/week6/problem_1.py

@@ -1,10 +1,42 @@
 from linetools.spectra.xspectrum1d import XSpectrum1D
-
 from linetools.guis import xspecgui as ltxsg
+import numpy as np
 
-# fname = '~/Downloads/J0929+4644_172_157_spec.fits'
-fname = '~/Downloads/J0943+0531_106_34_redspec.fits'
 
-sp = XSpectrum1D.from_file(fname)
+def run_gui():
+    fname = '~/Downloads/J0929+4644_172_157_spec.fits'
+    # fname = '~/Downloads/J0943+0531_106_34_redspec.fits'
 
-ltxsg.main(sp)
+    sp = XSpectrum1D.from_file(fname)
+
+    ltxsg.main(sp)
+
+
+# run_gui()
+
+z = [0.01657, 0.22833]
+sII_6716 = [583.826, 18.8138]
+sII_6731 = [459.247, 14.4956]
+oIII_5007 = [14772.4, 16.6357]
+H_beta = [2712.44, 38.7836]
+nII_6583 = [452.39, 83.1522]
+H_alpha = [9469.3, 150.604]
+
+for i in [0, 1]:
+    print(f"Galaxy {i + 1}")
+    sII_ratio = sII_6716[i] / sII_6731[i]
+    O3HB = oIII_5007[i] / H_beta[i]
+    N2 = nII_6583[i] / H_alpha[i]
+    O3N2 = np.log10(O3HB / N2)
+
+    if O3N2 < 1.9:
+        oxy_abund = 8.73 - (0.32 * O3N2)
+    else:
+        oxy_abund = 8.92 + (0.57 * np.log10(N2))
+
+    print(f"  Redshift = {z[i]:1.3f}")
+    print(f"  sII ratio = {sII_ratio:1.2f}")
+    print(f"  O3HB = {O3HB:1.2f}")
+    print(f"  N2 = {N2:1.2f}")
+    print(f"  O3N2 = {O3N2:1.2f}")
+    print(f"  Oxygen Abundance = {oxy_abund:1.2f}")

### 2022-11-08 541_ism/final_project/levels.py
add option for showing all energy levels

@@ -24,8 +24,8 @@ params = {'figure.figsize': (12, 8),
 plt.rcParams.update(params)
 
 
-def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=True, show_n_labels=True,
-                       fig=None, ax=None, show=True, figsize=(6, 10)):
+def plot_energy_levels(spec_terms, transitions=[], title=None, show_term_labels=True, show_n_labels=True,
+                       show_all_levels=False, fig=None, ax=None, show=True, figsize=(6, 10)):
     """Plot an energy level diagram
 
     Parameters
@@ -41,6 +41,8 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
         Whether to label each energy level with its term, by default True
     show_n_labels : `bool`, optional
         Whether to label each energy level with its n, by default True
+    show_all_levels : `bool`, optional
+        Whether to show every energy level (or only those relevant for transitions), by default False
     fig : `Figure`, optional
         A matplotlib Figure on which to plot, by default it will be automatically create
     ax : Axis, optional
@@ -60,7 +62,7 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
         fig, ax = plt.subplots(figsize=figsize)
 
     # work out the max n used in the transitions and sort the transitions
-    max_n = max(transitions)[0]
+    max_n = max(transitions)[0] if transitions != [] and not show_all_levels else len(spec_terms)
     transitions = sorted(transitions, key=lambda x: (x[1], x[0]), reverse=False)
 
     # create some dummy x values (1 for each transition plus some padding on the sides)
@@ -100,6 +102,8 @@ def plot_energy_levels(spec_terms, transitions, title=None, show_term_labels=Tru
 
     # loop over transitions
     for x, transition in enumerate(transitions):
+        if transitions == []:
+            break
         t_flag, _ = apply_selection_rules(spec_terms[transition[0] - 1], spec_terms[transition[1] - 1])
         linestyle = "-" if t_flag == 0 else ("dashed" if t_flag == 1 else "dotted")
 

### 2022-11-08 541_ism/final_project/configurations.py
off by 1 error!

@@ -92,7 +92,7 @@ def parse_electrons(input_string):
         while input_string[i] in ["I", "V"] and (-i) < len(input_string):
             i -= 1
         n_elec = electrons_from_element(input_string[:i + 1])
-        n_ion = _roman_to_int(input_string[i + 1:])
+        n_ion = _roman_to_int(input_string[i + 1:]) - 1
 
     # otherwise just lookup the element
     else:

### 2022-11-08 541_ism/week6/problem_2.py
start ISM w6p2

@@ -1,0 +1,7 @@
+from scipy.optimize import fsolve
+import numpy as np
+
+def func(T):
+    return T * (0.86 + 0.54 * (T / 1e4)**(0.37)) - 3/2 * 32000
+
+print(fsolve(func, 32000))

### 2022-11-13 541_ism/final_project/configurations.py
absolute path

@@ -15,7 +15,7 @@ def read_elements():
         List of elements, each has (n_electron, symbol, name)
     """
     element_list = []
-    with open("data/elements.csv") as elements:
+    with open("/home/tomwagg/Documents/uw-grad-classes/541_ism/final_project/data/elements.csv") as elements:
         element_list = [line.rstrip("\n").split("\t") for line in elements]
     return np.transpose(element_list)
 

### 2022-11-13 541_ism/week6/solutions.tex
start ISM w6p2
finish week6 problems

@@
     \end{center}
-    \textbf{Density regimes:} Both of these galaxies are in a low density regime since the SII ratios are both close to 1.44 (which we calculated in the last homework).
+    \textbf{Density regimes:} Both of these galaxies are in a low density regime since the SII ratios are both close to 1.44 (which we calculated in the last homework).\\
 
-    \noindent \textbf{Spectrum Noise:} The second galaxy is noisier because it is at a higher redshift and therefore all fluxes are lower.
+    \noindent \textbf{Spectrum Noise:} The second galaxy is noisier because it is at a higher redshift and therefore all fluxes are lower.\\
 
@@
 
+\question{\textbf{2. HII Heating and Cooling Balance}}
+
+\question{2a. Heating}
+\answer{
+    The main source of heating is the photoionisation.
+The energy of the photons is transferred into ionising the hydrogen and the leftover kinetic energy that the electrons gains leads to heating of the gas.
+The heating rate due to this can be written as follows
+    \begin{equation}
+        \boxed{\Gamma_{\rm ion, H} = \underbrace{\vphantom{\frac{3}{2} k_B T_{\rm init}} n_e n_p}_{(1)} \underbrace{\vphantom{\frac{3}{2} k_B T_{\rm init}} \alpha_B(T)}_{(2)} \underbrace{\frac{3}{2} k_B T_{\rm init}}_{(3)}}
+    \end{equation}
+    Where (1) shows the density of ``things that could be ionised'', i.e.
+the hydrogen, and the density of electrons.
+(2) is the total recombination coefficient which has the temperature dependence and indicate the rate at which recombination occurs.
+(3) is the average initial kinetic energy.
+}
+
+\clearpage
+
+\question{2b. Cooling}
+\answer{
+    A large source of cooling is recombination, this cooling rate can be written as follows. We can also convert $\beta_H(T)$ to $\alpha_B(T)$ using the expression given in the question.
+    \begin{align}
+        \Lambda_{\rm recomb, H} &= n_e n_p k_B T \beta_H(T)\\
+        \Aboxed{ \Lambda_{\rm recomb, H} &= 0.86 n_e n_p k_B T \alpha_B(T) }
+    \end{align}
+    $\beta_H(T)$ is the likelihood that an electron will be reabsorbed per energy (and includes the maxwellian distribution), $k_B T$ gives the energy based on the temperature and the densities indicate how often we find electrons to be reabsorbed and protons with which to recombine.
+}
+
+\question{2c. Balance}
+\answer{
+    Now we can equate the heating rate from part (a) and the cooling rate from part b (plus the free-free emission) and reduce the equation.
+    \begin{align}
+        n_e n_p \alpha_B(T) \cdot \frac{3}{2} k_B T_{\rm init} &= 0.86 n_e n_p k_B T \alpha_B(T) + 0.54 n_e n_p \alpha_B(T) \qty(\frac{T}{10^4 \unit{K}})^{0.37} k_B T \\
+        \alpha_B(T) \cdot \frac{3}{2} T_{\rm init} &= 0.86 T \alpha_B(T) + 0.54 \alpha_B(T) \qty(\frac{T}{10^4 \unit{K}})^{0.37} T \\
+        \Aboxed{\frac{3}{2} T_{\rm init} &= T \qty[0.86  + 0.54 \qty(\frac{T}{10^4 \unit{K}})^{0.37}]}
+    \end{align}
+}
+
+\question{2d. Solve it!}
+\answer{
+    Using \href{https://docs.scipy.org/doc/scipy/reference/generated/scipy.optimize.fsolve.html}{\texttt{scipy.optimize.fsolve}}, I found that the equilibrium temperature is
+    \begin{equation}
+        \boxed{ T = 28918 \unit{K} }
+    \end{equation}
+}
+
+\question{2e. Metal Lines}
+\answer{
+    Metal line transitions would lead to addition cooling and produce an extra term for the cooling side of the balance equation.
+Each potential metal transition allows a different photon energy to absorbed and then re-emit.
+Many of the forbidden line transitions can then produce photons that escape easily and lead to cooling.
+
+    This will additionally work differently in different density regimes.
+At low density there is a stronger ($n^2$) dependence on number density, but at higher density this weakens to a linear scaling with $n$.
+}
+
+\clearpage
+
+\question{Bonus round - OII cooling}
+\answer{
+    Congratulations - by making it this far through the problem set you've unlocked a secret bonus round!
+Let's investigate how the cooling from a particular OII transition could change things in a cloud with a nonzero metallicity.
+
+    \noindent Imagine that we have an HII region with solar abundance of oxygen (but somehow no other metals) and let's just focus on a particular transition that Jess indicated would be important.
+Additionally, let's just operate in the low density regime (which is maybe reasonable at this temperature?
+Either way I'm doing it to help the algebra :p).\\
+
+    \noindent Okay now let's rebalance our heating and cooling with an additional term for the low-density metal line cooling from OII.
+    \begin{equation}
+        \frac{3}{2} T_{\rm init} = T \qty[0.86 + 0.54 \qty(\frac{T}{10^4 \unit{K}})^{0.37}] + \underbrace{f_{O/H} \gamma_{lu}(T) \frac{h \nu_{ul}}{k_B \alpha_B(T)}}_{\text{this is the new bit}}
+    \end{equation}
+    Nice!
+Now we need to define all of these delightfully delectable terms.
+$f_{O/H}$ is the relative number density of oxygen to hydrogen in the Sun, looking this up online\footnote{\url{http://hyperphysics.phy-astr.gsu.edu/hbase/Tables/suncomp.html}} I found that
+    \begin{equation}
+        f_{O/H} = 0.043 / 91.2 = 4.7 \times 10^{-4}
+    \end{equation}
+    It would now be useful to consider which transition we are focussing on (it's about to be relevant), so we can use my final project to work out the spectroscopic terms of OII and plot up its energy level diagram (with some transitions grabbed from Draine):
+    \begin{center}
+        \includegraphics[width=0.8\textwidth]{OII_levels.pdf}
+    \end{center}
+    We're going to focus on the second transition shown here with a wavelength of $\lambda = 3278.8 \unit{\AA}$. Draine gives (in Table F.4) that, for this transition,
+    \begin{equation}
+        \Omega_{ul} = 0.803 T_4^{0.023 - 0.008 \ln T_4},
+    \end{equation}
+    where $T_4 = T / 10^4 \unit{K}$. In lecture 5 (towards the end) we related $\Omega_{ul}$ to $\gamma_{ul}$ as
+    \begin{equation}
+        \gamma_{ul} = \frac{\Omega_{ul}}{g_u} \frac{h^2}{(2 \pi m_e)^{3/2} (k_B T)^{1/2}}
+    \end{equation}
+    And in this case $g_u = 2 (5/2) + 1 = 6$. Okay hang in there, just two more terms (and they're easier!). Given the transition wavelength, we immediately have the energy:
+    \begin{equation}
+        h \nu_{ul} = \frac{h c}{\lambda_{ul}} = 3.33 \unit{eV}
+    \end{equation}
+    And we are given an expression for $\alpha_B(T)$ in the question
+    \begin{equation}
+        \alpha_B(T) = 4 \times 10^{-13} \unit{cm^3}{s^{-1}} T_4^{-0.73}
+    \end{equation}
+    And that's everything we need! We can summarise at the new cooling rate from OII
+    \begin{align}
+        \Lambda_{\rm OII} &= 4.7 \times 10^{-4} \qty(0.803 T_4^{0.023 - 0.008 \ln T_4}) \frac{1}{6} \frac{h^2}{(2 \pi m_e)^{3/2} (k_B T)^{1/2}} \frac{h c}{\lambda k_B} \cdot 2.5 \times 10^{12} \unit{cm^{-3}}{s} T_4^{0.73} \\
+        \Lambda_{\rm OII} &= \qty[1.57 \times 10^{8} \frac{h^3 c}{\lambda (2 \pi m_e)^{3/2} k_B^{3/2}}\cdot \unit{cm^{-3}}{s}] \cdot T_4^{0.753 - 0.008 \ln T_4} T^{-1/2}
+    \end{align}\\
+
+    \noindent Now it \emph{should} be as easy as plugging this into the same Python function as before...but sadly \textbf{there seems to be a mistake somewhere} because it crashes.
+When we were doing this at the end of class, we just added units to the $\Omega_{ul}$ terms which I don't think is valid given the conversion that I found in the lecture notes.
+This allowed us to find a temperature of a couple thousand Kelvin.
+
+    In reality, if I plug in this expression, the cooling seems to be so large that there is no positive temperature solution. I'd be very intrigued to hear what I did wrong here!
+}
+
 \end{document}

### 2022-11-13 541_ism/week6/problem_2.py

@@ -1,7 +1,53 @@
 from scipy.optimize import fsolve
 import numpy as np
+import astropy.constants as const
+import astropy.units as u
+import matplotlib.pyplot as plt
 
-def func(T):
-    return T * (0.86 + 0.54 * (T / 1e4)**(0.37)) - 3/2 * 32000
+import sys
+sys.path.append("../final_project")
+from atom import Atom
 
-print(fsolve(func, 32000))
+# relative solar abundance of oxygen vs. hydrogen
+# http://hyperphysics.phy-astr.gsu.edu/hbase/Tables/suncomp.html
+fO_H = 0.043 / 91.2
+
+# energy emitted from the [OII 3727] transition
+wavelength = 3728.8 * u.Angstrom
+hnu = (const.h * const.c / wavelength)
+
+print(f"h nu = {hnu.to(u.eV):1.2f}")
+
+
+# total recombination coefficient
+def alpha(T4):
+    return 4e-13 * (u.cm**3 / u.s) * (T4)**(-0.73)
+
+
+def HII_cooling(T, T_init):
+    """Function for fsolve - for HII cooling a pure HII region"""
+    T4 = T / 1e4
+    return T * (0.86 + 0.54 * (T4)**(0.37)) - 3/2 * T_init
+
+
+def HII_cooling_with_OII(T, T_init):
+    """Function for fsolve - for HII cooling with a particular OII line"""
+    T4 = T / 1e4
+
+    oII_term = (T4**(0.753 - 0.008 * np.log(T4)) * (T * u.K)**(-1/2)
+                * 1.57e8 * u.cm**(-3) * u.s * (const.h**3 * const.c /
+                (wavelength * (2 * np.pi * const.m_e)**(3/2) * (const.k_B)**(3/2))))
+
+    return T * (0.86 + 0.54 * (T4)**(0.37)) + oII_term.to(u.K).value - 3/2 * T_init
+
+
+print(fsolve(HII_cooling, 32000, args=(32000)))
+print(fsolve(HII_cooling_with_OII, 32000, args=(32000)))
+
+fig, ax = Atom("OII").plot_energy_levels([(2, 1, 3276.0 * u.Angstrom),
+                                          (3, 1, 3278.8 * u.Angstrom),
+                                          (4, 1, 2470.3 * u.Angstrom),
+                                          (5, 1, 2470.2 * u.Angstrom)],
[…]

### 2022-11-15 541_ism/week6/problem_2.py
try changing to subsolar abundance

@@ -10,7 +10,7 @@ from atom import Atom
 
 # relative solar abundance of oxygen vs. hydrogen
 # http://hyperphysics.phy-astr.gsu.edu/hbase/Tables/suncomp.html
-fO_H = 0.043 / 91.2
+fO_H = 0.078 / 91.2 / 10
 
 # energy emitted from the [OII 3727] transition
 wavelength = 3728.8 * u.Angstrom
@@ -35,7 +35,7 @@ def HII_cooling_with_OII(T, T_init):
     T4 = T / 1e4
 
     oII_term = (T4**(0.753 - 0.008 * np.log(T4)) * (T * u.K)**(-1/2)
-                * 1.57e8 * u.cm**(-3) * u.s * (const.h**3 * const.c /
+                * 0.1 * 1.57e8 * u.cm**(-3) * u.s * (const.h**3 * const.c /
                 (wavelength * (2 * np.pi * const.m_e)**(3/2) * (const.k_B)**(3/2))))
 
     return T * (0.86 + 0.54 * (T4)**(0.37)) + oII_term.to(u.K).value - 3/2 * T_init

### 2022-11-15 541_ism/week7/solutions.tex
add first week7 problem for ISM

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{array}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\psetnum}{7}
\newcommand{\class}{ASTR 541 - Interstellar Medium}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Week \psetnum}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\newcolumntype{C}{>{$}c<{$}}

\begin{document}

\tomtitle{}

\noindent For reference, if you'd ever like to see the code that I've used to get my answers to these, \href{https://github.com/TomWagg/uw-grad-classes/tree/main/541_ism}{here's a link to my GitHub repo}! (\#astropy.units for life)\\

\question{\textbf{1. HI 21cm Radio Observations}}

\question{1a. On/off Line}
\answer{
    On-line means taking a measurement at the frequency the line (in this case, the HI 21cm line)\footnote{Not to be confused with ``online'', which is a term that the kids use when they're referring to this new-fangled interweb thing.}, whereas an off-line measurement is taken at a different frequency, outside of the line profile of the target line.

    \noindent Tobin put it nicely as on-line being like tuning into the 21.11cm FM radio station and off-line as listening to the static between stations.
    \begin{align}
        F_{\rm on-line} &= F_{\nu, 0}\, e^{-\tau_\nu},\qquad \text{a.k.a. ``You're listening to 21.11cm FM!''} \\
        F_{\rm off-line} &= F_{\nu, 0}\,,\qquad\,\,\,\,\,\quad \text{a.k.a. ``Hsssttttt''}
    \end{align}
}

\question{1b. Optical Depth Limit}
\answer{
    We are given that the change in flux density is below 1\%. We can use this to solve for what optical depth this gives.
    \begin{align}
        F_{\rm off-line} - F_{\rm on-line} &< 0.01 F_{\rm off-line}\\
        1 - e^{-\tau_\nu} &< 0.01\\
        e^{-\tau_\nu} &> 0.99\\
        \Aboxed{ \tau_\nu &> 0.01 }
    \end{align}
}

\question{1c. Temperature Limit}
\answer{
    The equation in the problem gives that
    \begin{equation}
        \tau_\nu = 0.552 \qty(\frac{100 \unit{K}}{T_{\rm spin}}) \qty(\frac{\dv{N_{\rm HI}}{\nu}}{\frac{10^{20} \unit{cm^{-2}}}{\unit{km}{s^{-1}}}}),
    \end{equation}
    which can be quite easily transformed to get $T_{\rm spin}$ as a function of optical depth
    \begin{equation}
        T_{\rm spin} = 55.2 \unit{K}\cdot \tau_\nu \qty(\frac{\dv{N_{\rm HI}}{\nu}}{\frac{10^{20} \unit{cm^{-2}}}{\unit{km}{s^{-1}}}}),
    \end{equation}
    We are given $\dd{N_{\rm HI}} / \dd{\nu}$ in the question and so plugging this in with the optical depth limit we find that the upper limit on the temperature is
    \begin{equation}
        \boxed{ T_{\rm spin} > 824 \unit{K} }
    \end{equation}
}

\question{1d. More absorption}
\answer{
    Repeating parts (b) and (c) but instead using a change of 50\% instead, we find that
    \begin{equation}
        \boxed{ T_{\rm spin} = 12 \unit{K} }
    \end{equation}
}

\question{1e. Spin temperature and absorption}
\answer{
    We note that the trend is \textbf{a higher spin temperature results in a lower detectability of HI-21cm in absorption}. A quick answer to this is that the emissivity is temperature independent whilst the absorptivity is inversely proportional to temperature and thus the rate of absorption declines with increased temperatures.

    More conceptually, an increased spin temperature is the same as an increase excitation temperature, implying that there are a larger number of collisions. This increases number of collisions affects the level populations of HI and increases the rate of emission. As this emission increases, it cancels more absorption and thus decreases the detectability of the absorption.
}

\end{document}

### 2022-11-15 541_ism/week7/problem_1.py

@@ -1,0 +1,21 @@
+import numpy as np
+import astropy.units as u
+import matplotlib.pyplot as plt
+
+def does_stuff(frac):
+    tau_nu = -np.log(1 - frac)
+    T_spin = 0.552 * 100 * u.K / tau_nu * (3 / 20)
+    return tau_nu, T_spin
+
+fracs = [0.01, 0.5]
+
+for frac in fracs:
+    print(f"< {frac} fractional change")
+    tau_nu, T_spin = does_stuff(frac)
+    print(f"  Optical depth < {tau_nu:1.2f}")
+    print(f"  Spin temperature > {T_spin:1.0f}")
+
+
+frac_range = np.logspace(-5, 0, 1000)
+plt.loglog(frac_range, does_stuff(frac_range)[0])
+plt.show()

### 2022-11-18 541_ism/week7/solutions.tex
finish week 7 problems for ISM

@@
 
+\question{\textbf{2. Broadening}}
+
+\question{2a. Optically Thin Solution}
+\answer{
+    We have a system of two equations with two unknowns, let's do some rearranging. I'm also going to refer to the FWHM as $w$ for brevity.
+    \begin{align}
+        \frac{w_1^2}{8 \ln 2} &= \sigma_{v, {\rm turb}}^2 + \frac{k_B T}{M_1},\qquad \frac{w_2^2}{8 \ln 2} = \sigma_{v, {\rm turb}}^2 + \frac{k_B T}{M_2} \\
+        \frac{w_1^2}{8 \ln 2} - \frac{k_B T}{M_1} &= \frac{w_2^2}{8 \ln 2} - \frac{k_B T}{M_2} \\
+        \frac{w_1^2}{8 \ln 2} - \frac{w_2^2}{8 \ln 2} &= T \qty(\frac{k_B}{M_1} - \frac{k_B}{M_2}) \\
+        T &= \frac{w_1^2 - w_2^2}{8 \ln 2} \qty(\frac{k_B}{M_1} - \frac{k_B}{M_2})^{-1}
+    \end{align}
+    We can plug in $w_1 = 10 \unit{km}{s^{-1}}$, $w_2 = 5 \unit{km}{s^{-1}}$, $M_1 = 1 \unit{amu}$ and $M_2 = 23 \unit{amu}$ to find that the value for an optically thin solution are
+    \begin{equation}
+        \boxed{ T = 1713 \unit{K},\qquad \sigma_{v, {\rm turb}} = 1.97 \unit{km}{s^{-1}} }
+    \end{equation}
+}
+
+\question{2b. What if it's not optically thin?}
+\answer{
+    The FHWM is a larger when it is saturated and so this represents \textbf{an upper limit on the FWHM}.
+    \begin{equation}
+        \sigma_{v, {\rm turb}}^2 + \frac{k_B T}{M} \le \sigma_v^2
+    \end{equation}
+    For a similar reason the inferred \textbf{$\sigma_{v, {\rm turb}}$ is an upper limit}.
+However, if you hold the turbulence fixed and assume the FWHM is an upper limit, this gives that \textbf{the temperature is a lower limit}.
+}
+
 \end{document}

### 2022-11-18 541_ism/week7/problem_2.py

@@ -1,0 +1,15 @@
+import numpy as np
+import astropy.units as u
+import astropy.constants as const
+
+w1 = 10 * u.km / u.s
+w2 = 5 * u.km / u.s
+M1 = 1 * const.m_p
+M2 = 23 * const.m_p
+
+T = (((w1**2 - w2**2) / (8 * np.log(2))) * (const.k_B / M1 - const.k_B / M2)**(-1)).decompose()
+
+sigma_turb = np.sqrt(w1**2 / (8 * np.log(2)) - const.k_B * T / M1)
+
+print(f"Temperature: {T.to(u.K):1.0f}")
+print(f"Turblence: {sigma_turb.to(u.km / u.s):1.2f}")

### 2022-12-05 541_ism/week8/solutions.tex
week 8 stuff

--- 2022-11-15 541_ism/week7/solutions.tex
@@
 
-\newcommand{\psetnum}{7}
+\newcommand{\psetnum}{8}
 \newcommand{\class}{ASTR 541 - Interstellar Medium}
@@
 
-\question{1a. On/off Line}
-\answer{
-    On-line means taking a measurement at the frequency the line (in this case, the HI 21cm line)\footnote{Not to be confused with ``online'', which is a term that the kids use when they're referring to this new-fangled interweb thing.}, whereas an off-line measurement is taken at a different frequency, outside of the line profile of the target line.
-
-    \noindent Tobin put it nicely as on-line being like tuning into the 21.11cm FM radio station and off-line as listening to the static between stations.
-    \begin{align}
-        F_{\rm on-line} &= F_{\nu, 0}\, e^{-\tau_\nu},\qquad \text{a.k.a. ``You're listening to 21.11cm FM!''} \\
-        F_{\rm off-line} &= F_{\nu, 0}\,,\qquad\,\,\,\,\,\quad \text{a.k.a. ``Hsssttttt''}
-    \end{align}
-}
-
-\question{1b. Optical Depth Limit}
-\answer{
-    We are given that the change in flux density is below 1\%. We can use this to solve for what optical depth this gives.
-    \begin{align}
-        F_{\rm off-line} - F_{\rm on-line} &< 0.01 F_{\rm off-line}\\
-        1 - e^{-\tau_\nu} &< 0.01\\
-        e^{-\tau_\nu} &> 0.99\\
-        \Aboxed{ \tau_\nu &> 0.01 }
-    \end{align}
-}
-
-\question{1c. Temperature Limit}
-\answer{
-    The equation in the problem gives that
-    \begin{equation}
-        \tau_\nu = 0.552 \qty(\frac{100 \unit{K}}{T_{\rm spin}}) \qty(\frac{\dv{N_{\rm HI}}{\nu}}{\frac{10^{20} \unit{cm^{-2}}}{\unit{km}{s^{-1}}}}),
-    \end{equation}
-    which can be quite easily transformed to get $T_{\rm spin}$ as a function of optical depth
-    \begin{equation}
-        T_{\rm spin} = 55.2 \unit{K}\cdot \tau_\nu \qty(\frac{\dv{N_{\rm HI}}{\nu}}{\frac{10^{20} \unit{cm^{-2}}}{\unit{km}{s^{-1}}}}),
-    \end{equation}
-    We are given $\dd{N_{\rm HI}} / \dd{\nu}$ in the question and so plugging this in with the optical depth limit we find that the upper limit on the temperature is
-    \begin{equation}
-        \boxed{ T_{\rm spin} > 824 \unit{K} }
-    \end{equation}
-}
-
-\question{1d. More absorption}
-\answer{
-    Repeating parts (b) and (c) but instead using a change of 50\% instead, we find that
-    \begin{equation}
-        \boxed{ T_{\rm spin} = 12 \unit{K} }
-    \end{equation}
-}
-
-\question{1e. Spin temperature and absorption}
-\answer{
-    We note that the trend is \textbf{a higher spin temperature results in a lower detectability of HI-21cm in absorption}.
-A quick answer to this is that the emissivity is temperature independent whilst the absorptivity is inversely proportional to temperature and thus the rate of absorption declines with increased temperatures.
-
-    More conceptually, an increased spin temperature is the same as an increase excitation temperature, implying that there are a larger number of collisions.
-This increases number of collisions affects the level populations of HI and increases the rate of emission.
-As this emission increases, it cancels more absorption and thus decreases the detectability of the absorption.
-}
-
 \end{document}

### 2022-12-05 541_ism/week8/problem_2.py

@@ -1,0 +1,15 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import astropy.units as u
+
+
+def column_density(ew_rest, osc_strength, wavelength):
+    return 1.13e17 * ew_rest.to(u.mAngstrom).value / (osc_strength * wavelength.to(u.Angstrom).value**2) * u.cm**(-2)
+
+
+labels = ["SiII 1304", "SiII 1260", "SiII 1193", "SiII 1190"]
+wavelengths = [1304.3702, 1260.4221, 1193.2897, 1190.4158] * u.Angstrom
+osc_strengths = [0.0863, 1.18, 0.582, 0.292]
+ew_rests = [210.945, 504.717, 434.75, 236.652 + 111.967] * u.mAngstrom
+
+log_column_density_1304 = 14.3786

### 2022-12-05 541_ism/week9/solutions.tex
ISM week 9 problems all done

--- 2022-11-15 541_ism/week7/solutions.tex
@@
 
-\newcommand{\psetnum}{7}
+\newcommand{\psetnum}{9}
 \newcommand{\class}{ASTR 541 - Interstellar Medium}
@@
 
-\question{\textbf{1. HI 21cm Radio Observations}}
+\question{\textbf{1. Molecules}}
 
-\question{1a. On/off Line}
+\question{Part 1: Rate Equations}
 \answer{
-    On-line means taking a measurement at the frequency the line (in this case, the HI 21cm line)\footnote{Not to be confused with ``online'', which is a term that the kids use when they're referring to this new-fangled interweb thing.}, whereas an off-line measurement is taken at a different frequency, outside of the line profile of the target line.
-
-    \noindent Tobin put it nicely as on-line being like tuning into the 21.11cm FM radio station and off-line as listening to the static between stations.
+    Each rate equation uses the rate coefficient and the number density of each reactant on the left hand side of the equations.
+The exception is the third where the number density of photons has been accounted for.
     \begin{align}
-        F_{\rm on-line} &= F_{\nu, 0}\, e^{-\tau_\nu},\qquad \text{a.k.a. ``You're listening to 21.11cm FM!''} \\
-        F_{\rm off-line} &= F_{\nu, 0}\,,\qquad\,\,\,\,\,\quad \text{a.k.a. ``Hsssttttt''}
+        \mathcal{R}_{\rm ra} &= k_{\rm ra} n_H n_e \\
+        \mathcal{R}_{\rm ad} &= k_{\rm ad} n_H n_{H^{-}} \\
+        \mathcal{R}_{\rm pd} &= \zeta_{\rm pd} n_{H^{-}}
     \end{align}
@@
 
-\question{1b. Optical Depth Limit}
+\question{Part 2: Rate balance}
 \answer{
-    We are given that the change in flux density is below 1\%. We can use this to solve for what optical depth this gives.
+    Now we can balance these equations and solve for the ratio of ionised to neutral hydrogen.
     \begin{align}
-        F_{\rm off-line} - F_{\rm on-line} &< 0.01 F_{\rm off-line}\\
-        1 - e^{-\tau_\nu} &< 0.01\\
-        e^{-\tau_\nu} &> 0.99\\
-        \Aboxed{ \tau_\nu &> 0.01 }
+        \mathcal{R}_{\rm ra} &= \mathcal{R}_{\rm ad} + \mathcal{R}_{\rm pd} \\
+        k_{\rm ra} n_H n_e &= k_{\rm ad} n_H n_{H^{-}} + \zeta_{\rm pd} n_{H^{-}} \\
+        \frac{n_{H^{-}}}{n_H} &= \frac{k_{\rm ra} n_e}{k_{\rm ad} n_H + \zeta_{\rm pd}} \\
+        \Aboxed{ \frac{n_{H^{-}}}{n_H} &= 1.36 \times 10^{-11} }
     \end{align}
@@
 
-\question{1c. Temperature Limit}
+\question{Part 3: Fraction of associated detachment}
 \answer{
-    The equation in the problem gives that
+    We can find the fraction out of total detachment as follows (luckily the H- cancels so we can do this easily!)
     \begin{equation}
-        \tau_\nu = 0.552 \qty(\frac{100 \unit{K}}{T_{\rm spin}}) \qty(\frac{\dv{N_{\rm HI}}{\nu}}{\frac{10^{20} \unit{cm^{-2}}}{\unit{km}{s^{-1}}}}),
-    \end{equation}
-    which can be quite easily transformed to get $T_{\rm spin}$ as a function of optical depth
-    \begin{equation}
-        T_{\rm spin} = 55.2 \unit{K}\cdot \tau_\nu \qty(\frac{\dv{N_{\rm HI}}{\nu}}{\frac{10^{20} \unit{cm^{-2}}}{\unit{km}{s^{-1}}}}),
-    \end{equation}
-    We are given $\dd{N_{\rm HI}} / \dd{\nu}$ in the question and so plugging this in with the optical depth limit we find that the upper limit on the temperature is
-    \begin{equation}
-        \boxed{ T_{\rm spin} > 824 \unit{K} }
+        \boxed{ \frac{\mathcal{R}_{\rm ad}}{\mathcal{R}_{\rm ad} + \mathcal{R}_{\rm pd}} = \frac{k_{\rm ad} n_H}{k_{\rm ad} n_H + \zeta_{\rm pd}} = 0.14 }
     \end{equation}
@@
 
-\question{1d. More absorption}
+\pagebreak
+
+\question{Part 4: Compare to dust}
 \answer{
-    Repeating parts (b) and (c) but instead using a change of 50\% instead, we find that
+    We can plug in the numbers given in the question to find
     \begin{equation}
-        \boxed{ T_{\rm spin} = 12 \unit{K} }
+        \mathcal{R}_{\rm H^{-}} \equiv k_{\rm ad} \frac{n_{H^{-}}}{n_H} = 1.77 \times 10^{-20} \unit{cm^3}{s^{-1}}
     \end{equation}
+    We can take a ratio and find
+    \begin{equation}
+        \boxed{ \frac{\mathcal{R}_{\rm H^{-}}}{\mathcal{R}_{\rm dc}} \approx 1700 }
+    \end{equation}
+    So the dust is much stronger!
 }
 
-\question{1e. Spin temperature and absorption}
+\question{Part 5: What temperature do we need}
 \answer{
-    We note that the trend is \textbf{a higher spin temperature results in a lower detectability of HI-21cm in absorption}.
-A quick answer to this is that the emissivity is temperature independent whilst the absorptivity is inversely proportional to temperature and thus the rate of absorption declines with increased temperatures.
-
-    More conceptually, an increased spin temperature is the same as an increase excitation temperature, implying that there are a larger number of collisions.
-This increases number of collisions affects the level populations of HI and increases the rate of emission.
-As this emission increases, it cancels more absorption and thus decreases the detectability of the absorption.
+    Now we can add a temperature dependence and solve for the temperature at which they are equal
+    \begin{align}
+        1.77 \times 10^{-20} T_2^{0.67} &= 3 \times 10^{-17} \\
+        \Aboxed{ T &= 6.6 \times 10^{6} \unit{K} }
+    \end{align}
+    This is not physically plausible because we'll never get neutral hydrogen in high enough quantities at these temperatures.
+Our assumption of a constant $n_{\rm H} = 30 \unit{cm^{-3}}$ cannot be valid.
 }

### 2022-12-05 541_ism/week9/problem_1.py

@@ -1,0 +1,22 @@
+import numpy as np
+import astropy.units as u
+
+k_ra = 1.9e-16 * u.cm**3 / u.s
+k_ad = 1.3e-9 * u.cm**3 / u.s
+zeta_pd = 2.4e-7 * u.s**(-1)
+
+n_H = 30 * u.cm**(-3)
+n_e = 0.02 * u.cm**(-3)
+
+ratio = (k_ra * n_e) / (k_ad * n_H + zeta_pd)
+
+print(f"n_H-/ n_H: {ratio:1.2e}")
+
+fraction_of_ad = k_ad * n_H / (k_ad * n_H + zeta_pd)
+print(f"Fraction of associated detachments {fraction_of_ad:1.2f}")
+
+R_hminus = k_ad * ratio
+print(R_hminus)
+
+R_ratio = ((3e-17 * u.cm**3 / u.cm**(-1)) / R_hminus).value
+print(f"Required temperature {R_ratio**(1/0.67) * 100 * u.K:1.2e}")

### 2022-12-06 541_ism/week8/solutions.tex
ism week 8 problem 1

@@
 
-\question{\textbf{1. HI 21cm Radio Observations}}
+\question{\textbf{1. Fun with the SiIV Doublet}}
+
+\question{1a. An exercise in counting}
+\answer{
+    We have copies of each component of the SiIV doublet on graph paper.
+We can therefore use the grid to estimate the ratio of equivalent widths.
+No doubt this grid has physical meaning and units...but it'll just cancel I guess so I'm going to ignore that and just count boxes :D
+
+    I find that the number of boxes for the 1402 is 14.5, and for the 1393 is 24. This gives that the ratio is
+    \begin{equation}
+        \boxed{ \frac{W_{1393}}{W_{1402}} = 1.66 }
+    \end{equation}
+}
+
+\question{1b. Curve of Growth Regime}
+\answer{
+    We would expect this ratio to be 2 if we were in the linear regime.
+Given that it is a metal line, we can assume that it's not in the damped regime and so this means that it must be in the \textbf{flat} regime.
+This means that we must have some saturation in our measurement, which will lead to an underestimate of the column density!!
+}
+
+\question{1c. $\tau_0$ constraints}
+\answer{
+    If we assume that my measurement of the equivalent widths was perfect (doubtful), the equation in the question gives that
+    \begin{align}
+        \frac{W_{1393}}{W_{1402}} &= \qty[1 + \ln(f_{1393} \lambda_{1393}/f_{1402} \lambda_{1402}) / \ln(\tau_0 / \ln 2)]^{1/2} \\
+        1.66 &= \qty[1 + 0.694 / \ln(\tau_0 / \ln 2)]^{1/2} \\
+        \ln(\tau_0 / \ln 2) &= 0.4 \\
+        \Aboxed{ \tau_0 &= 1.033 }
+    \end{align}
+}
+
+\question{1d. AODM (not Another Ol' Dark Matter)}
+\answer{
+    I should actually write down that this means Apparent Optical Depth Method or I'll forget haha.
+Given that our line is saturated and has an optical depth above 1, this method will only give a \textbf{lower limit} on the column density
+    \begin{align}
+        N &= 1.13 \times 10^{7} \unit{cm^{-2}} \frac{W_{\lambda} (\unit{m\angstrom})}{f_{\lambda} \lambda^2(\unit{\angstrom})} \\
+        N_{1393} &> 5.51 \times 10^{12} \unit{cm^{-2}} \\
+        \Aboxed{ N_{1402} &> 6.62 \times 10^{12} \unit{cm^{-2}} }
+    \end{align}
+    where we've used that $W_{1393} = 50 \unit{m\angstrom}$ (from the question) and that $W_{1402} = 30.2 \unit{m\angstrom}$ (using the ratio we found).
+}
+
+\question{1e. Cloudy with a chance of Silicon}
+\answer{
+    We find a lower limit on the abundance of silicon as
+    \begin{equation}
+        12 + \log_{10}({\rm Si / H}) > 12 + \log_{10}\qty(\frac{10 \cdot 6.62 \times 10^{12}}{10^{19}}) > 6.82 \\
+    \end{equation}
+    This is lower than solar abundance and so this cloud likely formed from more metal-poor gas.
+}
+
+\question{1f. FWHM is back}
+\answer{
+    We can calculate $\sigma_{v, \rm turb}$ as
+    \begin{align}
+        \sigma_{v, \rm turb} &= \qty[\frac{\rm FWHM^2}{8 \ln 2} - \frac{k_B T}{M}]^{1/2} \\
+        \sigma_{v, \rm turb} &= \qty[\frac{\rm (40 \unit{km}{s^{-1}})^2}{8 \ln 2} - \frac{k_B \cdot 10000 \unit{K}}{1 \unit{amu}}]^{1/2} \\
+        \Aboxed{ \sigma_{v, \rm turb} &= 14.35 \unit{km}{s^{-1}} }
+    \end{align}
+    If we assume that the temperature is the same, we would assume that the FWHM of the Silicon would be
+    \begin{equation}
+        {\rm FWHM}_{Si} = \sqrt{8 \ln 2 \qty[\sigma_{v, \rm turb}^2 - \frac{k_B T}{28 \unit{amu}}]} = 34 \unit{km}{s^{-1}}
+    \end{equation}
+    If we instead measure $50 \unit{km}{s^{-1}}$ then it may be possible that our assumption of equal temperatures was not valid.
+}
 

### 2022-12-06 541_ism/week8/problem_1.py

@@ -1,0 +1,44 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import astropy.units as u
+import astropy.constants as const
+
+ratio = 24 / 14.5
+
+
+f1393 = 0.5280
+f1402 = 0.2620
+
+l1393 = 1393.755 * u.Angstrom
+l1402 = 1402.770 * u.Angstrom
+
+W1393 = 50 * u.mAngstrom
+W1402 = W1393 / ratio
+
+factor = np.log(f1393 * l1393 / (f1402 * l1402))
+tau_0 = np.e**(factor / (ratio**2 - 1)) * np.log(2)
+
+
+def column_density(ew, f, l):
+    return 1.13e17 * ew.to(u.mAngstrom).value /\
+        (f * l.to(u.Angstrom).value**2) * u.cm**(-2)
+
+
+print(f"tau_0 = {tau_0:1.2f}")
+
+N_1393 = column_density(W1393, f1393, l1393)
+N_1402 = column_density(W1402, f1402, l1402)
+print(f"N_1393 = {N_1393:1.2e}")
+print(f"N_1402 = {N_1402:1.2e}")
+
+abundance = 12 + np.log10(N_1402.value * 10 / 1e19)
+print(f"abundance = {abundance:1.2f}")
+
+FWHM = 40 * u.km / u.s
+T = 10000 * u.K
+M = 1 * const.m_p
+
+sigma_vturb = np.sqrt(FWHM**2 / (8 * np.log(2)) - const.k_B * T / M)
+print(sigma_vturb)
+
+print((sigma_vturb**2 + const.k_B * T / (M * 28))**(0.5) * np.sqrt(8 * np.log(2)))

### 2022-12-06 541_ism/week8/solutions.tex
ism week 8 all done

@@
 
+\question{\textbf{2. The Grand Return of Xspec GUI}}
+
+\question{2a. Favourite Systems}
+\answer{
+    \begin{center}
+        \begin{tabular}{c|c|c}
+            Redshift, $z$ & Shows SiIV? & Shows OVI? \\
+            \hline\hline
+            0.1470 & \checkmark & $\times$ \\
+            0.2055 & \checkmark & \checkmark \\
+            0.2180 & $\times$ & \checkmark \\
+            0.35566 & $\times$ & $\times$
+        \end{tabular}
+    \end{center}
+    In the cases in which there was no absorption, it is either because (i) the ratio of EWs is reversed, implying another species is interfering sneakily (ii) the line is redshifted outside of the observed wavelength range (iii) there was just no change in normalised flux.\\
+
+    \noindent \textbf{My favourite system is at $\mathbf{z = 0.35566}$}, its lack of both SiIV and OVI absorption lends it an air of mystery and wonder that is unparalleled amongst the other systems.
+Alas, I have doubts about its continued involvement in the remainder of the problem, so long $z = 0.35566$!
+}
+
+\question{2b. OVI Doublet}
+\answer{
+    Using the GUI I find that the OVI doublet EW ratios are
+    \begin{center}
+        \boxed{\begin{tabular}{c|c}
+            Redshift, $z$ & OVI Ratio \\
+            \hline\hline
+            0.2055 & 1.30 \\
+            0.2180 & 1.58
+        \end{tabular}}
+    \end{center}
+    Since both of these ratios are less than 2 (the expected ratio for the linear regime) and there is no evidence of strong lorentzian wings, this implies that both are in the \textbf{flat regime of the curve of growth}.
+}
+
+\question{1c. Curve of Growth}
+\answer{
+    Here's my curve of growth!
+We can see that the Silicon tends to be closer to the linear regime but it swerves off towards the flat part.
+The first HI line is possibily still in the flat part but after that is seems to rise following the square root regime.
+    \begin{center}
+        \includegraphics[width=0.8\textwidth]{cog.pdf}
+    \end{center}
+    Absorption line saturation can be contextualised as diverging from the linear part of the curve of growth and heading into the flat part I think.
+}
+
 \end{document}

### 2022-12-06 541_ism/week8/problem_2.py

@@ -2,14 +2,59 @@ import numpy as np
 import matplotlib.pyplot as plt
 import astropy.units as u
 
+plt.rc('font', family='serif')
+plt.rcParams['text.usetex'] = False
+fs = 24
+
+# update various fontsizes to match
+params = {'figure.figsize': (12, 8),
+          'legend.fontsize': 0.7 * fs,
+          'axes.labelsize': fs,
+          'xtick.labelsize': 0.7 * fs,
+          'ytick.labelsize': 0.7 * fs,
+          'axes.linewidth': 1.1,
+          'xtick.major.size': 7,
+          'xtick.minor.size': 4,
+          'ytick.major.size': 7,
+          'ytick.minor.size': 4}
+plt.rcParams.update(params)
 
 def column_density(ew_rest, osc_strength, wavelength):
     return 1.13e17 * ew_rest.to(u.mAngstrom).value / (osc_strength * wavelength.to(u.Angstrom).value**2) * u.cm**(-2)
 
 
-labels = ["SiII 1304", "SiII 1260", "SiII 1193", "SiII 1190"]
-wavelengths = [1304.3702, 1260.4221, 1193.2897, 1190.4158] * u.Angstrom
-osc_strengths = [0.0863, 1.18, 0.582, 0.292]
-ew_rests = [210.945, 504.717, 434.75, 236.652 + 111.967] * u.mAngstrom
+labels = ["SiII 1304", "SiII 1260", "SiII 1193", "SiII 1190", "HI 1215", "HI 1025", "HI 972", "HI 949"]
+wavelengths = [1304.3702, 1260.4221, 1193.2897, 1190.4158, 1215.67, 1025.7222, 972.5367, 949.7430] * u.Angstrom
+osc_strengths = [0.0863, 1.18, 0.582, 0.292, 0.4164, 0.07914, 0.02901, 0.01395]
+ew_rests = [210.945, 504.717, 434.75, 236.652 + 111.967, 1109.33, 735.994, 641.365, 601.016] * u.mAngstrom
 
 log_column_density_1304 = 14.3786
+
+Ns = np.concatenate((np.repeat(column_density(ew_rests[0], osc_strengths[0], wavelengths[0]), 4),
+                     np.repeat(10**(18.4) * u.cm**(-2), 4)))
+
+calc_Ns = column_density(ew_rests, osc_strengths, wavelengths).to(u.cm**(-2)).value
[…]

### 2023-02-09 597_rubin_lsst/HW6/solutions.tex
finish 597 HW6

\documentclass[12pt, letterpaper, twoside]{article}
\usepackage{nopageno,epsfig, amsmath, amssymb}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{empheq}
\usepackage{wrapfig}

\usepackage[letterpaper,
            margin=0.8in]{geometry}

\newcommand{\psetnum}{6}
\newcommand{\class}{ASTR 597A - Rubin Observatory and LSST}

\newcommand{\tomtitle}{
    \noindent {\LARGE \fontfamily{cmr}\selectfont \textbf{\class}} \hfill \\[1\baselineskip]
    \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
}

\title{\class : Week \psetnum}
\author{\textbf{Tom Wagg}}

\newcommand{\question}[1]{{\noindent \it #1}}
\newcommand{\answer}[1]{
    \par\noindent\rule{\textwidth}{0.4pt}#1\vspace{0.5cm}
}
\newcommand{\todo}[1]{{\color{red}\begin{center}TODO: #1\end{center}}}

% custom function for adding units
\makeatletter
\newcommand{\unit}[1]{%
    \,\mathrm{#1}\checknextarg}
\newcommand{\checknextarg}{\@ifnextchar\bgroup{\gobblenextarg}{}}
\newcommand{\gobblenextarg}[1]{\,\mathrm{#1}\@ifnextchar\bgroup{\gobblenextarg}{}}
\makeatother

\newcommand{\avg}[1]{\left\langle #1 \right\rangle}
\newcommand{\angstrom}{\mbox{\normalfont\AA}}
\allowdisplaybreaks

\begin{document}

\tomtitle{}

\question{\textbf{1. Definition of H}}
\answer{
    Given that $\alpha = 0$ and both $r$ and $\Delta$ must be $1 \unit{au}$, this means that the Observer would need to be in the same location as the Sun (uh oh).
    \begin{center}
        \includegraphics[width=0.5\textwidth]{asteroid_hw_diagram.png}
    \end{center}
}

\question{\textbf{2. Magnitude as a function of distances}}
\answer{
    We currently have an expression for the apparent magnitude as a function of flux. We therefore need an expression of flux in terms of $r$ and $\Delta$. First, let's consider the flux incident on the asteroid. For this we can write our usual expression as
    \begin{align}
        f_{\text{onto-asteroid}} = \frac{L_{\rm \odot}}{4 \pi r^2}
    \end{align}
    Now the asteroid reflects some fraction of this light based on its albedo, $a$. However, this light is emitted in all directions and so we must account for this spreading with a similar factor to the above
    \begin{equation}
        f(r, \Delta) \propto \frac{L_{\rm \odot}}{4 \pi r^2} \cdot \frac{a}{4 \pi \Delta^2},
    \end{equation}
    note that this is only proportional since in reality we'd probably need to integrate over the surface of the asteroid to get the units to work out. But we're about to take a ratio so let's not worry about that. We know that $f_0$ needs to be the value where we get $H$ so
    \begin{equation}
        f_0 = f(1 \unit{au}, 1 \unit{au}) = \frac{L_{\rm \odot} a}{16 \pi^2 \unit{au^2}}
    \end{equation}
    Therefore, we can get the ratio of these two and write the expression for magnitude as
    \begin{equation}
        \boxed{ m = H - 2.5 \log_{10} \qty(\frac{\unit{au^2}}{r^2 \Delta^2}) }
    \end{equation}
}

\clearpage

\question{\textbf{3. Limits}}
\answer{
    First, we are now taking $r = \Delta + 1 \unit{au}$, so we can write the expression for magnitude as
    \begin{equation}
        m = H - 2.5 \log_{10} \qty(\frac{\unit{au^2}}{(\Delta + 1 \unit{au})^2 \Delta^2})
    \end{equation}
    In the case where $\Delta$ is very large, we can ignore the addition of $1 \unit{au}$ and write
    \begin{equation}
        m \approx H - 2.5 \log_{10} \qty(\frac{\unit{au^2}}{\Delta^4})
    \end{equation}
    And so as $\Delta \to \infty$ we find that the magnitudes scales as
    \begin{equation}
        \boxed{ \lim_{\Delta \to \infty} m \propto - \log_{10} \qty( \frac{1}{\Delta^4} ) }
    \end{equation}
    So we can see that the magnitude is scaling with 1/distance$^4$ instead of just 1/distance$^2$ :o That's going to make detecting asteroids rather difficult! The key point here is that the light is \textit{reflected} and so you get this ``spreading out'' of light twice which really attenuates the flux.
}

\end{document}

### 2023-02-09 597_rubin_lsst/HW6/solutions.tex
add collaborators

@@
     \noindent {\large \fontfamily{cmr}\selectfont Problem Set \psetnum \hfill \textsc{Tom Wagg}}\\[0.5\baselineskip]
-    {\fontfamily{cmr}\selectfont \textit{\today}}\\[2\baselineskip]
+    {\fontfamily{cmr}\selectfont \textit{\today} \hfill Collaborators: David \& Tobin}\\[2\baselineskip]
 }
